### 2016-01-28 doc/doc.tex
adding a short document explaining how to use the emacs mode
git-svn-id: https://svn.divms.uiowa.edu/repos/clc/projects/lambda-encodings/implementations/cedille@22482 bd178f5f-7bea-4014-844f-495ca19b8805

\documentclass{article}

\usepackage{url}
\usepackage{parskip}
\usepackage{fullpage}

\begin{document}

\title{Documentation for Cedille}

\author{Aaron Stump}

\maketitle

\section{Basic setup}

Cedille is used via the \verb|cedille-mode| emacs mode.\footnote{How
  to install this mode is not covered in this document.}  This mode is
based on Carl Olson's \verb|structured-editing| (\verb|se|) mode,
which provides a generic, language-independent (minor) mode for
navigating through structured text.  If you open a \verb|.ced| file,
Cedille mode will be invoked, and a request will be made to
\verb|cedille.sh| to parse the file.  This \verb|cedille.sh| wrapper
script will invoke the actual \verb|cedille| executable on the file.
When the \verb|cedille| executable runs on a file like \verb|nat.ced|,
it will produce a file called \verb|nat.cede| containing all the
typing information for \verb|nat.ced|.  It will also do this for any
files that are imported by \verb|nat.ced|.  (Note that currently, any
imported files must be in the same directory.)  The \verb|cedille.sh|
script will send the contents of the \verb|.cede| file (e.g.,
\verb|nat.cede|) back to the emacs mode.  Then Carl's code will parse
this information to construct a parse tree, which can then be
navigated within the emacs mode, as described more below.

\section{Entering and exiting navigation mode}

To enter navigation mode for a buffer holding a \verb|.ced| file, you
type ``Meta-s'', where Meta is likely bound to the Alt key on your
keyboard.  In a pinch, the Escape key always works for Meta.  Once in
navigation mode, you will see the \emph{mode line} at the bottom of
the buffer for the \verb|.ced| file display ``Cedille navi'' instead
of just ``Cedille''.  To exit the mode, you can type ``q'', or
``Control-g'' will also work.  You cannot modify the text while in
navi mode.  You can only traverse through it and view type
information.

To get some rudimentary help on the navi mode commands, you can type
``h'' while in navi mode.

\section{Structured navigation}

The basic idea of \verb|se-mode| is to allow the user to navigate
through text following the structure of a parse tree for that text.
The parse tree information is generated by the external tool, in this
case the \verb|cedille| executable, in the form of \emph{spans}.  A
span consists of a name, a starting character position in the file, an
ending character position, and additional information.  The nested
structure of spans is what \verb|se-mode| uses to reconstruct the
parse tree for the text. The basic navigation commands are the
following:

\begin{itemize}
\item ``p'' selects the innermost span which starts at point
  (``point'' is the character position at which the cursor is
  currently located in the buffer).  If a span is already selected,
  then select its parent span (i.e., go up in the parse tree).

\item ``n'' selects the span corresponding to the first child of 
  the parse tree.

\item ``f'' select the span corresponding to the next sibling (to the
  right) of the node in the parse tree whose span is currently
  displayed.

\item ``b'' select the span corresponding to the previous sibling (to the left) of the node in the parse tree whose span is currently displayed.

\item ``a'' select the first span out of all the siblings of the currently
 displayed span.

\item ``e'' select the last span out of all the currently displayed span's siblings.

\end{itemize}

\section{The info buffer}

For each file \verb|X.ced| being viewed in navi mode, the Cedille
emacs mode will create an info buffer, with the name
\verb|*cedille-info-X*|.  To display this buffer, you type ``i'' in
navi mode.  To hide it, you also type ``i''.  This buffer shows
information associated (by the \verb|cedille| executable) with the
currently selected span of the buffer.

\section{Errors}

Parse errors in Cedille can be hard to debug, and due to the
home-grown parser technology (\verb|gratr|) that I am using, a file
with parse errors can be very slow to process by the \verb|cedille|
executable.  If there are parse errors, they will be displayed in a
small buffer when navi mode is activated.  They must be fixed before
navigation features will work.

Assuming there are no parse errors, you will frequently have type
errors to deal with in your Cedille code.  To see the next type error,
type ``r'' in navi mode.  To see the previous one, type ``R''.  If
there are no errors, you will see ``No errors'' displayed when you
type ``r'' or ``R''.

\end{document}

### 2018-02-12 parser/test/Tests.hs
Haskell parser integration. Previous parser is still available through "make cedille-old" directive. Note the equality type, lifting types, let/in and comments syntax have changed (more detail about the syntax changes can be found at parser/README.md). In the newlib folder are the contents of lib folder exported to the new syntax.
git-svn-id: https://svn.divms.uiowa.edu/repos/clc/projects/lambda-encodings/implementations/cedille@29231 bd178f5f-7bea-4014-844f-495ca19b8805

@@ -1,0 +1,208 @@
+{-# LANGUAGE OverloadedStrings #-}
+module Main(main) where
+
+import Prelude
+import Test.HUnit
+import CedilleTypes
+import CedilleLexer hiding (main)
+import CedilleParser hiding (main)
+import qualified CedilleCommentsLexer as C
+import Control.Arrow
+import Control.Monad
+import System.IO  
+import qualified System.Exit as Exit
+import Data.Text (Text, pack)
+
+test_comments_str = unlines [
+    " aaaaaa            " -- 20 characters per row
+  , " ddk 4adjf30  ,,   "
+  , " {- comments       "
+  , "    more           "
+  , "               -}  "
+  , " dadfdf            "
+  , " aaaa -- comments  "
+  , " fffffffffffffff   " ]
+
+test_lexer_comments = TestCase (assertEqual "test lexer comments "
+                        (C.File (C.Entity (C.EntityComment "42" "98") (C.Entity (C.EntityComment "127" "140") C.EndEntity)))
+                        (C.scanComments (pack test_comments_str)))
+  
+
+removeLexerPositions :: Either a [Token] -> Either a [TokenClass]
+removeLexerPositions = right (map (\ (Token _ tc) -> tc))
+
+test_lexer_str = "αβγδ" -- greek letters
+
+test_lexer = TestCase (assertEqual "test lexer 0"
+                        (Prelude.Right [TVar "αβγδ"])
+                        (removeLexerPositions (scanner test_lexer_str)))
+                        
+
+test_parser_str = unlines [
+  "import ../hello/world . %comments",
+  "import module . ",
+  "%comments 2 ",
+  "module αβγδ . "
+  ]
+
+test_parser_module = TestCase (assertEqual "test parser 0"
+                               (Prelude.Right (File "0" ImportsStart "αβγδ" ParamsNil CmdsStart "34"))
+                               (runAlex test_parser_str $ cedilleParser))
+
+-- test_atype_hole_str = "  ●"
+-- test_atype_hole = TestCase (assertEqual "test atype Hole"
+--                        (Prelude.Right (TpHole (AlexPn 2 1 3)))
+--                        (runAlex test_atype_hole_str $ atypes))
+
[…]

### 2018-02-12 src/CedilleTypes.hs

@@ -1,0 +1,173 @@
+module CedilleTypes where
+
+import Prelude hiding(Num)
+import CedilleLexer
+import Data.Text(Text)
+
+type Num = Text
+type Fpth = Text
+type Var = Text
+type Bvar = Text
+type Qvar = Text
+type Qkvar = Text
+type Kvar = Text
+type PosInfo = Text
+
+data Arg = TermArg Term | TypeArg Type
+     deriving (Show,Eq)
+
+data Args = ArgsCons Arg Args | ArgsNil PosInfo
+     deriving (Show,Eq)
+     
+data ArrowType = ErasedArrow | UnerasedArrow
+     deriving (Show,Eq)
+     
+data Binder = All | Pi
+     deriving (Show,Eq)
+     
+data Cmd =
+       DefKind PosInfo Kvar Params Kind PosInfo
+     | DefTermOrType DefTermOrType PosInfo
+     | ImportCmd Imprt
+     deriving (Show,Eq)
+
+data Cmds =
+        CmdsNext Cmd Cmds
+      | CmdsStart 
+      deriving (Show,Eq)
+
+data Decl =
+     Decl PosInfo PosInfo Bvar Tk PosInfo
+     deriving (Show,Eq)
+
+data DefTermOrType =
+       DefTerm PosInfo Var MaybeCheckType Term
+     | DefType PosInfo Var Kind Type
+     deriving (Show,Eq)
+     
+data Imports =
+       ImportsNext Imprt Imports
+     | ImportsStart
+     deriving (Show,Eq)
+
+data Imprt =
+     Import PosInfo Fpth OptAs Args PosInfo
+     deriving (Show,Eq)
+
+data Kind =
+       KndArrow Kind Kind
+     | KndParens PosInfo Kind PosInfo
+     | KndPi PosInfo PosInfo Bvar Tk Kind
+     | KndTpArrow Type Kind
+     | KndVar PosInfo Qkvar Args
+     | Star PosInfo
+     deriving (Show,Eq)
+
+data Lam = ErasedLambda | KeptLambda
+     deriving (Show,Eq)
+
+data LeftRight = Both | Left | Right
+     deriving (Show,Eq)
+
+data LiftingType =
+       LiftArrow LiftingType LiftingType
+     | LiftParens PosInfo LiftingType PosInfo
+     | LiftPi PosInfo Bvar Type LiftingType
+     | LiftStar PosInfo
[…]

### 2018-02-14 docs/cedille-core/spec.tex
spec for cedille core
git-svn-id: https://svn.divms.uiowa.edu/repos/clc/projects/lambda-encodings/implementations/cedille@29272 bd178f5f-7bea-4014-844f-495ca19b8805

\documentclass{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{url}
\usepackage{proof}
\usepackage{stmaryrd}
\usepackage{parskip}
\usepackage{fullpage}

\newcommand{\choice}[0]{\zeta}
\newcommand{\elcap}[0]{\cap}
\newcommand{\abs}[4]{{#1}\, #2\! : \! #3.\, #4}
\newcommand{\absu}[3]{{#1}\, #2.\, #3}
\newcommand{\iolamp}[0]{\ensuremath{\iota\lambda P 2}}
\mathchardef\mhyph="2D % Define a "math hyphen"
\begin{document}

\newcommand{\interp}[1]{\llbracket #1 \rrbracket} 
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{corollary}[theorem]{Corollary}

\title{Syntax and Typing for Core Cedille}

\author{Aaron Stump}

\maketitle

\section{Motivation}

This note proposes a syntax for annotated terms and type-computation
rules for them, for a core version of Cedille.  Cedille should be
easily translatable to core Cedille, though we have neither proved nor
implemented this yet (we are planning to implement in the next
month or so).

The goal is to have a version of Cedille for which typing can be
implemented straightforwardly, in a small, trustworthy checker.
Cedille as we are implementing it for interactive development of
proofs has a lot of features that are not needed in a lower-level
format intended just for confirming typings.  The checker for core
Cedille does not need to produce span information for the benefit of
an IDE, as our full implementation of Cedille does.  Furthermore, it
is acceptable for core Cedille to require more annotations on terms,
to make type checking easier to implement.

There is one downside of the approach taken here: there will be
notably more calls to check definitional equality of types, and hence
checking time might be a bit slower.  This is because when checking
(for instance) an application of $f : A \to B$ to $a : A'$, we must
check that $A$ and $A'$ are definitionally equal.  In bidirectional
checking, we synthesize the type $A \to B$ for $f$ , and then check
the argument $a$ against $A$.  The latter check will generally
decompose $A$, but could be quite a bit cheaper than testing
definitional equality of $A$ with some other (possibly identical!)
type.

\subsection{A few wrinkles}

There are a few places where the system below differs from what we
have developed so far in papers and/or our implementation (which is
usually ahead of the papers):

\begin{enumerate}
  \item The system below uses Curry-style type-level
    $\lambda$-abstractions (so no type with the bound variable), while
    Cedille as we have developed it so far uses Church-style ones (a
    type for the bound variable is an essential part of the
    expression).  We believe we know how to support Curry-style
    type-level $\lambda$-abstractions now; we did not know how to
    give the semantics for them before.

  \item Below we use $\beta\eta$-reduction, while our papers
    just talk about $\beta$-reduction.  The use of $\eta$ seems
    somewhat essential to some of what we are doing, so this is an
    important addition.  We think it should be straightforward to
    add $\eta$ to our semantics.

  \item The system below omits lifting, which is needed in our
    approach for large eliminations.  We have an approach to lifting
    in~\cite{stump17a}, but the way it is done there is somewhat
    complicated, and furthermore we believe that it is incomplete: there
    are some lifting terms (with free variables) that should be semantically
    equivalent but which are not definitionally equal.  So we are
    anticipating needing to rework lifting, which will likely not be
    an easy theoretical job (though probably not complicating the implementation
    of Core Cedille very much).  So I have left out lifting for now.

  \item Note that full Cedille currently does not allow $\forall$ at the kind level,
    and hence does not support erased $\lambda$-abstractions at the type level.  While
    perhaps we know enough now to support this feature semantically, we have not
    seen any need for it and hence omit it below.

    \end{enumerate}

\section{Syntax}

Figure~\ref{fig:syntax} gives the syntax for terms of Core Cedille, in
the style of pure type systems~\cite{B92}: we have just one syntactic
category of terms, and we rely on the type system to distinguish
terms, types, and kinds (and a sole superkind $\Box$).  The constructs
are listed with some short comments to give a hint to the reader of
what they mean; the typing rules below should clarify the meaning
further.  Also, the constructs are listed in the figure in order from
highest precedence (most aggressively binding arguments) to lowest.
This ordering is a little different in a couple places from current
Cedille, but we have changed our parser recently and so may redo our
operator order to match this.  I believe this order is a little more
natural than the one we are currently implementing.  By the way, we
also plan to add support to our implementation of (full) Cedille for a
couple forms below that are currently not legal Cedille syntax
(dependent introductions, the particular form of $\rho$-terms).  In
the typing rules below, we will also use this definition:
\[
\textit{sorts}\ \mathcal{S} := \{ \star , \Box \}
\]
\noindent We will use $s$ as a metavariable ranging over $\mathcal{S}$.

\begin{figure}
\[
\begin{array}{llll}
  \textit{variables}\ x & \ & \ &\ \\
  \textit{annotated terms}\ t & ::= & x\ & \textnormal{use of a term variable} \\
  \ &\ & \star&\textnormal{kind for types}\\
  \ &\ & \Box&\textnormal{sole superkind} \\
  \ &\ & t.1 & \textnormal{project first view of a dependent intersection}\\
  \ &\ & t.2 & \textnormal{project second view of a dependent intersection}\\
  \ &\ & \beta\{t\} &\textnormal{proof of a true equation, where the proof erases to }t  \\
  \ &\ & \varsigma\ t &\textnormal{symmetry of equality}  \\
  \ &\ & t\ t' & \textnormal{application of term to term}\\
  \ &\ & t\ \mhyph t' & \textnormal{application of a term to an erased argument}\\
  \ &\ & \rho\ t\ @\ x.t'\ \mhyph\ t''&\textnormal{equality elimination by type-guided rewriting}  \\ 
  \ &\ & \abs{\forall}{x}{t}{t'}\ &\textnormal{implicit product (quantify over erased argument)} \\
  \ &\ & \abs{\Pi}{x}{t}{t'}\ &\textnormal{explicit product (usual }\Pi\textnormal{-type)} \\
  \ &\ & \abs{\iota}{x}{T}{T'}\ &\textnormal{dependent intersection} \\
  \ &\ & \abs{\lambda}{x}{t}{t'} & \textnormal{usual }\lambda\textnormal{-abstraction} \\
  \ &\ & \abs{\Lambda}{x}{t}{t'} & \textnormal{erased }\lambda\textnormal{-abstraction}\\
  \ &\ & [ t , t' \ @\ x.t'' ] & \textnormal{introduce dependent intersection}\\
  \ &\ & \phi\ t\ \mhyph\ t'\ \{t''\} & \textnormal{when }t\textnormal{ proves }t'\textnormal{ and }t''
                                   \textnormal{ are equal, erase to }t'' \\
  \ & \ & [ x = t : t' ]\ \mhyph\ t'' & \textnormal{let }x\textnormal{ equal }t\textnormal{ of type }t'\textnormal{ in }t'' \\
  \ &\ & \{ t \simeq t' \} &\textnormal{equality between terms}
\end{array}
  \]
  \caption{Sytax for core Cedille}
  \label{fig:syntax}
\end{figure}

\section{Erasure}

When comparing terms for definitional equality, the rules below will
compare erased terms, using the erasure function defined in
Figure~\ref{fig:eraser}.

\begin{figure}
  \[
  \begin{array}{lll}
    |x| & = & x \\
    |\star| & = & \star \\
    |\Box| & = & \Box \\
    |t.1| & = & |t| \\
    |t.2| & = & |t| \\
    |\beta\{t\}| & = & |t|\\
    |\varsigma\ t| & = & |t| \\
    |t\ t'| & = & |t|\ |t'| \\
    |t\ \mhyph t'| & = & |t| \\
    |\rho\ t\ @\ x.t'\ \mhyph\ t''| & = & |t''| \\
    |\abs{\forall}{x}{t}{t'}| & = & \abs{\forall}{x}{|t|}{|t'|}\\
    |\abs{\Pi}{x}{t}{t'}| & = & \abs{\Pi}{x}{|t|}{|t'|}\\
    |\abs{\iota}{x}{T}{T'}| & = & \abs{\iota}{x}{|t|}{|t'|} \\
    |\abs{\lambda}{x}{t}{t'}| & = &  \absu{\lambda}{x}{|t'|} \\
    |\abs{\Lambda}{x}{t}{t'}| & = &  |t'| \\ 
    |[ t , t' \ @\ x.t'' ]| & = & |t| \\ 
    |\phi\ t\ \mhyph\ t'\ \{t''\}| & = & |t''| \\
    |[ x = t : t' ]\ \mhyph\ t'' | & = & (\absu{\lambda}{x}{|t''|})\ |t|\\
    |\{ t \simeq t' \}|| & = & \{ |t| \simeq |t'| \}
  \end{array}
  \]
  \caption{Erasure for annotated terms}
  \label{fig:eraser}
\end{figure}

\section{Typing}

The type-checking algorithm for Core Cedille is presented as ``almost'' algorithmic typing rules (more on this shortly),
in Figure~\ref{fig:rules}.  The rules are almost algorithmic because we must understand a few conventions for applying the rules:

\begin{itemize}
\item The rules are applied bottom-up, and a judgment $\Gamma\vdash t : t'$ represents a call to compute
  a type $t'$ given a context $\Gamma$ and a term $t$.
\item Some of the premises of the rules state that a computed type should have a certain form (e.g., be a $\Pi$-type).  It may happen, though, that the computed type $\beta$-reduces to a $\Pi$-type but is not literally one.  To handle this case, one should head-normalize the computed types for premises that require a specific form.
\item Where two premises use the same meta-variable, one must check definitional equality of the terms in question (for example, the domain-part of a $\Pi$-type and the type of an argument, in the application typing rule).
  The definitional equality relation, which we can denote $\Gamma \vdash t =_{\beta\eta} t'$,
  is the usual relation of $\beta\eta$-equivalence of
terms in pure untyped lambda calculus, extended congruentially for the
typing constructs $\forall$, $\Pi$, $\iota$, and $\simeq$, and also
extended to make use of let-definitions $x = t : t'$ contained in
$\Gamma$ (by replacing $x$ with $t$ when checking definitional
equality).  This same mechanism can be used for global definitions, as
well.
\end{itemize}

\begin{figure}
  \[
  \begin{array}{lll}
    \infer{\Gamma\vdash \star : \Box}{\ } &
    \infer{\Gamma\vdash\abs{\Pi}{x}{t}{t'} : s}{\Gamma \vdash t : \star \quad \Gamma,x:t\vdash t' : s} &
    \infer{\Gamma\vdash\abs{\forall}{x}{t}{t'} : \star}{\Gamma \vdash t : s \quad \Gamma,x:t\vdash t' : \star} 
    \\ \\
    \infer{\Gamma, x : t \vdash x : t}{\ } &
    \infer{\Gamma\vdash t\ t' : [t'/x]t_2}{\Gamma\vdash t : \abs{\Pi}{x}{t_1}{t_2} & \Gamma\vdash t : t_1} &
    \infer{\Gamma\vdash\abs{\iota}{x}{t}{t'} : \star}{\Gamma \vdash t : \star \quad \Gamma,x:t\vdash t' : \star} 
    \\ \\
    \infer{\Gamma\vdash t\ \mhyph t' : [t'/x]t_2}{\Gamma\vdash t : \abs{\forall}{x}{t_1}{t_2} & \Gamma\vdash t : t_1} &
    \infer{\Gamma\vdash\abs{\lambda}{x}{t}{t'} : \abs{\Pi}{x}{t}{t''}}{\Gamma,x:t\vdash t':t'' \quad \Gamma \vdash \abs{\Pi}{x}{t}{t''} : s}  &
    \infer{\Gamma\vdash\abs{\Lambda}{x}{t}{t'} : \abs{\forall}{x}{t}{t''}}{\Gamma,x:t\vdash t':t'' \quad \Gamma \vdash \abs{\forall}{x}{t}{t''} : s} \\ \\
    \infer{\Gamma\vdash t.1 : t_1}{\Gamma\vdash t : \abs{\iota}{x}{t_1}{t_2}} &
    \infer{\Gamma\vdash t.2 : [t/x]t_1}{\Gamma\vdash t : \abs{\iota}{x}{t_1}{t_2}} &    
    \infer{\Gamma\vdash [t,t'\ @\ x.t_2] : \abs{\iota}{x}{t_1}{t_2}}{\Gamma\vdash t : t_1 \quad \Gamma\vdash t' : [t/x]t_2}
    \\ \\
    \infer{\Gamma\vdash t \simeq t' : \star}{\textit{FV}(t\ t')\subseteq \textit{dom}(\Gamma) } &
    \infer{\Gamma\vdash \beta\{t\} : \{ t' \simeq t'\}}{\ } &
    \infer{\Gamma\vdash \rho\ t\ @\ x.t'\ \mhyph\ t'' : [t_2/x]t'}{\Gamma \vdash t : \{ t_1 \simeq t_2 \} \quad \Gamma\vdash t'' : [t_1/x]t'}
    \\ \\
    \infer{\Gamma\vdash \varsigma\ t : \{ t_2 \simeq t_1\}}{\Gamma\vdash t : \{t_1 \simeq t_2\}} &
    \infer{\Gamma\vdash \phi\ t\ \mhyph\ t_1\ \{t_2\} : t'}{\Gamma \vdash t : \{t_1 \simeq t_2\} \quad \Gamma\vdash t_1 : t'} &
    \infer{\Gamma\vdash [ x = t_1 : t' ]\ \mhyph\ t_2 : t''}{\Gamma\vdash t_1 : t' \quad \Gamma, x = t_1 : t' \vdash t_2 : t''} 
  \end{array}
  \]
  \caption{Type-checking rules for Core Cedille}
  \label{fig:rules}
\end{figure}

\section{Conclusion}

For further examples and intuition on how some of these constructs are
used, personal consultation and comparison with published, submitted,
and planned papers will be necessary.  Also, please note that I wrote
out these rules in a short period of time and so it is possible there
are typos (so please ask me if anything seems weird).

\bibliographystyle{plain}
\bibliography{biblio}

\end{document}

### 2018-02-14 docs/cedille-core/spec.tex
some tweaks
git-svn-id: https://svn.divms.uiowa.edu/repos/clc/projects/lambda-encodings/implementations/cedille@29278 bd178f5f-7bea-4014-844f-495ca19b8805

@@
 \section{Syntax}
+\label{sec:syntax}
 
@@
 \begin{array}{llll}
-  \textit{variables}\ x & \ & \ &\ \\
+  \textit{term variables}\ u & \ & \ &\ \\
+  \textit{type variables}\ X & \ & \ &\ \\
+  \textit{kind variables}\ k & \ & \ &\ \\
+  \textit{variables}\ x & ::= & u\ |\ X\ |\ k &\ \\  
+  \textit{pure terms}\ p & ::= & u\ |\ p\ p'\ |\ \absu{\lambda}{u}{p} &\ \\
   \textit{annotated terms}\ t & ::= & x\ & \textnormal{use of a term variable} \\
@@
   \ & \ & [ x = t : t' ]\ \mhyph\ t'' & \textnormal{let }x\textnormal{ equal }t\textnormal{ of type }t'\textnormal{ in }t'' \\
-  \ &\ & \{ t \simeq t' \} &\textnormal{equality between terms}
+  \ &\ & \{ p \simeq p' \} &\textnormal{equality between pure terms}
 \end{array}
@@
 \end{figure}
+
+\subsection{Syntactic check for being a term}
+\label{sec:termcheck}
+
+It is important that we only form equations between terms, because
+Cedille's semantics does not support forming equations except where
+both sides of the equation are terms.  Terms can be syntactically
+distinguished from types and kinds as long as we use different
+syntactic categories for term variables, type variables, and kind
+variables.  That is why the syntax of Figure~\ref{fig:syntax}
+distinguishes these.  The typing rule below for forming equations
+requires the sides to be pure (i.e., unannotated) terms.
 
@@
 well.
+\item Note that the formation rules for $\forall$- and $\Pi$-abstractions use
+  a premise $\textit{Var}(x,s)$ to check
+  that the variable $x$ is a legal form of variable given the kind $s$.  This
+  judgment is defined by these rules:
+  \[
+  \begin{array}{lll}
+    \infer{\textit{Var}(u,\star)}{\ } & \ \ \ & \infer{\textit{Var}(X,\Box)}{\ }
+  \end{array}
+  \]
+
 \end{itemize}
@@
     \infer{\Gamma\vdash \star : \Box}{\ } &
-    \infer{\Gamma\vdash\abs{\Pi}{x}{t}{t'} : s}{\Gamma \vdash t : \star \quad \Gamma,x:t\vdash t' : s} &
-    \infer{\Gamma\vdash\abs{\forall}{x}{t}{t'} : \star}{\Gamma \vdash t : s \quad \Gamma,x:t\vdash t' : \star} 
+    \infer{\Gamma\vdash\abs{\Pi}{x}{t}{t'} : s'}{\Gamma \vdash t : s \quad \Gamma,x:t\vdash t' : s' \quad \textit{Var}(x,s)} &
+    \infer{\Gamma\vdash\abs{\forall}{x}{t}{t'} : \star}{\Gamma \vdash t : s \quad \Gamma,x:t\vdash t' : \star \quad \textit{Var}(x,s)} 
     \\ \\
@@
     \infer{\Gamma\vdash t\ t' : [t'/x]t_2}{\Gamma\vdash t : \abs{\Pi}{x}{t_1}{t_2} & \Gamma\vdash t : t_1} &
-    \infer{\Gamma\vdash\abs{\iota}{x}{t}{t'} : \star}{\Gamma \vdash t : \star \quad \Gamma,x:t\vdash t' : \star} 
+    \infer{\Gamma\vdash\abs{\iota}{u}{t}{t'} : \star}{\Gamma \vdash t : \star \quad \Gamma,x:t\vdash t' : \star} 
     \\ \\
-    \infer{\Gamma\vdash t\ \mhyph t' : [t'/x]t_2}{\Gamma\vdash t : \abs{\forall}{x}{t_1}{t_2} & \Gamma\vdash t : t_1} &
+    \infer{\Gamma\vdash t.2 : [t/x]t_1}{\Gamma\vdash t : \abs{\iota}{x}{t_1}{t_2}} &    
     \infer{\Gamma\vdash\abs{\lambda}{x}{t}{t'} : \abs{\Pi}{x}{t}{t''}}{\Gamma,x:t\vdash t':t'' \quad \Gamma \vdash \abs{\Pi}{x}{t}{t''} : s}  &
@@
     \infer{\Gamma\vdash t.1 : t_1}{\Gamma\vdash t : \abs{\iota}{x}{t_1}{t_2}} &
-    \infer{\Gamma\vdash t.2 : [t/x]t_1}{\Gamma\vdash t : \abs{\iota}{x}{t_1}{t_2}} &    
-    \infer{\Gamma\vdash [t,t'\ @\ x.t_2] : \abs{\iota}{x}{t_1}{t_2}}{\Gamma\vdash t : t_1 \quad \Gamma\vdash t' : [t/x]t_2}
+    \infer{\Gamma\vdash t\ \mhyph t' : [t'/x]t_2}{\Gamma\vdash t : \abs{\forall}{x}{t_1}{t_2} & \Gamma\vdash t : t_1} &
+    \infer{\Gamma\vdash [t,t'\ @\ x.t_2] : \abs{\iota}{x}{t_1}{t_2}}{\Gamma\vdash t : t_1 \quad \Gamma\vdash t' : [t/x]t_2 \quad \Gamma \vdash \abs{\iota}{x}{t_1}{t_2} : \star}
     \\ \\
-    \infer{\Gamma\vdash t \simeq t' : \star}{\textit{FV}(t\ t')\subseteq \textit{dom}(\Gamma) } &
-    \infer{\Gamma\vdash \beta\{t\} : \{ t' \simeq t'\}}{\ } &
+    \infer{\Gamma\vdash \beta\{t\} : \{ t' \simeq t'\}}{\Gamma\vdash \{ t' \simeq t' \} : \star } &
+    \infer{\Gamma\vdash \{ p \simeq p'\} : \star}{\textit{FV}(p\ p')\subseteq \textit{dom}(\Gamma)} &
     \infer{\Gamma\vdash \rho\ t\ @\ x.t'\ \mhyph\ t'' : [t_2/x]t'}{\Gamma \vdash t : \{ t_1 \simeq t_2 \} \quad \Gamma\vdash t'' : [t_1/x]t'}
@@
     \infer{\Gamma\vdash \phi\ t\ \mhyph\ t_1\ \{t_2\} : t'}{\Gamma \vdash t : \{t_1 \simeq t_2\} \quad \Gamma\vdash t_1 : t'} &
-    \infer{\Gamma\vdash [ x = t_1 : t' ]\ \mhyph\ t_2 : t''}{\Gamma\vdash t_1 : t' \quad \Gamma, x = t_1 : t' \vdash t_2 : t''} 
+    \infer{\Gamma\vdash [ k = t_1 : \Box ]\ \mhyph\ t_2 : t''}{\Gamma\vdash t_1 : \Box \quad
+                                                               \Gamma, k = t_1 : t' \vdash t_2 : t''} 
+    \\ \\
+    \multicolumn{2}{l}{
+    \infer{\Gamma\vdash [ x = t_1 : t' ]\ \mhyph\ t_2 : t''}{\Gamma\vdash t_1 : t' \quad \Gamma \vdash t' : s \quad \textit{Var}(x,s) \quad
+      \Gamma, x = t_1 : t' \vdash t_2 : t''} } &
+    \     
   \end{array}
@@
 
+\appendix 
+\section{Change log}
+
+\begin{itemize}
+\item Feb. 14, 2018:
+  \begin{itemize}
+  \item added another premise to the rule for introducing dependent intersections, to
+    make sure that the dependent intersection itself if typable.
+  \item Equations are now required to be only between pure
+    (unannotated) terms.  This is a syntactic check, which is now
+    described in Section~\ref{sec:termcheck}.  It requires different
+    lexical classes for term, type, and kind variables.  In turn, this
+    requires that the typing rules check, with a new helper judgment
+    $\textit{Var}(x,s)$, that we are using the proper class of
+    variable when we introduce a local variable.
+
+  \end{itemize}
+\end{itemize}
+
 \end{document}

### 2018-03-26 docs/cedille-core/spec.tex
about to move the spec for Cedille core to github, so pushing some old edits
git-svn-id: https://svn.divms.uiowa.edu/repos/clc/projects/lambda-encodings/implementations/cedille@29940 bd178f5f-7bea-4014-844f-495ca19b8805

@@
   \textit{pure terms}\ p & ::= & u\ |\ p\ p'\ |\ \absu{\lambda}{u}{p} &\ \\
-  \textit{annotated terms}\ t & ::= & x\ & \textnormal{use of a term variable} \\
+  \textit{annotated terms}\ t & ::= & x\ & \textnormal{use of a variable} \\
   \ &\ & \star&\textnormal{kind for types}\\
@@
   \ &\ & t.2 & \textnormal{project second view of a dependent intersection}\\
-  \ &\ & \beta\{t\} &\textnormal{proof of a true equation, where the proof erases to }t  \\
+  \ &\ & \beta\ t'\{t\} &\textnormal{proof of }t' \simeq t'\textnormal{, where the proof erases to }t  \\
   \ &\ & \varsigma\ t &\textnormal{symmetry of equality}  \\
@@
     |t.2| & = & |t| \\
-    |\beta\{t\}| & = & |t|\\
+    |\beta\ t'\{t\}| & = & |t|\\
     |\varsigma\ t| & = & |t| \\
@@
     |\abs{\iota}{x}{T}{T'}| & = & \abs{\iota}{x}{|t|}{|t'|} \\
-    |\abs{\lambda}{x}{t}{t'}| & = &  \absu{\lambda}{x}{|t'|} \\
+    |\abs{\lambda}{u}{t}{t'}| & = &  \absu{\lambda}{u}{|t'|} \\
+    |\abs{\lambda}{X}{t}{t'}| & = &  \abs{\lambda}{X}{|t|}{|t'|} \\
     |\abs{\Lambda}{x}{t}{t'}| & = &  |t'| \\ 
@@
     \\ \\
-    \infer{\Gamma\vdash t.2 : [t/x]t_1}{\Gamma\vdash t : \abs{\iota}{x}{t_1}{t_2}} &    
+    \infer{\Gamma\vdash t.2 : [t/x]t_2}{\Gamma\vdash t : \abs{\iota}{x}{t_1}{t_2}} &    
     \infer{\Gamma\vdash\abs{\lambda}{x}{t}{t'} : \abs{\Pi}{x}{t}{t''}}{\Gamma,x:t\vdash t':t'' \quad \Gamma \vdash \abs{\Pi}{x}{t}{t''} : s}  &
-    \infer{\Gamma\vdash\abs{\Lambda}{x}{t}{t'} : \abs{\forall}{x}{t}{t''}}{\Gamma,x:t\vdash t':t'' \quad \Gamma \vdash \abs{\forall}{x}{t}{t''} : s} \\ \\
+    \infer{\Gamma\vdash \varsigma\ t : \{ t_2 \simeq t_1\}}{\Gamma\vdash t : \{t_1 \simeq t_2\}}
+
+\\ \\
     \infer{\Gamma\vdash t.1 : t_1}{\Gamma\vdash t : \abs{\iota}{x}{t_1}{t_2}} &
     \infer{\Gamma\vdash t\ \mhyph t' : [t'/x]t_2}{\Gamma\vdash t : \abs{\forall}{x}{t_1}{t_2} & \Gamma\vdash t : t_1} &
-    \infer{\Gamma\vdash [t,t'\ @\ x.t_2] : \abs{\iota}{x}{t_1}{t_2}}{\Gamma\vdash t : t_1 \quad \Gamma\vdash t' : [t/x]t_2 \quad \Gamma \vdash \abs{\iota}{x}{t_1}{t_2} : \star}
-    \\ \\
-    \infer{\Gamma\vdash \beta\{t\} : \{ t' \simeq t'\}}{\Gamma\vdash \{ t' \simeq t' \} : \star } &
-    \infer{\Gamma\vdash \{ p \simeq p'\} : \star}{\textit{FV}(p\ p')\subseteq \textit{dom}(\Gamma)} &
     \infer{\Gamma\vdash \rho\ t\ @\ x.t'\ \mhyph\ t'' : [t_2/x]t'}{\Gamma \vdash t : \{ t_1 \simeq t_2 \} \quad \Gamma\vdash t'' : [t_1/x]t'}
     \\ \\
-    \infer{\Gamma\vdash \varsigma\ t : \{ t_2 \simeq t_1\}}{\Gamma\vdash t : \{t_1 \simeq t_2\}} &
-    \infer{\Gamma\vdash \phi\ t\ \mhyph\ t_1\ \{t_2\} : t'}{\Gamma \vdash t : \{t_1 \simeq t_2\} \quad \Gamma\vdash t_1 : t'} &
+    \infer{\Gamma\vdash \beta\ t'\{t\} : \{ t' \simeq t'\}}{\Gamma\vdash \{ t' \simeq t' \} : \star } &
+    \infer{\Gamma\vdash \{ p \simeq p'\} : \star}{\textit{FV}(p\ p')\subseteq \textit{dom}(\Gamma)} &
      \infer{\Gamma\vdash [ k = t_1 : \Box ]\ \mhyph\ t_2 : t''}{\Gamma\vdash t_1 : \Box \quad
@@
     \multicolumn{2}{l}{
-    \infer{\Gamma\vdash [ x = t_1 : t' ]\ \mhyph\ t_2 : t''}{\Gamma\vdash t_1 : t' \quad \Gamma \vdash t' : s \quad \textit{Var}(x,s) \quad
+    \infer{\Gamma\vdash [t,t'\ @\ x.t_2] : \abs{\iota}{x}{t_1}{t_2}}{\Gamma\vdash t : t_1 \quad \Gamma\vdash t' : [t/x]t_2 \quad \Gamma \vdash \abs{\iota}{x}{t_1}{t_2} : \star \quad |t| = |t'|}} &  
+  \infer{\Gamma\vdash \phi\ t\ \mhyph\ t_1\ \{t_2\} : t'}{\Gamma \vdash t : \{t_1 \simeq t_2\} \quad \Gamma\vdash t_1 : t'}  
+    \\ \\
+
+    \multicolumn{2}{l}{
+    \infer{\Gamma\vdash [ x = t_1 ]\ \mhyph\ t_2 : t''}{\Gamma\vdash t_1 : t' \quad \Gamma \vdash t' : s \quad \textit{Var}(x,s) \quad
       \Gamma, x = t_1 : t' \vdash t_2 : t''} } &
    \ 
+    \\ \\
+    \multicolumn{3}{l}{
+      \infer{\Gamma\vdash\abs{\Lambda}{x}{t}{t'} : \abs{\forall}{x}{t}{t''}}{\Gamma,x:t\vdash t':t'' \quad x\not\in\textit{FV}(|t'|)\quad \Gamma \vdash \abs{\forall}{x}{t}{t''} : s}}
   \end{array}
@@
   \end{itemize}
+\item Feb. 14, 2018, pm:
+  \begin{itemize}
+  \item Corrected glaring omission of the requirement for typing $[t,t'\ @\ x.T]$, that $|t| = |t'|$.  It is fine if one implements
+    this check instead as $|t| =_{\beta\eta} |t'|$, simply to save writing a separate function for testing $\alpha$-equivalence. 
+  \end{itemize}
+\item Feb. 16:
+  \begin{itemize}
+  \item removed type annotation on definiens in a let-expression.
+  \item Changed syntax $\beta\{t\}$ to $\beta\ t'\{t\}$.
+    \end{itemize}
+\item Feb. 20: fix silly typo noticed by Victor, in the rule for $t.2$.  Also, erase $\lambda x$ differently from $\lambda X$.
+
+\item Feb. 21: fixed rule for $\Lambda$-terms: we must check that the bound variable is not free in the erasure of the body.
 \end{itemize}

### 2018-03-26 docs/src/compile-docs.sh
New beta-reduction buffer commands, progress updates from backend, updated docs
git-svn-id: https://svn.divms.uiowa.edu/repos/clc/projects/lambda-encodings/implementations/cedille@29950 bd178f5f-7bea-4014-844f-495ca19b8805

@@ -1,7 +1,7 @@
 #!/bin/bash
 # to be run in docs/src/ because of the relative path
 
-makeinfo -o ../info/cedille-info-main.info cedille-main.texi
+makeinfo -o ../info/cedille-info-main.info cedille-info-main.texi
 
 # for later, generating more forms of the documentation
 # makeinfo -o ../html --html cedille-main.texi

### 2018-06-09 core/CedilleCore.hs
Fixed several elaboration corner-case bugs, some issues with Cedille Core substitution, a bug with rewriting (qualification hid it except for some very rare cases), and type-to-term applications inside type-guided rewrite (using '@') being printed without parentheses when they should have been: before, types like 'Q (S n)' were being printed 'Q S n'. Elaboration is almost fully functional; the only remaining issue to my knowledge in elaborating any error-free file without lifting is that rewrites can make types incorrect. This is the same problem we have with kind-checking meta variables.
git-

@@ -60,28 +60,3 @@ checkFile c fs fp =
 main = getArgs >>= \ as -> case as of
   (fp : []) -> checkFile emptyCtxt emptyTrie fp >> return ()
   _ -> putStrLn "Run with the name of a single file"
-
-{-
---splitStr :: String -> String -> [String]
-splitStr s [] = s : []
-splitStr s (' ' : cs) = s : splitStr "" cs
-splitStr s (c : cs) = splitStr (s ++ (c : [])) cs
-
---joinStr :: [String] -> String
-joinStr [] = ""
-joinStr (s : []) = s
-joinStr (s : ss) = s ++ " " ++ joinStr ss
-
---handleInput :: Ctxt -> [String] -> IO ()
-handleInput c ("check" : fp : []) = checkFile emptyCtxt emptyTrie fp >>= mainWithCtxt . fst
-handleInput c ("lookup" : "term" : "hnf" : v : []) = putStrLn (maybe "Term not defined" show (ctxtLookupTermVar c v)) >> mainWithCtxt c
-handleInput c ("lookup" : "type" : "hnf" : v : []) = putStrLn (maybe "Type not defined" show (ctxtLookupTypeVar c v)) >> mainWithCtxt c
-handleInput c ("lookup" : "term" : "type" : v : []) = putStrLn (maybe "Term not defined" show (ctxtLookupVarType c v)) >> mainWithCtxt c
-handleInput c ("lookup" : "type" : "kind" : v : []) = putStrLn (maybe "Type not defined" show (ctxtLookupVarKind c v)) >> mainWithCtxt c
-handleInput c ("quit" : []) = return ()
-handleInput c _ = putStrLn "Unknown command" >> mainWithCtxt c
-
-mainWithCtxt c = getLine >>= handleInput c . splitStr ""
-
-main = mainWithCtxt emptyCtxt
--}

### 2018-06-09 core/Check.hs

@@ -61,7 +61,7 @@ synthTerm' c (TmLambda v tp tm) =
 synthTerm' c (TmLambdaE v tk tm) =
   errIfCtxtBinds c v >>
   errIf (freeInTerm v (eraseTerm tm))
-    ("Implicit variable occurs free in its body: " ++ v) >>
+    ("Implicit variable occurs free in its body") >>
   synthTpKd c tk >>
   let tk' = hnfeTpKd c tk in
   synthTerm (ctxtDefTpKd c v tk') tm >>= Right . TpAll v tk'
@@ -85,12 +85,15 @@ synthTerm' c (TmAppTp tm tp) =
     _ -> err "Expected the head of an application to synthesize a kind-forall type"
 synthTerm' c (TmIota tm tm' v tp) =
   errIfCtxtBinds c v >>
-  errIfNot (convTerm c (eraseTerm tm) (eraseTerm tm')) ("in an iota pair, " ++ show (hnfeTerm c tm) ++ " != " ++ show (hnfeTerm c tm')) >>
   synthTerm c tm >>= \ ltp ->
   synthTerm c tm' >>= \ rtp ->
   checkType (ctxtDefTerm c v (Nothing, Just ltp)) tp >>
-  let tp' = eraseType tp in
-  errIfNot (convType (ctxtInternalDef c v (Left (hnfeTerm c tm))) tp' rtp) "Inconvertible types in an iota pair" >>
+  let tp' = eraseType tp
+      tm'' = eraseTerm tm
+      tm''' = eraseTerm tm'
+      htm'' = hnfTerm c tm'' in
+  errIfNot (convTerm' c htm'' (hnfTerm c tm''')) ("In an iota pair, " ++ show htm'' ++ " != " ++ show (hnfTerm c tm''')) >>
+  errIfNot (convType (ctxtInternalDef c v (Left htm'')) tp' rtp) "Inconvertible types in an iota pair" >>
   Right (TpIota v ltp tp')
 synthTerm' c (IotaProj1 tm) = synthTerm c tm >>= \ tp -> case hnfType c tp of
   (TpIota v tp tp') -> Right (hnfType c tp)
@@ -106,8 +109,8 @@ synthTerm' c (Sigma tm) = synthTerm c tm >>= \ tp -> case tp of
   (TpEq ltm rtm) -> Right (TpEq rtm ltm)
   _ -> err "Expected to synthesize an equational type from the body of a sigma term"
 synthTerm' c (Delta tp tm) =
-  doRename' c "x" $ \ x ->
[…]

### 2018-06-09 core/Ctxt.hs

@@ -20,16 +20,20 @@ emptyTrie = Node Nothing []
 trieLookup (Node a ts) (c : cs) = calLookup ts c >>= \ t -> trieLookup t cs
 trieLookup (Node a ts) [] = a
 
-trieInsert (Node a ts) (c : cs) a' = maybe
-  (Node a ((c, trieInsert emptyTrie cs a') : ts))
-  (\ t -> Node a (calInsert ts c (trieInsert t cs a')))
+trieInsert' (Node a ts) (c : cs) a' = maybe
+  (Node a ((c, trieInsert' emptyTrie cs a') : ts))
+  (\ t -> Node a (calInsert ts c (trieInsert' t cs a')))
   (calLookup ts c)
-trieInsert (Node a ts) [] a' = Node (Just a') ts
+trieInsert' (Node a ts) [] a' = Node a' ts
+
+trieInsert t s a = trieInsert' t s (Just a)
 
 trieMember t s = maybe False (\ _ -> True) (trieLookup t s)
 
 trieAt t "" = t
 trieAt (Node a ts) (c : cs) = maybe emptyTrie (\ t -> trieAt t cs) (calLookup ts c)
+
+trieRemove t s = trieInsert' t s Nothing
 
 trieMappings = h [] "" where
   h acc path (Node Nothing cs) = h' path acc cs
@@ -55,6 +59,7 @@ varMapRep m k = case trieLookup m k of
   Nothing -> k
 
 ctxtClearExternals (Ctxt es is vs rs) = Ctxt emptyTrie is vs rs
+ctxtOnlySubst (Ctxt es is vs rs) = Ctxt emptyTrie emptyTrie vs rs
 ctxtShowAll (Ctxt es is vs _) rs = Ctxt es is vs rs
 ctxtShown (Ctxt es is vs rs) = rs
 
@@ -113,11 +118,11 @@ ctxtLookupInternalType (Ctxt es is vs rs) v = case trieLookup is (varMapRep vs v
   
 --ctxtDef :: Ctxt -> Var -> ExternalDef -> Ctxt
 ctxtDef c "_" d = c
-ctxtDef (Ctxt es is vs rs) v d = Ctxt (trieInsert es v d) is vs (trieInsert rs v ())
+ctxtDef (Ctxt es is vs rs) v d = Ctxt (trieInsert es v d) is (trieRemove vs v) (trieInsert rs v ())
 
[…]

### 2018-06-09 core/Norm.hs

@@ -36,24 +36,24 @@ eraseKind (KdPi v tk kd) = KdPi v (eraseTpKd tk) (eraseKind kd)
 eraseTpKd (TpKdTp tp) = TpKdTp (eraseType tp)
 eraseTpKd (TpKdKd kd) = TpKdKd (eraseKind kd)
 
-tkIsType (TpKdTp _) = True
-tkIsType (TpKdKd _) = False
+tpKdIsType (TpKdTp _) = True
+tpKdIsType (TpKdKd _) = False
 
 --substTerm :: Ctxt -> PureTerm -> PureTerm
-substTerm c (PureVar v) = maybe (PureVar (ctxtRep c v)) id (ctxtLookupInternalTerm c v)
+substTerm c (PureVar v) = maybe (PureVar (ctxtRep c v)) (substTerm (ctxtOnlySubst c)) (ctxtLookupInternalTerm c v)
 substTerm c (PureApp tm tm') = PureApp (substTerm c tm) (substTerm c tm')
-substTerm c (PureLambda v tm) = doRename' c v $ \ v' ->
+substTerm c (PureLambda v tm) = freshVar c v $ \ v' ->
   PureLambda v' (substTerm (ctxtRename c v v') tm)
 {-
 --substType :: Ctxt -> PureType -> PureType
 substType c (TpVar v) = maybe (TpVar (ctxtRep c v)) id (ctxtLookupTypeVar c v)
-substType c (TpLambda v tk tp) = doRename' c v $ \ v' ->
+substType c (TpLambda v tk tp) = freshVar c v $ \ v' ->
   TpLambda v' (substTpKd c tk) (substType (ctxtRename c v v') tp)
-substType c (TpAll v tk tp) = doRename' c v $ \ v' ->
+substType c (TpAll v tk tp) = freshVar c v $ \ v' ->
   TpAll v' (substTpKd c tk) (substType (ctxtRename c v v') tp)
-substType c (TpPi v tp tp') = doRename' c v $ \ v' ->
+substType c (TpPi v tp tp') = freshVar c v $ \ v' ->
   TpPi v' (substType c tp) (substType (ctxtRename c v v') tp')
-substType c (TpIota v tp tp') = doRename' c v $ \ v' ->
+substType c (TpIota v tp tp') = freshVar c v $ \ v' ->
   TpIota v' (substType c tp) (substType (ctxtRename c v v') tp')
[…]

### 2018-06-09 core/ToString.hs

@@ -52,7 +52,7 @@ termToString b (IotaProj2 tm) = termToString True tm ++ " . 2"
 termToString b (Beta pt pt') = "β < " ++ pureTermToString True pt ++ " > { " ++ show pt' ++ " }"
 termToString b (Sigma tm) = "ς " ++ termToString True tm
 termToString b (Delta tp tm) = "δ " ++ typeToString True tp ++ " - " ++ termToString True tm
-termToString b (Rho tm v tp tm') = "ρ " ++ termToString True tm ++ " @ " ++ v ++ " " ++ typeToString True tp ++ " - " ++ show tm'
+termToString b (Rho tm v tp tm') = "ρ " ++ termToString True tm ++ " @ " ++ v ++ " . " ++ typeToString True tp ++ " - " ++ show tm'
 termToString b (Phi tm tm' tm'') = "φ " ++ termToString True tm ++ " - " ++ termToString True tm' ++ " { " ++ show tm'' ++ " }"
 
 typeToString b (TpVar v) = v

### 2018-06-12 src/CedilleTypes.hs
Internal support for erased import term arguments.
git-svn-id: https://svn.divms.uiowa.edu/repos/clc/projects/lambda-encodings/implementations/cedille@30777 bd178f5f-7bea-4014-844f-495ca19b8805

@@ -13,7 +13,7 @@ type Qkvar = Text
 type Kvar = Text
 type PosInfo = Text
 
-data Arg = TermArg Term | TypeArg Type
+data Arg = TermArg MaybeErased Term | TypeArg Type
      deriving (Show,Eq)
 
 data Args = ArgsCons Arg Args | ArgsNil
@@ -37,7 +37,7 @@ data Cmds =
       deriving (Show,Eq)
 
 data Decl =
-     Decl PosInfo PosInfo Bvar Tk PosInfo
+     Decl PosInfo PosInfo MaybeErased Bvar Tk PosInfo
      deriving (Show,Eq)
 
 data DefTermOrType =

### 2018-06-12 docs/semantics/paper.tex
adding document with semantics of Cedille
git-svn-id: https://svn.divms.uiowa.edu/repos/clc/projects/lambda-encodings/implementations/cedille@30779 bd178f5f-7bea-4014-844f-495ca19b8805

\documentclass{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{url}
\usepackage{comment}
\usepackage{proof}
\usepackage{stmaryrd}
\usepackage{parskip}
\usepackage{fullpage}

\newcommand{\choice}[0]{\zeta}
\newcommand{\elcap}[0]{\cap}
\newcommand{\abs}[4]{{#1}\, #2\! : \! #3.\, #4}
\newcommand{\absu}[3]{{#1}\, #2.\, #3}
\mathchardef\mhyph="2D % Define a "math hyphen"
\newcommand{\interp}[1]{\llbracket #1 \rrbracket} 
\newcommand{\tpcheck}[0]{\Leftarrow}
\newcommand{\tpsynth}[0]{\Rightarrow}
\newcommand{\cbe}[0]{c\beta\eta}
\newcommand{\utp}[0]{\mathcal{U}}
\newcommand{\startcase}[1]{\vspace{#1} \noindent\textbf{\underline{Case:}}}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{observation}[theorem]{Observation}

\bibliographystyle{plain}

\begin{document}

\title{Syntax and Semantics of Cedille}

\author{Aaron Stump \\
  Computer Science \\
  The University of Iowa \\
\texttt{aaron-stump@uiowa.edu}}

\date{}

\maketitle

\section{Introduction}

The type theory of Cedille is called the Calculus of Dependent Lambda
Eliminations (CDLE).  This document presents the version of CDLE as of
June 1, 2018.  We have made many changes from the first paper on CDLE~\cite{stump17},
mostly in the form of dropping constructs we discovered (to our
surprise) could be derived~\cite{stump18}. I have also omitted
\emph{lifting} -- a technique for large eliminations with lambda
encodings -- in this document's version of CDLE.  Some uses of lifting
can be simulated other ways within the system, though the limits of
this are still under investigation.  We also include a construct
$\delta$, for deriving a contradiction from a proof that
lambda-encoded true equals lambda-encoded false.  This also
compensates somewhat for the lack of lifting.

At a high level, CDLE is an extrinsic (i.e., Curry-style) type theory
extending the Calculus of Constructions with three additional
constructs, which allow deriving induction principles within the
theory, for inductive datatypes.  The goal is to support usual
idioms of dependently typed programming and proving as in Agda or
similar tools, but using pure lambda encodings for all data, and
requiring a much smaller core theory.

The current Cedille implementation of CDLE extends the system
described below with a number of features intended to make programming
in the system more convenient and with less redundancy.  These
features all compile away to a slightly simplified version of the
theory presented in this document, called Cedille Core, described
here: \url{https://github.com/astump/cedille-core-spec}.

\section{Classification Rules}

The classification rules are given in
Figures~\ref{fig:superknd},~\ref{fig:knd}, and~\ref{fig:tp}.  For
brevity, we take these figures as implicitly specifying the syntax of
kinds $\kappa$, types $T$, and annotated terms $t$; these may use term
variables $x$ and type variables $X$, which we assume come from
distinct sets.  So terms and types are syntactically distinguished.
The typing rules (Figure~\ref{fig:tp}) are
bidirectional~\cite{pierce+00}, while the kinding and superkinding
rules (Figure~\ref{fig:knd} and~\ref{fig:superknd}) are only
synthesizing.  We write $\Leftrightarrow$ to range over
$\{\tpcheck,\tpsynth\}$. We follow the syntax of our implementation
Cedille, which distinguishes application of a term or type $e$ to a
type ($e \cdot T$), from application to a term ($e\ t$), and
application to an erased term argument ($e\ \mhyph t$).  The rules are
intended, with a few points of nondeterminism, to be read bottom-up
(in a standard way; cf.~\cite{peytonjones07}) as an algorithm for
computing a classifier from a context and an expression ($\tpsynth$)
or checking an expression against a classifier in context
($\tpcheck$).

The classification rules refer to an erasure function, defined in
Figure~\ref{fig:eraser}.  The type theory is \emph{extrinsic} (aka,
Curry-style), and hence we only consider erasures $|t|$ of terms when
testing for $\beta\eta$-equivalence.  This is done by the conversion
relation $T\cong T'$, whose central rules are given in
Figure~\ref{fig:conv}.  That figure omits the various congruence rules
needed to equate bigger expressions by equating subexpressions.  The
main ideas of conversion shown in the figure are to use
$\beta$-equivalence at the type level, and $\beta\eta$-equivalence of
erased terms at the term level.

\begin{figure}
  \[
  \begin{array}{llll}
    \infer{\Gamma \vdash \star}{\ } &
    \infer{\Gamma\vdash\abs{\Pi}{x}{T}{\kappa}}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash\kappa} &
    \infer{\Gamma\vdash\abs{\Pi}{X}{\kappa'}{\kappa}}{\Gamma \vdash \kappa' & \Gamma,X:\kappa'\vdash\kappa}
  \end{array}
  \]
  \caption{Rules for checking that a kind is well-formed ($\Gamma \vdash \kappa$)}
  \label{fig:superknd}
\end{figure}

\begin{figure}
  \[
  \begin{array}{ll}
    \infer{\Gamma \vdash X \tpsynth \kappa}{(X : \kappa) \in \Gamma} &
    \infer{\Gamma\vdash \abs{\forall}{X}{\kappa}{T} \tpsynth \star}{\Gamma \vdash \kappa & \Gamma,X:\kappa\vdash T \tpsynth \star} \\ \\
    \infer{\Gamma\vdash\abs{\forall}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} &
    \infer{\Gamma\vdash\abs{\Pi}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} \\ \\
    \infer{\Gamma\vdash\abs{\lambda}{x}{T}{T'} \tpsynth \abs{\Pi}{x}{T}{\kappa}}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T'\tpsynth\kappa} &
    \infer{\Gamma\vdash\abs{\lambda}{X}{\kappa}{T'} \tpsynth \abs{\Pi}{X}{\kappa}{\kappa'}}{\Gamma \vdash \kappa & \Gamma,X:\kappa\vdash T'\tpsynth\kappa'} \\ \\
    \infer{\Gamma\vdash T\ t \tpsynth [t/x]\kappa}{\Gamma\vdash T \tpsynth \abs{\Pi}{x}{T'}{\kappa} & \Gamma\vdash t \tpcheck T'} &
    \infer{\Gamma\vdash T\cdot T' \tpsynth [T'/X]\kappa}{\Gamma\vdash T \tpsynth \abs{\Pi}{X}{\kappa'}{\kappa} & \Gamma\vdash T' \tpsynth \kappa' & \kappa \cong \kappa'} \\ \\
    \infer{\Gamma\vdash\abs{\iota}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} &
    \infer{\Gamma\vdash \{ t \simeq t' \} : \star}{\textit{FV}(t\ t')\subseteq\textit{dom}(\Gamma)}
  \end{array}
  \]
  \caption{Rules for synthesizing a kind for a type ($\Gamma \vdash T \tpsynth \kappa$)}
  \label{fig:knd}
\end{figure}

\begin{figure}
  \[
  \begin{array}{ll}
    \infer{\Gamma\vdash x\tpsynth T}{(x : t)\in\Gamma} &
    \infer{\Gamma\vdash t\tpcheck T'}{\Gamma\vdash t \tpcheck T & T' \leadsto^*_\beta T} \\ \\    
    \infer{\Gamma\vdash t\tpsynth T'}{\Gamma\vdash t \tpsynth T & T \leadsto^*_\beta T'} &
    \infer{\Gamma\vdash t\tpcheck T}{\Gamma\vdash t\tpsynth T' & T' \cong T} \\ \\    
    \infer{\Gamma\vdash \absu{\lambda}{x}{t} \tpcheck \abs{\Pi}{x}{T}{T'}}{\Gamma,x:T\vdash t\tpcheck T'} &
    \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]T}{\Gamma\vdash t \tpsynth \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\

    \infer{\Gamma\vdash \absu{\Lambda}{X}{t} \tpcheck \abs{\forall}{X}{\kappa}{T}}{\Gamma,X:\kappa\vdash t \tpcheck T} &
    \infer{\Gamma\vdash t \cdot T' \tpsynth [T'/X]T}
          {\Gamma\vdash t \tpsynth \abs{\forall}{X}{\kappa}{T} & \Gamma\vdash T' \tpcheck\kappa} \\ \\

    \infer{\Gamma\vdash \absu{\Lambda}{x}{t} \tpcheck \abs{\forall}{x}{T'}{T}}{\Gamma,x:T'\vdash t \tpcheck T & x\not\in\textit{FV}(|t|)} &
    \infer{\Gamma\vdash t\ \mhyph t' \tpsynth [t'/x]T}{\Gamma\vdash t \tpsynth \abs{\forall}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\

    \infer{\Gamma\vdash [ t , t' ] \tpcheck \abs{\iota}{x}{T}{T'}}
          {\Gamma\vdash t \tpcheck T & \Gamma\vdash t' \tpcheck [t/x]T' & |t| =_{\beta\eta} |t'|} &
    \infer{\Gamma\vdash t.1 \tpsynth T}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} \\ \\
    \infer{\Gamma\vdash t.2 \tpsynth [t.1/x]T'}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} &

    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t \simeq t \}}{\Gamma\vdash \textit{FV}(t)\subseteq \textit{dom}(\Gamma)}  \\ \\    
    \infer{\Gamma\vdash \delta\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
    \infer{\Gamma\vdash \rho\ t'\ \mhyph\ t \Leftrightarrow [t_2/x]T}
          {\Gamma\vdash t' \tpsynth t_1 \simeq t_2 & \Gamma \vdash t \Leftrightarrow [t_1/x]T} \\ \\
    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpcheck T'}
          {\Gamma\vdash T\tpcheck \star & \Gamma\vdash t \tpcheck T & T \cong T'} &
    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpsynth T}
          {\Gamma\vdash T\tpcheck \star & \Gamma\vdash t \tpsynth T' & T \cong T'} \\ \\
    \infer{\Gamma\vdash \phi\ t\ \mhyph\ t'\{t''\} \Leftrightarrow T}
          {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}  & \
  \end{array}
  \]
\caption{Rules for checking a term against a well-kinded type ($\Gamma \vdash t \tpcheck T$)
           and synthesizing a type for a term ($\Gamma \vdash t \tpsynth T$)}
\label{fig:tp}
\end{figure}

\begin{figure}
  \[
  \begin{array}{lllllll}
    |x| & = & x &\ &
    |\abs{\lambda}{x}{t'}{t}| & = & \absu{\lambda}{x}{|t|} \\
    |t\ t'| & = & |t|\ |t'| &\ &
    |t\cdot T| & = & |t| \\
    |\absu{\Lambda}{x}{t'}{t}| & = & |t| &\ &
    |t\ \mhyph t'| & = & |t| \\
    |[t , t']| & = & |t| &\ &
    |t.1| & = & |t| \\
    |t.2| & = & |t| &\ &
    |\beta\{t\}| & = & |t|\\
    |\delta\ t| & = & |t|&\ &
    |\rho\ t\ \mhyph\ t'| & = & |t'| \\
    |\phi\ t\ \mhyph\ t'\ \{t''\}| & = & |t''| &\ &
    |\chi\ T\ \mhyph\ t'| & = & |t'|
  \end{array}
  \]
  \caption{Erasure for annotated terms}
  \label{fig:eraser}
\end{figure}

\begin{figure}
  \[
  \begin{array}{ll}
    \infer{T \cong T'}{T \leadsto^*_\beta T_1 & T' \leadsto^*_\beta T_2 & T_1\cong^t T_2}  & 
    \infer{T \cong T'}{T \cong^t T'} \\ \\
    \infer{T\ t \cong^t T\ t'}{T \cong^t T' & |t| =_{\beta\eta} |t'|} &
    \infer{\{ t_1 \simeq t_2 \} \cong^t \{ t_1'\ \simeq t_2' \}}{|t_1| =_{\beta\eta} |t_1'| & |t_2| =_{\beta\eta} |t_2'|}
  \end{array}
  \]
  \caption{Non-congruence rules for conversion}
  \label{fig:conv}
\end{figure}

\subsection{Overview of the constructs}
\label{sec:overview}

CDLE has as a subsystem the extrinsic Calculus of
Constructions (CC).  We have dependent types
$\abs{\Pi}{x}{T}{T'}$ and kinds $\abs{\Pi}{x}{T}{\kappa}$, as well as
term- and type-level quantification over (possibly higher-kinded)
types $\abs{\forall}{X}{\kappa}{T}$ and
$\abs{\Pi}{X}{\kappa}{\kappa'}$.  We use $\forall$ when the
corresponding argument will be erased, and $\Pi$ when it will be
retained.  Since we do not erase term or type arguments from
type-level applications, we thus write $\abs{\Pi}{X}{\kappa}{\kappa'}$
instead of $\abs{\forall}{X}{\kappa}{\kappa'}$.  We write $\lambda$ to
correspond to $\Pi$ and $\Lambda$ to correspond to $\forall$.  As noted
above, application to a type is denoted with center dot ($\cdot$).

To Curry-style CC, CDLE adds: implicit products, introduced orginially
by Miquel~\cite{miquel01}; a primitive equality type $\{ t \simeq
t'\}$; and dependent intersection types $\abs{\iota}{x}{T}{T'}$,
introduced by Kopylov~\cite{kopylov03}.  Implicit products are used
for erased arguments to functions, found also in systems like Agda
(cf.~\cite{mishraLinger08}).  Dependent intersections are a rather
exotic construct allowing us to assign type $\abs{\iota}{x}{T'}{T}$ to
erased term $t$ when we can assign $T'$ to $t$, and also assign
$[t/x]T$ to $t$.  For an annotated introduction form, we write
$[t,t'$], where $t$ checks against type $T'$, $t'$ checks against
$[t/x]T$, and $t$ and $t'$ have identical (i.e., $\alpha$-equivalent)
erasures.  Dependent intersections thus enable a controlled form of
self-reference in the type.  Previous work showed how to use this to
derive induction for Church-encoded natural numbers~\cite{stump18}.
We will see below further uses of this construct.

The typing rules include conversion checks in a few places; e.g., as
standardly, when switching from checking to synthesizing mode.  Two
rules near the top of Figure~\ref{fig:tp} state that one may (nondeterministically)
$\beta$-reduce the type one is synthesizing or checking, before proceeding.
This allows reduction to head-normal form,
to match the form of type required by other rules.
Finally, we include the construct $\chi\ T\ \mhyph\ t$ to change the
synthesized or checked type $T'$ to $T$, if $T \cong T'$.  This may
be necessary to get the type into a specific form for purposes of rewriting
with the $\rho$ construct.

Finally, we have modified
the rules for equality types $\{ t \simeq t' \}$ so that we require
nothing of $t$ and $t'$ except that the set $\textit{dom}(\Gamma)$ of variables
declared by $\Gamma$ includes their free variables $\textit{FV}(t\ t')$.  Further modifications
over the version of CDLE in~\cite{stump18} are:
\begin{itemize}
\item To prove $\{ t \simeq t\}$, one now writes $\beta\{t'\}$,
  with the critical idea that $|\beta\{t\}|$ erases to $|t|$.  We call this the \textbf{Kleene trick},
  because it goes back to Kleene's numeric realizability, which accepts any number $n$ as a realizer
  of a true equation.  Here, we accept any closed term $t$ as a realizer
  of $\{ t \simeq t\}$.  This means that in Cedille, any such term -- even otherwise untypable
  terms, non-normalizing terms, etc. -- have type $\{ t \simeq t\}$ for any term $t$.
\item The $\rho$ construct allows one to rewrite occurrences of $t_1$ to $t_2$ in the synthesized or
  checked type, where $t_1$ and $t_2$ are provably equal.  In the Cedille implementation, we rewrite
  all matching occurrences.  This may be compared to \verb|rewrite| in Agda, except that it may be
  applied anywhere, not just as part of pattern matching~\cite{agda}.
\item We adopt a strong form of Nuprl's \textbf{direct computation rules}~\cite{constable+86}:
  If we have a term $t'$ of type $T$ and a proof $t$ that $\{ t' \simeq t''\}$, then we may conclude that
  $t''$ has type $T$ by writing the annotated term $\phi\ t\ \mhyph\ t'\{t''\}$, which
  erases to $t''$.
\item Where the previous version of CDLE uses $\beta$-equivalence for (erased) terms, we here adopt $\beta\eta$-equivalence.  This
  allows us to observe in many cases that retyping functions are actually $\beta\eta$-equivalent to $\absu{\lambda}{x}{x}$.
  While $\beta\eta$-equivalence takes more work to incorporate into intrinsic 
  type theory~\cite{geuvers92},
  it raises no difficulties for our extrinsic one.
\item In this version, we add an explicit axiom $\delta$ saying that Church-encoded boolean \emph{true} is different from
  \emph{false}.  In the first version of CDLE, such an axiom was derivable from \emph{lifting}, a construct allowing
  simply typable terms to be lifted to the type level~\cite{stump17}.  We omit lifting in this new version of CDLE, because while
  sound, lifting as defined in that previous work is complicated and appears to be incomplete.  Developing a new
  form of lifting remains to future work.
\end{itemize}

The equality type remains \textbf{intensional}: we equate terms iff they are $\beta\eta$-equal.

\subsection{Semantics and metatheory}

Figure~\ref{fig:semtp} gives a realizability semantics for types and
kinds, following the semantics given in the previous papers on
CDLE~\cite{stump18,stump17}.  Details of this semantics are presented
further in Section~\ref{sec:snd} below.  Using the semantics and the
definition in Figure~\ref{fig:semctxt} of $\interp{\Gamma}$, we can
prove the following theorem:
\begin{theorem}[Soundness]
\label{thm:snd}
Suppose $(\sigma,\rho)\in\interp{\Gamma}$.  Then we have:
\begin{enumerate}
\item If $\Gamma\vdash \kappa$, then $\interp{\kappa}_{\sigma,\rho}$ is defined.
\item If $\Gamma\vdash T \tpsynth \kappa$, then $\interp{T}_{\sigma,\rho}\in\interp{\kappa}_{\sigma,\rho}$.
\item If $\Gamma\vdash t \tpsynth T$ then $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho}\in \mathcal{R}$.
\item If $\Gamma\vdash t \tpcheck T$ and $\interp{T}_{\sigma,\rho}\in \mathcal{R}$, then
    $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho}\in \mathcal{R}$.
\item If $T \cong T'$ or $T \cong^t T'$ and $\interp{T}_{\sigma,\rho}$ and $\interp{T'}_{\sigma,\rho}$ are both defined, then they are equal.
\end{enumerate}
\end{theorem}

An easy corollary, by the semantics of $\forall$-types, is then:

\begin{theorem}[Logical consistency]
\label{thm:consis}
  There is no term $t$ such that $\vdash t : \abs{\forall}{X}{\star}{X}$.
\end{theorem}

It may worry some readers that we have:
\begin{observation}
  There are typable terms $t$ which fail to normalize.
\end{observation}

Defining \verb|Top| to be $\{\absu{\lambda}{x}{x} \simeq
\absu{\lambda}{x}{x}\}$, we may assign \verb|Top| to any closed term \verb|t|,
including non-normalizing ones.  In our annotated syntax, we write \verb|β{t}|.
Even without this, the presence of
$\delta$ in combination with $\phi$ allows us to type non-normalizing
terms assuming an erased argument $x$ of type $\{ \textit{tt} \simeq
\textit{ff} \}$ for Church-encoded booleans \textit{tt} and
\textit{ff}.  For example, $\delta\ x$ has type
\verb|{λ x . x ≃ λ x . x x}|, and with $\phi$ we can use this to type
$\Omega$ by changing the typed term \verb|id · True id|, where
\verb|True| is \verb|∀ X : ★ . X ➔ X|. But failure of normalization
does not impinge on Theorem~\ref{thm:consis}.   Extensional Martin-L\"of type theory (MLTT) is
also non-normalizing, for a very similar reason, but fact
does not contradict its logical soundness~\cite{dybjer16}.  In CDLE, the guarantees one gets about the
behavior of terms are expressed almost entirely in their types.  If
the types are weak, then not much is guaranteed; but stronger types
can guarantee properties like normalization.

Given the lack of normalization, several checks in the typing rules --
for things like $t =_{\beta\eta} t'$ -- are formally undecidable.  In
practice, we simply impose a bound on the number of steps of reduction,
and thus restore formal decidability (we are checking ``typable within
a given budget'').  In practice, the same is done for Coq and Agda,
where type checking is decidable but, in general, infeasible (since one
may write astronomically slow terminating functions).

Finally, in line with ideas recently advocated by Dreyer, we
do not concern ourselves with syntactic
type preservation~\cite{dreyer18}, noting instead that by construction,
semantic types $\interp{T}_{\sigma,\rho}$ are preserved by $\beta\eta$-reduction:

  \begin{theorem}[Semantic type preservation]
    If $t \leadsto_{\beta\eta} t'$ and $t\in\interp{T}_{\sigma,\rho}$, then $t'\in\interp{T}_{\sigma,\rho}$.
    \end{theorem}

  Confluence of $\beta\eta$-reduction for (erased)
  terms is nothing other than confluence of untyped lambda calculus.
  This is because, as easily verified by inspecting
  Figure~\ref{fig:eraser}, the erasure function maps annotated terms
  $t$ to terms $|t|$ of pure untyped lambda calculus.
\begin{comment}
  \begin{lemma}
    If $t$ is an annotated term of CDLE, then $|t|$ is a term of pure untyped lambda calculus.
    \end{lemma}
\end{comment}

\begin{figure}
\[
\begin{array}{lll}
\interp{X}_{\sigma,\rho} & = & \rho(X) \\ 
\interp{\Pi x : T_1. T_2}_{\sigma,\rho} & = & 
    [\{ \lambda x.t\ |\ \forall E\in\interp{T_1}_{\sigma,\rho}.\\
\ &\ &\ \ \ \  [[\choice(E)/x]t]_{\cbe}\in\interp{T_2}_{\sigma[x\mapsto \choice(E)],\rho} \ \wedge\ t = |t|\}]_{\cbe}
 \\
\interp{\forall X:\kappa.T}_{\sigma,\rho} & = & 
  \elcap \{ \interp{T}_{\sigma,\rho[X\mapsto S]} |\ S\in\interp{\kappa}_{\sigma,\rho} \}  \\ 
\interp{\forall x:T.T'}_{\sigma,\rho} & = & 
  \elcap_\star \{ \interp{T'}_{\sigma[x\mapsto \choice(E)],\rho}\ |\ E\in\interp{T}_{\sigma,\rho} \} \\ 
\interp{\iota x:T.T'}_{\sigma,\rho} & = & \{ E\in\interp{T}_{\sigma,\rho} |\ E \in \interp{T'}_{\sigma[x\mapsto \choice(E)],\rho} \} \\ 
\interp{\lambda X:\kappa.T}_{\sigma,\rho} & = & (S\in\interp{\kappa}_{\sigma,\rho} \mapsto \interp{T}_{\sigma,\rho[X\mapsto S]}) 
\\ 
\interp{\lambda x:T.T'}_{\sigma,\rho} & = & 
    (E\in\interp{T}_{\sigma,\rho} \mapsto \interp{T'}_{\sigma[x\mapsto \choice(E)],\rho}) 
\\ 
\interp{T\ T'}_{\sigma,\rho} & = & \interp{T}_{\sigma,\rho}(\interp{T'}_{\sigma,\rho})
\\ 
\interp{T\ t}_{\sigma,\rho} & = & \interp{T}_{\sigma,\rho}([\sigma |t|]_{\cbe})
\\
\interp{t \simeq t'}_{\sigma,\rho} & = & [\{ t''\ |\ \sigma |t| =_{\beta\eta} \sigma |t'|\ \wedge\ t'' = |t''| \}]_{\cbe} \\
\ &\ &\ \ \ \textnormal{ if }\textit{FV}(t\ t')\subseteq\textit{dom}(\sigma) 
\\
\interp{\star}_{\sigma,\rho} & = & \mathcal{R} \\ 
\interp{\Pi x:T.\kappa}_{\sigma,\rho} & = & 
(E\in\interp{T}_{\sigma,\rho} \to \interp{\kappa}_{\sigma[x\mapsto \choice(E)],\rho}),\\
\ &\ &\ \ \ \textnormal{ if }\interp{T}_{\sigma,\rho}\in\mathcal{R} \\
\interp{\Pi x:\kappa.\kappa'}_{\sigma,\rho} & = & (S\in\interp{\kappa}_{\sigma,\rho} \to \interp{\kappa}_{\sigma,\rho[X\mapsto S]}) \\
\elcap_\star X & = & \left\{\begin{array}{l}
                                         \, \negthinspace\elcap X, \textnormal{ if } X\neq\emptyset\\
                                         \, \negthinspace[\mathcal{L}]_{\cbe},\textnormal{ otherwise}
                                       \end{array}\right. \\
\end{array}
\]
\caption{Semantics for types and kinds}
\label{fig:semtp}
\end{figure}

\begin{figure}
\[
\begin{array}{lll}
(\sigma\uplus[x\mapsto t],\rho)\in\interp{\Gamma,x:T} & \Leftrightarrow & (\sigma,\rho)\in\interp{\Gamma} \ \wedge\ 
 [t]_{\cbe}\in\interp{T}_{\sigma,\rho}\in\mathcal{R}\ \wedge\ t = |t| \\
(\sigma,\rho\uplus[X\mapsto S])\in\interp{\Gamma,X:\kappa} & \Leftrightarrow & (\sigma,\rho)\in\interp{\Gamma} \ \wedge\ 
S\in\interp{\kappa}_{\sigma,\rho} \\
(\emptyset,\emptyset)\in\interp{\cdot}
\end{array}
\]
\caption{Semantics of typing contexts $\Gamma$}
\label{fig:semctxt}
\end{figure}

\subsection{Some details about the semantics and the proof of Theorem~\ref{thm:snd}}
\label{sec:snd}

Following the development in~\cite{stump17}, we work with
set-theoretic partial functions for the semantics of higher-kinded
types.  Types are interpreted as $\beta\eta$-closed sets of closed
terms. Let $\mathcal{L}$ be the set of closed terms of pure lambda calculus
(differently from~\cite{stump17}, we include all terms at this point,
even non-normalizing ones).  We
write $=_{\cbe}$ for standard $\beta\eta$-equivalence of pure lambda calculus, restricted to
closed terms; and $[t]_{\cbe}$ for $\{ t'\ |\ t =_{\cbe} t'\}$.  This
is extended to sets $S$ of terms by writing $[S]_{\cbe}$ for
$\{[t]_{\cbe}\ |\ t\in S\}$.  In a few places we write
$\textit{nf}(t)$ for the (unique) $\beta\eta$-normal form of term $t$,
if it has one.  If (in our meta-language) we affirm a statement
involving application of a partial function, then it is to be
understood that that application is defined.

\begin{definition}[Reducibility candidates]
  $\mathcal{R} := \{ [S]_{\cbe}\ |\ S\subseteq \mathcal{L} \}$.
\end{definition}

Throughout the development we find it convenient to use a
\textbf{choice function} $\choice$.  Given any nonempty set $E$ of
terms, $\choice$ returns some element of $E$.  Note that if $a \in A
\in \mathcal{R}$, then $a$ is a nonempty set of terms of pure lambda
calculus; it can also happen that $A \in\mathcal{R}$ is empty.  The
proof of Theorem~\ref{thm:snd} (see appendix) is then a straightforward adaptation of~\cite{stump17}.

\textbf{Acknowledgments.}  This work was partially supported by the US
NSF support under award 1524519, and US DoD support under award
FA9550-16-1-0082 (MURI program).

%% Bibliography
\bibliography{biblio}

%% Appendix
\appendix

\section{Proof of Theorem~\ref{thm:snd}}

First a few lemmas (easy proofs omitted):

\begin{lemma}
  $\interp{\kappa}_{\sigma,\rho}$ is nonempty if defined.
\end{lemma}

\begin{lemma}
\label{lem:choice}
If $E$ is nonempty, then $[\choice(E)]_{\cbe} = E$
\end{lemma}

\begin{lemma}
  The set $\mathcal{R}$ ordered by subset forms a complete lattice,
  with greatest element $[\mathcal{L}]_{\cbe}$ and greatest lower bound
  of a nonempty set of elements given by
  intersection.  Also, $\emptyset$ is the least element.
\end{lemma}

\begin{lemma}[Term substitution and interpretation]
\label{lem:termsubstinterp}
If $t' =_{\cbe} \sigma |t|$, then: 
\begin{itemize}
\item $\interp{T}_{\sigma[x\mapsto t'],\rho} = \interp{[t/x]T}_{\sigma,\rho}$
\item $\interp{\kappa}_{\sigma[x\mapsto t'],\rho} = \interp{[t/x]\kappa}_{\sigma,\rho}$
\end{itemize}
\end{lemma}

\begin{lemma}[Type substitution and interpretation] 
\label{lem:tpsubstinterp}
\begin{itemize}
\item $\interp{T}_{\sigma,\rho[X\mapsto\interp{T'}_{\sigma,\rho}]} = \interp{[T'/X]T}_{\sigma,\rho}$
\item $\interp{\kappa}_{\sigma,\rho[X\mapsto\interp{T'}_{\sigma,\rho}]} = \interp{[T'/X]\kappa}_{\sigma,\rho}$
\end{itemize}
\end{lemma}

\begin{lemma}
  \label{lem:interppres}
  If $T \leadsto^*_{\beta} T'$ and $\interp{T}_{\sigma,\rho}$ is defined, then $\interp{T'}_{\sigma,\rho}$ is also defined and equals $\interp{T}_{\sigma,\rho}$.
\end{lemma}
\begin{proof}
  This follows by induction on the reduction derivation, making use of the previous substitution lemmas.
  \end{proof}

\begin{proof}[Soundness (Theorem~\ref{thm:snd})]
  The following proof is adapted from~\cite{stump17}.  It proceeds by
  mutual induction on the assumed typing, kinding, or superkinding
  derivation, for each part of the lemma.  We prove the parts
  successively.

\subsection{Proof of part (1)}

\startcase{.2cm}
\[
\infer{\Gamma \vdash \star }{\ }
\]
$\interp{\star}_{\sigma,\rho}$ is just $\mathcal{R}$, which is
defined.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash\abs{\Pi}{x}{T}{\kappa}}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash\kappa} 
\]
By the IH, $\interp{T}_{\sigma,\rho}\in\mathcal{R}$, and so
$\interp{\Pi x : T.\, \kappa}_{\sigma,\rho}$ is
$(E\in\interp{T}_{\sigma,\rho} \to \interp{\kappa}_{\sigma[x\mapsto  \choice(E)],\rho})$.
The latter quantity is defined if for all
$E\in\interp{T}_{\sigma,\rho}$, $\interp{\kappa}_{\sigma[x\mapsto \choice(E)],\rho})$ is, too.  Since
$\interp{T}_{\sigma,\rho}\in\mathcal{R}$, every element $E$ of
$\interp{T}_{\sigma,\rho}$ is nonempty, as noted above, 
so $\choice(E)$ is defined.  We may apply the IH to the second
premise, since
$(\sigma[x\mapsto\choice(E)],\rho)\in\interp{\Gamma,x:T}$, because $E\in\interp{T}_{\sigma,\rho}$ (by assumption)
and $[\choice(E)]_{\cbe} = E$.  This gives definedness of the semantics
of the $\Pi$-kind.

\startcase{.2cm}
\[
   \infer{\Gamma\vdash\abs{\Pi}{X}{\kappa'}{\kappa}}{\Gamma \vdash \kappa' & \Gamma,X:\kappa'\vdash\kappa}
\]
We must show $(S\in\interp{\kappa}_{\sigma,\rho} \to \interp{\kappa}_{\sigma,\rho[X\mapsto S]})$ is defined.
This is true if $\interp{\kappa}_{\sigma,\rho}$ is defined, which is the case by
the IH applied to the first premise; and if for all
$S\in\interp{\kappa}_{\sigma,\rho}$,
$\interp{\kappa}_{\sigma,\rho[X\mapsto S]}$ is defined.  The latter is
true by the IH applied to the second premise.

\subsection{Proof of part (2)}

\startcase{.2cm}
\[
\infer{\Gamma \vdash X \tpsynth \kappa}{(X : \kappa) \in \Gamma} 
\]
From the definition of $\interp{\Gamma}$, we obtain
$\rho(x)\in\interp{\kappa}_{\sigma,\rho}$.

\startcase{.2cm}
\[
   \infer{\Gamma\vdash\abs{\Pi}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} 
\]
We must show $\interp{\Pi x:T.T'}_{\sigma,\rho}\in\mathcal{R}$. The
semantics defines $\interp{\Pi x:T.T'}_{\sigma,\rho}$ to be
$[A]_{\cbe}$ for a certain $A$, where if $A$ is defined, then
$A\subseteq\mathcal{L}$.  So it suffices to shown definedness. By the IH
for the first premise, $\interp{T}_{\sigma,\rho}\in\mathcal{R}$.  This
means that if $E\in\interp{T}_{\sigma,\rho}$, $\choice(E)$ is defined.
We can then apply the IH to the second premise, since
$\sigma[x\mapsto\choice(E)]\in\interp{\Gamma,x:T}$, to obtain
definedness of $\interp{T'}_{\sigma[x\mapsto\choice(E),\rho}$.

\startcase{.2cm}
\[
   \infer{\Gamma\vdash\abs{\forall}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} 
\]
By the IH for the second premise, $\interp{T_2}_{\sigma[x\mapsto  \choice(E)],\rho}\in\mathcal{R}$, for every
$E\in\interp{T_1}_{\sigma,\rho}$ where
$\interp{T_1}_{\sigma,\rho}\in\mathcal{R}$.  By the IH for the first
premise, we indeed have $\interp{T_1}_{\sigma,\rho}\in\mathcal{R}$.
So if $\interp{T_1}_{\sigma,\rho}$ is non-empty, then the intersection of all the sets
$\interp{T_2}_{\sigma[x\mapsto \choice(E)],\rho}$ where $E\in\interp{T_1}_{\sigma,\rho}$ is a
reducibility candidate, since each of those sets is.  By the semantics
of $\forall$-types quantifying over terms, this is sufficient.  If $\interp{T_1}_{\sigma,\rho}$ is
empty, then the interpretation of the $\forall$-type is $[\mathcal{L}]_{\cbe}$ by the definition
of $\elcap_\star$, and this is in $\mathcal{R}$.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \abs{\forall}{X}{\kappa}{T} \tpsynth \star}{\Gamma \vdash \kappa & \Gamma,X:\kappa\vdash T \tpsynth \star} 
\]
Similarly to the previous case: by the IH for the second premise,
$\interp{T_2}_{\sigma,\rho[X\mapsto S}\in\mathcal{R}$, for every
$S\in\interp{\kappa}_{\sigma,\rho}$.  By the IH part for the first
premise, $\interp{\kappa}_{\sigma,\rho}$ is defined.  So the
intersection of all the sets $\interp{T_2}_{\sigma,\rho[X\mapsto S]}$
where $S\in\interp{\kappa}_{\sigma,\rho}$ is a reducibility candidate,
since each of those sets is.  The intersection is nonempty, since $\interp{\kappa}_{\sigma,\rho}$ is (as stated in a lemma above).
By the semantics of $\forall$-types
quantifying over types, this is sufficient.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash\abs{\iota}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} 
\]
The set $\interp{\iota x:T.T'}_{\sigma,\rho}$ is explicitly defined to
be a subset of $\interp{T}_{\sigma,\rho}$, which is in $\mathcal{R}$,
by the IH applied to the first premise.  Since for any
$A\subseteq\mathcal{L}$, $[A]_{\cbe}$ is in $\mathcal{R}$, to show that
$\interp{\iota x:T.T'}_{\sigma,\rho}$ is also in $\mathcal{R}$ it suffices
to show definedness of $\interp{T'}_{\sigma[x\mapsto \choice(E)],\rho}\}$
(which is used in the predicate picking out the
particular subset of $\interp{T}_{\sigma,\rho}$), for
$E\in\interp{T}_{\sigma,\rho}$.  For such $E$, $\choice(E)$ is defined
(since $\interp{T}_{\sigma,\rho}\in\mathcal{R}$ and hence $E\in\interp{T}_{\sigma,\rho}$ is nonempty)
and in $E$,
so $\sigma[x\mapsto\choice(E)]\in\interp{\Gamma,x:T}$.  So
by the IH for the second premise,
$\interp{T'}_{\sigma[x\mapsto\choice(E),\rho]}$ is defined.

\startcase{.2cm}
\[
 \infer{\Gamma\vdash\abs{\lambda}{x}{T}{T'} \tpsynth \abs{\Pi}{x}{T}{\kappa}}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T'\tpsynth\kappa}
\]
By the semantics, $\interp{\lambda x:T.T'}_{\sigma,\rho}$ is
$(E\in\interp{T}_{\sigma,\rho} \mapsto \interp{T'}_{\sigma[x\mapsto
  \choice(E)],\rho})$.  We must show that this (meta-level) function
is in $\interp{\Pi x:T.\kappa}_{\sigma,\rho}$.  By the semantics of
kinds, the latter quantity, if defined, is
$(E\in\interp{T}_{\sigma,\rho} \to_{\cbe}
\interp{\kappa}_{\sigma[x\mapsto \choice(E)],\rho})$.
By the IH for the first premise, $\interp{T}_{\sigma,\rho}\in\mathcal{R}$.
So we must just show that for any $E\in\interp{T}_{\sigma,\rho}$,
$\interp{T'}_{\sigma[x\mapsto
  \choice(E)],\rho}\in\interp{\kappa}_{\sigma[x\mapsto
  \choice(E)],\rho}$.  But this follows by the IH for the second
premise.

\startcase{.2cm}
\[
\infer{\Gamma\vdash\abs{\lambda}{X}{\kappa}{T'} \tpsynth \abs{\Pi}{X}{\kappa}{\kappa'}}
      {\Gamma \vdash \kappa & \Gamma,X:\kappa\vdash T'\tpsynth\kappa'} 
\]
This case is an easier version of the previous one.  It suffices to
assume an arbitrary $S\in\interp{\kappa}_{\sigma,\rho}$ and show
$\interp{T'}_{\sigma,\rho[X\mapsto S]}\in\interp{\kappa'}_{\sigma,\rho[X\mapsto S]}$.  But this follows
by the IH applied to the second premise.  And we have definedness of
$\interp{\kappa}_{\sigma,\rho}$ by the IH for the first premise.

\startcase{.2cm}
\[
   \infer{\Gamma\vdash T\ t \tpsynth [t/x]\kappa}{\Gamma\vdash T \tpsynth \abs{\Pi}{x}{T'}{\kappa} & \Gamma\vdash t \tpcheck T'} 
\]
By the IH for the first premise,
$\interp{T}_{\sigma,\rho}\in\interp{\Pi x:T'.\kappa}_{\sigma,\rho}$.
By the semantics of $\Pi$-kinds, this means that
$\interp{T}_{\sigma,\rho}$ is a function which given any
$E\in\interp{T'}_{\sigma,\rho}$, will produce a result in
$\interp{\kappa}_{\sigma[x\mapsto \choice(E)],\rho}$.  By the
semantics of type applications, $\interp{T\ t}_{\sigma,\rho}$ is equal
to $\interp{T}_{\sigma,\rho}([\sigma |t|]_{\cbe})$.  This is defined,
since $[\sigma |t|]_{\cbe}\in\interp{T'}_{\sigma,\rho}$, by the IH for
the second premise; note that $\interp{T'}_{\sigma,\rho}$ is defined
since otherwise $\interp{\Pi x:T'.\kappa}_{\sigma,\rho}$ would not be defined.
The result of applying the function is thus indeed
in $\interp{[t/x]\kappa}_{\sigma,\rho}$, since by
Lemma~\ref{lem:termsubstinterp}, this equals
$\interp{\kappa}_{\sigma[x\mapsto \choice([\sigma |t|]_{\cbe})],\rho}$
(the codomain of the function being applied).

\startcase{.2cm}
\[
\infer{\Gamma\vdash T\cdot T' \tpsynth [T'/X]\kappa}
      {\Gamma\vdash T \tpsynth \abs{\Pi}{X}{\kappa'}{\kappa} & \Gamma\vdash T' \tpsynth \kappa' & \kappa \cong \kappa'} 
\]
By the IH applied to the first premise,
$\interp{T}_{\sigma,\rho}\in\interp{\abs{\Pi}{X}{\kappa'}{\kappa}}_{\sigma,\rho}$.
By the semantics of $\Pi$-kinds, this means that for any
$S\in\interp{\kappa'}_{\sigma,\rho}$, $\interp{T}_{\sigma,\rho}\ S$ is in
$\interp{\kappa}_{\sigma,\rho[X\mapsto S]}$.  By the IH for the second premise, we have 
$\interp{T'}\interp{\kappa'}_{\sigma,\rho}$, and by the IH for the third premise,
we have $\interp{\kappa}_{\sigma,\rho} = \interp{\kappa'}_{\sigma,\rho}$.  So
we get $\interp{T}_{\sigma,\rho}(\interp{T'}_{\sigma,\rho})\in \interp{\kappa}_{\sigma,\rho[X\mapsto \interp{T'}_{\sigma,\rho}]}$,
which suffices by Lemma~\ref{lem:tpsubstinterp}.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \{ t \simeq t' \} : \star}{\textit{FV}(t\ t')\subseteq\textit{dom}(\Gamma)}
\]
Either $\sigma |t| =_{\cbe} \sigma |t'|$ or not.  Either way, the interpretation is defined and in $\mathcal{R}$, since
$\textit{FV}(t\ t')\subseteq\textit{dom}(\sigma)$ (as an easy consequence of $(\sigma,\rho)\in\interp{\Gamma}$).

\subsection{Proof of parts (3) and (4)}

\startcase{.2cm}
\[
    \infer{\Gamma\vdash x\tpsynth T}{(x : t)\in\Gamma} 
\]
This follows from the definition of $\interp{\Gamma}$.

\startcase{.2cm}
\[
 \infer{\Gamma\vdash t\tpcheck T'}{\Gamma\vdash t \tpcheck T & T' \leadsto^*_\beta T} 
\]
We are assuming $\interp{T'}_{\sigma,\rho}$ is defined, since this is a checking judgment. The desired result then
follows from Lemma~\ref{lem:interppres}.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash t\tpsynth T'}{\Gamma\vdash t \tpsynth T & T \leadsto^*_\beta T'} 
\]
This also follows from Lemma~\ref{lem:interppres} and the induction hypothesis for the first premise,
which implies $\interp{T}_{\sigma,\rho}\in\mathcal{R}$ (and hence defined).

\startcase{.2cm}
\[
\infer{\Gamma\vdash t\tpcheck T}{\Gamma\vdash t\tpsynth T' & T' \cong T} 
\]
By the IH applied to the first premise, we have 
$[\sigma t]_{\cbe}\in\interp{T'}_{\sigma,\rho}\in\mathcal{R}$. By assumption, $\interp{T}_{\sigma,\rho}\in\mathcal{R}$,
and so by the IH applied to the second premise, we have $[\sigma t]_{\cbe}\in\interp{T'}_{\sigma,\rho} = \interp{T}_{\sigma,\rho}$.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \absu{\lambda}{x}{t} \tpcheck \abs{\Pi}{x}{T}{T'}}{\Gamma,x:T\vdash t\tpcheck T'} 
\]
To show $[\sigma \lambda x.t]_{\cbe}\in\interp{\Pi  x:T.T'}_{\sigma,\rho}$ (noting that the latter is defined
and in $\mathcal{R}$ by assumption), it suffices to assume an arbitrary
$E\in\interp{T}_{\sigma,\rho}$, and show
$[[\choice(E)/x]\sigma t]_{\cbe}\in\interp{T'}_{\sigma[x\mapsto\choice(E)],\rho}$.  By the IH,
we have
$[\sigma[x\mapsto\choice(E)]t]_{\cbe}\in\interp{T'}_{\sigma[x\mapsto\choice(E)],\rho}$.
But $[\sigma[x\mapsto\choice(E)]t]_{\cbe} = [[\choice(E)/x]\sigma t]_{\cbe}$,
so this is sufficient.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]T}{\Gamma\vdash t \tpsynth \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} 
\]
By the IH applied to the first premise, $[\sigma
t]_{\cbe}\in\interp{\Pi x:T'.T}_{\sigma,\rho}\in\mathcal{R}$.  This
means that there exists a $\lambda$-abstraction $\lambda x.\hat{t}$
such that $\lambda x.\hat{t} =_{\cbe} \sigma t$, by the semantics of $\Pi$-types.
Furthermore, for any $E\in\interp{T'}_{\sigma,\rho}$,
$[[\choice(E)/x]\hat{t}]_{\cbe}\in\interp{T}_{\sigma[x\mapsto\choice(E)],\rho}$.
By the IH applied to the second premise, $[\sigma t']_{\cbe}\in\interp{T'}_{\sigma,\rho}$,
so we can instantiate the quantifier in the previous formula to obtain
\[
 [[\choice([\sigma t']_{\cbe})/x]\hat{t}]_{\cbe}\in\interp{T}_{\sigma[x\mapsto\choice([\sigma t']_{\cbe})],\rho}
\]
By Lemma~\ref{lem:termsubstinterp}, this is equivalent to
\[
 [[\choice([\sigma t']_{\cbe})/x]\hat{t}]_{\cbe}\in\interp{[t'/x]T_2}_{\sigma,\rho}
\]
Since $\sigma (t\ t') =_{\cbe} (\lambda x.\hat{t})\ \sigma t' =_{\cbe} [[\choice([\sigma t']_{\cbe})/x]\hat{t}$,
this is sufficient.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \absu{\Lambda}{X}{t} \tpcheck \abs{\forall}{X}{\kappa}{T}}{\Gamma,X:\kappa\vdash t \tpcheck T} 
\]
By the IH, $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho[X\mapsto S]}$, for all $S\in\interp{\kappa}_{\sigma,\rho}$.
This is sufficient to prove $[\sigma |\absu{\Lambda}{X}{t}|]_{\cbe}\in\interp{\forall X:\kappa.T}_{\sigma,\rho}$, by the semantics
of $\forall$-types and definition of erasure.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash t \cdot T' \tpsynth [T'/X]T}
          {\Gamma\vdash t \tpsynth \abs{\forall}{X}{\kappa}{T} & \Gamma\vdash T' \tpcheck\kappa} 
\]
By the semantics of $\forall$-types and the IH applied to the first
premise, we have $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho[X\mapsto
  S]}$, for all $S\in\interp{\kappa}_{\sigma,\rho}$.  Since
$\interp{T'}_{\sigma,\rho}\in\interp{\kappa}_{\sigma,\rho}$ by the IH
applied to the second premise, we can derive $[\sigma t]_{\cbe}\in\interp{T}_{\sigma,\rho[X\mapsto \interp{T'}_{\sigma,\rho}]}$.
By Lemma~\ref{lem:tpsubstinterp},
this is equivalent to the required $[\sigma |t|]_{\cbe}\in\interp{[T'/X]T}_{\sigma,\rho}$,
using also the definition of erasure.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \absu{\Lambda}{x}{t} \tpcheck \abs{\forall}{x}{T'}{T}}{\Gamma,x:T'\vdash t \tpcheck T & x\not\in\textit{FV}(|t|)} 
\]

By the IH applied to the first premise, we have
$[\sigma[x\mapsto\choice(E)]
t]_{\cbe}\in\interp{T'}_{\sigma[x\mapsto\choice(E)],\rho}$, for any
$E\in\interp{T}_{\sigma,\rho}$.  This is because
$\interp{T}_{\sigma,\rho}\in\mathcal{R}$, since $\interp{\abs{\forall}{x}{T'}{T}}_{\sigma,\rho}$ is in $\mathcal{R}$ and
hence defined, by assumption.  Since $x\not\in\textit{FV}(t)$, we know
$[[\sigma[x\mapsto\choice(E)]t]_{\cbe} = [\sigma t]_{\cbe}$.  By the
semantics of $\forall$-types and definition of erasure, this suffices to show the desired
conclusion.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash t\ \mhyph t' \tpsynth [t'/x]T}{\Gamma\vdash t \tpsynth \abs{\forall}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'}
\]
The result follows easily by the IH applied to the premises, the
semantics of $\forall$-types, definition of erasure, and Lemma~\ref{lem:termsubstinterp}.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash [ t , t' ] \tpcheck \abs{\iota}{x}{T}{T'}}
          {\Gamma\vdash t \tpcheck T & \Gamma\vdash t' \tpcheck [t/x]T' & |t| =_{\beta\eta} |t'|} 
\]
By the IH, we have $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho}$ and
$[\sigma |t|]_{\cbe}\in\interp{[t/x]T'}_{\sigma,\rho}$.  By
Lemma~\ref{lem:termsubstinterp}, the latter is equivalent to
$[\sigma t]_{\cbe}\in\interp{T'}_{\sigma[x\mapsto\choice([\sigma t]_{\cbe}),\rho}$.  These two facts about $[\sigma t]_{\cbe}$ are
sufficient, by the semantics of $\iota$-types, for the desired
conclusion, using also the fact (from the third premise) that $\sigma|t| =_{\cbe} \sigma|t'|$.

\startcase{.2cm}
\[
   \infer{\Gamma\vdash t.1 \tpsynth T}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} 
\]
The desired conclusion follows easily from the IH and the semantics of $\iota$-types.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash t.2 \tpsynth [t.1/x]T'}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} 
\]
Similar to the previous case, using Lemma~\ref{lem:termsubstinterp}.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t \simeq t \}}{\Gamma\vdash \textit{FV}(t)\subseteq \textit{dom}(\Gamma)}  
\]
$[\sigma|t'|]_{\cbe}\in\interp{\{ t \simeq t \}}_{\sigma,\rho}$ follows directly from the semantics of equality types.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \delta\ t \tpcheck T}{\Gamma\vdash t\tpsynth \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  
\]
By the semantics of equality types, $[\sigma|t'|]_{\cbe}$ cannot be in the interpretation of the equation in the premise,
since the two terms in question are closed and not $\beta\eta$-equal.  By the IH applied to the first premise, however,
$[\sigma|t'|]_{\cbe}$ is in the interpretation of that equation.  This is a contradiction.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \rho\ t'\ \mhyph\ t \Leftrightarrow [t_2/x]T}
          {\Gamma\vdash t' \tpsynth t_1 \simeq t_2 & \Gamma \vdash t \Leftrightarrow [t_1/x]T} 
\]
By the IH applied to the first premise, $\sigma|t_1| =_{\beta\eta} \sigma|t_2|$.  The result then follows by the IH applied to the second premise,
and Lemma~\ref{lem:termsubstinterp}.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpcheck T'}
          {\Gamma\vdash T\tpcheck \star & \Gamma\vdash t \tpcheck T & T \cong T'} 
\]
Using the IH for the first premise and the assumption that $\interp{T'}_{\sigma,\rho}$ is in $\mathcal{R}$ and hence defined,
we can apply the IH to the third premise to get $\interp{T}_{\sigma,\rho} = \interp{T'}_{\sigma,\rho}$.  Using this and the IH for second premise,
we get the desired conclusion, using also the definition of erasure.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpsynth T}
          {\Gamma\vdash T\tpcheck \star & \Gamma\vdash t \tpsynth T' & T \cong T'} 
\]
By the IH applied to the second premise, we have $[\sigma|t|]_{\cbe}\in\interp{T'}_{\sigma,\rho}\in\mathcal{R}$.
Using definedness of $\interp{T'}_{\sigma,\rho}$ and the IH applied to the first premise, we can apply the
IH to the third premise to get $\interp{T}_{\sigma,\rho} = \interp{T'}_{\sigma,\rho}$, from which the desired
conclusion follows by definition of erasure.

\startcase{.2cm}
\[
    \infer{\Gamma\vdash \phi\ t\ \mhyph\ t'\{t''\} \Leftrightarrow T}
          {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}  
\]
By the IH for the first premise, $\sigma|t'| =_{\cbe} \sigma|t''|$.  By the IH for the second
premise, $[\sigma|t'|]_{\cbe}\in\interp{T}_{\sigma,\rho}$.  This suffices for the desired conclusion,
using also the definition of erasure ($|\phi\ t\ \mhyph\ t'\{t''\}| = |t''|$).

\subsection*{Proof of part (5) }

\startcase{.2cm}
\[
    \infer{T \cong T'}{T \leadsto^*_\beta T_1 & T' \leadsto^*_\beta T_2 & T_1\cong^t T_2}  
\]
By Lemma~\ref{lem:interppres}, we have
\[
\begin{array}{lll}
  \interp{T}_{\sigma,\rho} & = & \interp{T_1}_{\sigma,\rho}\\
  \interp{T'}_{\sigma,\rho} & = & \interp{T_2}_{\sigma,\rho}
\end{array}
\]
By the IH for the third premise, we have $\interp{T_1}_{\sigma,\rho} = \interp{T_2}_{\sigma,\rho}$,
which suffices.

\startcase{.2cm}
\[
    \infer{T \cong T'}{T \cong^t T'} 
\]
By the IH.

\startcase{.2cm}
\[
    \infer{T\ t \cong^t T'\ t'}{T \cong^t T' & |t| =_{\beta\eta} |t'|}
\]
By the semantics, $\interp{T\ t}_{\sigma,\rho} =
\interp{T}_{\sigma,\rho}([\sigma|t|]_{\cbe})$.  By the second premise
and the IH for the first premise, this equals
$\interp{T'}_{\sigma,\rho}([\sigma|t'|]_{\cbe})$, as required.

\startcase{.2cm}
\[
    \infer{\{ t_1 \simeq t_2 \} \cong^t \{ t_1'\ \simeq t_2' \}}{|t_1| =_{\beta\eta} |t_1'| & |t_2| =_{\beta\eta} |t_2'|}
\]
This follows easily from the premises and the semantics of equality types.

\end{proof}
\end{document}

### 2018-06-12 docs/semantics/arxiv-version.tex
arxiv version
git-svn-id: https://svn.divms.uiowa.edu/repos/clc/projects/lambda-encodings/implementations/cedille@30780 bd178f5f-7bea-4014-844f-495ca19b8805

--- 2018-06-12 docs/semantics/paper.tex
@@
 \newtheorem{observation}[theorem]{Observation}
-
-\bibliographystyle{plain}
 
@@
 
-%% Bibliography
-\bibliography{biblio}
+\begin{thebibliography}{10}
+
+\bibitem{constable+86}
+Robert~L. Constable, Stuart~F. Allen, Mark Bromley, Rance Cleaveland, J.~F.
+  Cremer, R.~W. Harper, Douglas~J. Howe, Todd~B. Knoblock, N.~P. Mendler,
+  Prakash Panangaden, James~T. Sasaki, and Scott~F. Smith.
+\newblock {\em Implementing mathematics with the Nuprl proof development
+  system}.
+\newblock Prentice Hall, 1986.
+
+\bibitem{dreyer18}
+Derek Dreyer.
+\newblock {The Type Soundness Theorem That You Really Want to Prove (and Now
+  You Can)}.
+\newblock Milner Award Lecture, delivered at Principles of Programming
+  Languages (POPL), 2018.
+
+\bibitem{dybjer16}
+Peter Dybjer and Erik Palmgren.
+\newblock {Intuitionistic Type Theory}.
+\newblock In Edward~N. Zalta, editor, {\em The Stanford Encyclopedia of
+  Philosophy}. Metaphysics Research Lab, Stanford University, winter 2016
+  edition, 2016.
+
+\bibitem{geuvers92}
+Herman Geuvers.
+\newblock {The Church-Rosser Property for beta-eta-reduction in Typed
+  lambda-Calculi}.
+\newblock In {\em Proceedings of the Seventh Annual Symposium on Logic in
+  Computer Science {(LICS} '92), Santa Cruz, California, USA, June 22-25,
+  1992}, pages 453--460. {IEEE} Computer Society, 1992.
+
+\bibitem{kopylov03}
+Alexei Kopylov.
+\newblock Dependent intersection: {A} new way of defining records in type
+  theory.
+\newblock In {\em 18th {IEEE} Symposium on Logic in Computer Science {(LICS)}},
+  pages 86--95, 2003.
+
+\bibitem{agda}
+\mbox{The Agda development team}.
+\newblock {\em Agda}, 2018.
+\newblock Version 2.5.3.
+
+\bibitem{miquel01}
+Alexandre Miquel.
+\newblock {The Implicit Calculus of Constructions Extending Pure Type Systems
+  with an Intersection Type Binder and Subtyping}.
+\newblock In Samson Abramsky, editor, {\em {Typed Lambda Calculi and
+  Applications}}, volume 2044 of {\em Lecture Notes in Computer Science}, pages
+  344--359. Springer, 2001.
+
+\bibitem{mishraLinger08}
+Nathan Mishra{-}Linger and Tim Sheard.
+\newblock {Erasure and Polymorphism in Pure Type Systems}.
+\newblock In Roberto~M. Amadio, editor, {\em Foundations of Software Science
+  and Computational Structures, 11th International Conference, {FOSSACS} 2008,
+  Held as Part of the Joint European Conferences on Theory and Practice of
+  Software, {ETAPS} 2008, Budapest, Hungary, March 29 - April 6, 2008.
+  Proceedings}, volume 4962 of {\em Lecture Notes in Computer Science}, pages
+  350--364. Springer, 2008.
+
+\bibitem{peytonjones07}
+Simon Peyton~Jones, Dimitrios Vytiniotis, Stephanie Weirich, and Mark Shields.
+\newblock {Practical Type Inference for Arbitrary-rank Types}.
+\newblock {\em J. Funct. Program.}, 17(1):1--82, January 2007.
+
+\bibitem{pierce+00}
+Benjamin~C. Pierce and David~N. Turner.
+\newblock Local type inference.
+\newblock {\em {ACM} Trans. Program. Lang. Syst.}, 22(1):1--44, 2000.
+
+\bibitem{stump17}
+Aaron Stump.
+\newblock {The Calculus of Dependent Lambda Eliminations}.
+\newblock {\em J. Funct. Program.}, 27:e14, 2017.
+
+\bibitem{stump18}
+Aaron Stump.
+\newblock {From Realizability to Induction via Dependent Intersection}, 2018.
+\newblock in press.
+
+\end{thebibliography}
 

### 2019-09-18 docs/semantics/paper.tex
docs: semantics: add termination guarantee
also: fix typos, update delta syntax

@@
     \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t \simeq t \}}{\Gamma\vdash \textit{FV}(t)\subseteq \textit{dom}(\Gamma)}  \\ \\    
-    \infer{\Gamma\vdash \delta\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
+    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
     \infer{\Gamma\vdash \rho\ t'\ \mhyph\ t \Leftrightarrow [t_2/x]T}
@@
     |\beta\{t\}| & = & |t|\\
-    |\delta\ t| & = & |t|&\ &
+    |\delta\ \mhyph\ t| & = & |t|&\ &
     |\rho\ t\ \mhyph\ t'| & = & |t'| \\
@@
 \absu{\lambda}{x}{x}\}$, we may assign \verb|Top| to any closed term \verb|t|,
-including non-normalizing ones.  In our annotated syntax, we write \verb|β{t}|.
-Even without this, the presence of
-$\delta$ in combination with $\phi$ allows us to type non-normalizing
-terms assuming an erased argument $x$ of type $\{ \textit{tt} \simeq
-\textit{ff} \}$ for Church-encoded booleans \textit{tt} and
-\textit{ff}.  For example, $\delta\ x$ has type
-\verb|{λ x . x ≃ λ x . x x}|, and with $\phi$ we can use this to type
-$\Omega$ by changing the typed term \verb|id · True id|, where
-\verb|True| is \verb|∀ X : ★ . X ➔ X|. But failure of normalization
-does not impinge on Theorem~\ref{thm:consis}.   Extensional Martin-L\"of type theory (MLTT) is
-also non-normalizing, for a very similar reason, but fact
-does not contradict its logical soundness~\cite{dybjer16}.  In CDLE, the guarantees one gets about the
-behavior of terms are expressed almost entirely in their types.  If
-the types are weak, then not much is guaranteed; but stronger types
-can guarantee properties like normalization.
+including non-normalizing ones. In our annotated syntax, we write
+\texttt{\(\beta\)\{t\}}. Even without this, the presence of $\delta$ in
+combination with $\phi$ allows us to type non-normalizing terms assuming an
+erased argument $x$ of type $\{ \textit{tt} \simeq \textit{ff} \}$ for
+Church-encoded booleans \textit{tt} and \textit{ff}. For example, $\delta\ \mhyph x$
+has type \(\{\absu{\lambda}{x}{x} \simeq \absu{\lambda}{x}{x\ x}\}\), and with
+$\phi$ we can use this to type $\Omega$ by changing the typed term
+\(\texttt{id}\ \cdot\texttt{True}\ \texttt{id}\), where \verb|True| is
+\(\abs{\forall}{X}{\star}{X \to X}\). But failure of normalization does not
+impinge on Theorem~\ref{thm:consis}. Extensional Martin-L\"of type theory (MLTT)
+is also non-normalizing, for a very similar reason, but fact does not contradict
+its logical soundness~\cite{dybjer16}. In CDLE, the guarantees one gets about
+the behavior of terms are expressed almost entirely in their types. If the types
+are weak, then not much is guaranteed; but stronger types can guarantee
+properties like normalization, as demonstrated by the following theorem:
 
-Given the lack of normalization, several checks in the typing rules --
+\begin{theorem}[Call-by-name normalization of functions]
+  \label{thm:cedille-termination}
+  Suppose \(\Gamma\vdash t : T\), $t$ is closed, and there exists a closed term
+  $t'$ which erases to \(\lambda x.\ x\) and whose type is $T \to \Pi x : T_1.\
+  T_2$ for some $T_1, T_2$. Then $|t|$ is call-by-name normalizing.
+  \end{theorem}
+
+Given the lack of normalization in general, several checks in the typing rules --
 for things like $t =_{\beta\eta} t'$ -- are formally undecidable.  In
@@
 \[
-    \infer{\Gamma\vdash \delta\ t \tpcheck T}{\Gamma\vdash t\tpsynth \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  
+    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpsynth \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  
 \]
@@
 \end{proof}
+
+\section{Proof of Theorem~\ref{thm:cedille-termination}}
+\begin{proof}
+Theorem \ref{thm:snd} implies that since
+$t'\ t$ is closed and of type $\Pi x:T_1.\ T_2$, we have
+$[|t'\ t|]_{c\beta\eta}\in \interp{\Pi x:T_1.\ T_2}_{\cdot,\rho}$,
+where $[|t'\ t|]_{c\beta\eta}$ is the set of closed terms which are
+$\beta\eta$-equivalent to $|t'\ t|$; and
+$(\cdot,\rho)\in\interp{\Gamma}$ gives interpretations $\cdot$ for
+term- and $\rho$ for type-variables in $\Gamma$.  By the semantics of
+types defined in Figure \ref{fig:semtp}, the
+interpretation of a $\Pi$-type consists of sets of the form
+$[\lambda x.\,t']_{c\beta\eta}$.  So we have that $|t'\ t|$ is $\beta\eta$-equivalent to
+$\lambda x.\, t'$ for some $x, t'$.  
+Since $|t'\ t|$ is $\beta$-equivalent to $t$, we know $t =_{\beta\eta} \lambda x.\, t'$.
+It is then an easy consequence of the
+standardization theorem for untyped lambda calculus that $|t|$ is
+call-by-name normalizing (cf.~\cite{Kashima2000}).
+
+\end{proof}
+
 \end{document}

### 2020-01-15 src/CedilleTypes.hs
syntax for first-class algebras complete, plus a few tests
parser/test/alg.ced has a few tests for the new syntax, which are
passing.

@@ -52,11 +52,15 @@ data Def = DefTerm PosInfo Var (Maybe Type) Term | DefType PosInfo Var Kind Type
 
 data Guide = Guide PosInfo Var Type deriving Show
 
-data Case = Case PosInfo Var [CaseArg] Term deriving Show
+data Pattern = Pattern PosInfo Var [CaseArg] deriving Show
+
+data Case = Case Pattern  Term deriving Show
 
 data CaseArg = CaseArg CaseArgSym PosInfo Var deriving Show
 
 data TpKd = Tkt Type | Tkk Kind deriving Show
+
+type AlgMotive = (Maybe (PosInfo , Var), Pattern, Type)
 
 data Type =
     TpAbs PosInfo MaybeErased PosInfo Var TpKd Type
@@ -67,6 +71,7 @@ data Type =
   | TpApp Type Type
   | TpAppt Type Term
   | TpArrow Type MaybeErased Type
+  | TpAlg PosInfo AlgMotive
   | TpEq PosInfo Term Term PosInfo
   | TpHole PosInfo
   | TpLam PosInfo PosInfo Var TpKd Type
@@ -94,8 +99,13 @@ data Term =
   | Rho PosInfo RhoHnf (Maybe [Num]) Term (Maybe Guide) Term
   | VarSigma PosInfo Term
   | Theta PosInfo Theta Term [Lterm]
-  | Mu PosInfo PosInfo Var Term (Maybe Type) PosInfo Cases PosInfo
+  | Mu PosInfo PosInfo Var (Maybe AlgMotive) PosInfo Cases PosInfo
   | Sigma PosInfo (Maybe Term) Term (Maybe Type) PosInfo Cases PosInfo
+  | Fold Term Term PosInfo
+  
+  -- combine Mu and Fold for better type inference
+  | MuFold Term PosInfo Var (Maybe AlgMotive) PosInfo Cases PosInfo 
+
   | Var PosInfo Var
   deriving Show
 

### 2020-01-15 src/CedilleTypes.hs
changed the ast to FoldMu instead of MuFold
to better reflect the order of terms in the syntax

@@ -104,7 +104,7 @@ data Term =
   | Fold Term Term PosInfo
   
   -- combine Mu and Fold for better type inference
-  | MuFold Term PosInfo Var (Maybe AlgMotive) PosInfo Cases PosInfo 
+  | FoldMu Term PosInfo Var (Maybe AlgMotive) PosInfo Cases PosInfo 
 
   | Var PosInfo Var
   deriving Show

### 2020-01-28 docs/semantics/arxiv-version.tex
Update docs to reference Agda version 2.5.4 (ticket #25)
docs: semantics: remove some differences between ArXiV and default version

@@
     \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t \simeq t \}}{\Gamma\vdash \textit{FV}(t)\subseteq \textit{dom}(\Gamma)}  \\ \\    
-    \infer{\Gamma\vdash \delta\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
+    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
     \infer{\Gamma\vdash \rho\ t'\ \mhyph\ t \Leftrightarrow [t_2/x]T}
-          {\Gamma\vdash t' \tpsynth t_1 \simeq t_2 & \Gamma \vdash t \Leftrightarrow [t_1/x]T} \\ \\
+          {\Gamma\vdash t' \tpsynth \{ t_1 \simeq t_2 \} & \Gamma \vdash t \Leftrightarrow [t_1/x]T} \\ \\
     \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpcheck T'}
@@
     |t\cdot T| & = & |t| \\
-    |\absu{\Lambda}{x}{t'}{t}| & = & |t| &\ &
+    |\abs{\Lambda}{x}{t'}{t}| & = & |t| &\ &
     |t\ \mhyph t'| & = & |t| \\
@@
 \textit{ff} \}$ for Church-encoded booleans \textit{tt} and
-\textit{ff}.  For example, $\delta\ x$ has type
+\textit{ff}.  For example, $\delta\ \mhyph x$ has type
 \verb|{λ x . x ≃ λ x . x x}|, and with $\phi$ we can use this to type
@@
 does not impinge on Theorem~\ref{thm:consis}.   Extensional Martin-L\"of type theory (MLTT) is
-also non-normalizing, for a very similar reason, but fact
+also non-normalizing, for a very similar reason, but this fact
 does not contradict its logical soundness~\cite{dybjer16}.  In CDLE, the guarantees one gets about the
@@
 the types are weak, then not much is guaranteed; but stronger types
-can guarantee properties like normalization.
-
-Given the lack of normalization, several checks in the typing rules --
+can guarantee properties like normalization, as demonstrated by the following
+theorem:
+
+\begin{theorem}[Call-by-name normalization of functions]
+  \label{thm:cedille-termination}
+  Suppose \(\Gamma\vdash t : T\), $t$ is closed, and there exists a closed term
+  $t'$ which erases to \(\lambda x.\ x\) and whose type is $T \to \Pi x : T_1.\
+  T_2$ for some $T_1, T_2$. Then $|t|$ is call-by-name normalizing.
+  \end{theorem}
+
+Given the lack of normalization in general, several checks in the typing rules --
 for things like $t =_{\beta\eta} t'$ -- are formally undecidable.  In
@@
 \newblock {\em Agda}, 2018.
-\newblock Version 2.5.3.
+\newblock Version 2.5.4.
 

### 2020-03-03 docs/semantics/paper.tex
Fixed typo in typing rules
This fixes #140
docs: semantics: fix erasure of implicit abstraction

@@
     \infer{\Gamma\vdash \rho\ t'\ \mhyph\ t \Leftrightarrow [t_2/x]T}
-          {\Gamma\vdash t' \tpsynth t_1 \simeq t_2 & \Gamma \vdash t \Leftrightarrow [t_1/x]T} \\ \\
+          {\Gamma\vdash t' \tpsynth \{ t_1 \simeq t_2 \} & \Gamma \vdash t \Leftrightarrow [t_1/x]T} \\ \\
     \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpcheck T'}
@@
 impinge on Theorem~\ref{thm:consis}. Extensional Martin-L\"of type theory (MLTT)
-is also non-normalizing, for a very similar reason, but fact does not contradict
+is also non-normalizing, for a very similar reason, but this fact does not contradict
 its logical soundness~\cite{dybjer16}. In CDLE, the guarantees one gets about

### 2020-10-18 docs/semantics/arxiv-version.tex
docs: update arxiv version

@@
 \absu{\lambda}{x}{x}\}$, we may assign \verb|Top| to any closed term \verb|t|,
-including non-normalizing ones.  In our annotated syntax, we write \verb|β{t}|.
-Even without this, the presence of
-$\delta$ in combination with $\phi$ allows us to type non-normalizing
-terms assuming an erased argument $x$ of type $\{ \textit{tt} \simeq
-\textit{ff} \}$ for Church-encoded booleans \textit{tt} and
-\textit{ff}.  For example, $\delta\ \mhyph x$ has type
-\verb|{λ x . x ≃ λ x . x x}|, and with $\phi$ we can use this to type
-$\Omega$ by changing the typed term \verb|id · True id|, where
-\verb|True| is \verb|∀ X : ★ . X ➔ X|. But failure of normalization
-does not impinge on Theorem~\ref{thm:consis}.   Extensional Martin-L\"of type theory (MLTT) is
-also non-normalizing, for a very similar reason, but this fact
-does not contradict its logical soundness~\cite{dybjer16}.  In CDLE, the guarantees one gets about the
-behavior of terms are expressed almost entirely in their types.  If
-the types are weak, then not much is guaranteed; but stronger types
-can guarantee properties like normalization, as demonstrated by the following
-theorem:
+including non-normalizing ones. In our annotated syntax, we write
+\texttt{\(\beta\)\{t\}}. Even without this, the presence of $\delta$ in
+combination with $\phi$ allows us to type non-normalizing terms assuming an
+erased argument $x$ of type $\{ \textit{tt} \simeq \textit{ff} \}$ for
+Church-encoded booleans \textit{tt} and \textit{ff}. For example, $\delta\ \mhyph x$
+has type \(\{\absu{\lambda}{x}{x} \simeq \absu{\lambda}{x}{x\ x}\}\), and with
+$\phi$ we can use this to type $\Omega$ by changing the typed term
+\(\texttt{id}\ \cdot\texttt{True}\ \texttt{id}\), where \verb|True| is
+\(\abs{\forall}{X}{\star}{X \to X}\). But failure of normalization does not
+impinge on Theorem~\ref{thm:consis}. Extensional Martin-L\"of type theory (MLTT)
+is also non-normalizing, for a very similar reason, but this fact does not contradict
+its logical soundness~\cite{dybjer16}. In CDLE, the guarantees one gets about
+the behavior of terms are expressed almost entirely in their types. If the types
+are weak, then not much is guaranteed; but stronger types can guarantee
+properties like normalization, as demonstrated by the following theorem:
 
@@
 
+\bibitem{Kashima2000}
+  Ryo Kashima.
+  \newblock {A Proof of the Standardization Theorem in \(\lambda\)-Calculus}.
+  \newblock Available from the author's web page, 2000.
+
 \bibitem{kopylov03}

### 2020-12-22 docs/semantics/paper.tex
docs: semantics: small fixes
- in beta, FV of t' must be declared in the context
- delta erases to \ x. x
- typo in delta case of logical soundness

@@
 
-    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t \simeq t \}}{\Gamma\vdash \textit{FV}(t)\subseteq \textit{dom}(\Gamma)}  \\ \\    
+    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t \simeq t \}}{\Gamma\vdash \textit{FV}(t\ t')\subseteq \textit{dom}(\Gamma)}  \\ \\    
     \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
@@
     |\beta\{t\}| & = & |t|\\
-    |\delta\ \mhyph\ t| & = & |t|&\ &
+    |\delta\ \mhyph\ t| & = & \absu{\lambda}{x}{x}&\ &
     |\rho\ t\ \mhyph\ t'| & = & |t'| \\
@@
 \]
-By the semantics of equality types, $[\sigma|t'|]_{\cbe}$ cannot be in the interpretation of the equation in the premise,
+By the semantics of equality types, $[\sigma|t|]_{\cbe}$ cannot be in the interpretation of the equation in the premise,
 since the two terms in question are closed and not $\beta\eta$-equal.  By the IH applied to the first premise, however,
-$[\sigma|t'|]_{\cbe}$ is in the interpretation of that equation.  This is a contradiction.
+$[\sigma|t|]_{\cbe}$ is in the interpretation of that equation.  This is a contradiction.
 

### 2020-12-22 docs/semantics/paper.tex
docs: semantics: more small changes
- update authorship
- elaborate on bidirectional type inference
- introduce call-by-name normalization in type conversion

@@
 
-\author{Aaron Stump \\
+\author{Aaron Stump, Chris Jenkins \\
   Computer Science \\
@@
 Eliminations (CDLE).  This document presents the version of CDLE as of
-June 1, 2018.  We have made many changes from the first paper on CDLE~\cite{stump17},
+December 1, 2020.  We have made many changes from the first paper on CDLE~\cite{stump17},
 mostly in the form of dropping constructs we discovered (to our
@@
 synthesizing.  We write $\Leftrightarrow$ to range over
-$\{\tpcheck,\tpsynth\}$. We follow the syntax of our implementation
+$\{\tpcheck,\tpsynth\}$, and when this symbol occurs multiple times in a rule,
+it is intended that such occurrences be read the same way (i.e., either all as
+\(\tpcheck\) or all as \(\tpsynth\)).
+We follow the syntax of our implementation
 Cedille, which distinguishes application of a term or type $e$ to a
 type ($e \cdot T$), from application to a term ($e\ t$), and
-application to an erased term argument ($e\ \mhyph t$).  The rules are
-intended, with a few points of nondeterminism, to be read bottom-up
-(in a standard way; cf.~\cite{peytonjones07}) as an algorithm for
-computing a classifier from a context and an expression ($\tpsynth$)
-or checking an expression against a classifier in context
-($\tpcheck$).
+application to an erased term argument ($e\ \mhyph t$, in which case \(e\) must
+be a term).
+The rules are intended, with a few points of nondeterminism, to be read
+bottom-up (in a standard way;
+c.f.~\cite{peytonjones07,Pfe01_Lecture-Notes-on-Bidirectional-Type-Checking}) as
+an algorithm for synthesizing a classifier from a context and an expression
+($\tpsynth$) (\emph{type synthesis}) or checking an expression against a classifier in context
+($\tpcheck$) (\emph{type checking}).
+Specifically, nondeterminism arises in the two rules allowing both synthesized
+and checked types to be reduced, written \(T \leadsto^*_\beta T'\), so that type
+constructors required by other rules may be revealed.
+The reduction relation \(\leadsto_\beta\) is call-by-name normalization, as this
+is sufficient to reduce types to weak head normal form.
 
@@
 Curry-style), and hence we only consider erasures $|t|$ of terms when
-testing for $\beta\eta$-equivalence.  This is done by the conversion
-relation $T\cong T'$, whose central rules are given in
+testing for $\beta\eta$-equivalence.  This applies also to the conversion
+relation $T\cong T'$ on types and \(\kappa \cong \kappa'\) on kinds, whose central rules are given in
 Figure~\ref{fig:conv}.  That figure omits the various congruence rules
@@
 erased terms at the term level.
+Note in particular that for conversion we check full \(\beta\) and
+\(\beta\eta\)-equivalence, not just equivalence with respect to call-by-name
+normalization.
 
@@
 $[t,t'$], where $t$ checks against type $T'$, $t'$ checks against
-$[t/x]T$, and $t$ and $t'$ have identical (i.e., $\alpha$-equivalent)
+$[t/x]T$, and $t$ and $t'$ have $\alpha\beta\eta$-equivalent
 erasures.  Dependent intersections thus enable a controlled form of
@@
 The typing rules include conversion checks in a few places; e.g., as
-standardly, when switching from checking to synthesizing mode.  Two
-rules near the top of Figure~\ref{fig:tp} state that one may (nondeterministically)
-$\beta$-reduce the type one is synthesizing or checking, before proceeding.
-This allows reduction to head-normal form,
-to match the form of type required by other rules.
-Finally, we include the construct $\chi\ T\ \mhyph\ t$ to change the
-synthesized or checked type $T'$ to $T$, if $T \cong T'$.  This may
-be necessary to get the type into a specific form for purposes of rewriting
-with the $\rho$ construct.
+standardly, when switching from checking to synthesizing mode.  As mentioned
+earlier, two rules near the top of Figure~\ref{fig:tp} state that one may
+(nondeterministically) $\beta$-reduce the type one is synthesizing or checking,
+before proceeding.
+This allows reduction to weak head normal form, to match the form of type
+required by other rules.
+Finally, we include the construct $\chi\ T\ \mhyph\ t$ which allows a term whose
+type can only be checked to be ascribed a user-provided type so that the whole
+expression may occur where type synthesis is expected.
 
-Finally, we have modified
-the rules for equality types $\{ t \simeq t' \}$ so that we require
+We have modified the rules for equality types $\{ t \simeq t' \}$ so that we require
 nothing of $t$ and $t'$ except that the set $\textit{dom}(\Gamma)$ of variables
@@
 \item To prove $\{ t \simeq t\}$, one now writes $\beta\{t'\}$,
-  with the critical idea that $|\beta\{t\}|$ erases to $|t|$.  We call this the \textbf{Kleene trick},
-  because it goes back to Kleene's numeric realizability, which accepts any number $n$ as a realizer
+  with the critical idea that $|\beta\{t'\}|$ erases to $|t'|$.  We call this the \textbf{Kleene trick},
+  because it goes back to Kleene's numeric
+  realizability~\cite{Kle65_Classical-Extensions-of-Intuitionistic-Mathematics},
+  which accepts any number $n$ as a realizer 
   of a true equation.  Here, we accept any closed term $t$ as a realizer
@@
 \item In this version, we add an explicit axiom $\delta$ saying that Church-encoded boolean \emph{true} is different from
-  \emph{false}.  In the first version of CDLE, such an axiom was derivable from \emph{lifting}, a construct allowing
+  \emph{false}.
+  In the implementation of Cedille, this is generalized to the rule where the
+  proof \(t\) synthesizes an equation \(\{t_1 \simeq t_2\}\) in which \(|t_1|\)
+  and \(|t_2|\) are separable using the \emph{B\"ohm-out algorithm}~\cite{BDPR79_Bohm-Algorithm}.
+
+  In the first version of CDLE, such an axiom was derivable from \emph{lifting}, a construct allowing
   simply typable terms to be lifted to the type level~\cite{stump17}.  We omit lifting in this new version of CDLE, because while
@@
 Finally, in line with ideas recently advocated by Dreyer, we
-do not concern ourselves with syntactic
-type preservation~\cite{dreyer18}, noting instead that by construction,
-semantic types $\interp{T}_{\sigma,\rho}$ are preserved by $\beta\eta$-reduction:
+are less concerned with syntactic
+type preservation as we are with \emph{semantic} type
+preservation~\cite{dreyer18}.
+Note that by construction, semantic types $\interp{T}_{\sigma,\rho}$ are
+preserved by $\beta\eta$-reduction:
 

### 2020-12-22 docs/semantics/paper.tex
docs: semantics: fix typo in type-to-type application

@@
     \infer{\Gamma\vdash T\ t \tpsynth [t/x]\kappa}{\Gamma\vdash T \tpsynth \abs{\Pi}{x}{T'}{\kappa} & \Gamma\vdash t \tpcheck T'} &
-    \infer{\Gamma\vdash T\cdot T' \tpsynth [T'/X]\kappa}{\Gamma\vdash T \tpsynth \abs{\Pi}{X}{\kappa'}{\kappa} & \Gamma\vdash T' \tpsynth \kappa' & \kappa \cong \kappa'} \\ \\
+    \infer{\Gamma\vdash T_1 \cdot T_2 \tpsynth [T_2/X]\kappa_1}{\Gamma\vdash T_1 \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1} & \Gamma\vdash T_2 \tpsynth \kappa_2' & \kappa_2 \cong \kappa_2'} \\ \\
     \infer{\Gamma\vdash\abs{\iota}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} &
@@
 \[
-\infer{\Gamma\vdash T\cdot T' \tpsynth [T'/X]\kappa}
-      {\Gamma\vdash T \tpsynth \abs{\Pi}{X}{\kappa'}{\kappa} & \Gamma\vdash T' \tpsynth \kappa' & \kappa \cong \kappa'} 
+  \infer{\Gamma\vdash T_1 \cdot T_2 \tpsynth [T_2/X]\kappa_1}{\Gamma\vdash T_1 \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1} & \Gamma\vdash T_2 \tpsynth \kappa_2' & \kappa_2 \cong \kappa_2'}
 \]
 By the IH applied to the first premise,
-$\interp{T}_{\sigma,\rho}\in\interp{\abs{\Pi}{X}{\kappa'}{\kappa}}_{\sigma,\rho}$.
+$\interp{T_1}_{\sigma,\rho}\in\interp{\abs{\Pi}{X}{\kappa_2}{\kappa_1}}_{\sigma,\rho}$.
 By the semantics of $\Pi$-kinds, this means that for any
-$S\in\interp{\kappa'}_{\sigma,\rho}$, $\interp{T}_{\sigma,\rho}\ S$ is in
-$\interp{\kappa}_{\sigma,\rho[X\mapsto S]}$.  By the IH for the second premise, we have 
-$\interp{T'}\interp{\kappa'}_{\sigma,\rho}$, and by the IH for the third premise,
-we have $\interp{\kappa}_{\sigma,\rho} = \interp{\kappa'}_{\sigma,\rho}$.  So
-we get $\interp{T}_{\sigma,\rho}(\interp{T'}_{\sigma,\rho})\in \interp{\kappa}_{\sigma,\rho[X\mapsto \interp{T'}_{\sigma,\rho}]}$,
+$S\in\interp{\kappa_2}_{\sigma,\rho}$, $\interp{T_1}_{\sigma,\rho}\ S$ is in
+$\interp{\kappa_1}_{\sigma,\rho[X\mapsto S]}$.  By the IH for the second premise, we have 
+$\interp{T_2} \in \interp{\kappa_2'}_{\sigma,\rho}$, and by the IH for the third premise,
+we have $\interp{\kappa_2}_{\sigma,\rho} = \interp{\kappa_2'}_{\sigma,\rho}$.  So
+we get $\interp{T_1}_{\sigma,\rho}(\interp{T_2}_{\sigma,\rho})\in \interp{\kappa_1}_{\sigma,\rho[X\mapsto \interp{T'}_{\sigma,\rho}]}$,
 which suffices by Lemma~\ref{lem:tpsubstinterp}.

### 2020-12-22 docs/semantics/paper.tex
docs: semantics: changes to several typing rules
- use \varphi instead of \phi
- use \rho with guide
- remove occurences of types being checked against kinds
- only one version of chi

@@
 
-    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t \simeq t \}}{\Gamma\vdash \textit{FV}(t\ t')\subseteq \textit{dom}(\Gamma)}  \\ \\    
+    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t \simeq t \}}{\Gamma\vdash \textit{FV}(t')\subseteq \textit{dom}(\Gamma)}  \\ \\    
     \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
-    \infer{\Gamma\vdash \rho\ t'\ \mhyph\ t \Leftrightarrow [t_2/x]T}
-          {\Gamma\vdash t' \tpsynth \{ t_1 \simeq t_2 \} & \Gamma \vdash t \Leftrightarrow [t_1/x]T} \\ \\
-    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpcheck T'}
-          {\Gamma\vdash T\tpcheck \star & \Gamma\vdash t \tpcheck T & T \cong T'} &
-    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpsynth T}
-          {\Gamma\vdash T\tpcheck \star & \Gamma\vdash t \tpsynth T' & T \cong T'} \\ \\
-    \infer{\Gamma\vdash \phi\ t\ \mhyph\ t'\{t''\} \Leftrightarrow T}
-          {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}  & \
+    \infer{\Gamma \vdash \rho\ t'\ @x.T\ \mhyph\ t \tpsynth [t_2/x]T}
+          {
+          \begin{array}{cc}
+            \Gamma \vdash t' \tpsynth \{t_1 \simeq t_2\}
+            & \quad \Gamma \vdash [t_2/x]T \tpsynth \star
+            \\ \Gamma \vdash t \tpsynth T'
+            & T' \cong [t_1/x]T
+          \end{array}
+          } \\ \\
+    \infer{\Gamma \vdash \chi\ T\ \mhyph t\ \tpsynth T}
+          {\Gamma \vdash T \tpsynth \star & \Gamma \vdash t \tpcheck T} &
+    \infer{\Gamma\vdash \varphi\ t\ \mhyph\ t'\{t''\} \Leftrightarrow T}
+          {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}
   \end{array}
@@
     |\delta\ \mhyph\ t| & = & \absu{\lambda}{x}{x}&\ &
-    |\rho\ t\ \mhyph\ t'| & = & |t'| \\
-    |\phi\ t\ \mhyph\ t'\ \{t''\}| & = & |t''| &\ &
+    |\rho\ t\ @x.T\ \mhyph\ t'| & = & |t'| \\
+    |\varphi\ t\ \mhyph\ t'\ \{t''\}| & = & |t''| &\ &
     |\chi\ T\ \mhyph\ t'| & = & |t'|
@@
   of $\{ t \simeq t\}$.  This means that in Cedille, any such term -- even otherwise untypable
-  terms, non-normalizing terms, etc. -- have type $\{ t \simeq t\}$ for any term $t$.
-\item The $\rho$ construct allows one to rewrite occurrences of $t_1$ to $t_2$ in the synthesized or
-  checked type, where $t_1$ and $t_2$ are provably equal.  In the Cedille implementation, we rewrite
-  all matching occurrences.  This may be compared to \verb|rewrite| in Agda, except that it may be
-  applied anywhere, not just as part of pattern matching~\cite{agda}.
+  terms, non-normalizing terms, etc. -- have type $\{ t \simeq t\}$ for any term
+  $t$.
+\item The \(\rho\) construct allows one to rewrite occurrences of \(t_1\) to
+  \(t_2\) in the synthesized type of the second subexpression.
+  The version presented here requires a ``guide'' \(@x.T\), where \(x\)
+  indicates the occurrences of \(t_1\) to substitute.
+  \([t_2/x]T\) must be a well-kinded type, but it is not required that
+  \([t_1/x]T\) also be, only that \(T'\) and \([t_1/x]T\) are convertible modulo
+  erasure --- it is not guaranteed that \(t_1\) and \(t_2\) will have the same
+  type, or will even be typeable.
+  In Cedille, the guide is optional and the construct may be used to rewrite a
+  contextually given type; a heuristic is used to produce a resulting type that
+  is well-kinded whose details are beyond the scope of this document.
 \item We adopt a strong form of Nuprl's \textbf{direct computation rules}~\cite{constable+86}:
   If we have a term $t'$ of type $T$ and a proof $t$ that $\{ t' \simeq t''\}$, then we may conclude that
-  $t''$ has type $T$ by writing the annotated term $\phi\ t\ \mhyph\ t'\{t''\}$, which
+  $t''$ has type $T$ by writing the annotated term $\varphi\ t\ \mhyph\ t'\{t''\}$, which
   erases to $t''$.
@@
 \texttt{\(\beta\)\{t\}}. Even without this, the presence of $\delta$ in
-combination with $\phi$ allows us to type non-normalizing terms assuming an
+combination with $\varphi$ allows us to type non-normalizing terms assuming an
 erased argument $x$ of type $\{ \textit{tt} \simeq \textit{ff} \}$ for
@@
 has type \(\{\absu{\lambda}{x}{x} \simeq \absu{\lambda}{x}{x\ x}\}\), and with
-$\phi$ we can use this to type $\Omega$ by changing the typed term
+$\varphi$ we can use this to type $\Omega$ by changing the typed term
 \(\texttt{id}\ \cdot\texttt{True}\ \texttt{id}\), where \verb|True| is
@@
     \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpcheck T'}
-          {\Gamma\vdash T\tpcheck \star & \Gamma\vdash t \tpcheck T & T \cong T'} 
+          {\Gamma\vdash T\tpsynth \star & \Gamma\vdash t \tpcheck T & T \cong T'} 
 \]
@@
     \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpsynth T}
-          {\Gamma\vdash T\tpcheck \star & \Gamma\vdash t \tpsynth T' & T \cong T'} 
+          {\Gamma\vdash T\tpsynth \star & \Gamma\vdash t \tpsynth T' & T \cong T'} 
 \]
@@
 \[
-    \infer{\Gamma\vdash \phi\ t\ \mhyph\ t'\{t''\} \Leftrightarrow T}
+    \infer{\Gamma\vdash \varphi\ t\ \mhyph\ t'\{t''\} \Leftrightarrow T}
           {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}  
@@
 premise, $[\sigma|t'|]_{\cbe}\in\interp{T}_{\sigma,\rho}$.  This suffices for the desired conclusion,
-using also the definition of erasure ($|\phi\ t\ \mhyph\ t'\{t''\}| = |t''|$).
+using also the definition of erasure ($|\varphi\ t\ \mhyph\ t'\{t''\}| = |t''|$).
 

### 2020-12-22 docs/semantics/paper.tex
docs: semantics: syntactic kind preservation for types under CBN normalization

@@
 
+We make a concession to syntactic classifier preservation in the case of types
+and kinds.
+During type inference, types may be reduced using a call-by-name operational
+semantics to reveal type constructors.
+With the removal of lifting from CDLE, terms cannot compute types and so no
+terms need to be reduced during this process.
+
+\begin{theorem}[Syntactic kind preservation]
+  \label{thm:syntactic-kind-pres}
+  If \(\Gamma \vdash T \tpsynth \kappa\) and \(T \leadsto_{\beta} T'\) then
+  \(\Gamma \vdash T' \tpsynth \kappa'\) for some \(\kappa'\) such that \(\kappa
+  \cong \kappa'\).
+\end{theorem}
+
 \begin{figure}
@@
 
+\section{Proof of Theorem~\ref{thm:syntactic-kind-pres}}
+\label{sec:proof-syntactic-kind-pres}
+
+First a few lemmas (easy proofs omitted):
+\begin{lemma}
+  \label{lem:inference-validity}
+  \ \\
+  \begin{itemize}
+  \item If \(\Gamma \vdash T \tpsynth \kappa\) then \(\Gamma \vdash \kappa\)
+
+  \item If \(\Gamma \vdash t \ T\) then \(\Gamma \vdash T \tpsynth
+    \kappa\) for some \(\kappa\)
+  \end{itemize}
+\end{lemma}
+
+\begin{lemma}
+  \label{lem:subst-kind}
+  \ \\
+  \begin{itemize}
+  \item If \(\Gamma,x:T \vdash T \tpsynth \kappa\) and \(\Gamma \vdash t
+    \tpcheck T\) then \(\Gamma \vdash [t/x]T \tpsynth [t/x]\kappa\)
+
+  \item If \(\Gamma,X:\kappa_2 \vdash T_1 \tpsynth \kappa_1\) and \(\Gamma
+    \vdash T_2 \tpsynth \kappa_2\) then \(\Gamma \vdash [T_2/X]T_1 \tpsynth [T_2/X]\kappa_1\)
+  \end{itemize}
+\end{lemma}
+
+\begin{lemma}
+  \label{lem:tpcheck-conv}
+  If \(\Gamma \vdash t \tpcheck T_1\) where \(\Gamma \vdash T_1 \tpsynth
+  \kappa_1\), and we have \(\Gamma \vdash T_2 \tpsynth \kappa_2\) with \(T_1
+  \cong T_2\) and \(\kappa_1 \cong \kappa_2\), then \(\Gamma \vdash t \tpcheck T_2\).
+\end{lemma}
+
+\begin{proof}[Proof (of Theorem~\ref{thm:syntactic-kind-pres})]
+  We may rule out the cases where \(T\) is a variable or formed by a type
+  constructor, as this would contradict the assumption that \(T \leadsto_{\beta}
+  T'\) for some \(T'\) (recall that \(\leadsto_\beta\) is call-by-name reduction
+  for types).
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma\vdash T\ t \tpsynth [t/x]\kappa_1}{\Gamma\vdash T \tpsynth
+      \abs{\Pi}{x}{T_1}{\kappa_1} & \Gamma\vdash t \tpcheck T_1}
+  \]
+
+  There are two subcases to consider for the derivation of \(T\ t \leadsto_\beta
+  T'\). In the first case, we have \(T \leadsto_\beta T''\) for some \(T''\) (so
+  \(T' = T''\ t\)). By the IH, we have \(\Gamma \vdash T'' \tpsynth
+  \abs{\Pi}{x}{T_1'}{\kappa_1'}\) for some \(T_1' \cong T_1\) and \(\kappa_1'
+  \cong \kappa_1\) (we have by assumption the kind of \(T''\) must be
+  convertible with \(\abs{\Pi}{x}{T_1}{\kappa_1}\)). By
+  Lemma~\ref{lem:tpcheck-conv}, \(\Gamma \vdash t \tpcheck T_1'\). By the rule,
+  \(\Gamma \vdash T''\ t \tpsynth [t/x]\kappa_1'\), and the synthesized kind is
+  clearly convertible with \([t/x]\kappa_1\).
+
+  In the other subcase, the subject of kinding is of the form
+  \((\abs{\lambda}{x}{T_1}{T''})\ t\) and our assumed reduction is to
+  \([t/x]T''\). By inversion of the kinding derivation, we have \(\Gamma,x:T_1
+  \vdash T'' \tpsynth \kappa_1\). By Lemma~\ref{lem:subst-kind}, we have
+  \(\Gamma \vdash [t/x]T'' \tpsynth [t/x]\kappa_1\).
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma\vdash T_1 \cdot T_2 \tpsynth [T_2/X]\kappa_1}{\Gamma\vdash T_1
+      \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1} & \Gamma\vdash T_2 \tpsynth
+      \kappa_2' & \kappa_2 \cong \kappa_2'}
+  \]
+
+  There are two subcases to consider for the derivation of \(T_1 \cdot T_2
+  \leadsto_\beta T'\). In the first subcase, we have \(T_1 \leadsto_\beta
+  T_1''\) for some \(T_1''\) (so \(T' = T_1'' \cdot T_2\)). By the IH, we have
+  \(\Gamma \vdash T_1'' \tpsynth \abs{\Pi}{X}{\kappa_2''}{\kappa_1''}\) for some
+  \(\kappa_2'' \cong \kappa_2\) and \(\kappa_1'' \cong \kappa_1\) (we have by
+  assumption the kind of \(T_1''\) must be convertible with
+  \(\abs{\Pi}{X}{\kappa_2}{\kappa_1}\)). By transitivity we can conclude
+  \(\kappa_2'' \cong \kappa_2\). By the rule, we have \(\Gamma \vdash T_1''
+  \cdot T_2 \tpsynth [T_2/X]\kappa_1''\), and the synthesized kind is clearly
+  convertible with \([T_2/X]\kappa_1\).
+
+  In the other subcase, the subject of kinding is of the form
+  \((\abs{\lambda}{X}{\kappa_2}{T_1''}) \cdot T_2\) and our assumed reduction is
+  to \([T_2/X]T_1''\). By inversion of the kinding derivation, we have
+  \(\Gamma,X:\kappa_2 \vdash T_1'' \tpsynth \kappa_1\). By
+  Lemma~\ref{lem:subst-kind}, we have \(\Gamma \vdash [T_2/X]T_1'' \tpsynth
+  [T_2/X]\kappa_1\).
+
+\end{proof}
 \end{document}
+
+% LocalWords:  nondeterministically Kleene's realizability realizer kinded CDLE
+% LocalWords:  Cedille subcases subcase kinding

### 2020-12-23 docs/semantics/paper.tex
docs: semantics: add subtheorem for kinds in soundness
docs: semantics: more typos

@@
           } \\ \\
-    \infer{\Gamma \vdash \chi\ T\ \mhyph t\ \tpsynth T}
+    \infer{\Gamma \vdash \chi\ T\ \mhyph\ t\ \tpsynth T}
           {\Gamma \vdash T \tpsynth \star & \Gamma \vdash t \tpcheck T} &
-    \infer{\Gamma\vdash \varphi\ t\ \mhyph\ t'\{t''\} \Leftrightarrow T}
+    \infer{\Gamma\vdash \varphi\ t\ \mhyph\ t'\ \{t''\} \Leftrightarrow T}
           {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}
@@
   If we have a term $t'$ of type $T$ and a proof $t$ that $\{ t' \simeq t''\}$, then we may conclude that
-  $t''$ has type $T$ by writing the annotated term $\varphi\ t\ \mhyph\ t'\{t''\}$, which
+  $t''$ has type $T$ by writing the annotated term $\varphi\ t\ \mhyph\ t'\ \{t''\}$, which
   erases to $t''$.
@@
 \item If $T \cong T'$ or $T \cong^t T'$ and $\interp{T}_{\sigma,\rho}$ and $\interp{T'}_{\sigma,\rho}$ are both defined, then they are equal.
+\item If \(\kappa \cong \kappa'\) and \(\interp{\kappa}_{\sigma,\rho}\) and
+  \(\interp{\kappa'}_{\sigma,\rho}\) are both defined, then they are equal.
 \end{enumerate}
@@
 \\
-\interp{t \simeq t'}_{\sigma,\rho} & = & [\{ t''\ |\ \sigma |t| =_{\beta\eta} \sigma |t'|\ \wedge\ t'' = |t''| \}]_{\cbe} \\
+\interp{\{t \simeq t'\}}_{\sigma,\rho} & = & [\{ t''\ |\ \sigma |t| =_{\beta\eta} \sigma |t'|\ \wedge\ t'' = |t''| \}]_{\cbe} \\
 \ &\ &\ \ \ \textnormal{ if }\textit{FV}(t\ t')\subseteq\textit{dom}(\sigma) 
@@
 This follows easily from the premises and the semantics of equality types.
+
+\subsection*{Proof of Part (6)}
+
+The convertibility relation for kinds consists entirely of congruential rules
+for quantification over types and kinds.
 

### 2020-12-23 docs/semantics/paper.tex
docs: semantics: rules and soundness proof
- update beta typing rule
- in soundness, fix cases for beta, rho, chi

@@
     \infer{\Gamma\vdash t.2 \tpsynth [t.1/x]T'}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} &
-
-    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t \simeq t \}}{\Gamma\vdash \textit{FV}(t')\subseteq \textit{dom}(\Gamma)}  \\ \\    
+    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t_1 \simeq t_2 \}}
+          {\Gamma\vdash \textit{FV}(t'\ t_1\ t_2)\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}  \\ \\    
     \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
@@
 \begin{itemize}
-\item To prove $\{ t \simeq t\}$, one now writes $\beta\{t'\}$,
-  with the critical idea that $|\beta\{t'\}|$ erases to $|t'|$.  We call this the \textbf{Kleene trick},
-  because it goes back to Kleene's numeric
+\item To prove $\{ t_1 \simeq t_2\}$ for definitionally equal terms (that is,
+  terms that are \(\beta\eta\)-equivalent modulo erasure), one now writes $\beta\{t'\}$,
+  with the critical idea that $|\beta\{t'\}|$ erases to (a possibly unrelated) $|t'|$.
+  We call this the \textbf{Kleene trick}, because it goes back to Kleene's numeric
   realizability~\cite{Kle65_Classical-Extensions-of-Intuitionistic-Mathematics},
   which accepts any number $n$ as a realizer 
-  of a true equation.  Here, we accept any closed term $t$ as a realizer
-  of $\{ t \simeq t\}$.  This means that in Cedille, any such term -- even otherwise untypable
-  terms, non-normalizing terms, etc. -- have type $\{ t \simeq t\}$ for any term
-  $t$.
+  of a true equation.  Here, we accept any term $t'$ as a realizer of $\{ t_1
+  \simeq t_2\}$, provided the free variables of \(t'\) are declared in the
+  context.
+  This means that in Cedille, any such term --- even otherwise untypable
+  terms, non-normalizing terms, etc. --- prove trivially true equations.
 \item The \(\rho\) construct allows one to rewrite occurrences of \(t_1\) to
@@
 \[
-    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t \simeq t \}}{\Gamma\vdash \textit{FV}(t)\subseteq \textit{dom}(\Gamma)}  
-\]
-$[\sigma|t'|]_{\cbe}\in\interp{\{ t \simeq t \}}_{\sigma,\rho}$ follows directly from the semantics of equality types.
+  \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t_1 \simeq t_2 \}}
+  {\Gamma\vdash \textit{FV}(t'\ t_1\ t_2)\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}
+\]
+Since by assumption \(|t_1| =_{\beta\eta} |t_2|\) it follows that \(\sigma|t_1|
+=_{\beta\eta} \sigma|t_2|\), and also by assumption it follows that
+\(\textit{FV}(t_1\ t_2) \subseteq \textit{dom}(\sigma)\).
+Also, we see \(\sigma|t'|\) is a closed term.
+So, $[\sigma|t'|]_{\cbe}\in\interp{\{ t_1 \simeq t_2 \}}_{\sigma,\rho}$ follows
+directly from the semantics of equality types.
 
@@
 \[
-    \infer{\Gamma\vdash \rho\ t'\ \mhyph\ t \Leftrightarrow [t_2/x]T}
-          {\Gamma\vdash t' \tpsynth t_1 \simeq t_2 & \Gamma \vdash t \Leftrightarrow [t_1/x]T} 
-\]
-By the IH applied to the first premise, $\sigma|t_1| =_{\beta\eta} \sigma|t_2|$.  The result then follows by the IH applied to the second premise,
-and Lemma~\ref{lem:termsubstinterp}.
-
-\startcase{.2cm}
-\[
-    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpcheck T'}
-          {\Gamma\vdash T\tpsynth \star & \Gamma\vdash t \tpcheck T & T \cong T'} 
-\]
-Using the IH for the first premise and the assumption that $\interp{T'}_{\sigma,\rho}$ is in $\mathcal{R}$ and hence defined,
-we can apply the IH to the third premise to get $\interp{T}_{\sigma,\rho} = \interp{T'}_{\sigma,\rho}$.  Using this and the IH for second premise,
-we get the desired conclusion, using also the definition of erasure.
-
-\startcase{.2cm}
-\[
-    \infer{\Gamma\vdash \chi\ T\ \mhyph\ t \tpsynth T}
-          {\Gamma\vdash T\tpsynth \star & \Gamma\vdash t \tpsynth T' & T \cong T'} 
-\]
-By the IH applied to the second premise, we have $[\sigma|t|]_{\cbe}\in\interp{T'}_{\sigma,\rho}\in\mathcal{R}$.
-Using definedness of $\interp{T'}_{\sigma,\rho}$ and the IH applied to the first premise, we can apply the
-IH to the third premise to get $\interp{T}_{\sigma,\rho} = \interp{T'}_{\sigma,\rho}$, from which the desired
-conclusion follows by definition of erasure.
+  \infer{\Gamma \vdash \rho\ t'\ @x.T\ \mhyph\ t \tpsynth [t_2/x]T}
+  {
+    \begin{array}{cc}
+      \Gamma \vdash t' \tpsynth \{t_1 \simeq t_2\}
+      & \quad \Gamma \vdash [t_2/x]T \tpsynth \star
+      \\ \Gamma \vdash t \tpsynth T'
+      & T' \cong [t_1/x]T
+    \end{array}
+  }
+\]
+By the IH applied to the first premise in the first row, $\sigma|t_1| =_{\beta\eta}
+\sigma|t_2|$.
+By the IH applied to the first premise in the second row,
+\([\sigma|t|]_{c\beta\eta} \in \interp{T'}_{\sigma.\rho}\).
+By the IH applied to the second premise in the second row,
+\(\interp{T'}_{\sigma,\rho} = \interp{[t_1/x]T}_{\sigma,\rho}\).
+The result then follows by applying Lemma~\ref{lem:termsubstinterp}.
+
+\startcase{.2cm}
+\[
+  \infer{\Gamma \vdash \chi\ T\ \mhyph\ t\ \tpsynth T}
+  {\Gamma \vdash T \tpsynth \star & \Gamma \vdash t \tpcheck T}
+\]
+We apply the IH for the first premise to get that
+$\interp{T}_{\sigma,\rho}$ is in $\mathcal{R}$ and hence defined.
+Using this, we apply the IH on the second premise to get \([\sigma|t|]_{\cbe}
+\in \interp{T}_{\sigma,\rho}\), which by the definition of erasure is what we
+must show.
 

### 2020-12-24 docs/semantics/paper.tex
docs: semantics: some cases for substitution

@@
 \newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{corollary}[theorem]{Corollary}
 \newtheorem{observation}[theorem]{Observation}
@@
   \begin{array}{ll}
-    \infer{\Gamma\vdash x\tpsynth T}{(x : t)\in\Gamma} &
+    \infer{\Gamma\vdash x\tpsynth T}{(x : T)\in\Gamma} &
     \infer{\Gamma\vdash t\tpcheck T'}{\Gamma\vdash t \tpcheck T & T' \leadsto^*_\beta T} \\ \\    
@@
     \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t_1 \simeq t_2 \}}
-          {\Gamma\vdash \textit{FV}(t'\ t_1\ t_2)\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}  \\ \\    
+          {\textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}  \\ \\    
     \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
@@
 \label{fig:tp}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \[
+    \begin{array}{lll}
+      \infer{\vdash \cdot}{}
+      & \infer{
+        \vdash \Gamma,x:T
+        }{\Gamma \vdash T \tpsynth \star}
+      & \infer{
+        \vdash \Gamma,X:\kappa
+        }{
+        \Gamma \vdash \kappa
+        }
+    \end{array}
+  \]
+  \caption{Rules for checking a context is well-formed}
+  \label{fig:ctxt}
 \end{figure}
@@
 erased argument $x$ of type $\{ \textit{tt} \simeq \textit{ff} \}$ for
-Church-encoded booleans \textit{tt} and \textit{ff}. For example, $\delta\ \mhyph x$
-has type \(\{\absu{\lambda}{x}{x} \simeq \absu{\lambda}{x}{x\ x}\}\), and with
+Church-encoded booleans \textit{tt} and \textit{ff}. For example, $\delta\
+\mhyph\ x$ can be checked against type \(\{\absu{\lambda}{x}{x} \simeq
+\absu{\lambda}{x}{x\ x}\}\), and with 
 $\varphi$ we can use this to type $\Omega$ by changing the typed term
@@
   \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t_1 \simeq t_2 \}}
-  {\Gamma\vdash \textit{FV}(t'\ t_1\ t_2)\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}
+  {\textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}
 \]
@@
 First a few lemmas (easy proofs omitted):
+
 \begin{lemma}
-  \label{lem:inference-validity}
+  \label{lem:subst-class}
+  Below, each statement seperately universally quantifies over meta-variables,
+  and it is assumed that typing contexts occuring in assumed derivations are
+  well-formed.
+  We also assume types in type checking judgments are well-formed under the
+  given context.
+  \begin{enumerate}
+  \item \textbf{Kinds:}
+    \begin{itemize}
+    \item If \(\Gamma_1,x:T,\Gamma_2 \vdash \kappa\) and \(\Gamma_1 \vdash t
+      \tpcheck T\) then \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x] \kappa\)
+
+    \item If \(\Gamma_1,X:\kappa',\Gamma_2 \vdash \kappa\) and \(\Gamma_1 \vdash
+      T \tpsynth \kappa'\) then \(\Gamma_1,[T/X]\Gamma_2 \vdash [T/X]\kappa\)
+    \end{itemize}
+
+  \item \textbf{Types}
+    \begin{itemize}
+    \item If \(\Gamma_1,x:T',\Gamma_2 \vdash T \tpsynth \kappa\) and \(\Gamma_1 \vdash t
+      \tpsynth T'\) then \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]T \tpsynth [t/x]\kappa\)
+
+    \item If \(\Gamma_1,X:\kappa_2,\Gamma_2 \vdash T_1 \tpsynth \kappa_1\) and \(\Gamma_1
+      \vdash T_2 \tpsynth \kappa_2\) then \(\Gamma_1,[T_2/X]\Gamma_2 \vdash [T_2/X]T_1 \tpsynth [T_2/X]\kappa_1\)
+    \end{itemize}
+
+  \item \textbf{Terms:}
+    \begin{itemize}
+    \item If \(\Gamma_1,x:T',\Gamma_2 \vdash t \Leftrightarrow T\) and
+      \(\Gamma_1 \vdash t' \tpsynth T'\) then \(\Gamma_1,[t'/x]\Gamma_2 \vdash [t'/x]t
+      \Leftrightarrow [t'/x]T\)
+
+    \item If \(\Gamma_1,X:\kappa,\Gamma_2 \vdash t \Leftrightarrow T'\) and
+      \(\Gamma_1 \vdash T \tpsynth \kappa\) then \(\Gamma_1,[T/X]\Gamma_2 \vdash
+      [T/X]t \Leftrightarrow [T/X]T'\)
+    \end{itemize}
+  \end{enumerate}
+\end{lemma}
+\begin{proof}
+  By mutual induction on the assumed derivations.
+  We only show a few interesting cases.
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma_1,X:\kappa,\Gamma_2 \vdash X_1 \tpsynth \kappa_1}{(X_1 : \kappa_1) \in \Gamma_1,X:\kappa,\Gamma_2}
+  \]
+
+  We have two cases.
+  If \(X_1 = X\), then \(\kappa_1 = \kappa\) and by assumption \(\Gamma_1 \vdash
+  T \tpsynth \kappa\), and the desired result holds by weakening.
+  Otherwise, either \((X_1 : \kappa_1) \in \Gamma_1\) and \(X \notin
+  \textit{FV}(\kappa_1)\), or else \((X_1 : \kappa_1) \in \Gamma_2\).
+  Either way, we have \(\Gamma_1,[T/X]\Gamma_2 \vdash X_1 \tpsynth
+  [T/X]\kappa_1\).
+
+  \startcase{.2cm}
+  \[ \infer{\Gamma_1,x:T,\Gamma_2\vdash T_2\ t_1 \tpsynth [t_1/x_1]\kappa
+    }{
+      \Gamma_1,x:T,\Gamma_2\vdash T_2 \tpsynth \abs{\Pi}{x_1}{T_1}{\kappa}
+      & \Gamma_1,x:T,\Gamma_2\vdash t_1 \tpcheck T_1
+    }
+  \]
+
+  By the IH, \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]T_2 \tpsynth
+  \abs{\Pi}{x_1}{[t/x]T_1}{[t/x]\kappa}\) and \(\Gamma_1,[t/x]\Gamma_2 \vdash
+  [t/x]t_1 \tpcheck [t/x]T_1\).
+  By the rule, we have \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]T_2\ [t/x]t_1
+  \tpsynth [[t/x]t_1/x_1][t/x]\kappa\), where the synthesized kind is equal to
+  the desired \([t/x][t_1/x_1]\kappa\).
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,x:T,\Gamma_2\vdash \{ t_1 \simeq t_2 \} : \star
+    }{
+      \textit{FV}(t_1\ t_2)\subseteq\textit{dom}(\Gamma_1,x:T,\Gamma_2)
+    }
+  \]
+  It suffices to show that \(\textit{FV}([t/x]t_1\ [t/x]t_2) \subseteq
+  \textit{dom}(\Gamma_1,[t/x]\Gamma_2)\).
+  It is clear \(x\) is not a free variable of this expression, that the free
+  variables of \(t\) are declared in \(\Gamma_1\), and by assumption the other
+  free variables of it are declared in \(\Gamma_1,[t/x]\Gamma_2\).
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma_1,x:T,\Gamma_2 \vdash x_1\tpsynth T_1}{(x_1 : T_1) \in \Gamma_1,x:T,\Gamma_2}
+  \]
+  We elaborate on the case where \(x_1 = x\).
+  It is important that we assumed that \(\Gamma_1 \vdash t \tpsynth T\) (as
+  opposed to having its type checked), as we may now replace the given rule with
+  this assumed derivation without changing the definition of substitution.
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma_1,x:T,\Gamma_2\vdash \beta\{t'\} \tpcheck \{ t_1 \simeq t_2 \}}
+    {\textit{FV}(t')\subseteq \textit{dom}(\Gamma_1,x:T,\Gamma_2) \quad |t_1| =_{\beta\eta} |t_2|}
+  \]
+
+  From our assumptions we may conclude \(\textit{FV}([t/x]t) \subseteq
+  \textit{dom}(\Gamma_1,[t/x]\Gamma_2)\), and from the second premise that
+  \(|[t/x]t_1| =_{\beta\eta} |[t/x]t_2|\).
+  Thus, \(\Gamma_1,[t/x]\Gamma_2\ \vdash \beta\{[t/x]t'\}\tpcheck \{[t/x]t_1
+  \simeq [t/x]t_2\}\).
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma_1,x:T,\Gamma_2 \vdash \rho\ t''\ @x_1.T_1\ \mhyph\ t' \tpsynth [t_2/x_1]T_1}
+    {
+      \begin{array}{cc}
+        \Gamma_1,x:T,\Gamma_2 \vdash t'' \tpsynth \{t_1 \simeq t_2\}
+        & \quad \Gamma_1,x:T,\Gamma_2 \vdash [t_2/x_1]T_1 \tpsynth \star
+        \\ \Gamma_1,x:T,\Gamma_2 \vdash t' \tpsynth T'
+        & T' \cong [t_1/x_1]T_1
+      \end{array}
+    }
+  \]
+  From the IH, we have that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]t'' \tpsynth \{[t/x]t_1
+  \simeq [t/x]t_2\}\), that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x][t_2/x]T_1
+  \tpsynth \star\), and that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]t' \tpsynth
+  [t/x]T'\).
+  From the last premise, we may conclude that \([t/x]T' \cong
+  [t/x][t_1/x_1]T_1\).
+  Applying the rule and permuting substitutions gives us the desired result
+  (note \([t/x][t_1/x_1]T_1 = [[t/x]t_1/x_1]T_1\)).
+\end{proof}
+
+\begin{corollary}
   \ \\
   \begin{itemize}
-  \item If \(\Gamma \vdash T \tpsynth \kappa\) then \(\Gamma \vdash \kappa\)
-
-  \item If \(\Gamma \vdash t \ T\) then \(\Gamma \vdash T \tpsynth
-    \kappa\) for some \(\kappa\)
+  \item If \(\vdash \Gamma_1,x:T,\Gamma_2\) and \(\Gamma_1 \vdash t \tpsynth T\)
+    then \(\vdash \Gamma_1,[t/x]\Gamma_2\)
+
+  \item If \(\vdash \Gamma_1,X:\kappa,\Gamma_2\) and \(\Gamma_1 \vdash T
+    \tpsynth \kappa\) then \(\vdash \Gamma_1,[T/X]\Gamma_2\).
   \end{itemize}
-\end{lemma}
-
-\begin{lemma}
-  \label{lem:subst-kind}
-  \ \\
-  \begin{itemize}
-  \item If \(\Gamma,x:T \vdash T \tpsynth \kappa\) and \(\Gamma \vdash t
-    \tpcheck T\) then \(\Gamma \vdash [t/x]T \tpsynth [t/x]\kappa\)
-
-  \item If \(\Gamma,X:\kappa_2 \vdash T_1 \tpsynth \kappa_1\) and \(\Gamma
-    \vdash T_2 \tpsynth \kappa_2\) then \(\Gamma \vdash [T_2/X]T_1 \tpsynth [T_2/X]\kappa_1\)
-  \end{itemize}
-\end{lemma}
+\end{corollary}
+\begin{proof}
+  By induction on the assumed derivation, appealing to
+  Lemma~\ref{lem:subst-class} at each step.
+\end{proof}
 
@@
   \([t/x]T''\). By inversion of the kinding derivation, we have \(\Gamma,x:T_1
-  \vdash T'' \tpsynth \kappa_1\). By Lemma~\ref{lem:subst-kind}, we have
+  \vdash T'' \tpsynth \kappa_1\). By Lemma~\ref{lem:subst-class}, we have
   \(\Gamma \vdash [t/x]T'' \tpsynth [t/x]\kappa_1\).
@@
   \(\Gamma,X:\kappa_2 \vdash T_1'' \tpsynth \kappa_1\). By
-  Lemma~\ref{lem:subst-kind}, we have \(\Gamma \vdash [T_2/X]T_1'' \tpsynth
+  Lemma~\ref{lem:subst-class}, we have \(\Gamma \vdash [T_2/X]T_1'' \tpsynth
   [T_2/X]\kappa_1\).

### 2020-12-25 docs/semantics/paper.tex
docs: semantics: minor
- context formation
- types can't have their kinds checked

@@
 The classification rules are given in
-Figures~\ref{fig:superknd},~\ref{fig:knd}, and~\ref{fig:tp}.  For
-brevity, we take these figures as implicitly specifying the syntax of
-kinds $\kappa$, types $T$, and annotated terms $t$; these may use term
-variables $x$ and type variables $X$, which we assume come from
+Figures~\ref{fig:superknd},~\ref{fig:knd}, and~\ref{fig:tp}, with
+Figure~\ref{fig:ctxt} giving the context formation rules.
+For brevity, we take these figures as implicitly specifying the syntax of
+contexts \(\Gamma\), kinds $\kappa$, types $T$, and annotated terms $t$; these
+may use term variables $x$ and type variables $X$, which we assume come from
 distinct sets.  So terms and types are syntactically distinguished.
@@
     \infer{\Gamma\vdash\abs{\iota}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} &
-    \infer{\Gamma\vdash \{ t \simeq t' \} : \star}{\textit{FV}(t\ t')\subseteq\textit{dom}(\Gamma)}
+    \infer{\Gamma\vdash \{ t \simeq t' \} \tpsynth \star}{\textit{FV}(t\ t')\subseteq\textit{dom}(\Gamma)}
   \end{array}
@@
     \infer{\Gamma\vdash t \cdot T' \tpsynth [T'/X]T}
-          {\Gamma\vdash t \tpsynth \abs{\forall}{X}{\kappa}{T} & \Gamma\vdash T' \tpcheck\kappa} \\ \\
+          {\Gamma\vdash t \tpsynth \abs{\forall}{X}{\kappa}{T} & \Gamma\vdash T' \tpsynth \kappa' & \kappa'\cong\kappa} \\ \\
 

### 2020-12-25 docs/semantics/paper.tex
docs: semantics: lemma for convertible kinds in contexts

@@
 \begin{lemma}
+  \label{lem:ctxt-conv-class}
+  Let \(\kappa_1,\kappa_1'\) be kinds such that \(\kappa_1 \cong \kappa_1'\).
+  \begin{itemize}
+  \item If \(\Gamma_1,X:\kappa_1,\Gamma_2 \vdash \kappa_2\) and \(\Gamma_1
+    \vdash \kappa_1'\) then \(\Gamma_1,X:\kappa_1',\Gamma_2 \vdash \kappa_2\)
+
+  \item If \(\Gamma_1,X:\kappa_1,\Gamma_2 \vdash T \tpsynth \kappa_2\) and
+    \(\Gamma_1 \vdash \kappa_1'\) then \(\Gamma_1,X:\kappa_1',\Gamma_2 \vdash T
+    \tpsynth \kappa_2\) for some \(\kappa_2' \cong \kappa_2\)
+
+  \item If \(\Gamma_1,X:\kappa_1,\Gamma_2 \vdash t \Leftrightarrow T\) and
+    \(\Gamma_1 \vdash \kappa_1'\) then \(\Gamma_1,X:\kappa_1',\Gamma_2 \vdash t
+    \Leftrightarrow T\).
+  \end{itemize}
+\end{lemma}
+
+\begin{proof}
+  By induction on the assumed derivation.
+  We show a few interesting cases.
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash X \tpsynth \kappa}{(X : \kappa) \in \Gamma_1,X_1:\kappa_1,\Gamma_2}
+  \]
+  We have two subcases to consider.
+  If \(X = X_1\), then by the rule we obtain \(\Gamma_1,X_1:\kappa_1',\Gamma_2
+  \vdash X_1 \tpsynth \kappa_1'\), as desired.
+  Otherwise, by the rule we obtain \(\Gamma_1,X_1:\kappa_1',\Gamma_2
+  \vdash X \tpsynth \kappa\).
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash \abs{\lambda}{X}{\kappa}{T'} \tpsynth
+      \abs{\Pi}{X}{\kappa}{\kappa'}
+    }{
+      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash \kappa
+      & \Gamma_1,X_1:\kappa_1,\Gamma_2,X:\kappa \vdash  T' \tpsynth \kappa'
+    }
+  \]
+  By the IH on the first premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash
+  \kappa\).
+  By the IH on the second premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2,X:\kappa
+  \vdash T' \tpsynth \kappa''\) for some \(\kappa'' \cong \kappa'\).
+  By the rule, \(\Gamma_1,X_1\kappa_1',\Gamma_2 \vdash
+  \abs{\lambda}{X}{\kappa}{T'} \tpsynth \abs{\Pi}{X}{\kappa}{\kappa''}\), which
+  is convertible with the original kind synthesized for this type.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash T \cdot T' \tpsynth
+      [T'/X]\kappa
+    }{
+      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash T \tpsynth
+      \abs{\Pi}{X}{\kappa'}{\kappa}
+      & \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash T' \tpsynth \kappa''
+      & \kappa' \cong \kappa''}
+  \]
+  By the IH on the first premise, \(\Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash T
+  \tpsynth \abs{\Pi}{X}{\kappa_2'}{\kappa_2}\) for some \(\kappa_2' \cong
+  \kappa'\), \(\kappa_2 \cong \kappa\).
+  By the IH on the second premise, \(\Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash T'
+  \tpsynth \kappa_2''\) for some \(\kappa_2'' \cong \kappa''\).
+  By transitivity, \(\kappa_2' \cong \kappa_2''\).
+  By the rule, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash T \cdot T' \tpsynth
+  [T'/X]\kappa_2\), with this kind congruent to \([T'/X]\kappa\).
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \cdot T' \tpsynth [T'/X]T
+    }{
+      \Gamma_1,X_1:\kappa_1,\Gamma_2\vdash t \tpsynth
+      \abs{\forall}{X}{\kappa}{T}
+      & \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash T' \tpsynth \kappa'
+      & \kappa'\cong\kappa
+    }
+  \]
+  By the IH on the first premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t
+  \tpsynth \abs{\forall}{X}{\kappa}{T}\).
+  By the IH on the second premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash T'
+  \tpsynth \kappa''\) for some \(\kappa'' \cong \kappa'\).
+  With the third premise and transitivity of congruence, we have \(\kappa''
+  \cong \kappa\).
+  We apply the rule and conclude.
+\end{proof}
+
+\begin{corollary}
+  \label{cor:ctxt-conv-class}
+  If \(\vdash \Gamma_1,X:\kappa_1,\Gamma_1\) and \(\Gamma_1 \vdash \kappa_1'\)
+  with \(\kappa_1' \cong \kappa\) then \(\vdash \Gamma_1,X:\kappa_1',\Gamma_2\).
+\end{corollary}
+
+\begin{lemma}
   \label{lem:subst-class}

### 2020-12-25 docs/semantics/paper.tex
docs: semantics: modify substitution for bidirectional rules

@@
 \(\tpcheck\) or all as \(\tpsynth\)).
+Since term variables synthesize their types but the usual notion of substitution may
+replace these with terms whose types can only be checked, we introduce a
+short-hand notation: let \([t/x]^T\) stand for \([\chi\ T\ \mhyph\ t/x]\),
+where the \(\chi\) construct gives an explicit type annotation (see
+Section~\ref{sec:overview}).
+
 We follow the syntax of our implementation
@@
     \infer{\Gamma\vdash\abs{\lambda}{X}{\kappa}{T'} \tpsynth \abs{\Pi}{X}{\kappa}{\kappa'}}{\Gamma \vdash \kappa & \Gamma,X:\kappa\vdash T'\tpsynth\kappa'} \\ \\
-    \infer{\Gamma\vdash T\ t \tpsynth [t/x]\kappa}{\Gamma\vdash T \tpsynth \abs{\Pi}{x}{T'}{\kappa} & \Gamma\vdash t \tpcheck T'} &
+    \infer{\Gamma\vdash T\ t \tpsynth [t/x]^{T'}\kappa}{\Gamma\vdash T \tpsynth \abs{\Pi}{x}{T'}{\kappa} & \Gamma\vdash t \tpcheck T'} &
     \infer{\Gamma\vdash T_1 \cdot T_2 \tpsynth [T_2/X]\kappa_1}{\Gamma\vdash T_1 \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1} & \Gamma\vdash T_2 \tpsynth \kappa_2' & \kappa_2 \cong \kappa_2'} \\ \\
@@
   \[
+    \begin{array}{c}
   \begin{array}{ll}
@@
     \infer{\Gamma\vdash \absu{\lambda}{x}{t} \tpcheck \abs{\Pi}{x}{T}{T'}}{\Gamma,x:T\vdash t\tpcheck T'} &
-    \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]T}{\Gamma\vdash t \tpsynth \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\
+    \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynth \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\
 
@@
     \infer{\Gamma\vdash \absu{\Lambda}{x}{t} \tpcheck \abs{\forall}{x}{T'}{T}}{\Gamma,x:T'\vdash t \tpcheck T & x\not\in\textit{FV}(|t|)} &
-    \infer{\Gamma\vdash t\ \mhyph t' \tpsynth [t'/x]T}{\Gamma\vdash t \tpsynth \abs{\forall}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\
+    \infer{\Gamma\vdash t\ \mhyph t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynth \abs{\forall}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\
 
     \infer{\Gamma\vdash [ t , t' ] \tpcheck \abs{\iota}{x}{T}{T'}}
-          {\Gamma\vdash t \tpcheck T & \Gamma\vdash t' \tpcheck [t/x]T' & |t| =_{\beta\eta} |t'|} &
+          {\Gamma\vdash t \tpcheck T & \Gamma\vdash t' \tpcheck [t/x]^TT' & |t| =_{\beta\eta} |t'|} &
     \infer{\Gamma\vdash t.1 \tpsynth T}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} \\ \\
-    \infer{\Gamma\vdash t.2 \tpsynth [t.1/x]T'}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} &
+    \infer{\Gamma\vdash t.2 \tpsynth [t.1/x]^TT'}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} &
     \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t_1 \simeq t_2 \}}
@@
             \Gamma \vdash t' \tpsynth \{t_1 \simeq t_2\}
-            & \quad \Gamma \vdash [t_2/x]T \tpsynth \star
+            & \quad \Gamma \vdash [\beta\{t_2\}/x]^{\texttt{Top}}\ T \tpsynth \star
             \\ \Gamma \vdash t \tpsynth T'
@@
           {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}
+  \end{array}
+      \\ \\
+      \texttt{Top} = \{\absu{\lambda}{x}{x} \simeq \absu{\lambda}{x}{x} \}
     \end{array}
@@
   terms, non-normalizing terms, etc. --- prove trivially true equations.
+  Put another way, any trivially true equation type in CDLE is a suitable type
+  to classify all untyped lambda calculus terms.
+  We define such a type \texttt{Top} at the bottom of Figure~\ref{fig:tp}, which
+  is used in the typing of the next construct we discuss, \(\rho\).
 \item The \(\rho\) construct allows one to rewrite occurrences of \(t_1\) to
@@
   indicates the occurrences of \(t_1\) to substitute.
-  \([t_2/x]T\) must be a well-kinded type, but it is not required that
+  \([t_2/x]^{\texttt{Top}}\ T\) must be a well-kinded type, but it is not required that
   \([t_1/x]T\) also be, only that \(T'\) and \([t_1/x]T\) are convertible modulo
@@
   is well-kinded whose details are beyond the scope of this document.
+
+  It might appear at first that the requirement that the guide \(T\) be written
+  such that the resulting substitution is well-kinded is too strict.
+  For example, what if we are trying to rewrite \(n\) in \(X\ n\), where \(X\)
+  is a predicate on (Church-encoded) naturals?
+  The next construct, \(\varphi\), gives us the ability to recover the typing
+  information associated to \(n\).
 \item We adopt a strong form of Nuprl's \textbf{direct computation rules}~\cite{constable+86}:

### 2020-12-26 docs/semantics/paper.tex
docs: semantics: minor
- typed substitution (and fix rho)
- fix a case in semantic soundness proof (types can't have their kinds checked)

@@
     \infer{\Gamma\vdash x\tpsynth T}{(x : T)\in\Gamma} &
-    \infer{\Gamma\vdash t\tpcheck T'}{\Gamma\vdash t \tpcheck T & T' \leadsto^*_\beta T} \\ \\    
+    \infer{\Gamma\vdash t\tpcheck T'}{\Gamma\vdash t \tpcheck T & T' \leadsto^*_\beta T'} \\ \\
     \infer{\Gamma\vdash t\tpsynth T'}{\Gamma\vdash t \tpsynth T & T \leadsto^*_\beta T'} &
@@
     \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
-    \infer{\Gamma \vdash \rho\ t'\ @x.T\ \mhyph\ t \tpsynth [t_2/x]T}
+    \infer{\Gamma \vdash \rho\ t'\ @x.T\ \mhyph\ t \tpsynth [\beta\{t_2\}/x]^{\texttt{Top}}\ T}
           {
@@
   indicates the occurrences of \(t_1\) to substitute.
-  \([t_2/x]^{\texttt{Top}}\ T\) must be a well-kinded type, but it is not required that
+  \([\beta\{t_2\}/x]^{\texttt{Top}}\ T\) must be a well-kinded type, but it is not required that
   \([t_1/x]T\) also be, only that \(T'\) and \([t_1/x]T\) are convertible modulo
@@
 \[
-   \infer{\Gamma\vdash T\ t \tpsynth [t/x]\kappa}{\Gamma\vdash T \tpsynth \abs{\Pi}{x}{T'}{\kappa} & \Gamma\vdash t \tpcheck T'} 
+   \infer{\Gamma\vdash T\ t \tpsynth [t/x]^{T'}\ \kappa}{\Gamma\vdash T \tpsynth
+     \abs{\Pi}{x}{T'}{\kappa} & \Gamma\vdash t \tpcheck T'}  
 \]
@@
 \[
-    \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]T}{\Gamma\vdash t \tpsynth \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} 
+    \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]^{T'} T}{\Gamma\vdash t \tpsynth \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} 
 \]
@@
     \infer{\Gamma\vdash t \cdot T' \tpsynth [T'/X]T}
-          {\Gamma\vdash t \tpsynth \abs{\forall}{X}{\kappa}{T} & \Gamma\vdash T' \tpcheck\kappa} 
+    {\Gamma\vdash t \tpsynth \abs{\forall}{X}{\kappa}{T}
+      & \Gamma\vdash T' \tpsynth \kappa'
+      & \kappa' \cong \kappa
+    } 
 \]
@@
 premise, we have $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho[X\mapsto
-  S]}$, for all $S\in\interp{\kappa}_{\sigma,\rho}$.  Since
-$\interp{T'}_{\sigma,\rho}\in\interp{\kappa}_{\sigma,\rho}$ by the IH
-applied to the second premise, we can derive $[\sigma t]_{\cbe}\in\interp{T}_{\sigma,\rho[X\mapsto \interp{T'}_{\sigma,\rho}]}$.
+  S]}$, for all $S\in\interp{\kappa}_{\sigma,\rho}$.
+By applying the IH twice, once to the second premise and once to the third, we
+have, we have \(\interp{T'}_{\sigma,\rho} \in \interp{\kappa}_{\sigma,\rho}\).
+So, can derive $[\sigma t]_{\cbe}\in\interp{T}_{\sigma,\rho[X\mapsto \interp{T'}_{\sigma,\rho}]}$.
 By Lemma~\ref{lem:tpsubstinterp},
@@
 \[
-    \infer{\Gamma\vdash t\ \mhyph t' \tpsynth [t'/x]T}{\Gamma\vdash t \tpsynth \abs{\forall}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'}
+    \infer{\Gamma\vdash t\ \mhyph t' \tpsynth [t'/x]^{T'}\ T}{\Gamma\vdash t \tpsynth \abs{\forall}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'}
 \]
@@
     \infer{\Gamma\vdash [ t , t' ] \tpcheck \abs{\iota}{x}{T}{T'}}
-          {\Gamma\vdash t \tpcheck T & \Gamma\vdash t' \tpcheck [t/x]T' & |t| =_{\beta\eta} |t'|} 
+          {\Gamma\vdash t \tpcheck T & \Gamma\vdash t' \tpcheck [t/x]^{T}\ T' & |t| =_{\beta\eta} |t'|} 
 \]
@@
 \section{Proof of Theorem~\ref{thm:cedille-termination}}
-\begin{proof}
+\begin{proof}[Proof (of Theorem~\ref{thm:cedille-termination})]
 Theorem \ref{thm:snd} implies that since
@@
 call-by-name normalizing (cf.~\cite{Kashima2000}).
-
 \end{proof}
@@
   well-formed.
-  We also assume types in type checking judgments are well-formed under the
-  given context.
   \begin{enumerate}
@@
       \(\Gamma_1 \vdash t' \tpsynth T'\) then \(\Gamma_1,[t'/x]\Gamma_2 \vdash [t'/x]t
-      \Leftrightarrow [t'/x]T\)
+      \tpsynth [t'/x]T\)
 
@@
       \(\Gamma_1 \vdash T \tpsynth \kappa\) then \(\Gamma_1,[T/X]\Gamma_2 \vdash
-      [T/X]t \Leftrightarrow [T/X]T'\)
+      [T/X]t \tpsynth [T/X]T'\)
     \end{itemize}
@@
   By mutual induction on the assumed derivations.
-  We only show a few interesting cases.
+  We only show a few interesting cases, and we omit type annotations on
+  substitutions when these are clear from context.
 
@@
   \startcase{.2cm}
-  \[ \infer{\Gamma_1,x:T,\Gamma_2\vdash T_2\ t_1 \tpsynth [t_1/x_1]\kappa
+  \[ \infer{\Gamma_1,x:T,\Gamma_2\vdash T_2\ t_1 \tpsynth [t_1/x_1]^{T_1}\kappa
     }{
@@
   \[
-    \infer{\Gamma_1,x:T,\Gamma_2 \vdash \rho\ t''\ @x_1.T_1\ \mhyph\ t' \tpsynth [t_2/x_1]T_1}
+    \infer{\Gamma_1,x:T,\Gamma_2 \vdash \rho\ t''\ @x_1.T_1\ \mhyph\ t' \tpsynth [\beta\{t_2\}/x_1]^{\texttt{Top}}\ T_1}
     {
@@
         \Gamma_1,x:T,\Gamma_2 \vdash t'' \tpsynth \{t_1 \simeq t_2\}
-        & \quad \Gamma_1,x:T,\Gamma_2 \vdash [t_2/x_1]T_1 \tpsynth \star
+        & \quad \Gamma_1,x:T,\Gamma_2 \vdash [\beta\{t_2\}/x_1]^{\texttt{Top}}\ T_1 \tpsynth \star
         \\ \Gamma_1,x:T,\Gamma_2 \vdash t' \tpsynth T'
@@
   From the IH, we have that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]t'' \tpsynth \{[t/x]t_1
-  \simeq [t/x]t_2\}\), that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x][t_2/x]T_1
+  \simeq [t/x]t_2\}\), that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x][\beta\{t_2\}/x]T_1
   \tpsynth \star\), and that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]t' \tpsynth
@@
   Applying the rule and permuting substitutions gives us the desired result
-  (note \([t/x][t_1/x_1]T_1 = [[t/x]t_1/x_1]T_1\)).
+  (note for example that \([t/x][t_1/x_1]T_1 = [[t/x]t_1/x_1]T_1\)).
 \end{proof}

### 2020-12-26 docs/semantics/paper.tex
docs: semantics: in CBN normalization, make context empty
The original can be recovered by weakening

@@
   \label{thm:cedille-termination}
-  Suppose \(\Gamma\vdash t : T\), $t$ is closed, and there exists a closed term
-  $t'$ which erases to \(\lambda x.\ x\) and whose type is $T \to \Pi x : T_1.\
-  T_2$ for some $T_1, T_2$. Then $|t|$ is call-by-name normalizing.
+  Suppose \(\cdot \vdash t \tpsynth T\) and there exists a closed term \(t'\)
+  such that \(\cdot \vdash t \tpsynth T \to \abs{\Pi}{x}{T_1}{T_2}\) for some
+  \(T_1,T_2\) and \(|t'| = \absu{\lambda}{x}{x}\).
+  Then \(|t|\) is call-by-name normalizing.
 \end{theorem}

### 2020-12-26 docs/semantics/paper.tex
docs: semantics: prove judgment validity (aka agreement)

@@
   \cong \kappa'\).
+\end{theorem}
+
+From this theorem and a few other lemmas (see
+Appendix~\ref{sec:proof-syntactic-kind-pres}), we can show the \emph{validity}
+(or \emph{agreement}) of the judgments comprising CDLE.
+
+\begin{theorem}[Judgment validity]
+  \label{thm:judge-valid}
+  If \(\vdash \Gamma\) then:
+  \begin{itemize}
+  \item if \(\Gamma \vdash T \tpsynth \kappa\) then \(\Gamma \vdash \kappa\)
+
+  \item if \(\Gamma \vdash t \tpsynth T\) then \(\Gamma \vdash T \tpsynth \star\)
+  \end{itemize}
 \end{theorem}
@@
 
-\section{Proof of Theorem~\ref{thm:syntactic-kind-pres}}
+\section{Proof of Theorems~\ref{thm:syntactic-kind-pres} and \ref{thm:judge-valid}}
 \label{sec:proof-syntactic-kind-pres}
@@
   for types).
-
-  \startcase{.2cm}
-  \[
-    \infer{\Gamma\vdash T\ t \tpsynth [t/x]\kappa_1}{\Gamma\vdash T \tpsynth
+  We omit type annotations from substitutions when they are clear from the context.
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma\vdash T\ t \tpsynth [t/x]^{T_1}\kappa_1}{\Gamma\vdash T \tpsynth
       \abs{\Pi}{x}{T_1}{\kappa_1} & \Gamma\vdash t \tpcheck T_1}
@@
   [T_2/X]\kappa_1\).
-
 \end{proof}
+
+\begin{proof}[Proof (of Theorem~\ref{thm:judge-valid})]
+  By induction on the assumed derivation.
+  We show a few interesting cases.
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma \vdash X \tpsynth \kappa}{(X : \kappa) \in \Gamma}
+  \]
+  By an easy inductive argument on the assumption \(\vdash \Gamma\) and
+  weakening, we obtain \(\Gamma \vdash \kappa\).
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma\vdash\abs{\lambda}{x}{T}{T'} \tpsynth \abs{\Pi}{x}{T}{\kappa}}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T'\tpsynth\kappa}
+  \]
+  By assumption, the first premise, and the context formation rules, we have
+  \(\vdash \Gamma,x:T\).
+  By the IH, we have \(\Gamma,x:T \vdash \kappa\).
+  Thus we obtain \(\Gamma \vdash \abs{\Pi}{x}{T}{\kappa}\).
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma\vdash T_1 \cdot T_2 \tpsynth [T_2/X]\kappa_1}{\Gamma\vdash T_1 \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1} & \Gamma\vdash T_2 \tpsynth \kappa_2' & \kappa_2 \cong \kappa_2'}
+  \]
+
+  By the IH on the first premise, we have \(\Gamma \vdash
+  \abs{\Pi}{X}{\kappa_2}{\kappa_1}\), and by inversion this gives us \(\Gamma
+  \vdash \kappa_2\), which yields \(\vdash \Gamma,X:\kappa_2\), and
+  \(\Gamma,X:\kappa_2 \vdash \kappa_1\).
+  By the IH on the second premise, \(\Gamma \vdash \kappa_2'\).
+  By Lemma~\ref{lem:ctxt-conv-class} using the third premise,
+  \(\Gamma,X:\kappa_2' \vdash \kappa_1\).
+  By Lemma~\ref{lem:subst-class}, \(\Gamma \vdash [T_2/X]\kappa_1\).
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma\vdash t\tpsynth T'}{\Gamma\vdash t \tpsynth T & T' \leadsto^*_\beta T'}
+  \]
+  By the IH, \(\Gamma \vdash T \tpsynth \star\).
+  By generalizing Theorem~\ref{thm:syntactic-kind-pres} to \(\leadsto^*_\beta\),
+  we have \(\Gamma \vdash T' \tpsynth \kappa'\) for some \(\kappa' \cong
+  \star\).
+  This means that in fact \(\kappa' = \star\), as desired.
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynth \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'}
+  \]
+  By the IH on the first premise, \(\Gamma \vdash \abs{\Pi}{x}{T'}{T} \tpsynth
+  \star\).
+  By inversion of this, we obtain both \(\Gamma \vdash T' \tpsynth \star\), from
+  which it follows by the \(\chi\) rule and second premise that \(\Gamma \vdash
+  \chi\ T'\ \mhyph\ t \tpsynth T'\), and that \(\Gamma,x:T' \vdash T \tpsynth
+  \star\), from which we obtain \(\Gamma \vdash [t/x]^{T'}\ T \tpsynth \star\)
+  using Lemma~\ref{lem:subst-class}
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma \vdash \rho\ t'\ @x.T\ \mhyph\ t \tpsynth [\beta\{t_2\}/x]^{\texttt{Top}}\ T}
+    {
+      \begin{array}{cc}
+        \Gamma \vdash t' \tpsynth \{t_1 \simeq t_2\}
+        & \quad \Gamma \vdash [\beta\{t_2\}/x]^{\texttt{Top}}\ T \tpsynth \star
+        \\ \Gamma \vdash t \tpsynth T'
+        & T' \cong [t_1/x]T
+      \end{array}
+    }    
+  \]
+  The desired result is given by the second assumption (reading first left to right, then top to bottom).
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma \vdash \chi\ T\ \mhyph\ t\ \tpsynth T}
+    {\Gamma \vdash T \tpsynth \star & \Gamma \vdash t \tpcheck T}
+  \]
+  Give to us by assumption (the first premise).
+\end{proof}
+
 \end{document}

### 2020-12-30 docs/semantics/paper.tex
docs: semantics: rework of inference rules
- make deterministic, needed for the well-foundedness arguments in future proof
- \rho requires one more piece of information in the anntation, needed for
  re-typeability when the type of a term variable in context is exchanged with
  a convertible type
- make \varphi more permissive, since CBN reduction in types can't promise a
  synthesized equation mentions a type
- fix some erasure rules
- update semantic soundness proof with new rules

@@
 \newcommand{\tpsynth}[0]{\Rightarrow}
+\newcommand{\tpsynthleads}[0]{\ensuremath{\overset{\leadsto_\beta^*}{\Rightarrow}}}
 \newcommand{\cbe}[0]{c\beta\eta}
@@
 \newcommand{\startcase}[1]{\vspace{#1} \noindent\textbf{\underline{Case:}}}
-
+\newcommand{\startsubcase}[1]{\vspace{#1} \noindent\textbf{\underline{Subcase:}}}
 \newtheorem{theorem}{Theorem}
@@
 mostly in the form of dropping constructs we discovered (to our
-surprise) could be derived~\cite{stump18}. I have also omitted
+surprise) could be derived~\cite{stump18}. We have also omitted
 \emph{lifting} -- a technique for large eliminations with lambda
@@
 constructs, which allow deriving induction principles within the
-theory, for inductive datatypes.  The goal is to support usual
+theory for inductive datatypes.  The goal is to support usual
 idioms of dependently typed programming and proving as in Agda or
@@
 distinct sets.  So terms and types are syntactically distinguished.
-The typing rules (Figure~\ref{fig:tp}) are
-bidirectional~\cite{pierce+00}, while the kinding and superkinding
-rules (Figure~\ref{fig:knd} and~\ref{fig:superknd}) are only
-synthesizing.  We write $\Leftrightarrow$ to range over
-$\{\tpcheck,\tpsynth\}$, and when this symbol occurs multiple times in a rule,
-it is intended that such occurrences be read the same way (i.e., either all as
-\(\tpcheck\) or all as \(\tpsynth\)).
-Since term variables synthesize their types but the usual notion of substitution may
-replace these with terms whose types can only be checked, we introduce a
-short-hand notation: let \([t/x]^T\) stand for \([\chi\ T\ \mhyph\ t/x]\),
-where the \(\chi\) construct gives an explicit type annotation (see
-Section~\ref{sec:overview}).
-
 We follow the syntax of our implementation
@@
 be a term).
-The rules are intended, with a few points of nondeterminism, to be read
-bottom-up (in a standard way;
-c.f.~\cite{peytonjones07,Pfe01_Lecture-Notes-on-Bidirectional-Type-Checking}) as
-an algorithm for synthesizing a classifier from a context and an expression
-($\tpsynth$) (\emph{type synthesis}) or checking an expression against a classifier in context
-($\tpcheck$) (\emph{type checking}).
-Specifically, nondeterminism arises in the two rules allowing both synthesized
-and checked types to be reduced, written \(T \leadsto^*_\beta T'\), so that type
-constructors required by other rules may be revealed.
-The reduction relation \(\leadsto_\beta\) is call-by-name normalization, as this
-is sufficient to reduce types to weak head normal form.
-
-The classification rules refer to an erasure function, defined in
-Figure~\ref{fig:eraser}.  The type theory is \emph{extrinsic} (aka,
-Curry-style), and hence we only consider erasures $|t|$ of terms when
-testing for $\beta\eta$-equivalence.  This applies also to the conversion
-relation $T\cong T'$ on types and \(\kappa \cong \kappa'\) on kinds, whose central rules are given in
-Figure~\ref{fig:conv}.  That figure omits the various congruence rules
-needed to equate bigger expressions by equating subexpressions.  The
-main ideas of conversion shown in the figure are to use
-$\beta$-equivalence at the type level, and $\beta\eta$-equivalence of
-erased terms at the term level.
-Note in particular that for conversion we check full \(\beta\) and
-\(\beta\eta\)-equivalence, not just equivalence with respect to call-by-name
-normalization.
+Note that center dot (\(\cdot\)) is also used to denote the empty type context;
+since the usage for typing contexts always has the symbol occur to the left of
+the turnstile (\(\vdash\)), no confusion should arise from overloading notation.
+
+The typing rules (Figure~\ref{fig:tp}) are
+bidirectional~\cite{pierce+00}, while the kinding and superkinding
+rules (Figure~\ref{fig:knd} and~\ref{fig:superknd}) are unidirectional
+(i.e., synthesizing only).
+We write $\Leftrightarrow$ to range over
+$\{\tpcheck,\tpsynth\}$, and when this symbol occurs multiple times in a rule,
+it is intended that such occurrences be read the same way (i.e., read the
+occurrences as either all \(\tpcheck\) or all \(\tpsynth\)).
+The rules are intended to be read bottom-up as an algorithm (in a standard way,
+c.f.~\cite{peytonjones07,Pfe01_Lecture-Notes-on-Bidirectional-Type-Checking})
+for synthesizing a classifier from a context and an expression (\emph{type
+  synthesis}, \(\tpsynth\)) or checking an expression against a classifier in a
+context (\emph{type checking}, \(\tpcheck\)).
+Since variables have their type synthesized, and since we sometimes wish to
+substitute a variable with a term whose type can only be checked, we define a
+shorthand notation: \([t/x]^T\) means \([\chi\ T\ \mhyph\ t/x]\), where \(\chi\)
+is the construct for explicit type annotations (see Section~\ref{sec:overview}).
+
+In terms, when the type of an introduction form is checked or the type of an
+elimination form is synthesized, we use call-by-name normalization for types
+\(\leadsto_\beta^*\) to put types in weak head normal form, revealing type
+constructors.
+We abbreviate the conjuction ``term \(t\) synthesizes some type, and that type
+call-by-name reduces to another type'' with the symbol \(\tpsynthleads\),
+defined formally at the top of Figure~\ref{fig:tp}.
+When a redex (reducible expression) occurs such that the argument to a type is a
+term, such as \((\abs{\lambda}{x}{T_1}{T_2})\ t_1\), the reduction uses
+substitution with annotations to \([t_1/x]^{T_1}\ T_2\).
+
+Call-by-name reduction is \emph{not} what underpins the
+congruence relation \(\cong\) for types or terms, which is full
+\(\beta\)-equivalence (for types) and \(\beta\eta\)-equivalence (for terms)
+modulo erasure of annotations in terms.
+The erasure operation is defined in Figure~\ref{fig:eraser}, and is essentially
+the \emph{extraction} function for the Implicit Calculus of Constructions given
+by~\cite{BB08_ICC-as-a-Programming-Language} adapted to CDLE.
+To understand the role of erasure, recall that the type theory is \emph{extrinsic} (a.k.a.\
+Curry-style), and hence we only consider erasures \(|t|\) of terms when testing
+convertibility.
+This is lifted to the conversion relation on types \(T \cong T'\) and kinds
+\(\kappa \cong \kappa'\), whose central rules are given in
+Figure~\ref{fig:conv}.
+That figure omits the various congruence rules needed to equate bigger
+expressions by equating subexpressions (the convertibility rules for kinds
+consists entirely of congruence rules).
 
@@
   \[
-  \begin{array}{llll}
+  \begin{array}{ccc}
     \infer{\Gamma \vdash \star}{\ } &
@@
   \[
-  \begin{array}{ll}
+    \begin{array}{c}
+      [t/x]^T\ =\ [\chi\ T\ \mhyph\ t/x]
+      \\ \\ \\
+  \begin{array}{cc}
     \infer{\Gamma \vdash X \tpsynth \kappa}{(X : \kappa) \in \Gamma} &
@@
   \end{array}
+    \end{array}
   \]
@@
     \begin{array}{c}
-  \begin{array}{ll}
+      \Gamma \vdash t \tpsynthleads T\ =\ \exists T'.\ \Gamma
+      \vdash t \tpsynth T' \land T' \leadsto_\beta^* T
+      \\ \\ \\
+  \begin{array}{cc}
     \infer{\Gamma\vdash x\tpsynth T}{(x : T)\in\Gamma} &
-    \infer{\Gamma\vdash t\tpcheck T'}{\Gamma\vdash t \tpcheck T & T' \leadsto^*_\beta T'} \\ \\
-    \infer{\Gamma\vdash t\tpsynth T'}{\Gamma\vdash t \tpsynth T & T \leadsto^*_\beta T'} &
     \infer{\Gamma\vdash t\tpcheck T}{\Gamma\vdash t\tpsynth T' & T' \cong T} \\ \\
-    \infer{\Gamma\vdash \absu{\lambda}{x}{t} \tpcheck \abs{\Pi}{x}{T}{T'}}{\Gamma,x:T\vdash t\tpcheck T'} &
-    \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynth \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\
-
-    \infer{\Gamma\vdash \absu{\Lambda}{X}{t} \tpcheck \abs{\forall}{X}{\kappa}{T}}{\Gamma,X:\kappa\vdash t \tpcheck T} &
+    \infer{\Gamma\vdash \absu{\lambda}{x}{t} \tpcheck T}{T \leadsto_\beta^* \abs{\Pi}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t\tpcheck T_2} &
+    \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynthleads \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\
+
+    \infer{\Gamma\vdash \absu{\Lambda}{X}{t} \tpcheck T'}
+          {T' \leadsto_\beta^* \abs{\forall}{X}{\kappa}{T} & \Gamma,X:\kappa\vdash t \tpcheck T} &
     \infer{\Gamma\vdash t \cdot T' \tpsynth [T'/X]T}
-          {\Gamma\vdash t \tpsynth \abs{\forall}{X}{\kappa}{T} & \Gamma\vdash T' \tpsynth \kappa' & \kappa'\cong\kappa} \\ \\
-
-    \infer{\Gamma\vdash \absu{\Lambda}{x}{t} \tpcheck \abs{\forall}{x}{T'}{T}}{\Gamma,x:T'\vdash t \tpcheck T & x\not\in\textit{FV}(|t|)} &
-    \infer{\Gamma\vdash t\ \mhyph t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynth \abs{\forall}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\
-
-    \infer{\Gamma\vdash [ t , t' ] \tpcheck \abs{\iota}{x}{T}{T'}}
-          {\Gamma\vdash t \tpcheck T & \Gamma\vdash t' \tpcheck [t/x]^TT' & |t| =_{\beta\eta} |t'|} &
-    \infer{\Gamma\vdash t.1 \tpsynth T}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} \\ \\
-    \infer{\Gamma\vdash t.2 \tpsynth [t.1/x]^TT'}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} &
-    \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t_1 \simeq t_2 \}}
-          {\textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}  \\ \\    
-    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpcheck \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  &
-    \infer{\Gamma \vdash \rho\ t'\ @x.T\ \mhyph\ t \tpsynth [\beta\{t_2\}/x]^{\texttt{Top}}\ T}
+          {\Gamma\vdash t \tpsynthleads \abs{\forall}{X}{\kappa}{T} & \Gamma\vdash T' \tpsynth \kappa' & \kappa'\cong\kappa} \\ \\
+
+    \infer{\Gamma\vdash \absu{\Lambda}{x}{t} \tpcheck T}
+          {T \leadsto_\beta^* \abs{\forall}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t \tpcheck T_2 & x\not\in\textit{FV}(|t|)} &
+    \infer{\Gamma\vdash t\ \mhyph t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynthleads \abs{\forall}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\
+
+    \infer{\Gamma\vdash [ t_1 , t_2 ] \tpcheck T}
           {
            \begin{array}{cc}
-            \Gamma \vdash t' \tpsynth \{t_1 \simeq t_2\}
-            & \quad \Gamma \vdash [\beta\{t_2\}/x]^{\texttt{Top}}\ T \tpsynth \star
-            \\ \Gamma \vdash t \tpsynth T'
+             T \leadsto_\beta^* \abs{\iota}{x}{T_1}{T_1}
+             & \Gamma \vdash t_1 \tpcheck T_1
+             \\ \Gamma \vdash t_2 \tpcheck [t/x]^{T_1}\ T_2
+             & |t_1| =_{\beta\eta} |t_2|
+           \end{array}
+          } &
+    \infer{\Gamma\vdash t.1 \tpsynth T}{\Gamma\vdash t \tpsynthleads \abs{\iota}{x}{T}{T'}} \\ \\
+    \infer{\Gamma\vdash t.2 \tpsynth [t.1/x] T'}{\Gamma\vdash t \tpsynthleads \abs{\iota}{x}{T}{T'}} &
+    \infer{\Gamma\vdash \beta\{t'\} \tpcheck T}
+          {T \leadsto_\beta^* \{t_1 \simeq t_2\} \quad \textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}  \\ \\    
+    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}
+          {\Gamma\vdash t\tpcheck \{\absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}} &
+    \infer{\Gamma \vdash \rho\ t'\ @x\langle t_2 \rangle.T\ \mhyph\ t \tpsynth [t_2/x]\ T}
+          {
+          \begin{array}{ccc}
+            \Gamma \vdash t' \tpsynthleads \{t_1 \simeq t_2'\}
+            & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma)
+            & |t_2'| =_{\beta\eta} |t_2|
+            \\ \Gamma \vdash [t_2/x]\ T \tpsynth \star
+            & \Gamma \vdash t \tpsynth T'
             & T' \cong [t_1/x]T
@@
     \infer{\Gamma\vdash \varphi\ t\ \mhyph\ t'\ \{t''\} \Leftrightarrow T}
-          {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}
-  \end{array}
-      \\ \\
-      \texttt{Top} = \{\absu{\lambda}{x}{x} \simeq \absu{\lambda}{x}{x} \}
-    \end{array}
-  \]
-\caption{Rules for checking a term against a well-kinded type ($\Gamma \vdash t \tpcheck T$)
+          {
+            \begin{array}{ccc}
+              \Gamma\vdash t\tpsynthleads \{t_1\simeq t_2\}
+              &
+              & \Gamma\vdash t' \Leftrightarrow T
+              \\ |t_1| =_{\beta\eta} |t'|
+              & |t_2| =_{\beta\eta} |t''|
+              & \textit{FV}(t'') \subseteq \textit{dom}(\Gamma)
+            \end{array}
+          }
+  \end{array}
+    \end{array}
+  \]
+\caption{Rules for checking a term against a type ($\Gamma \vdash t \tpcheck T$)
            and synthesizing a type for a term ($\Gamma \vdash t \tpsynth T$)}
@@
     |x| & = & x &\ &
-    |\abs{\lambda}{x}{t'}{t}| & = & \absu{\lambda}{x}{|t|} \\
+    |\absu{\lambda}{x}{t}| & = & \absu{\lambda}{x}{|t|} \\
     |t\ t'| & = & |t|\ |t'| &\ &
     |t\cdot T| & = & |t| \\
-    |\abs{\Lambda}{x}{t'}{t}| & = & |t| &\ &
+    |\absu{\Lambda}{x}{t}| & = & |t| &\ &
     |t\ \mhyph t'| & = & |t| \\
@@
 type-level applications, we thus write $\abs{\Pi}{X}{\kappa}{\kappa'}$
-instead of $\abs{\forall}{X}{\kappa}{\kappa'}$.  We write $\lambda$ to
+instead of $\abs{\forall}{X}{\kappa}{\kappa'}$.
+For abstractions, we write $\lambda$ to
 correspond to $\Pi$ and $\Lambda$ to correspond to $\forall$.  As noted
@@
 for erased arguments to functions, found also in systems like Agda
-(cf.~\cite{mishraLinger08}).  Dependent intersections are a rather
-exotic construct allowing us to assign type $\abs{\iota}{x}{T'}{T}$ to
-erased term $t$ when we can assign $T'$ to $t$, and also assign
+(c.f.~\cite{mishraLinger08}).  Dependent intersections are a rather
+exotic construct allowing us to assign type $\abs{\iota}{x}{T_1}{T_2}$ to
+erased term $t$ when we can assign $T_1$ to $t$, and also assign
 $[t/x]T$ to $t$.  For an annotated introduction form, we write
-$[t,t'$], where $t$ checks against type $T'$, $t'$ checks against
-$[t/x]T$, and $t$ and $t'$ have $\alpha\beta\eta$-equivalent
+$[t_1,t_2$], where $t_1$ checks against type $T_1$, $t_2$ checks against
+$[t_1/x]^{T_1}T_2$, and $t_1$ and $t_2$ have $\beta\eta$-equivalent
 erasures.  Dependent intersections thus enable a controlled form of
@@
 
-The typing rules include conversion checks in a few places; e.g., as
-standardly, when switching from checking to synthesizing mode.  As mentioned
+The typing rules include conversion checks in a few places, e.g., as is
+standard when switching from checking to synthesizing mode.  As mentioned
 earlier, two rules near the top of Figure~\ref{fig:tp} state that one may
-(nondeterministically) $\beta$-reduce the type one is synthesizing or checking,
-before proceeding.
-This allows reduction to weak head normal form, to match the form of type
-required by other rules.
-Finally, we include the construct $\chi\ T\ \mhyph\ t$ which allows a term whose
-type can only be checked to be ascribed a user-provided type so that the whole
-expression may occur where type synthesis is expected.
-
-We have modified the rules for equality types $\{ t \simeq t' \}$ so that we require
-nothing of $t$ and $t'$ except that the set $\textit{dom}(\Gamma)$ of variables
-declared by $\Gamma$ includes their free variables $\textit{FV}(t\ t')$.  Further modifications
+(nondeterministically) call-by-name reduce the type one is synthesizing or
+checking, before proceeding.
+We also include the construct $\chi\ T\ \mhyph\ t$ which allows a term whose
+type can be checked to be ascribed a user-provided type so that the whole
+expression may occur where a term whose type can be synthesized is expected.
+
+We have modified the rules for equality types $\{ t_1 \simeq t_2 \}$ so that we require
+nothing of $t_1$ and $t_2$ except that the set $\textit{dom}(\Gamma)$ of variables
+declared by $\Gamma$ includes their free variables $\textit{FV}(t_1\ t_2)$.  Further modifications
 over the version of CDLE in~\cite{stump18} are:
@@
   terms that are \(\beta\eta\)-equivalent modulo erasure), one now writes $\beta\{t'\}$,
-  with the critical idea that $|\beta\{t'\}|$ erases to (a possibly unrelated) $|t'|$.
+  with the critical idea that $|\beta\{t'\}|$ erases to (the possibly unrelated) $|t'|$.
   We call this the \textbf{Kleene trick}, because it goes back to Kleene's numeric
@@
   context.
-  This means that in Cedille, any such term --- even otherwise untypable
+
+  The Kleene trick means that in Cedille, any such term --- even otherwise untypable
   terms, non-normalizing terms, etc. --- prove trivially true equations.
@@
   to classify all untyped lambda calculus terms.
-  We define such a type \texttt{Top} at the bottom of Figure~\ref{fig:tp}, which
-  is used in the typing of the next construct we discuss, \(\rho\).
 \item The \(\rho\) construct allows one to rewrite occurrences of \(t_1\) to
-  \(t_2\) in the synthesized type of the second subexpression.
-  The version presented here requires a ``guide'' \(@x.T\), where \(x\)
-  indicates the occurrences of \(t_1\) to substitute.
-  \([\beta\{t_2\}/x]^{\texttt{Top}}\ T\) must be a well-kinded type, but it is not required that
-  \([t_1/x]T\) also be, only that \(T'\) and \([t_1/x]T\) are convertible modulo
-  erasure --- it is not guaranteed that \(t_1\) and \(t_2\) will have the same
-  type, or will even be typeable.
+  \(t_2'\) in the synthesized type of its second subexpression.
+  The version presented here requires a type annotations (``guide'') \(@x\langle
+  t_2 \rangle.T\), where \(x\) indicates the occurrences of \(t_1\) to
+  substitute and \(t_2\) is a term, \(\beta\eta\)-equivalent to \(t_2'\), that
+  is used to replace \(x\).
+  There is no requirement that \(t_1\) and \(t_2\) will have the same type, or
+  even be typable.
+  However, \([t_2/x]\ T\) must be a well-kinded type and \([t_1/x]T\) (which
+  might be ill-kinded) must be convertible with \(T'\) modulo erasure.
   In Cedille, the guide is optional and the construct may be used to rewrite a
-  contextually given type; a heuristic is used to produce a resulting type that
-  is well-kinded whose details are beyond the scope of this document.
-
-  It might appear at first that the requirement that the guide \(T\) be written
-  such that the resulting substitution is well-kinded is too strict.
-  For example, what if we are trying to rewrite \(n\) in \(X\ n\), where \(X\)
-  is a predicate on (Church-encoded) naturals?
-  The next construct, \(\varphi\), gives us the ability to recover the typing
-  information associated to \(n\).
+  contextually given type; a heuristic, whose details are beyond the scope of
+  this document, is used to produce a resulting type that is well-kinded.
+
 \item We adopt a strong form of Nuprl's \textbf{direct computation rules}~\cite{constable+86}:
@@
 
-  In the first version of CDLE, such an axiom was derivable from \emph{lifting}, a construct allowing
-  simply typable terms to be lifted to the type level~\cite{stump17}.  We omit lifting in this new version of CDLE, because while
+  In the first version of CDLE, such an axiom was derivable from \emph{lifting},
+  a construct allowing terms with simple types to be lifted to the type
+  level~\cite{stump17}.
+  We omit lifting in this new version of CDLE, because while
   sound, lifting as defined in that previous work is complicated and appears to be incomplete.  Developing a new
@@
 \label{thm:consis}
-  There is no term $t$ such that $\vdash t : \abs{\forall}{X}{\star}{X}$.
+  There is no term $t$ such that $\cdot \vdash t : \abs{\forall}{X}{\star}{X}$.
 \end{theorem}
@@
 including non-normalizing ones. In our annotated syntax, we write
-\texttt{\(\beta\)\{t\}}. Even without this, the presence of $\delta$ in
-combination with $\varphi$ allows us to type non-normalizing terms assuming an
-erased argument $x$ of type $\{ \textit{tt} \simeq \textit{ff} \}$ for
-Church-encoded booleans \textit{tt} and \textit{ff}. For example, $\delta\
-\mhyph\ x$ can be checked against type \(\{\absu{\lambda}{x}{x} \simeq
-\absu{\lambda}{x}{x\ x}\}\), and with 
-$\varphi$ we can use this to type $\Omega$ by changing the typed term
-\(\texttt{id}\ \cdot\texttt{True}\ \texttt{id}\), where \verb|True| is
-\(\abs{\forall}{X}{\star}{X \to X}\). But failure of normalization does not
+\texttt{\(\beta\)\{t\}}.
+Even without this, the presence of \(\varphi\) allows us to type non-normalizing
+terms assuming an erased argument \(x\) of type \(\{\absu{\lambda}{x}{x} \simeq
+\absu{\lambda}{x}{x\ x}\}\) by changing the type of the term \(\texttt{id}\
+\cdot \texttt{True}\ \texttt{id}\), where \texttt{True} is
+\(\abs{\forall}{X}{\star}{X \to X}\).
+This would allow us to give the type \texttt{True} to \(\Omega = (\absu{\lambda}{x}{x\ x})\
+\absu{\lambda}{x}{x\ x}\).
+In general, we can use any inconsistent assumption to do this, and in the
+presence of \(\delta\) that includes all equations between two terms that are
+B\"ohm-separable.
+But, failure of normalization does not
 impinge on Theorem~\ref{thm:consis}. Extensional Martin-L\"of type theory (MLTT)
@@
   Suppose \(\cdot \vdash t \tpsynth T\) and there exists a closed term \(t'\)
-  such that \(\cdot \vdash t \tpsynth T \to \abs{\Pi}{x}{T_1}{T_2}\) for some
+  such that \(\cdot \vdash t' \tpsynth T \to \abs{\Pi}{x}{T_1}{T_2}\) for some
   \(T_1,T_2\) and \(|t'| = \absu{\lambda}{x}{x}\).
@@
 Given the lack of normalization in general, several checks in the typing rules --
-for things like $t =_{\beta\eta} t'$ -- are formally undecidable.  In
-practice, we simply impose a bound on the number of steps of reduction,
+for things like $|t| =_{\beta\eta} |t'|$ -- are formally undecidable.
+We simply impose a bound on the number of steps of reduction,
 and thus restore formal decidability (we are checking ``typable within
@@
   If \(\vdash \Gamma\) then:
-  \begin{itemize}
+  \begin{enumerate}
   \item if \(\Gamma \vdash T \tpsynth \kappa\) then \(\Gamma \vdash \kappa\)
@@
   \item if \(\Gamma \vdash t \tpsynth T\) then \(\Gamma \vdash T \tpsynth \star\)
-  \end{itemize}
+  \end{enumerate}
 \end{theorem}
@@
 is extended to sets $S$ of terms by writing $[S]_{\cbe}$ for
-$\{[t]_{\cbe}\ |\ t\in S\}$.  In a few places we write
-$\textit{nf}(t)$ for the (unique) $\beta\eta$-normal form of term $t$,
-if it has one.  If (in our meta-language) we affirm a statement
+$\{[t]_{\cbe}\ |\ t\in S\}$.  % In a few places we write
+% $\textit{nf}(t)$ for the (unique) $\beta\eta$-normal form of term $t$,
+% if it has one.
+If (in our meta-language) we affirm a statement
 involving application of a partial function, then it is to be
@@
 
+Note that Lemma~\ref{lem:termsubstinterp} also applies to typed substitution: if
+\(t' =_{\cbe} \sigma|t|\) then by erasure so does \(\chi\ T\ \mhyph\ t'\).
+
 \begin{lemma}[Type substitution and interpretation] 
   \label{lem:tpsubstinterp}
+  \ \\
   \begin{itemize}
@@
 From the definition of $\interp{\Gamma}$, we obtain
-$\rho(x)\in\interp{\kappa}_{\sigma,\rho}$.
+$\rho(X)\in\interp{\kappa}_{\sigma,\rho}$.
 
@@
 The result of applying the function is thus indeed
-in $\interp{[t/x]\kappa}_{\sigma,\rho}$, since by
-Lemma~\ref{lem:termsubstinterp}, this equals
+in $\interp{[t/x]^{T'}\ \kappa}_{\sigma,\rho}$, since \(|\chi\ T'\ \mhyph\ t| =
+|t|\) (recall the shorthand \([t/x]^{T'} = [\chi\ T'\ \mhyph\ t/x]\)), and with  
+Lemma~\ref{lem:termsubstinterp} we have that the interpretation equals
 $\interp{\kappa}_{\sigma[x\mapsto \choice([\sigma |t|]_{\cbe})],\rho}$
@@
 \[
-    \infer{\Gamma\vdash x\tpsynth T}{(x : t)\in\Gamma} 
+    \infer{\Gamma\vdash x\tpsynth T}{(x : T)\in\Gamma} 
 \]
@@
 
-\startcase{.2cm}
-\[
- \infer{\Gamma\vdash t\tpcheck T'}{\Gamma\vdash t \tpcheck T & T' \leadsto^*_\beta T} 
-\]
-We are assuming $\interp{T'}_{\sigma,\rho}$ is defined, since this is a checking judgment. The desired result then
-follows from Lemma~\ref{lem:interppres}.
-
-\startcase{.2cm}
-\[
-    \infer{\Gamma\vdash t\tpsynth T'}{\Gamma\vdash t \tpsynth T & T \leadsto^*_\beta T'} 
-\]
-This also follows from Lemma~\ref{lem:interppres} and the induction hypothesis for the first premise,
-which implies $\interp{T}_{\sigma,\rho}\in\mathcal{R}$ (and hence defined).
+% \startcase{.2cm}
+% \[
+%  \infer{\Gamma\vdash t\tpcheck T'}{\Gamma\vdash t \tpcheck T & T' \leadsto^*_\beta T} 
+% \]
+% We are assuming $\interp{T'}_{\sigma,\rho}$ is defined, since this is a checking judgment. The desired result then
+% follows from Lemma~\ref{lem:interppres}.
+
+% \startcase{.2cm}
+% \[
+%     \infer{\Gamma\vdash t\tpsynth T'}{\Gamma\vdash t \tpsynth T & T \leadsto^*_\beta T'} 
+% \]
+% This also follows from Lemma~\ref{lem:interppres} and the induction hypothesis for the first premise,
+% which implies $\interp{T}_{\sigma,\rho}\in\mathcal{R}$ (and hence defined).
 
@@
 By the IH applied to the first premise, we have 
-$[\sigma t]_{\cbe}\in\interp{T'}_{\sigma,\rho}\in\mathcal{R}$. By assumption, $\interp{T}_{\sigma,\rho}\in\mathcal{R}$,
-and so by the IH applied to the second premise, we have $[\sigma t]_{\cbe}\in\interp{T'}_{\sigma,\rho} = \interp{T}_{\sigma,\rho}$.
-
-\startcase{.2cm}
-\[
-    \infer{\Gamma\vdash \absu{\lambda}{x}{t} \tpcheck \abs{\Pi}{x}{T}{T'}}{\Gamma,x:T\vdash t\tpcheck T'} 
-\]
-To show $[\sigma \lambda x.t]_{\cbe}\in\interp{\Pi  x:T.T'}_{\sigma,\rho}$ (noting that the latter is defined
-and in $\mathcal{R}$ by assumption), it suffices to assume an arbitrary
-$E\in\interp{T}_{\sigma,\rho}$, and show
-$[[\choice(E)/x]\sigma t]_{\cbe}\in\interp{T'}_{\sigma[x\mapsto\choice(E)],\rho}$.  By the IH,
-we have
-$[\sigma[x\mapsto\choice(E)]t]_{\cbe}\in\interp{T'}_{\sigma[x\mapsto\choice(E)],\rho}$.
-But $[\sigma[x\mapsto\choice(E)]t]_{\cbe} = [[\choice(E)/x]\sigma t]_{\cbe}$,
+$[\sigma|t|]_{\cbe}\in\interp{T'}_{\sigma,\rho}\in\mathcal{R}$. By assumption, $\interp{T}_{\sigma,\rho}\in\mathcal{R}$,
+and so by the IH applied to the second premise, we have $[\sigma|t|]_{\cbe}\in\interp{T'}_{\sigma,\rho} = \interp{T}_{\sigma,\rho}$.
+
+\startcase{.2cm}
+\[
+  \infer{\Gamma\vdash \absu{\lambda}{x}{t} \tpcheck T}{T \leadsto_\beta^* \abs{\Pi}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t\tpcheck T_2}
+\]
+To show $[\sigma \lambda
+x.|t|]_{\cbe}\in\interp{\abs{\Pi}{x}{T_1}{T_2}}_{\sigma,\rho}$ (noting that the
+latter is defined and in $\mathcal{R}$ by assumption), it suffices to assume an
+arbitrary $E\in\interp{T_1}_{\sigma,\rho}$, and show
+$[[\choice(E)/x]\sigma|t|]_{\cbe}\in\interp{T_2}_{\sigma[x\mapsto\choice(E)],\rho}$.
+By the IH, we have
+$[\sigma[x\mapsto\choice(E)]|t|]_{\cbe}\in\interp{T_2}_{\sigma[x\mapsto\choice(E)],\rho}$.
+But $[\sigma[x\mapsto\choice(E)]t]_{\cbe} = [[\choice(E)/x]\sigma|t|]_{\cbe}$,
 so this is sufficient.
@@
 \[
-    \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]^{T'} T}{\Gamma\vdash t \tpsynth \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} 
+  \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynthleads \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\
 \]
 By the IH applied to the first premise, $[\sigma
-t]_{\cbe}\in\interp{\Pi x:T'.T}_{\sigma,\rho}\in\mathcal{R}$.  This
+|t|]_{\cbe}\in\interp{\Pi x:T'.T}_{\sigma,\rho}\in\mathcal{R}$.  This
 means that there exists a $\lambda$-abstraction $\lambda x.\hat{t}$
-such that $\lambda x.\hat{t} =_{\cbe} \sigma t$, by the semantics of $\Pi$-types.
+such that $\lambda x.\hat{t} =_{\cbe} \sigma |t|$, by the semantics of $\Pi$-types.
 Furthermore, for any $E\in\interp{T'}_{\sigma,\rho}$,
 $[[\choice(E)/x]\hat{t}]_{\cbe}\in\interp{T}_{\sigma[x\mapsto\choice(E)],\rho}$.
-By the IH applied to the second premise, $[\sigma t']_{\cbe}\in\interp{T'}_{\sigma,\rho}$,
+By the IH applied to the second premise, $[\sigma |t'|]_{\cbe}\in\interp{T'}_{\sigma,\rho}$,
 so we can instantiate the quantifier in the previous formula to obtain
 \[
- [[\choice([\sigma t']_{\cbe})/x]\hat{t}]_{\cbe}\in\interp{T}_{\sigma[x\mapsto\choice([\sigma t']_{\cbe})],\rho}
+ [[\choice([\sigma |t'|]_{\cbe})/x]\hat{t}]_{\cbe}\in\interp{T}_{\sigma[x\mapsto\choice([\sigma |t'|]_{\cbe})],\rho}
 \]
@@
 \[
- [[\choice([\sigma t']_{\cbe})/x]\hat{t}]_{\cbe}\in\interp{[t'/x]T_2}_{\sigma,\rho}
-\]
-Since $\sigma (t\ t') =_{\cbe} (\lambda x.\hat{t})\ \sigma t' =_{\cbe} [[\choice([\sigma t']_{\cbe})/x]\hat{t}$,
+ [[\choice([\sigma |t'|]_{\cbe})/x]\hat{t}]_{\cbe}\in\interp{[t'/x]T_2}_{\sigma,\rho}
+\]
+Since $\sigma |t\ t'| =_{\cbe} (\lambda x.\hat{t})\ \sigma |t'| =_{\cbe} [[\choice([\sigma |t'|]_{\cbe})/x]\hat{t}$,
 this is sufficient.
@@
 \[
-    \infer{\Gamma\vdash \absu{\Lambda}{X}{t} \tpcheck \abs{\forall}{X}{\kappa}{T}}{\Gamma,X:\kappa\vdash t \tpcheck T} 
+  \infer{\Gamma\vdash \absu{\Lambda}{X}{t} \tpcheck T'}
+  {T' \leadsto_\beta^* \abs{\forall}{X}{\kappa}{T} & \Gamma,X:\kappa\vdash t \tpcheck T}
 \]
@@
   \infer{\Gamma\vdash t \cdot T' \tpsynth [T'/X]T}
-    {\Gamma\vdash t \tpsynth \abs{\forall}{X}{\kappa}{T}
-      & \Gamma\vdash T' \tpsynth \kappa'
-      & \kappa' \cong \kappa
-    } 
+  {\Gamma\vdash t \tpsynthleads \abs{\forall}{X}{\kappa}{T} & \Gamma\vdash T' \tpsynth \kappa' & \kappa'\cong\kappa} \\ \\
 \]
@@
 have, we have \(\interp{T'}_{\sigma,\rho} \in \interp{\kappa}_{\sigma,\rho}\).
-So, can derive $[\sigma t]_{\cbe}\in\interp{T}_{\sigma,\rho[X\mapsto \interp{T'}_{\sigma,\rho}]}$.
+So, can derive $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho[X\mapsto \interp{T'}_{\sigma,\rho}]}$.
 By Lemma~\ref{lem:tpsubstinterp},
@@
 \[
-    \infer{\Gamma\vdash \absu{\Lambda}{x}{t} \tpcheck \abs{\forall}{x}{T'}{T}}{\Gamma,x:T'\vdash t \tpcheck T & x\not\in\textit{FV}(|t|)} 
-\]
-
-By the IH applied to the first premise, we have
+  \infer{\Gamma\vdash \absu{\Lambda}{x}{t} \tpcheck T}
+  {T \leadsto_\beta^* \abs{\forall}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t \tpcheck T_2 & x\not\in\textit{FV}(|t|)}
+\]
+
+By the IH applied to the second premise, we have
 $[\sigma[x\mapsto\choice(E)]
-t]_{\cbe}\in\interp{T'}_{\sigma[x\mapsto\choice(E)],\rho}$, for any
-$E\in\interp{T}_{\sigma,\rho}$.  This is because
-$\interp{T}_{\sigma,\rho}\in\mathcal{R}$, since $\interp{\abs{\forall}{x}{T'}{T}}_{\sigma,\rho}$ is in $\mathcal{R}$ and
-hence defined, by assumption.  Since $x\not\in\textit{FV}(t)$, we know
-$[[\sigma[x\mapsto\choice(E)]t]_{\cbe} = [\sigma t]_{\cbe}$.  By the
-semantics of $\forall$-types and definition of erasure, this suffices to show the desired
-conclusion.
-
-\startcase{.2cm}
-\[
-    \infer{\Gamma\vdash t\ \mhyph t' \tpsynth [t'/x]^{T'}\ T}{\Gamma\vdash t \tpsynth \abs{\forall}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'}
+|t|]_{\cbe}\in\interp{T_2}_{\sigma[x\mapsto\choice(E)],\rho}$, for any
+$E\in\interp{T_1}_{\sigma,\rho}$.
+This is because $\interp{T_1}_{\sigma,\rho}\in\mathcal{R}$, since
+$\interp{\abs{\forall}{x}{T_1}{T_2}}_{\sigma,\rho}$ is in $\mathcal{R}$ and 
+hence defined, by assumption.
+Since $x\not\in\textit{FV}(|t|)$, we know
+$[[\sigma[x\mapsto\choice(E)]|t|]_{\cbe} = [\sigma |t|]_{\cbe}$.
+By the semantics of $\forall$-types and definition of erasure, this suffices to
+show the desired conclusion.
+
+\startcase{.2cm}
+\[
+  \infer{\Gamma\vdash t\ \mhyph t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynthleads \abs{\forall}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\
 \]
@@
 \[
-    \infer{\Gamma\vdash [ t , t' ] \tpcheck \abs{\iota}{x}{T}{T'}}
-          {\Gamma\vdash t \tpcheck T & \Gamma\vdash t' \tpcheck [t/x]^{T}\ T' & |t| =_{\beta\eta} |t'|} 
-\]
-By the IH, we have $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho}$ and
-$[\sigma |t|]_{\cbe}\in\interp{[t/x]T'}_{\sigma,\rho}$.  By
-Lemma~\ref{lem:termsubstinterp}, the latter is equivalent to
-$[\sigma t]_{\cbe}\in\interp{T'}_{\sigma[x\mapsto\choice([\sigma t]_{\cbe}),\rho}$.  These two facts about $[\sigma t]_{\cbe}$ are
+  \infer{\Gamma\vdash [ t_1 , t_2 ] \tpcheck T}
+  {
+    \begin{array}{cc}
+      T \leadsto_\beta^* \abs{\iota}{x}{T_1}{T_1}
+      & \Gamma \vdash t_1 \tpcheck T_1
+      \\ \Gamma \vdash t_2 \tpcheck [t/x]^{T_1}\ T_2
+      & |t_1| =_{\beta\eta} |t_2|
+    \end{array}
+  }
+\]
+By the IH, we have $[\sigma |t_1|]_{\cbe}\in\interp{T_1}_{\sigma,\rho}$ and
+$[\sigma |t_2|]_{\cbe}\in\interp{[t_1/x]T_2}_{\sigma,\rho}$.
+By Lemma~\ref{lem:termsubstinterp}, the latter is equivalent to
+$[\sigma |t|]_{\cbe}\in\interp{T_2}_{\sigma[x\mapsto\choice([\sigma
+  |t|]_{\cbe}),\rho}$.
+These two facts about $[\sigma |t|]_{\cbe}$ are
 sufficient, by the semantics of $\iota$-types, for the desired
-conclusion, using also the fact (from the third premise) that $\sigma|t| =_{\cbe} \sigma|t'|$.
-
-\startcase{.2cm}
-\[
-   \infer{\Gamma\vdash t.1 \tpsynth T}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} 
+conclusion, using also the fact (from the fourth premise) that $\sigma|t| =_{\cbe} \sigma|t'|$.
+
+\startcase{.2cm}
+\[
+  \infer{\Gamma\vdash t.1 \tpsynth T}{\Gamma\vdash t \tpsynthleads \abs{\iota}{x}{T}{T'}}
 \]
@@
 \[
-    \infer{\Gamma\vdash t.2 \tpsynth [t.1/x]T'}{\Gamma\vdash t \tpsynth \abs{\iota}{x}{T}{T'}} 
-\]
-Similar to the previous case, using Lemma~\ref{lem:termsubstinterp}.
-
-\startcase{.2cm}
-\[
-  \infer{\Gamma\vdash \beta\{t'\} \tpcheck \{ t_1 \simeq t_2 \}}
-  {\textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}
-\]
-Since by assumption \(|t_1| =_{\beta\eta} |t_2|\) it follows that \(\sigma|t_1|
-=_{\beta\eta} \sigma|t_2|\), and also by assumption it follows that
+  \infer{\Gamma\vdash t.2 \tpsynth [t.1/x] T'}{\Gamma\vdash t \tpsynthleads \abs{\iota}{x}{T}{T'}}
+\]
+Similar to the previous case, additionally using Lemma~\ref{lem:termsubstinterp}.
+
+\startcase{.2cm}
+\[
+  \infer{\Gamma\vdash \beta\{t'\} \tpcheck T}
+  {T \leadsto_\beta^* \{t_1 \simeq t_2\} \quad \textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}
+\]
+By the third premise \(|t_1| =_{\beta\eta} |t_2|\) it follows that \(\sigma|t_1|
+=_{\beta\eta} \sigma|t_2|\), and also by the second premise and the fact that
+\(\sigma \in \interp{\Gamma}\) it follows that
 \(\textit{FV}(t_1\ t_2) \subseteq \textit{dom}(\sigma)\).
@@
 \[
-  \infer{\Gamma \vdash \rho\ t'\ @x.T\ \mhyph\ t \tpsynth [t_2/x]T}
+  \infer{\Gamma \vdash \rho\ t'\ @x\langle t_2 \rangle.T\ \mhyph\ t \tpsynth [t_2/x]\ T}
   {
-    \begin{array}{cc}
-      \Gamma \vdash t' \tpsynth \{t_1 \simeq t_2\}
-      & \quad \Gamma \vdash [t_2/x]T \tpsynth \star
-      \\ \Gamma \vdash t \tpsynth T'
+    \begin{array}{ccc}
+      \Gamma \vdash t' \tpsynthleads \{t_1 \simeq t_2'\}
+      & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma)
+      & |t_2'| =_{\beta\eta} |t_2|
+      \\ \Gamma \vdash [t_2/x]\ T \tpsynth \star
+      & \Gamma \vdash t \tpsynth T'
       & T' \cong [t_1/x]T
@@
 \]
-By the IH applied to the first premise in the first row, $\sigma|t_1| =_{\beta\eta}
-\sigma|t_2|$.
+By the IH applied to the first premise in the first row, $\sigma|t_1| =_{\cbe}
+\sigma|t_2'|$.
+With the second and third premise, and from the assumption \(\sigma \in
+\interp{\Gamma}\), we have \(\sigma|t_2'| =_{\cbe} \sigma|t_2|\) (and
+\(\sigma|t_2|\) is closed), so we obtain
+\([\sigma|t_1|]_{\cbe} = [\sigma|t_2|]_{\cbe}\).
+By the IH applied to the first premise of the second row,
+\(\interp{[t_2/x]T}_{\sigma,\rho} \in \mathcal{R}\) and so is defined.
 By the IH applied to the first premise in the second row,
-\([\sigma|t|]_{c\beta\eta} \in \interp{T'}_{\sigma.\rho}\).
+\([\sigma|t|]_{c\beta\eta} \in \interp{T'}_{\sigma,\rho}\).
 By the IH applied to the second premise in the second row,
@@
 \[
-    \infer{\Gamma\vdash \varphi\ t\ \mhyph\ t'\{t''\} \Leftrightarrow T}
-          {\Gamma\vdash t\tpsynth \{t'\simeq t''\} & \Gamma\vdash t' \Leftrightarrow T}  
-\]
-By the IH for the first premise, $\sigma|t'| =_{\cbe} \sigma|t''|$.  By the IH for the second
-premise, $[\sigma|t'|]_{\cbe}\in\interp{T}_{\sigma,\rho}$.  This suffices for the desired conclusion,
-using also the definition of erasure ($|\varphi\ t\ \mhyph\ t'\{t''\}| = |t''|$).
+  \infer{\Gamma\vdash \varphi\ t\ \mhyph\ t'\ \{t''\} \Leftrightarrow T}
+  {
+    \begin{array}{ccc}
+      \Gamma\vdash t\tpsynthleads \{t_1\simeq t_2\}
+      &
+      & \Gamma\vdash t' \Leftrightarrow T
+      \\ |t_1| =_{\beta\eta} |t'|
+      & |t_2| =_{\beta\eta} |t''|
+      & \textit{FV}(t'') \subseteq \textit{dom}(\Gamma)
+    \end{array}
+  }
+\]
+By the IH for the first premise, $\sigma|t_1| =_{\cbe} \sigma|t_2|$.
+Using the first and second premise of the second row, we obtain \(\sigma|t'|
+=_{\cbe} \sigma|t''|\).
+By the IH for the second premise,
+$[\sigma|t'|]_{\cbe}\in\interp{T}_{\sigma,\rho}$ (in the case of type synthesis,
+the IH also tells us \(\interp{T}_{\sigma,\rho}\) is defined).
+This suffices for the desired conclusion, using also the definition of erasure
+($|\varphi\ t\ \mhyph\ t'\{t''\}| = |t''|$).
 

### 2020-12-30 docs/semantics/paper.tex
docs: semantics: finish validity
- make Lemma 15 mutual with a similar proof for term variables in context
- update existing cases with new rules

@@
     \(\Gamma_1 \vdash \kappa_1'\) then \(\Gamma_1,X:\kappa_1',\Gamma_2 \vdash T
-    \tpsynth \kappa_2\) for some \(\kappa_2' \cong \kappa_2\)
-
-  \item If \(\Gamma_1,X:\kappa_1,\Gamma_2 \vdash t \Leftrightarrow T\) and
+    \tpsynth \kappa'_2\) for some \(\kappa_2' \cong \kappa_2\)
+
+  \item If \(T_3 \cong T_4\) and \(\Gamma_1,X:\kappa_1,\Gamma_2 \vdash t
+    \tpcheck T_3\) then \(\Gamma_1,X:\kappa_1',\Gamma_2 \vdash t \tpcheck T_4\)
+
+  \item If \(\Gamma_1,X:\kappa_1,\Gamma_2 \vdash t \tpsynth T\) and
     \(\Gamma_1 \vdash \kappa_1'\) then \(\Gamma_1,X:\kappa_1',\Gamma_2 \vdash t
-    \Leftrightarrow T\).
+    \tpsynth T\).
   \end{itemize}
+  Furthermore, the resulting typing derivations derivations have depths no
+  larger than the assumed ones.
 \end{lemma}
@@
 \begin{proof}
-  By induction on the assumed derivation.
-  We show a few interesting cases.
-
+  By mutual induction on the assumed derivation, and mutually with
+  Lemma~\ref{lem:ctxt-conv-class2}.
+  The measure, which omits the size of derivations of the convertibility
+  relation, is used to ensure that each mutually inductive call is well-founded.
+  We show the interesting cases.
+
+  % kinding
   \startcase{.2cm}
@@
   \vdash X \tpsynth \kappa\).
+  In both cases, the resulting derivation has the same depth (i.e., 1) as we
+  started with.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash\abs{\lambda}{x}{T}{T'} \tpsynth
+      \abs{\Pi}{x}{T}{\kappa}
+    }{
+      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash T \tpsynth \star
+      & \Gamma_1,X_1:\kappa_1,\Gamma_2 ,x:T\vdash T'\tpsynth\kappa
+    }
+  \]
+  By the IH on the first premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash T
+  \tpsynth \star\) at no greater depth (note \(\star\) is convertible only with
+  itself).
+  By the IH on the second premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2,x:T \vdash
+  T' \tpsynth \kappa'\) (at no greater depth) for some \(\kappa' \cong \kappa\).
+  By the rule, we have \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash
+  \abs{\lambda}{x}{T}{T'} \tpsynth \abs{\Pi}{x}{T}{\kappa'}\), which is
+  congruent with the original type in the conclusion.
+  We see the the measure is preserved.
 
@@
   \vdash T' \tpsynth \kappa''\) for some \(\kappa'' \cong \kappa'\).
-  By the rule, \(\Gamma_1,X_1\kappa_1',\Gamma_2 \vdash
+  By the rule, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash
   \abs{\lambda}{X}{\kappa}{T'} \tpsynth \abs{\Pi}{X}{\kappa}{\kappa''}\), which
   is convertible with the original kind synthesized for this type.
+  Since the depths of the premises for the resulting derivation are the same as
+  the depths for the premises of the assumed one, the measure is preserved.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash T\ t \tpsynth [t/x]^{T'}\kappa
+    }{
+      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash T \tpsynth \abs{\Pi}{x}{T'}{\kappa}
+      & \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \tpcheck T'
+    }
+  \]
+  By the IH on the first premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash T
+  \tpsynth \abs{\Pi}{x}{T''}{\kappa''}\) for some \(T'' \cong T'\) and
+  \(\kappa'' \cong \kappa\).
+  By the IH on the second premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t
+  \tpcheck T''\).
+  By the rule, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash T\ t \tpsynth
+  [t/x]^{T''}\kappa''\), convertible with the given type \([t/x]^{T'}\kappa\).
+  We see that the measure is preserved.
 
@@
   \tpsynth \kappa_2''\) for some \(\kappa_2'' \cong \kappa''\).
-  By transitivity, \(\kappa_2' \cong \kappa_2''\).
+  By transitivity and the third premise, \(\kappa_2' \cong \kappa_2''\).
   By the rule, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash T \cdot T' \tpsynth
-  [T'/X]\kappa_2\), with this kind congruent to \([T'/X]\kappa\).
+  [T'/X]\kappa_2\), with this kind convertible to \([T'/X]\kappa\) by
+  congruence.
+  We see the measure is preserved.
+
+  % checking
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t\tpcheck T_3
+    }{
+      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t\tpsynth T'
+      & T' \cong T_3
+    }
+  \]
+  By the IH on the first premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t
+  \tpsynth T'\).
+  By assumption and transitivity of congruence, \(T' \cong T_4\) (where \(T_4\)
+  is the type given to us in the proof).
+  By the rule, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t \tpcheck T_4\), and we
+  see the measure is preserved.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash \absu{\lambda}{x}{t} \tpcheck T_3
+    }{
+      T_3 \leadsto_\beta^* \abs{\Pi}{x}{T_1}{T_2}
+      & \Gamma_1,X_1:\kappa_1,\Gamma_2,x:T_1\vdash t\tpcheck T_2
+    }
+  \]
+  By the first premise, \(T_3 \cong \abs{\Pi}{x}{T_1}{T_2}\).
+  By assumption and transitivity of the convertibility relation, \(T_4 \cong
+  \abs{\Pi}{x}{T_1}{T_2}\).
+  This means \(T_4 \leadsto_\beta^* \abs{\Pi}{x}{T_1'}{T_2'}\) for some \(T_1'
+  \cong T_1\) and \(T_2' \cong T_2\).
+  By mutual induction with Lemma~\ref{lem:ctxt-conv-class2} on the second
+  premise, we have a derivation of \(\Gamma_1,X_1:\kappa_1,\Gamma_2,x:T_1'
+  \vdash t \tpcheck T_2'\) that is no deeper than that premise.
+  So, we are entitled to use the IH on this new derivation to obtain
+  \(\Gamma_1,X_1:\kappa_1',\Gamma_2,x:T_1' \vdash t \tpcheck T_2'\), which is
+  also no deeper.
+  By the rule, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash \absu{\lambda}{x}{t}
+  \tpcheck T_4\), and the measure is preserved.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,X_1:\kappa_1,\Gamma_2\vdash \absu{\Lambda}{X}{t} \tpcheck T_3
+    }{
+      T_3 \leadsto_\beta^* \abs{\forall}{X}{\kappa}{T}
+      & \Gamma_1,X_1:\kappa_1,\Gamma_2,X:\kappa\vdash t \tpcheck T
+    }
+  \]
+  By the first premise, \(T_3 \cong \abs{\forall}{X}{\kappa}{T}\), so by
+  assumption and transitivity of convertibility, \(T_4 \cong
+  \abs{\forall}{X}{\kappa}{T}\).
+  This means \(T_4 \leadsto_\beta^* \abs{\forall}{X}{\kappa'}{T'}\) for some
+  \(\kappa' \cong \kappa\) and \(T' \cong T\).
+  We apply the IH once on the second premise to obtain a derivation of
+  \(\Gamma_1,X_1:\kappa_1,\Gamma_2,X:\kappa' \vdash t \tpcheck T'\), and then
+  again to obtain \(\Gamma_1,X_1:\kappa_1',\Gamma_2,X:\kappa' \vdash t \tpcheck
+  T'\), noting this derivation's depth is no greater than the second premise.
+  By the rule, \(\Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash \absu{\Lambda}{X}{t}
+  \tpcheck T_4\), and the measure is preserved.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash \beta\{t'\} \tpcheck T_3
+    }{
+      T_3 \leadsto_\beta^* \{t_1 \simeq t_2\}
+      & \textit{FV}(t')\subseteq \textit{dom}(\Gamma_1,X_1:\kappa_1,\Gamma_2)
+      & |t_1| =_{\beta\eta} |t_2|
+    }
+  \]
+  From the first premise, assumption, and transitivity of convertibility, \(T_4
+  \cong \{t_1 \simeq t_2\}\).
+  This means that \(T_4 \leadsto_\beta^* \{t_3 \simeq t_4\}\) for some
+  \(t_3,t_4\) such that \(|t_3| =_{\beta\eta} |t_1|\) and \(|t_4| =_{\beta\eta}
+  |t_2|\).
+  From this, the third premise, and transitivity of convertibility, we obtain
+  \(|t_3| =_{\beta\eta} |t_4|\).
+  We also see from the second premise that \(\textit{FV}(t') \subseteq
+  \textit{dom}(\Gamma_1,X_1:\kappa_1',\Gamma_2)\).
+  So, we use the rule to conclude that \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash
+  \beta\{t'\} \tpcheck T_4\), and the depth (1) is preserved.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash [ t_1 , t_2 ] \tpcheck T_3
+    }{
+      \begin{array}{cc}
+        T_3 \leadsto_\beta^* \abs{\iota}{x}{T_1}{T_1}
+        & \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t_1 \tpcheck T_1
+        \\ \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t_2 \tpcheck [t/x]^{T_1}\ T_2
+        & |t_1| =_{\beta\eta} |t_2|
+      \end{array}
+    }
+  \]
+  From the first premise, assumption, and transitivity of the convertibility
+  relation, \(T_4 \cong \abs{\iota}{x}{T_1}{T_2}\).
+  This means that \(T_4 \leadsto_\beta^* \abs{\iota}{x}{T_1'}{T_2'}\) for some
+  \(T_1' \cong T_1\) and \(T_2' \cong T_2\).
+  From this last congruence, we have \([t_1/x]^{T_1'}T_2' \cong [t_1/x]^{T_1} T_2\).
+  By the IH on the second premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t_1
+  \tpcheck T_1'\).
+  By the IH on the first premise, second row, we have
+  \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t_2 \tpcheck [t/x]T_2'\).
+  We use the rule to conclude, noting that the measure is preserved.
+
+  % synthesis
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t\ t' \tpsynth [t'/x]^{T'}T
+    }{
+      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \tpsynthleads \abs{\Pi}{x}{T'}{T}
+      & \Gamma_1,X_1:\kappa_1,\Gamma_2\vdash t' \tpcheck T'
+    }
+  \]
+  By the IH on the first premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t
+  \tpsynthleads \abs{\Pi}{x}{T'}{T}\).
+  By the IH on the second premise, \(\Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t'
+  \tpcheck T'\).
+  By the rule, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t\ t' \tpsynth [t'/x]T\),
+  and the measure is preserved.
 
@@
     }{
-      \Gamma_1,X_1:\kappa_1,\Gamma_2\vdash t \tpsynth
+      \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \tpsynthleads
       \abs{\forall}{X}{\kappa}{T}
@@
   By the IH on the first premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t
-  \tpsynth \abs{\forall}{X}{\kappa}{T}\).
+  \tpsynthleads \abs{\forall}{X}{\kappa}{T}\).
   By the IH on the second premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash T'
-  \tpsynth \kappa''\) for some \(\kappa'' \cong \kappa'\).
-  With the third premise and transitivity of congruence, we have \(\kappa''
-  \cong \kappa\).
-  We apply the rule and conclude.
+  \tpsynth \kappa''\) for some \(\kappa'' \cong \kappa'\), from which we obtain
+  \(\kappa'' \cong \kappa\).
+  By the rule, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t \cdot T' \tpsynth
+  [T'/X]T\), and the measure is preserved.
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash \rho\ t'\ @x\langle t_2 \rangle.T\ \mhyph\ t \tpsynth [t_2/x]\ T}
+    {
+      \begin{array}{ccc}
+        \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t' \tpsynthleads \{t_1 \simeq t_2'\}
+        & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma_1,X_1:\kappa_1,\Gamma_2)
+        & |t_2'| =_{\beta\eta} |t_2|
+        \\ \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash [t_2/x]\ T \tpsynth \star
+        & \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \tpsynth T'
+        & T' \cong [t_1/x]T
+      \end{array}
+    }
+  \]
+  By the IH on the first premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t'
+  \tpsynthleads \{t_1 \simeq t_2'\}\).
+  By the IH on the second premise of the second row,
+  \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash [t_2/x]\ T \tpsynth \star\)
+  (\(\star\) is only convertible with itself).
+  We apply the IH to the second premise of the second row and use the rule to conclude
+  (\(\textit{dom}(\Gamma_1,X_1:\kappa_1,\Gamma_2) =
+  \textit{dom}(\Gamma_1,X_1:\kappa_1',\Gamma_2)\)).
+\end{proof}
+
+\begin{lemma}
+  \label{lem:ctxt-conv-class2}
+  Let \(T_1,T_2\) be types such that \(T_1 \cong T_2\).
+  \begin{itemize}
+  \item If \(\Gamma_1,x:T_1,\Gamma_2 \vdash \kappa\) then
+    \(\Gamma_1,x:T_2,\Gamma_2 \vdash \kappa\)
+
+  \item If \(\Gamma_1,x:T_1,\Gamma_2 \vdash T \tpsynth \kappa_1\) then
+    \(\Gamma_1,x:T_2,\Gamma_2 \vdash T \tpsynth \kappa_1\)
+
+  \item If \(T_3 \cong T_4\) and \(\Gamma_1,x:T_1,\Gamma_2 \vdash t \tpcheck T_3\) then
+    \(\Gamma_1,x:T_2,\Gamma_2 \vdash t \tpcheck T_4\)
+
+  \item If \(\Gamma_1,x:T_1,\Gamma_2 \vdash t \tpsynth T\) then
+    \(\Gamma_1,x:T_2,\Gamma_2 \vdash t \tpsynth T'\) for some \(T' \cong T\).
+
+  \end{itemize}
+  Furthermore, the resulting typing derivations derivations have depths no
+  larger than the assumed ones.
+\end{lemma}
+\begin{proof}
+  By mutual induction on the assumed derivation (specifically, its depth), and
+  mutually with Lemma~\ref{lem:ctxt-conv-class}.
+  The measure, which omits the size of derivations of the convertibility
+  relation, is used to ensure that each mutually inductive call is well-founded.
+  We show the interesting cases.
+
+  % kinding types
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,x_1:T_1,\Gamma_2 \vdash \abs{\lambda}{x}{T}{T'} \tpsynth \abs{\Pi}{x}{T}{\kappa}
+    }{
+      \Gamma_1,x_1:T_1,\Gamma_2 \vdash T \tpsynth \star
+      & \Gamma_1,x_1:T_1,\Gamma_2,x:T \vdash T' \tpsynth \kappa
+    }
+  \]
+  By the IH on the first premise, \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash T \tpsynth
+  \star\).
+  By the IH on the second premise, \(\Gamma_1,x_1:T_2,\Gamma_2,x:T \vdash T'
+  \tpsynth \kappa\).
+  We also have that the depths of the resulting typing derivations are no greater than
+  the those used to derive the premises, so we conclude using the rule.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,x_1:T_1,\Gamma_2\vdash\abs{\lambda}{X}{\kappa}{T'} \tpsynth
+      \abs{\Pi}{X}{\kappa}{\kappa'}
+    }{
+      \Gamma_1,x_1:T_1,\Gamma_2 \vdash \kappa & \Gamma_1,x_1:T_1,\Gamma_2,X:\kappa\vdash T'\tpsynth\kappa'
+    }
+  \]
+  By the IH on the first premise, \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash T \tpsynth
+  \kappa\).
+  By the IH on the second premise, \(\Gamma_1,x_1:T_2,\Gamma_2,X:\kappa \vdash
+  T' \tpsynth \kappa'\).
+  We also have that the depths of the resulting derivations are no greater than
+  the those used to derive the premises, so we conclude using the rule.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,x_1:T_1,\Gamma_2\vdash T\ t \tpsynth [t/x]^{T'}\kappa
+    }{
+      \Gamma_1,x_1:T_1,\Gamma_2 \vdash T \tpsynth \abs{\Pi}{x}{T'}{\kappa}
+      & \Gamma_1,x_1:T_1,\Gamma_2 \vdash t \tpcheck T'
+    }
+  \]
+  By the IH on the first premise, \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash T \tpsynth
+  \abs{\Pi}{x}{T'}{\kappa}\).
+  By the IH on the second premise, \(\Gamma_2,x_1:T_2,\Gamma_2 \vdash t \tpcheck
+  T'\).
+  We also have that the depths of the resulting derivations are no greater than
+  the those used to derive the premises, so we conclude using the rule.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,x_1:T_1,\Gamma_2 \vdash T \cdot T' \tpsynth [T'/X]\kappa
+    }{
+      \Gamma_1,x_1:T_1,\Gamma_2\vdash T \tpsynth \abs{\Pi}{X}{\kappa'}{\kappa}
+      & \Gamma_1,x_1:T_1,\Gamma_2 \vdash T_2 \tpsynth \kappa'' & \kappa'' \cong \kappa'
+    }
+  \]
+  By the IH on the first premise, \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash T \tpsynth
+  \abs{\Pi}{X}{\kappa'}{\kappa}\).
+  By the IH on the second premise, \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash T_2
+  \tpsynth \kappa''\).
+  We also have that the depths of the resulting derivations are no greater than
+  the those used to derive the premises, so we conclude using the rule, keeping
+  the third premise.
+
+  % type checking
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,x_1:T_1,\Gamma_2 \vdash t\tpcheck T
+    }{
+      \Gamma_1,x_1:T_1,\Gamma_2 \vdash t\tpsynth T'
+      & T' \cong T
+    }
+  \]
+  By the IH on the first premise, \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash t \tpsynth
+  T''\) for some \(T'' \cong T'\).
+  By transitivity of convertibility and the second premise, \(T'' \cong T\).
+  By transitivity again, \(T'' \cong T_4\) (where \(T_4 \cong T\) is given to
+  us).
+  We also have that the depths of the resulting derivations are no greater than
+  the those used to derive the premises, so we conclude using the rule.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,x_1:T_1,\Gamma_2 \vdash \absu{\lambda}{x}{t} \tpcheck T
+    }{
+      T \leadsto_\beta^* \abs{\Pi}{x}{T_5}{T_6}
+      & \Gamma_1,x_1:T_1,\Gamma_2,x:T_5\vdash t\tpcheck T_6
+    }
+  \]
+  From the first premise, we have \(T \cong \abs{\Pi}{x}{T_5}{T_6}\), so by
+  transitivity of convertibility and assumption, \(T_4 \cong
+  \abs{\Pi}{x}{T_5}{T_6}\).
+  This means that \(T_4 \leadsto_\beta^* \abs{\Pi}{x}{T_5'}{T_6'}\) for some
+  \(T_5' \cong T_5\) and \(T_6' \cong T_6\).
+  By one use of the IH on the first premise, we have
+  \(\Gamma_1,x_1:T_2,\Gamma_2,x:T_5 \vdash t \tpcheck T_6'\).
+  By another use, we have \(\Gamma_1,x_1:T_2,\Gamma_2,x:T_5' \vdash t \tpcheck
+  T_6'\) (this is well-founded, because the depth of the typing derivation this
+  second use is  applied to is no greater that the depth of the premise).
+  We also have that the depths of the resulting derivations are no greater than
+  the those used to derive the premises, so we conclude using the rule.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,x_1:T_1,\Gamma_2 \vdash \absu{\Lambda}{X}{t} \tpcheck T'
+    }{
+      T' \leadsto_\beta^* \abs{\forall}{X}{\kappa}{T}
+      & \Gamma_1,x_1:T_1,\Gamma_2,X:\kappa \vdash t \tpcheck T
+    }
+  \]
+  From the first premise, \(T' \cong \abs{\forall}{X}{\kappa}{T}\).
+  From transitivity of convertibility and assumption, \(T_4 \cong
+  \abs{\forall}{X}{\kappa}{T}\).
+  This means that \(T_4 \leadsto_\beta^* \abs{\forall}{X}{\kappa''}{T''}\) for
+  some \(\kappa'' \cong \kappa\) and \(T'' \cong T\).
+  By mutual induction with Lemma~\ref{lem:ctxt-conv-class} on the second premise,
+  \(\Gamma_1,x_1:T_1,\Gamma_2,X:\kappa'' \vdash t \tpcheck T\), and this
+  derivation is no deeper than the second premise.
+  By the IH, \(\Gamma_1,x_1:T_1,\Gamma_2,X:\kappa'' \vdash t \tpcheck T''\), and
+  this derivation is no deeper than the second premise.
+  We conclude using the rule, with the measure preserved.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,x_1:T_1,\Gamma_2 \vdash [ t , t' ] \tpcheck T_3
+    }{
+      \begin{array}{cc}
+        T_3 \leadsto_\beta^* \abs{\iota}{x}{T}{T'}
+        & \Gamma_1,x_1:T_1,\Gamma_2 \vdash t \tpcheck T
+        \\ \Gamma_1,x_1:T_1,\Gamma_2 \vdash t' \tpcheck [t/x]^{T}\ T'
+        & |t| =_{\beta\eta} |t|
+      \end{array}
+    }
+  \]
+  From the first premise, \(T_3 \cong \abs{\iota}{x}{T}{T'}\).
+  From transitivity of convertibility and assumption, \(T_4 \cong
+  \abs{\iota}{x}{T}{T'}\).
+  This means \(T_4 \leadsto \abs{\iota}{x}{T''}{T'''}\) for some \(T'' \cong T\)
+  and \(T''' \cong T'\).
+  By the IH, \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash t \tpcheck T''\) with depth no
+  greater than the second premise (read left to right, then top to bottom).
+  By the IH, \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash t' \tpcheck T'''\) with depth no
+  greater than the third premise.
+  We conclude using the rule, keeping the fourth premise as-is and noting the
+  measure is preserved.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,x_1:T_1,\Gamma_2 \vdash \beta\{t'\} \tpcheck T_3
+    }{
+      T_3 \leadsto_\beta^* \{t_1 \simeq t_2\}
+      \quad \textit{FV}(t')\subseteq \textit{dom}(\Gamma_1,x_1:T_1,\Gamma_2)
+      \quad |t_1| =_{\beta\eta} |t_2|
+    }
+  \]
+  From the first premise, \(T_3 \cong \{t_1 \simeq t_2\}\).
+  From transitivity of the convertibility relation and assumption, \(T_3 \cong
+  \{t_1 \simeq t_2\}\).
+  This means \(T_4 \leadsto_\beta^* \{t_3 \simeq t_4\}\) for some \(t_3,t_4\)
+  such that \(|t_3| =_{\beta\eta} |t_1|\) and \(|t_4| =_{\beta\eta} |t_2|\).
+  From the third premise and transitivity of convertibility, we have \(|t_3|
+  =_{\beta\eta} |t_4|\).
+  From the second premise, we obtain that \(\textit{FV}(t') \subseteq
+  \textit{dom}(\Gamma_1,x_1:T_2,\Gamma_2)\).
+  We conclude using the rule, noting that the depth remains 1.
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma_1,x_1:T_1,\Gamma_2 \vdash \varphi\ t\ \mhyph\ t'\ \{t''\} \tpcheck T_3}
+    {
+      \begin{array}{ccc}
+        \Gamma_1,x_1:T_1,\Gamma_2 \vdash t\tpsynthleads \{t_1\simeq t_2\}
+        &
+        & \Gamma_1,x_1:T_1,\Gamma_2\vdash t' \tpcheck T_3
+        \\ |t_1| =_{\beta\eta} |t'|
+        & |t_2| =_{\beta\eta} |t''|
+        & \textit{FV}(t'') \subseteq \textit{dom}(\Gamma_1,x_1:T_1,\Gamma_2)
+      \end{array}
+    }
+  \]
+  By the IH on the first premise (first row), \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash
+  t \tpsynth T\) (with no greater depth) for some \(T \cong \{t_1 \simeq t_2\}\).
+  This means that \(T \leadsto_\beta^* \{t_3 \simeq t_4\}\) for some \(t_3,t_4\)
+  such that \(|t_3| =_{\beta\eta} |t_1|\) and \(|t_4| =_{\beta\eta} |t_2|\).
+  Using the first and second premise of the second row and transitivity of
+  \(\beta\eta\)-equivalence, \(|t_3| =_{\beta\eta} |t'|\) and \(|t_4|
+  =_{\beta\eta} |t''|\).
+  By the IH on the second premise of the first row, \(\Gamma_1,x_1:T_2,\Gamma_2
+  \vdash t' \tpcheck T_4\) (with no greater depth, and where \(T_4\) is the type
+  given to us in the lemma and \(T_4 \cong T_3\)).
+  Also, we can see from the last premise of the second row that
+  \(\textit{FV}(t') \subseteq \textit{dom}(\Gamma_1,x_1:T_2,\Gamma_2)\).
+  So, we use the rule to conclude \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash \varphi\ t\
+  \mhyph\ t'\ \{t''\} \tpcheck T_4\).
+
+  % type synthesis
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,x_1:T_1,\Gamma_2 \vdash x \tpsynth T
+    }{
+      (x : T)\in\Gamma_1,x_1:T_1,\Gamma_2
+    }
+  \]
+  We have two subcases to consider.
+  If \(x = x_1\), then we use the rule to produce a derivation of
+  \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash x \tpsynth T_2\), where by assumption \(T_1 \cong T_2\).
+  Otherwise, we use the rule to produce a derivation of
+  \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash x \tpsynth T\).
+  In both cases, the measure (depth of 1) is preserved.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,x_1:T_1,\Gamma_2 \vdash t\ t' \tpsynth [t'/x]^{T'}T
+    }{
+      \Gamma_1,x_1:T_1,\Gamma_2 \vdash t \tpsynthleads \abs{\Pi}{x}{T'}{T}
+      & \Gamma_1,x_1:T_1,\Gamma_2 \vdash t' \tpcheck T'
+    }
+  \]
+  By the IH on the premise, \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash t \tpsynth T_3\)
+  for some \(T_3 \cong \abs{\Pi}{x}{T'}{T}\), and the depth of this typing
+  derivation is no greater than that of the first premise.
+  From this we obtain that \(T_3 \leadsto^*_\beta \abs{\Pi}{x}{T_4'}{T_4}\) for
+  some \(T_4' \cong T'\) and \(T_4 \cong T\).
+  By the IH on the last premise, \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash t' \tpcheck
+  T_4'\), and the depth of this derivation is no greater than that of the last premise.
+  We conclude with the rule to obtain \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash t\ t'
+  \tpsynth [t'/x]^{T_4'}\ T_4\), with the measure preserved (\([t'/x]^{T_4'}\
+  T_4\) is convertible with \([t'/x]^{T'}\ T\), since \(|\chi\ T_4'\ \mhyph\ t'|
+  =_{\beta\eta} |\chi\ T'\ \mhyph\ t'|\) by erasure and \(T \cong T_4\)).
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,x_1:T_1,\Gamma_2 \vdash t \cdot T' \tpsynth [T'/X]T
+    }{
+      \Gamma_1,x_1:T_1,\Gamma_2 \vdash t \tpsynthleads
+      \abs{\forall}{X}{\kappa}{T}
+      & \Gamma_1,x_1:T_1,\Gamma_2 \vdash T' \tpsynth \kappa'
+      & \kappa'\cong\kappa
+    }
+  \]
+
+  By the IH on the first premise, \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash t \tpsynth
+  T_3\) for some \(T_3 \cong \abs{\forall}{X}{\kappa}{T}\), with no greater
+  depth.
+  This means \(T_3 \leadsto_\beta^* \abs{\forall}{X}{\kappa''}{T''}\) for some
+  \(\kappa'' \cong \kappa\) and \(T'' \cong T\).
+  By the IH on the second premise, \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash T'
+  \tpsynth \kappa'\), with no greater depth.
+  By transitivity of convertibility, \(\kappa' \cong \kappa''\).
+  We use the rule to conclude \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash t \cdot T'
+  \tpsynth [T'/X]T''\), with the measure preserved and the synthesized type in
+  the conclusion convertible with \([T'/X]T\) (since \(T'' \cong T\)).
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma_1,x_1:T_1,\Gamma_2 \vdash \rho\ t'\ @x\langle t_2 \rangle.T\ \mhyph\ t \tpsynth [t_2/x]\ T}
+    {
+      \begin{array}{ccc}
+        \Gamma_1,x_1:T_1,\Gamma_2 \vdash t' \tpsynthleads \{t_1 \simeq t_2'\}
+        & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma)
+        & |t_2'| =_{\beta\eta} |t_2|
+        \\ \Gamma_1,x_1:T_1,\Gamma_2 \vdash [t_2/x]\ T \tpsynth \star
+        & \Gamma_1,x_1:T_1,\Gamma_2 \vdash t \tpsynth T'
+        & T' \cong [t_1/x]T
+      \end{array}
+    }
+  \]
+  By the IH on the first premise, \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash t \tpsynth
+  T_3\) (with no greater depth) for some \(T_3 \cong \{t_1 \simeq t_2'\}\).
+  This means \(T_3 \leadsto_\beta^* \{t_3 \simeq t_4\}\) for some \(t_3,t_4\)
+  such that \(|t_3| =_{\beta\eta} t_1\) and \(|t_4| =_{\beta\eta} |t_2'|\).
+  So we further obtain \(|t_4| =_{\beta\eta} |t_2|'\).
+  By the IH on the first premise of the second row, \(\Gamma_1,x_1:T_2,\Gamma_2
+  \vdash [t_2/x]T \tpsynth \star\) with no greater depth.
+  By the IH on the second premise of the second row, \(\Gamma_1,x_1:T_2,\Gamma_2
+  \vdash t \tpsynth T''\) (with no greater depth) for some \(T'' \cong T'\).
+  With the final premise of the second row and transitivity, we obtain \(T''
+  \cong [t_1/x]T\).
+  We conclude with the rule, preserving the measure.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma_1,x_1:T_1,\Gamma_2 \vdash \chi\ T\ \mhyph\ t\ \tpsynth T
+    }{
+      \Gamma_1,x_1:T_1,\Gamma_2 \vdash T \tpsynth \star
+      & \Gamma_1,x_1:T_1,\Gamma_2 \vdash t \tpcheck T
+    }
+  \]
+  By the IH on the first premise, \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash T \tpsynth
+  \star\) with no greater depth.
+  By the IH on the second premise \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash t \tpcheck
+  T\) at no greater depth.
+  We conclude using the rule, with the measure preserved.
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma_1,x_1:T_1,\Gamma_2\vdash \varphi\ t\ \mhyph\ t'\ \{t''\} \tpsynth T}
+    {
+      \begin{array}{ccc}
+        \Gamma_1,x_1:T_2,\Gamma_2\vdash t\tpsynthleads \{t_1\simeq t_2\}
+        &
+        & \Gamma_1,x_1:T_1,\Gamma_2\vdash t' \tpsynth T
+        \\ |t_1| =_{\beta\eta} |t'|
+        & |t_2| =_{\beta\eta} |t''|
+        & \textit{FV}(t'') \subseteq \textit{dom}(\Gamma)
+      \end{array}
+    }
+  \]
+
+  Similar to the checking case for \(\varphi\), only now after using the IH the
+  second premise of the first row synthesizes a type \(T' \cong T\).
+  The whole expression, then, synthesizes type \(T'\).
 \end{proof}
@@
   \label{cor:ctxt-conv-class}
-  If \(\vdash \Gamma_1,X:\kappa_1,\Gamma_1\) and \(\Gamma_1 \vdash \kappa_1'\)
-  with \(\kappa_1' \cong \kappa\) then \(\vdash \Gamma_1,X:\kappa_1',\Gamma_2\).
+  If \(\vdash \Gamma_1,X:\kappa_1,\Gamma_2\) and \(\Gamma_1 \vdash \kappa_1'\)
+  with \(\kappa_1' \cong \kappa_1\) then \(\vdash \Gamma_1,X:\kappa_1',\Gamma_2\).
+\end{corollary}
+
+\begin{corollary}
+  \label{cor:ctxt-conv-class2}
+  If \(\vdash \Gamma_1,x_1:T_1,\Gamma_2\) and \(\Gamma_1 \vdash T_2 \tpsynth
+  \star\) with \(T_1 \cong T_2\) then \(\vdash \Gamma_1,x_1:T_2,\Gamma_2\).
 \end{corollary}
@@
     \item If \(\Gamma_1,x:T,\Gamma_2 \vdash \kappa\) and \(\Gamma_1 \vdash t
-      \tpcheck T\) then \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x] \kappa\)
+      \tpsynth T\) then \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x] \kappa\)
 
@@
   Otherwise, either \((X_1 : \kappa_1) \in \Gamma_1\) and \(X \notin
-  \textit{FV}(\kappa_1)\), or else \((X_1 : \kappa_1) \in \Gamma_2\).
+  \textit{FV}(\kappa_1)\) (which we obtain from the assumption that the typing
+  context is well formed), or else \((X_1 : \kappa_1) \in \Gamma_2\). 
   Either way, we have \(\Gamma_1,[T/X]\Gamma_2 \vdash X_1 \tpsynth
@@
 
-  From our assumptions we may conclude \(\textit{FV}([t/x]t) \subseteq
+  From our assumptions we may conclude \(\textit{FV}([t/x]t') \subseteq
   \textit{dom}(\Gamma_1,[t/x]\Gamma_2)\), and from the second premise that
@@
   \[
-    \infer{\Gamma_1,x:T,\Gamma_2 \vdash \rho\ t''\ @x_1.T_1\ \mhyph\ t' \tpsynth [\beta\{t_2\}/x_1]^{\texttt{Top}}\ T_1}
+    \infer{\Gamma_1,x:T,\Gamma_2 \vdash \rho\ t''\ @x_1.T_1\ \mhyph\ t' \tpsynth [t_2/x_1] T_1}
     {
@@
         \Gamma_1,x:T,\Gamma_2 \vdash t'' \tpsynth \{t_1 \simeq t_2\}
-        & \quad \Gamma_1,x:T,\Gamma_2 \vdash [\beta\{t_2\}/x_1]^{\texttt{Top}}\ T_1 \tpsynth \star
+        & \quad \Gamma_1,x:T,\Gamma_2 \vdash [t_2/x_1]\ T_1 \tpsynth \star
         \\ \Gamma_1,x:T,\Gamma_2 \vdash t' \tpsynth T'
@@
   From the IH, we have that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]t'' \tpsynth \{[t/x]t_1
-  \simeq [t/x]t_2\}\), that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x][\beta\{t_2\}/x]T_1
+  \simeq [t/x]t_2\}\), that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x][t_2/x]T_1
   \tpsynth \star\), and that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]t' \tpsynth
@@
 
-\begin{lemma}
-  \label{lem:tpcheck-conv}
-  If \(\Gamma \vdash t \tpcheck T_1\) where \(\Gamma \vdash T_1 \tpsynth
-  \kappa_1\), and we have \(\Gamma \vdash T_2 \tpsynth \kappa_2\) with \(T_1
-  \cong T_2\) and \(\kappa_1 \cong \kappa_2\), then \(\Gamma \vdash t \tpcheck T_2\).
-\end{lemma}
-
+\subsection{Theorem~\ref{thm:syntactic-kind-pres}}
 \begin{proof}[Proof (of Theorem~\ref{thm:syntactic-kind-pres})]
@@
   \abs{\Pi}{x}{T_1'}{\kappa_1'}\) for some \(T_1' \cong T_1\) and \(\kappa_1'
-  \cong \kappa_1\) (we have by assumption the kind of \(T''\) must be
+  \cong \kappa_1\) (we have from that IH that kind of \(T''\) must be
   convertible with \(\abs{\Pi}{x}{T_1}{\kappa_1}\)). By
-  Lemma~\ref{lem:tpcheck-conv}, \(\Gamma \vdash t \tpcheck T_1'\). By the rule,
-  \(\Gamma \vdash T''\ t \tpsynth [t/x]\kappa_1'\), and the synthesized kind is
-  clearly convertible with \([t/x]\kappa_1\).
+  Lemma~\ref{lem:ctxt-conv-class2}, \(\Gamma \vdash t \tpcheck T_1'\).
+  By the rule, \(\Gamma \vdash T''\ t \tpsynth [t/x]\kappa_1'\), and the
+  synthesized kind is clearly convertible with \([t/x]\kappa_1\).
 
@@
   \(\Gamma \vdash T_1'' \tpsynth \abs{\Pi}{X}{\kappa_2''}{\kappa_1''}\) for some
-  \(\kappa_2'' \cong \kappa_2\) and \(\kappa_1'' \cong \kappa_1\) (we have by
-  assumption the kind of \(T_1''\) must be convertible with
-  \(\abs{\Pi}{X}{\kappa_2}{\kappa_1}\)). By transitivity we can conclude
-  \(\kappa_2'' \cong \kappa_2\). By the rule, we have \(\Gamma \vdash T_1''
-  \cdot T_2 \tpsynth [T_2/X]\kappa_1''\), and the synthesized kind is clearly
-  convertible with \([T_2/X]\kappa_1\).
+  \(\kappa_2'' \cong \kappa_2\) and \(\kappa_1'' \cong \kappa_1\) (we have from
+  the IH that the kind of \(T_1''\) must be convertible with
+  \(\abs{\Pi}{X}{\kappa_2}{\kappa_1}\)).
+  By transitivity we can conclude \(\kappa_2'' \cong \kappa_2'\).
+  By the rule, we have \(\Gamma \vdash T_1'' \cdot T_2 \tpsynth
+  [T_2/X]\kappa_1''\), and the synthesized kind is clearly convertible with
+  \([T_2/X]\kappa_1\).
 
@@
   \((\abs{\lambda}{X}{\kappa_2}{T_1''}) \cdot T_2\) and our assumed reduction is
-  to \([T_2/X]T_1''\). By inversion of the kinding derivation, we have
-  \(\Gamma,X:\kappa_2 \vdash T_1'' \tpsynth \kappa_1\). By
-  Lemma~\ref{lem:subst-class}, we have \(\Gamma \vdash [T_2/X]T_1'' \tpsynth
-  [T_2/X]\kappa_1\).
+  to \([T_2/X]T_1''\).
+  By inversion of the kinding derivation, we have \(\Gamma,X:\kappa_2 \vdash
+  T_1'' \tpsynth \kappa_1\).
+  By Lemma~\ref{lem:ctxt-conv-class}, we have \(\Gamma,X:\kappa_2' \vdash T_1''
+  \tpsynth \kappa_1'\) for some \(\kappa_1' \cong \kappa_1\), and from this and
+  Lemma~\ref{lem:subst-class} we have \(\Gamma \vdash [T_2/X]T_1'' \tpsynth
+  [T_2/X]\kappa_1'\).
+  This is clearly convertible with \([T_2/X]\kappa_1\), as desired.
 \end{proof}
 
+\begin{corollary}
+  \label{cor:syntactic-kind-pres}
+  If \(\Gamma \vdash T \tpsynth \kappa\) and \(T \leadsto_\beta^* T'\) then
+  \(\Gamma \vdash T' \tpsynth \kappa'\) for some \(\kappa' \cong \kappa\).
+\end{corollary}
+
+\subsection{Theorem~\ref{thm:judge-valid}}
+
 \begin{proof}[Proof (of Theorem~\ref{thm:judge-valid})]
-  By induction on the assumed derivation.
+  By induction on the assumed derivation, making implicit use of
+  Corollary~\ref{cor:syntactic-kind-pres} for the shorthand \(\Gamma \vdash t
+  \tpsynthleads T\) and mostly omitting annotated substitutions when these are
+  clear from context (we show the first one that occurs).
   We show a few interesting cases.
@@
   \[
+    \infer{\Gamma\vdash\abs{\lambda}{X}{\kappa}{T'} \tpsynth \abs{\Pi}{X}{\kappa}{\kappa'}}{\Gamma \vdash \kappa & \Gamma,X:\kappa\vdash T'\tpsynth\kappa'}
+  \]
+  By assumption, the first premise, and the context formation rules, we have
+  \(\vdash \Gamma,X:\kappa\).
+  From this and the IH on the second premise, we have \(\Gamma,X:\kappa \vdash
+  \kappa'\).
+  We can then conclude that \(\Gamma \vdash \abs{\Pi}{X}{\kappa}{\kappa'}\)
+
+  \startcase{.2cm}
+  \[
     \infer{\Gamma\vdash T_1 \cdot T_2 \tpsynth [T_2/X]\kappa_1}{\Gamma\vdash T_1 \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1} & \Gamma\vdash T_2 \tpsynth \kappa_2' & \kappa_2 \cong \kappa_2'}
   \]
-
   By the IH on the first premise, we have \(\Gamma \vdash
@@
   \[
-    \infer{\Gamma\vdash t\tpsynth T'}{\Gamma\vdash t \tpsynth T & T' \leadsto^*_\beta T'}
-  \]
-  By the IH, \(\Gamma \vdash T \tpsynth \star\).
-  By generalizing Theorem~\ref{thm:syntactic-kind-pres} to \(\leadsto^*_\beta\),
-  we have \(\Gamma \vdash T' \tpsynth \kappa'\) for some \(\kappa' \cong
-  \star\).
-  This means that in fact \(\kappa' = \star\), as desired.
-
-  \startcase{.2cm}
-  \[
-    \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynth \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'}
+    \infer{\Gamma\vdash T\ t \tpsynth [t/x]^{T'}\kappa}{\Gamma\vdash T \tpsynth \abs{\Pi}{x}{T'}{\kappa} & \Gamma\vdash t \tpcheck T'}
+  \]
+  By the IH on the first premise, \(\Gamma \vdash \abs{\Pi}{x}{T'}{\kappa}\).
+  By inversion, \(\Gamma \vdash T' \tpsynth \star\) and \(\Gamma,x:T' \vdash
+  \kappa\).
+  So, from the first of these and by context formation \(\vdash \Gamma,x:T'\).
+  From the second premise and the \(\chi\) rule, \(\Gamma \vdash \chi\ T\
+  \mhyph\ t \tpsynth T\).
+  Finally by Lemma~\ref{lem:subst-class}, \(\Gamma \vdash [t/x]^{T} \kappa\).
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynthleads \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'}
   \]
@@
   \star\).
-  By inversion of this, we obtain both \(\Gamma \vdash T' \tpsynth \star\), from
-  which it follows by the \(\chi\) rule and second premise that \(\Gamma \vdash
-  \chi\ T'\ \mhyph\ t \tpsynth T'\), and that \(\Gamma,x:T' \vdash T \tpsynth
-  \star\), from which we obtain \(\Gamma \vdash [t/x]^{T'}\ T \tpsynth \star\)
-  using Lemma~\ref{lem:subst-class}
-
-  \startcase{.2cm}
-  \[
-    \infer{\Gamma \vdash \rho\ t'\ @x.T\ \mhyph\ t \tpsynth [\beta\{t_2\}/x]^{\texttt{Top}}\ T}
+  By inversion of this, we obtain both \(\Gamma \vdash T' \tpsynth \star\), and
+  that \(\Gamma,x:T' \vdash T \tpsynth \star\).
+  From the first of these and the context formation rules, we have \(\vdash
+  \Gamma,x:T'\).
+  Finally, from Lemma~\ref{lem:subst-class} we have \(\Gamma \vdash [t'/x]T
+  \tpsynth \star\).
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma\vdash t \cdot T' \tpsynth [T'/X]T
+    }{
+      \Gamma\vdash t \tpsynthleads \abs{\forall}{X}{\kappa}{T} & \Gamma\vdash T'
+      \tpsynth \kappa' & \kappa'\cong\kappa
+    }
+  \]
+  From the IH on the first premise, \(\Gamma \vdash \abs{\forall}{X}{\kappa}{T}
+  \tpsynth \star\).
+  By inversion of this, we have both \(\Gamma \vdash \kappa\) and
+  \(\Gamma,X:\kappa \vdash T \tpsynth \star\).
+  From the first of these, context formation rules, and assumption we have
+  \(\vdash \Gamma,X:\kappa\).
+  By the IH on the second premise, we also have \(\Gamma \vdash \kappa'\), and
+  from this we have \(\vdash \Gamma,X:\kappa'\).
+  We combine this with the earlier derivation of \(\Gamma,X:\kappa \vdash T
+  \tpsynth \star\), the third premise, and Lemma~\ref{lem:ctxt-conv-class} to
+  obtain \(\Gamma,X:\kappa' \vdash T \tpsynth \star\).
+  Finally, we use Lemma~\ref{lem:subst-class} to obtain \(\Gamma \vdash [T'/X]T
+  \tpsynth \star\).
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma \vdash \rho\ t'\ @x\langle t_2 \rangle.T\ \mhyph\ t \tpsynth [t_2/x]\ T}
     {
-      \begin{array}{cc}
-        \Gamma \vdash t' \tpsynth \{t_1 \simeq t_2\}
-        & \quad \Gamma \vdash [\beta\{t_2\}/x]^{\texttt{Top}}\ T \tpsynth \star
-        \\ \Gamma \vdash t \tpsynth T'
+      \begin{array}{ccc}
+        \Gamma \vdash t' \tpsynthleads \{t_1 \simeq t_2'\}
+        & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma)
+        & |t_2'| =_{\beta\eta} |t_2|
+        \\ \Gamma \vdash [t_2/x]\ T \tpsynth \star
+        & \Gamma \vdash t \tpsynth T'
         & T' \cong [t_1/x]T
@@
   \]
-  The desired result is given by the second assumption (reading first left to right, then top to bottom).
+  The desired result is given by the first premise of the second row.
 
@@
   \]
-  Give to us by assumption (the first premise).
+  Give to us by the first premise.
 \end{proof}

### 2020-12-30 docs/semantics/paper.tex
docs: semantics: simplify \varphi rule

@@
           {
-            \begin{array}{ccc}
-              \Gamma\vdash t\tpsynthleads \{t_1\simeq t_2\}
-              &
+              \Gamma\vdash t\tpcheck \{t' \simeq t''\}
               & \Gamma\vdash t' \Leftrightarrow T
-              \\ |t_1| =_{\beta\eta} |t'|
-              & |t_2| =_{\beta\eta} |t''|
               & \textit{FV}(t'') \subseteq \textit{dom}(\Gamma)
-            \end{array}
           }
@@
     |\delta\ \mhyph\ t| & = & \absu{\lambda}{x}{x}&\ &
-    |\rho\ t\ @x.T\ \mhyph\ t'| & = & |t'| \\
+    |\rho\ t\ @x\langle t_2 \rangle.T\ \mhyph\ t'| & = & |t'| \\
     |\varphi\ t\ \mhyph\ t'\ \{t''\}| & = & |t''| &\ &
@@
   {
-    \begin{array}{ccc}
-      \Gamma\vdash t\tpsynthleads \{t_1\simeq t_2\}
-      &
+    \Gamma\vdash t\tpcheck \{t' \simeq t''\}
     & \Gamma\vdash t' \Leftrightarrow T
-      \\ |t_1| =_{\beta\eta} |t'|
-      & |t_2| =_{\beta\eta} |t''|
     & \textit{FV}(t'') \subseteq \textit{dom}(\Gamma)
-    \end{array}
-  }
-\]
-By the IH for the first premise, $\sigma|t_1| =_{\cbe} \sigma|t_2|$.
-Using the first and second premise of the second row, we obtain \(\sigma|t'|
-=_{\cbe} \sigma|t''|\).
+  }
+\]
+By the IH for the first premise, $\sigma|t'| =_{\cbe} \sigma|t''|$ (we can see
+that \(\textit{FV}(t'\ t'') \subseteq \Gamma\), so \(\interp{\{t' \simeq
+  t''\}}_{\sigma,\rho}\) is defined).
 By the IH for the second premise,
@@
   \[
-    \infer{\Gamma_1,x_1:T_1,\Gamma_2 \vdash \varphi\ t\ \mhyph\ t'\ \{t''\} \tpcheck T_3}
+    \infer{\Gamma_1,x_1:T_1,\Gamma_2\vdash \varphi\ t\ \mhyph\ t'\ \{t''\} \Leftrightarrow T_3}
     {
-      \begin{array}{ccc}
-        \Gamma_1,x_1:T_1,\Gamma_2 \vdash t\tpsynthleads \{t_1\simeq t_2\}
-        &
-        & \Gamma_1,x_1:T_1,\Gamma_2\vdash t' \tpcheck T_3
-        \\ |t_1| =_{\beta\eta} |t'|
-        & |t_2| =_{\beta\eta} |t''|
+      \Gamma_1,x_1:T_1,\Gamma_2\vdash t\tpcheck \{t' \simeq t''\}
+      & \Gamma_1,x_1:T_1,\Gamma_2\vdash t' \Leftrightarrow T_3
       & \textit{FV}(t'') \subseteq \textit{dom}(\Gamma_1,x_1:T_1,\Gamma_2)
-      \end{array}
     }
@@
   By the IH on the first premise (first row), \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash
-  t \tpsynth T\) (with no greater depth) for some \(T \cong \{t_1 \simeq t_2\}\).
-  This means that \(T \leadsto_\beta^* \{t_3 \simeq t_4\}\) for some \(t_3,t_4\)
-  such that \(|t_3| =_{\beta\eta} |t_1|\) and \(|t_4| =_{\beta\eta} |t_2|\).
-  Using the first and second premise of the second row and transitivity of
-  \(\beta\eta\)-equivalence, \(|t_3| =_{\beta\eta} |t'|\) and \(|t_4|
-  =_{\beta\eta} |t''|\).
-  By the IH on the second premise of the first row, \(\Gamma_1,x_1:T_2,\Gamma_2
-  \vdash t' \tpcheck T_4\) (with no greater depth, and where \(T_4\) is the type
-  given to us in the lemma and \(T_4 \cong T_3\)).
-  Also, we can see from the last premise of the second row that
-  \(\textit{FV}(t') \subseteq \textit{dom}(\Gamma_1,x_1:T_2,\Gamma_2)\).
-  So, we use the rule to conclude \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash \varphi\ t\
-  \mhyph\ t'\ \{t''\} \tpcheck T_4\).
+  t \tpcheck \{t' \simeq t''\}\) (with no greater depth).
+  If this is type synthesis, then by the second premise
+  \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash t' \tpsynth T_4\) for some \(T_4 \cong
+  T_3\).
+  If this is type checking, by the second premise \(\Gamma_1,x_1:T_2,\Gamma_2
+  \vdash t' \tpcheck T_4\) since \(T_4\) (given to us in this case) is
+  convertible with \(T_3\).
+  Either way, we use the rule to conclude (note
+  \(\textit{dom}(\Gamma_1,x_1:T_1,\Gamma_2) =
+  \textit{dom}(\Gamma_1,x_1:T_2,\Gamma_2)\)), and the measure is preserved.
 
@@
   We conclude using the rule, with the measure preserved.
-
-  \startcase{.2cm}
-  \[
-    \infer{\Gamma_1,x_1:T_1,\Gamma_2\vdash \varphi\ t\ \mhyph\ t'\ \{t''\} \tpsynth T}
-    {
-      \begin{array}{ccc}
-        \Gamma_1,x_1:T_2,\Gamma_2\vdash t\tpsynthleads \{t_1\simeq t_2\}
-        &
-        & \Gamma_1,x_1:T_1,\Gamma_2\vdash t' \tpsynth T
-        \\ |t_1| =_{\beta\eta} |t'|
-        & |t_2| =_{\beta\eta} |t''|
-        & \textit{FV}(t'') \subseteq \textit{dom}(\Gamma)
-      \end{array}
-    }
-  \]
-
-  Similar to the checking case for \(\varphi\), only now after using the IH the
-  second premise of the first row synthesizes a type \(T' \cong T\).
-  The whole expression, then, synthesizes type \(T'\).
 \end{proof}

### 2021-01-02 docs/semantics/paper.tex
docs: semantics:
- make rho checking
- give full listing of conversion rules
- minor revisions to logical consistency, normalization theorem

@@
 
-\usepackage{amsmath,amssymb,amsthm}
+\usepackage{amsmath,amssymb,amsthm,MnSymbol}
 \usepackage{url}
@@
 Call-by-name reduction is \emph{not} what underpins the
-congruence relation \(\cong\) for types or terms, which is full
+equivalence relation \(\cong\) for types and kinds, which is full
 \(\beta\)-equivalence (for types) and \(\beta\eta\)-equivalence (for terms)
@@
 This is lifted to the conversion relation on types \(T \cong T'\) and kinds
-\(\kappa \cong \kappa'\), whose central rules are given in
+\(\kappa \cong \kappa'\), whose rules are given in
 Figure~\ref{fig:conv}.
-That figure omits the various congruence rules needed to equate bigger
-expressions by equating subexpressions (the convertibility rules for kinds
-consists entirely of congruence rules).
 
@@
              & \Gamma \vdash t_1 \tpcheck T_1
-             \\ \Gamma \vdash t_2 \tpcheck [t/x]^{T_1}\ T_2
+             \\ \Gamma \vdash t_2 \tpcheck [t_1/x]^{T_1}\ T_2
              & |t_1| =_{\beta\eta} |t_2|
@@
           {\Gamma\vdash t\tpcheck \{\absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}} &
-    \infer{\Gamma \vdash \rho\ t'\ @x\langle t_2 \rangle.T\ \mhyph\ t \tpsynth [t_2/x]\ T}
+    \infer{\Gamma \vdash \rho\ t\ @x\langle t_1 \rangle.T'\ \mhyph\ t' \tpcheck T}
           {
           \begin{array}{ccc}
-            \Gamma \vdash t' \tpsynthleads \{t_1 \simeq t_2'\}
-            & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma)
-            & |t_2'| =_{\beta\eta} |t_2|
-            \\ \Gamma \vdash [t_2/x]\ T \tpsynth \star
-            & \Gamma \vdash t \tpsynth T'
-            & T' \cong [t_1/x]T
+            \Gamma \vdash t \tpsynthleads \{t_1' \simeq t_2\}
+            & \textit{FV}(t_1) \subseteq \textit{dom}(\Gamma)
+            & |t_1'| =_{\beta\eta} |t_1|
+            \\ \Gamma \vdash [t_1/x] T' \tpsynth \star
+            & \Gamma \vdash t' \tpcheck [t_1/x]T'
+            & [t_2/x]T' \cong T
           \end{array}
@@
     |\delta\ \mhyph\ t| & = & \absu{\lambda}{x}{x}&\ &
-    |\rho\ t\ @x\langle t_2 \rangle.T\ \mhyph\ t'| & = & |t'| \\
+    |\rho\ t\ @x\langle t_1 \rangle.T'\ \mhyph\ t'| & = & |t'| \\
     |\varphi\ t\ \mhyph\ t'\ \{t''\}| & = & |t''| &\ &
@@
   \[
-  \begin{array}{ll}
-    \infer{T \cong T'}{T \leadsto^*_\beta T_1 & T' \leadsto^*_\beta T_2 & T_1\cong^t T_2}  & 
-    \infer{T \cong T'}{T \cong^t T'} \\ \\
-    \infer{T\ t \cong^t T\ t'}{T \cong^t T' & |t| =_{\beta\eta} |t'|} &
-    \infer{\{ t_1 \simeq t_2 \} \cong^t \{ t_1'\ \simeq t_2' \}}{|t_1| =_{\beta\eta} |t_1'| & |t_2| =_{\beta\eta} |t_2'|}
+    \begin{array}{c}
+      \fbox{\(T_1 \cong T_2\)}
+      \quad \fbox{\(T_1 \cong^{\text{t}} T_2\)}
+      \\ \\
+      \infer{
+       T_1 \cong T_2
+      }{
+       T_1 \leadsto^*_\beta T_1' \nleadsto_\beta
+       \quad T_2 \leadsto^*_\beta T_2' \nleadsto_\beta
+       \quad T_1' \cong^{\text{t}} T_2'
+      }
+      \\ \\
+      \begin{array}{cc}
+        \infer{X \cong^{\text{t}} X}{}
+        &
+        \infer{
+         \abs{\forall}{X}{\kappa_1}{T_1} \cong^{\text{t}} \abs{\forall}{X}{\kappa_2}{T_2}
+        }{
+         \kappa_1 \cong \kappa_2
+         \quad  T_1 \cong T_2
+        }
+        \\ \\
+          \infer{
+           \abs{\forall}{x}{T_1}{T_1'} \cong^{\text{t}} \abs{\forall}{x}{T_2}{T_2'}
+          }{
+           T_1 \cong T_2
+           \quad  T_1' \cong T_2'
+          }
+        & 
+        \infer{
+         \abs{\Pi}{x}{T_1}{T_1'} \cong^{\text{t}} \abs{\Pi}{x}{T_2}{T_2'}
+        }{
+         T_1 \cong T_2
+         \quad  T_1' \cong T_2'
+        }
+        \\ \\ 
+          \infer{
+           \abs{\lambda}{x}{T_1}{T_1'} \cong^{\text{t}} \abs{\lambda}{x}{T_2}{T_2'}
+          }{
+           T_1 \cong T_2
+           \quad  T_1' \cong T_2'
+          }
+        & 
+        \infer{
+         \abs{\lambda}{X}{\kappa_1}{T_1} \cong^{\text{t}} \abs{\lambda}{X}{\kappa_2}{T_2}
+        }{
+         \kappa_1 \cong \kappa_2
+         \quad  T_1 \cong T_2
+        }
+        \\ \\
+          \infer{
+          \abs{\iota}{x}{T_1}{T_1'} \cong^{\text{t}} \abs{\iota}{x}{T_2}{T_2'}
+          }{
+          T_1 \cong T_2
+          \quad  T_1' \cong T_2'
+          }
+        & 
+        \infer{
+         T_1\ t_1 \cong^{\text{t}} T_2\ t_2
+        }{
+         T_1 \cong^{\text{t}} T_2 \quad |t_1| =_{\beta\eta} |t_2|
+        }
+        \\ \\
+          \infer{
+           T_1 \cdot T_1' \cong^{\text{t}} T_2 \cdot T_2'
+          }{
+           T_1 \cong^{\text{t}} T_2
+           \quad T_1' \cong T_2'
+          }
+        &
+        \infer{
+         \{ t_1 \simeq t_2 \} \cong^{\text{t}} \{ t_1'\ \simeq t_2' \}
+        }{
+         |t_1| =_{\beta\eta} |t_1'| \quad |t_2| =_{\beta\eta} |t_2'|
+        }
       \end{array}
-  \]
-  \caption{Non-congruence rules for conversion}
+      \\ \\ \fbox{\(\kappa_1 \cong \kappa_2\)}
+      \\ \\
+      \begin{array}{cc}
+        \infer{
+         \star \cong \star
+        }{}
+        & \infer{
+           \abs{\Pi}{x}{T_1}{\kappa_1} \cong \abs{\Pi}{x}{T_2}{\kappa_2}
+          }{
+           T_1 \cong T_2
+           \quad \kappa_1 \cong \kappa_2
+          }
+        \\ \\
+        \infer{
+         \abs{\Pi}{X}{\kappa_1}{\kappa_1'} \cong \abs{\Pi}{X}{\kappa_2}{\kappa_2'}
+        }{
+         \kappa_1 \cong \kappa_2
+         \quad \kappa_1' \cong \kappa_2'
+        }
+      \end{array}
+    \end{array}
+  \]
+  \caption{Conversion rules for classifiers}
   \label{fig:conv}
@@
   to classify all untyped lambda calculus terms.
-\item The \(\rho\) construct allows one to rewrite occurrences of \(t_1\) to
-  \(t_2'\) in the synthesized type of its second subexpression.
+\item The \(\rho\) construct allows one to rewrite occurrences of \(t_2\) in the
+  checked type of the \(T\) of the whole expression to \(t_1\) before checking
+  the type of the subexpression \(t'\).
   The version presented here requires a type annotations (``guide'') \(@x\langle
-  t_2 \rangle.T\), where \(x\) indicates the occurrences of \(t_1\) to
-  substitute and \(t_2\) is a term, \(\beta\eta\)-equivalent to \(t_2'\), that
-  is used to replace \(x\).
-  There is no requirement that \(t_1\) and \(t_2\) will have the same type, or
-  even be typable.
-  However, \([t_2/x]\ T\) must be a well-kinded type and \([t_1/x]T\) (which
-  might be ill-kinded) must be convertible with \(T'\) modulo erasure.
+  t_1 \rangle.T\).
+  The first subexpression \(t\) must synthesize (possibly after some
+  normalization) an equation type \(\{t_1' \simeq t_2\}\), and the user provided
+  term \(t_1\) must be \(\beta\eta\)-convertible (modulo erasure) to \(t_1'\).
+  Then, the user provided type \(T'\) is checked to have kind \(\star\) after
+  replacing occurrences of \(x\) with \(t_1\), and the subexpression \(t'\) is
+  checked against this type.
+  Finally, \([t_2/x]T'\) (which need not be well-kinded) must be convertible
+  with the expected type \(T\).
   In Cedille, the guide is optional and the construct may be used to rewrite a
@@
     $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho}\in \mathcal{R}$.
-\item If $T \cong T'$ or $T \cong^t T'$ and $\interp{T}_{\sigma,\rho}$ and $\interp{T'}_{\sigma,\rho}$ are both defined, then they are equal.
+\item If $T \cong T'$ or $T \cong^{\text{t}} T'$ and $\interp{T}_{\sigma,\rho}$ and $\interp{T'}_{\sigma,\rho}$ are both defined, then they are equal.
 \item If \(\kappa \cong \kappa'\) and \(\interp{\kappa}_{\sigma,\rho}\) and
@@
 \label{thm:consis}
-  There is no term $t$ such that $\cdot \vdash t : \abs{\forall}{X}{\star}{X}$.
+  There is no term $t$ such that $\cdot \vdash t \tpsynth \abs{\forall}{X}{\star}{X}$.
 \end{theorem}
@@
   \label{thm:cedille-termination}
-  Suppose \(\cdot \vdash t \tpsynth T\) and there exists a closed term \(t'\)
-  such that \(\cdot \vdash t' \tpsynth T \to \abs{\Pi}{x}{T_1}{T_2}\) for some
-  \(T_1,T_2\) and \(|t'| = \absu{\lambda}{x}{x}\).
+  Suppose \(\cdot \vdash t \tpsynth T\) and \(\cdot \vdash t' \tpsynth T \to
+  \abs{\Pi}{x}{T_1}{T_2}\), and furthermore that \(|t'| =
+  \absu{\lambda}{x}{x}\).
   Then \(|t|\) is call-by-name normalizing.
@@
 \[
-  \infer{\Gamma \vdash \rho\ t'\ @x\langle t_2 \rangle.T\ \mhyph\ t \tpsynth [t_2/x]\ T}
+  \infer{\Gamma \vdash \rho\ t\ @x\langle t_1 \rangle.T'\ \mhyph\ t' \tpcheck T}
   {
     \begin{array}{ccc}
-      \Gamma \vdash t' \tpsynthleads \{t_1 \simeq t_2'\}
-      & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma)
-      & |t_2'| =_{\beta\eta} |t_2|
-      \\ \Gamma \vdash [t_2/x]\ T \tpsynth \star
-      & \Gamma \vdash t \tpsynth T'
-      & T' \cong [t_1/x]T
+      \Gamma \vdash t \tpsynthleads \{t_1' \simeq t_2\}
+      & \textit{FV}(t_1) \subseteq \textit{dom}(\Gamma)
+      & |t_1'| =_{\beta\eta} |t_1|
+      \\ \Gamma \vdash [t_1/x] T' \tpsynth \star
+      & \Gamma \vdash t' \tpcheck [t_1/x]T'
+      & [t_2/x]T' \cong T
     \end{array}
@@
 \]
-By the IH applied to the first premise in the first row, $\sigma|t_1| =_{\cbe}
-\sigma|t_2'|$.
+By the IH applied to the first premise in the first row, $\sigma|t_1'| =_{\cbe}
+\sigma|t_2|$.
 With the second and third premise, and from the assumption \(\sigma \in
-\interp{\Gamma}\), we have \(\sigma|t_2'| =_{\cbe} \sigma|t_2|\) (and
-\(\sigma|t_2|\) is closed), so we obtain
+\interp{\Gamma}\), we have \(\sigma|t_1'| =_{\cbe} \sigma|t_1|\) (and
+\(\sigma|t_1|\) is closed), so we obtain
 \([\sigma|t_1|]_{\cbe} = [\sigma|t_2|]_{\cbe}\).
 By the IH applied to the first premise of the second row,
-\(\interp{[t_2/x]T}_{\sigma,\rho} \in \mathcal{R}\) and so is defined.
-By the IH applied to the first premise in the second row,
-\([\sigma|t|]_{c\beta\eta} \in \interp{T'}_{\sigma,\rho}\).
+\(\interp{[t_1/x]T}_{\sigma,\rho} \in \mathcal{R}\) and so is defined.
 By the IH applied to the second premise in the second row,
-\(\interp{T'}_{\sigma,\rho} = \interp{[t_1/x]T}_{\sigma,\rho}\).
+\([\sigma|t'|]_{c\beta\eta} \in \interp{[t_1/x]T'}_{\sigma,\rho}\).
+By the IH applied to the second premise in the second row,
+\(\interp{T}_{\sigma,\rho} = \interp{[t_2/x]T'}_{\sigma,\rho}\).
 The result then follows by applying Lemma~\ref{lem:termsubstinterp}.
@@
 \subsection*{Proof of part (5) }
-
-\startcase{.2cm}
-\[
-    \infer{T \cong T'}{T \leadsto^*_\beta T_1 & T' \leadsto^*_\beta T_2 & T_1\cong^t T_2}  
+We show the cases for the non-congruential rules of Figure~\ref{fig:conv}.
+
+\startcase{.2cm}
+\[
+  \infer{
+    T_1 \cong T_2
+  }{
+    T_1 \leadsto^*_\beta T_1' \nleadsto_\beta
+    \quad T_2 \leadsto^*_\beta T_2' \nleadsto_\beta
+    \quad T_1' \cong^{\text{t}} T_2'
+  }
 \]
@@
 \begin{array}{lll}
-  \interp{T}_{\sigma,\rho} & = & \interp{T_1}_{\sigma,\rho}\\
-  \interp{T'}_{\sigma,\rho} & = & \interp{T_2}_{\sigma,\rho}
+  \interp{T_1}_{\sigma,\rho} & = & \interp{T_1'}_{\sigma,\rho}\\
+  \interp{T_2}_{\sigma,\rho} & = & \interp{T_2'}_{\sigma,\rho}
 \end{array}
 \]
-By the IH for the third premise, we have $\interp{T_1}_{\sigma,\rho} = \interp{T_2}_{\sigma,\rho}$,
+By the IH for the third premise, we have $\interp{T_1'}_{\sigma,\rho} = \interp{T_2'}_{\sigma,\rho}$,
 which suffices.
@@
 \[
-    \infer{T \cong T'}{T \cong^t T'} 
-\]
-By the IH.
-
-\startcase{.2cm}
-\[
-    \infer{T\ t \cong^t T'\ t'}{T \cong^t T' & |t| =_{\beta\eta} |t'|}
-\]
-By the semantics, $\interp{T\ t}_{\sigma,\rho} =
-\interp{T}_{\sigma,\rho}([\sigma|t|]_{\cbe})$.  By the second premise
+    \infer{T_1\ t_1 \cong^{\text{t}} T_2\ t_2}{T_1 \cong^{\text{t}} T_1 & |t_1| =_{\beta\eta} |t_2|}
+\]
+By the semantics, $\interp{T_1\ t_1}_{\sigma,\rho} =
+\interp{T_1}_{\sigma,\rho}([\sigma|t_1|]_{\cbe})$.  By the second premise
 and the IH for the first premise, this equals
-$\interp{T'}_{\sigma,\rho}([\sigma|t'|]_{\cbe})$, as required.
-
-\startcase{.2cm}
-\[
-    \infer{\{ t_1 \simeq t_2 \} \cong^t \{ t_1'\ \simeq t_2' \}}{|t_1| =_{\beta\eta} |t_1'| & |t_2| =_{\beta\eta} |t_2'|}
+$\interp{T_2}_{\sigma,\rho}([\sigma|t_2|]_{\cbe})$, as required.
+
+\startcase{.2cm}
+\[
+    \infer{\{ t_1 \simeq t_2 \} \cong^{\text{t}} \{ t_1'\ \simeq t_2' \}}{|t_1| =_{\beta\eta} |t_1'| & |t_2| =_{\beta\eta} |t_2'|}
 \]
@@
   \[
+    \infer{\Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash \rho\ t\ @x\langle t_1 \rangle.T'\ \mhyph\ t' \tpcheck T_3}
+    {
+      \begin{array}{ccc}
+        \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \tpsynthleads \{t_1' \simeq t_2\}
+        & \textit{FV}(t_1) \subseteq \textit{dom}(\Gamma_1,X_1:\kappa_1,\Gamma_2)
+        & |t_1'| =_{\beta\eta} |t_1|
+        \\ \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash [t_1/x] T' \tpsynth \star
+        & \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \tpcheck [t_1/x]T'
+        & [t_2/x]T' \cong T_3
+      \end{array}
+    }
+  \]
+  By the IH on the first premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t'
+  \tpsynthleads \{t_1 \simeq t_2'\}\).
+  By the IH on the second premise of the second row,
+  \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash [t_1/x]\ T \tpsynth \star\)
+  (\(\star\) is only convertible with itself).
+  By the IH the second premise of the second row,
+  \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t \tpcheck [t_1/x]T'\).
+  By assumption, the third premise of the second row, and transitivity of
+  convertibility, we have \([t_2/x]T' \cong T_4\) (\(T_4\) is the type we must
+  check the entire expression against).
+  We use the rule to conclude (\(\textit{dom}(\Gamma_1,X_1:\kappa_1,\Gamma_2) =
+  \textit{dom}(\Gamma_1,X_1:\kappa_1',\Gamma_2)\)), noting that the measure is
+  preserved.
+
+  \startcase{.2cm}
+  \[
     \infer{
@@
   [T'/X]T\), and the measure is preserved.
-
-  \startcase{.2cm}
-  \[
-    \infer{\Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash \rho\ t'\ @x\langle t_2 \rangle.T\ \mhyph\ t \tpsynth [t_2/x]\ T}
-    {
-      \begin{array}{ccc}
-        \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t' \tpsynthleads \{t_1 \simeq t_2'\}
-        & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma_1,X_1:\kappa_1,\Gamma_2)
-        & |t_2'| =_{\beta\eta} |t_2|
-        \\ \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash [t_2/x]\ T \tpsynth \star
-        & \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \tpsynth T'
-        & T' \cong [t_1/x]T
-      \end{array}
-    }
-  \]
-  By the IH on the first premise, \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t'
-  \tpsynthleads \{t_1 \simeq t_2'\}\).
-  By the IH on the second premise of the second row,
-  \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash [t_2/x]\ T \tpsynth \star\)
-  (\(\star\) is only convertible with itself).
-  We apply the IH to the second premise of the second row and use the rule to conclude
-  (\(\textit{dom}(\Gamma_1,X_1:\kappa_1,\Gamma_2) =
-  \textit{dom}(\Gamma_1,X_1:\kappa_1',\Gamma_2)\)).
 \end{proof}
@@
 
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma_1,x_1:T_1,\Gamma_2 \vdash \rho\ t\ @x\langle t_1 \rangle.T'\ \mhyph\ t' \tpcheck T_3}
+    {
+      \begin{array}{ccc}
+        \Gamma_1,x_1:T_1,\Gamma_2 \vdash t \tpsynthleads \{t_1' \simeq t_2\}
+        & \textit{FV}(t_1) \subseteq \textit{dom}(\Gamma_1,x_1:T_1,\Gamma_2)
+        & |t_1'| =_{\beta\eta} |t_1|
+        \\ \Gamma_1,x_1:T_1,\Gamma_2 \vdash [t_1/x] T' \tpsynth \star
+        & \Gamma_1,x_1:T_1,\Gamma_2 \vdash t \tpcheck [t_1/x]T'
+        & [t_2/x]T' \cong T_3
+      \end{array}
+    }
+  \]
+  By the IH on the first premise, \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash t \tpsynth
+  T''\) (with no greater depth) for some \(T'' \cong \{t_1' \simeq t_2\}\).
+  That means \(T'' \leadsto_\beta^* \{t_3 \simeq t_4\}\) for some \(t_3,t_4\)
+  such that \(|t_3| =_{\beta\eta} |t_1'|\) and \(|t_4| =_{\beta\eta} |t_2|\).
+  So we further obtain that \(|t_3| =_{\beta\eta} |t_1|\).
+  By the IH on the first premise of the second row, \(\Gamma_1,x_1:T_2,\Gamma_2
+  \vdash [t_1/x]T \tpsynth \star\).
+  By the IH on the second premise of the second row, \(\Gamma,x_1:T_2,\Gamma_2
+  \vdash t \tpcheck [t_1/x]T'\).
+  By assumption, the third premise of the second row, and transitivity of
+  convertibility, we have \([t_2/x]T' \cong T_4\).
+  We conclude with the rule, preserving the measure.
+
   % type synthesis
@@
   \[
-    \infer{\Gamma_1,x_1:T_1,\Gamma_2 \vdash \rho\ t'\ @x\langle t_2 \rangle.T\ \mhyph\ t \tpsynth [t_2/x]\ T}
-    {
-      \begin{array}{ccc}
-        \Gamma_1,x_1:T_1,\Gamma_2 \vdash t' \tpsynthleads \{t_1 \simeq t_2'\}
-        & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma)
-        & |t_2'| =_{\beta\eta} |t_2|
-        \\ \Gamma_1,x_1:T_1,\Gamma_2 \vdash [t_2/x]\ T \tpsynth \star
-        & \Gamma_1,x_1:T_1,\Gamma_2 \vdash t \tpsynth T'
-        & T' \cong [t_1/x]T
-      \end{array}
-    }
-  \]
-  By the IH on the first premise, \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash t \tpsynth
-  T_3\) (with no greater depth) for some \(T_3 \cong \{t_1 \simeq t_2'\}\).
-  This means \(T_3 \leadsto_\beta^* \{t_3 \simeq t_4\}\) for some \(t_3,t_4\)
-  such that \(|t_3| =_{\beta\eta} t_1\) and \(|t_4| =_{\beta\eta} |t_2'|\).
-  So we further obtain \(|t_4| =_{\beta\eta} |t_2|'\).
-  By the IH on the first premise of the second row, \(\Gamma_1,x_1:T_2,\Gamma_2
-  \vdash [t_2/x]T \tpsynth \star\) with no greater depth.
-  By the IH on the second premise of the second row, \(\Gamma_1,x_1:T_2,\Gamma_2
-  \vdash t \tpsynth T''\) (with no greater depth) for some \(T'' \cong T'\).
-  With the final premise of the second row and transitivity, we obtain \(T''
-  \cong [t_1/x]T\).
-  We conclude with the rule, preserving the measure.
-
-  \startcase{.2cm}
-  \[
     \infer{
@@
       \(\Gamma_1 \vdash t' \tpsynth T'\) then \(\Gamma_1,[t'/x]\Gamma_2 \vdash [t'/x]t
-      \tpsynth [t'/x]T\)
+      \Leftrightarrow [t'/x]T\)
 
@@
       \(\Gamma_1 \vdash T \tpsynth \kappa\) then \(\Gamma_1,[T/X]\Gamma_2 \vdash
-      [T/X]t \tpsynth [T/X]T'\)
+      [T/X]t \Leftrightarrow [T/X]T'\)
     \end{itemize}
@@
   \[
-    \infer{\Gamma_1,x:T,\Gamma_2 \vdash \rho\ t''\ @x_1.T_1\ \mhyph\ t' \tpsynth [t_2/x_1] T_1}
+    \infer{\Gamma_1,x:T,\Gamma_2 \vdash \rho\ t''\ @x_1\langle t_1 \rangle.T_2\ \mhyph\ t' \tpcheck T_1}
     {
-      \begin{array}{cc}
-        \Gamma_1,x:T,\Gamma_2 \vdash t'' \tpsynth \{t_1 \simeq t_2\}
-        & \quad \Gamma_1,x:T,\Gamma_2 \vdash [t_2/x_1]\ T_1 \tpsynth \star
-        \\ \Gamma_1,x:T,\Gamma_2 \vdash t' \tpsynth T'
-        & T' \cong [t_1/x_1]T_1
+      \begin{array}{ccc}
+        \Gamma_1,x:T,\Gamma_2 \vdash t'' \tpsynthleads \{t_1' \simeq t_2\}
+        & \textit{FV}(t_1) \subseteq \textit{dom}(\Gamma_1,x:T,\Gamma_2)
+        & |t_1'| =_{\beta\eta} |t_1|
+        \\ \Gamma_1,x:T,\Gamma_2 \vdash [t_1/x_1] T_2 \tpsynth \star
+        & \Gamma_1,x:T,\Gamma_2 \vdash t' \tpcheck [t_1/x_1]T_2
+        & [t_2/x_1]T_2 \cong T_1
       \end{array}
     }
-  \]
-  From the IH, we have that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]t'' \tpsynth \{[t/x]t_1
-  \simeq [t/x]t_2\}\), that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x][t_2/x]T_1
-  \tpsynth \star\), and that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]t' \tpsynth
-  [t/x]T'\).
-  From the last premise, we may conclude that \([t/x]T' \cong
-  [t/x][t_1/x_1]T_1\).
+    % \infer{\Gamma_1,x:T,\Gamma_2 \vdash \rho\ t''\ @x_1.T_1\ \mhyph\ t' \tpsynth [t_2/x_1] T_1}
+    % {
+    %   \begin{array}{cc}
+    %     \Gamma_1,x:T,\Gamma_2 \vdash t'' \tpsynth \{t_1 \simeq t_2\}
+    %     & \quad \Gamma_1,x:T,\Gamma_2 \vdash [t_2/x_1]\ T_1 \tpsynth \star
+    %     \\ \Gamma_1,x:T,\Gamma_2 \vdash t' \tpsynth T'
+    %     & T' \cong [t_1/x_1]T_1
+    %   \end{array}
+    % }
+  \]
+  From the IH, we have that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]t'' \tpsynth \{[t/x]t_1'
+  \simeq [t/x]t_2\}\), that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x][t_1/x_1]T_2
+  \tpsynth \star\), and that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]t' \tpcheck
+  [t/x][t_1/x_1]T_2\).
+  From the last premise, we may conclude that \([t/x]T_1 \cong
+  [t/x][t_2/x_1]T_1\).
+  We also see from the premises that \(\textit{FV}([t/x]t_1) \subseteq
+  \textit{dom}(\Gamma_1,[t/x]\Gamma_2)\) and that \(|[t/x]t_1'| =_{\beta\eta}
+  |[t/x]t_1|\).
   Applying the rule and permuting substitutions gives us the desired result
-  (note for example that \([t/x][t_1/x_1]T_1 = [[t/x]t_1/x_1]T_1\)).
+  (note for example that \([t/x][t_1/x_1]T_1 = [[t/x]t_1/x_1][t/x]T_1\)).
 \end{proof}
@@
 
-  \startcase{.2cm}
-  \[
-    \infer{\Gamma \vdash \rho\ t'\ @x\langle t_2 \rangle.T\ \mhyph\ t \tpsynth [t_2/x]\ T}
-    {
-      \begin{array}{ccc}
-        \Gamma \vdash t' \tpsynthleads \{t_1 \simeq t_2'\}
-        & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma)
-        & |t_2'| =_{\beta\eta} |t_2|
-        \\ \Gamma \vdash [t_2/x]\ T \tpsynth \star
-        & \Gamma \vdash t \tpsynth T'
-        & T' \cong [t_1/x]T
-      \end{array}
-    }
-  \]
-  The desired result is given by the first premise of the second row.
+  % Type checking, not needed
+  % \startcase{.2cm}
+  % \[
+  %   \infer{\Gamma \vdash \rho\ t\ @x\langle t_1 \rangle.T'\ \mhyph\ t' \tpcheck T}
+  %   {
+  %     \begin{array}{ccc}
+  %       \Gamma \vdash t \tpsynthleads \{t_1' \simeq t_2\}
+  %       & \textit{FV}(t_1) \subseteq \textit{dom}(\Gamma)
+  %       & |t_1'| =_{\beta\eta} |t_1|
+  %       \\ \Gamma \vdash [t_1/x] T' \tpsynth \star
+  %       & \Gamma \vdash t' \tpcheck [t_1/x]T'
+  %       & [t_2/x]T' \cong T
+  %     \end{array}
+  %   }
+  % \]
+  % The desired result is given by the first premise of the second row.
 

### 2021-01-10 docs/semantics/paper.tex
docs: semantics: fix rho, again

@@
           {\Gamma\vdash t\tpcheck \{\absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}} &
-    \infer{\Gamma \vdash \rho\ t\ @x\langle t_1 \rangle.T'\ \mhyph\ t' \tpcheck T}
+    \infer{\Gamma \vdash \rho\ t\ @x\langle t_2 \rangle.T'\ \mhyph\ t' \tpcheck T}
           {
           \begin{array}{ccc}
-            \Gamma \vdash t \tpsynthleads \{t_1' \simeq t_2\}
-            & \textit{FV}(t_1) \subseteq \textit{dom}(\Gamma)
-            & |t_1'| =_{\beta\eta} |t_1|
-            \\ \Gamma \vdash [t_1/x] T' \tpsynth \star
-            & \Gamma \vdash t' \tpcheck [t_1/x]T'
-            & [t_2/x]T' \cong T
+            \Gamma \vdash t \tpsynthleads \{t_1 \simeq t_2'\}
+            & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma)
+            & |t_2'| =_{\beta\eta} |t_2|
+            \\ \Gamma \vdash [t_2/x] T' \tpsynth \star
+            & \Gamma \vdash t' \tpcheck [t_2/x]T'
+            & [t_1/x]T' \cong T
           \end{array}
@@
     |\delta\ \mhyph\ t| & = & \absu{\lambda}{x}{x}&\ &
-    |\rho\ t\ @x\langle t_1 \rangle.T'\ \mhyph\ t'| & = & |t'| \\
+    |\rho\ t\ @x\langle t_2 \rangle.T'\ \mhyph\ t'| & = & |t'| \\
     |\varphi\ t\ \mhyph\ t'\ \{t''\}| & = & |t''| &\ &
@@
   to classify all untyped lambda calculus terms.
-\item The \(\rho\) construct allows one to rewrite occurrences of \(t_2\) in the
-  checked type of the \(T\) of the whole expression to \(t_1\) before checking
+\item The \(\rho\) construct allows one to rewrite occurrences of \(t_1\) in the
+  checked type \(T\) of the whole expression to \(t_2\) before checking
   the type of the subexpression \(t'\).
   The version presented here requires a type annotations (``guide'') \(@x\langle
-  t_1 \rangle.T\).
+  t_2 \rangle.T\).
   The first subexpression \(t\) must synthesize (possibly after some
-  normalization) an equation type \(\{t_1' \simeq t_2\}\), and the user provided
-  term \(t_1\) must be \(\beta\eta\)-convertible (modulo erasure) to \(t_1'\).
-  Then, the user provided type \(T'\) is checked to have kind \(\star\) after
-  replacing occurrences of \(x\) with \(t_1\), and the subexpression \(t'\) is
+  normalization) an equation type \(\{t_1 \simeq t_2'\}\), and the user provided
+  term \(t_2\) must be \(\beta\eta\)-convertible (modulo erasure) to \(t_2'\).
+  The user provided type \(T'\) is then checked to have kind \(\star\) after
+  replacing occurrences of \(x\) with \(t_2\), and the subexpression \(t'\) is
   checked against this type.
-  Finally, \([t_2/x]T'\) (which need not be well-kinded) must be convertible
+  Finally, \([t_1/x]T'\) (which need not be well-kinded) must be convertible
   with the expected type \(T\).
@@
 \[
-  \infer{\Gamma \vdash \rho\ t\ @x\langle t_1 \rangle.T'\ \mhyph\ t' \tpcheck T}
+  \infer{\Gamma \vdash \rho\ t\ @x\langle t_2 \rangle.T'\ \mhyph\ t' \tpcheck T}
   {
     \begin{array}{ccc}
-      \Gamma \vdash t \tpsynthleads \{t_1' \simeq t_2\}
-      & \textit{FV}(t_1) \subseteq \textit{dom}(\Gamma)
-      & |t_1'| =_{\beta\eta} |t_1|
-      \\ \Gamma \vdash [t_1/x] T' \tpsynth \star
-      & \Gamma \vdash t' \tpcheck [t_1/x]T'
-      & [t_2/x]T' \cong T
+      \Gamma \vdash t \tpsynthleads \{t_1 \simeq t_2'\}
+      & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma)
+      & |t_2'| =_{\beta\eta} |t_2|
+      \\ \Gamma \vdash [t_2/x] T' \tpsynth \star
+      & \Gamma \vdash t' \tpcheck [t_2/x]T'
+      & [t_1/x]T' \cong T
     \end{array}
@@
 \]
-By the IH applied to the first premise in the first row, $\sigma|t_1'| =_{\cbe}
-\sigma|t_2|$.
+By the IH applied to the first premise in the first row, $\sigma|t_1| =_{\cbe}
+\sigma|t_2'|$.
 With the second and third premise, and from the assumption \(\sigma \in
-\interp{\Gamma}\), we have \(\sigma|t_1'| =_{\cbe} \sigma|t_1|\) (and
-\(\sigma|t_1|\) is closed), so we obtain
+\interp{\Gamma}\), we have \(\sigma|t_2'| =_{\cbe} \sigma|t_2|\) (and
+\(\sigma|t_2|\) is closed), so we obtain
 \([\sigma|t_1|]_{\cbe} = [\sigma|t_2|]_{\cbe}\).
 By the IH applied to the first premise of the second row,
-\(\interp{[t_1/x]T}_{\sigma,\rho} \in \mathcal{R}\) and so is defined.
+\(\interp{[t_2/x]T}_{\sigma,\rho} \in \mathcal{R}\) and so is defined.
 By the IH applied to the second premise in the second row,
-\([\sigma|t'|]_{c\beta\eta} \in \interp{[t_1/x]T'}_{\sigma,\rho}\).
+\([\sigma|t'|]_{c\beta\eta} \in \interp{[t_2/x]T'}_{\sigma,\rho}\).
 By the IH applied to the second premise in the second row,
-\(\interp{T}_{\sigma,\rho} = \interp{[t_2/x]T'}_{\sigma,\rho}\).
+\(\interp{T}_{\sigma,\rho} = \interp{[t_1/x]T'}_{\sigma,\rho}\).
 The result then follows by applying Lemma~\ref{lem:termsubstinterp}.
@@
   \[
-    \infer{\Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash \rho\ t\ @x\langle t_1 \rangle.T'\ \mhyph\ t' \tpcheck T_3}
+    \infer{\Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash \rho\ t\ @x\langle t_2 \rangle.T'\ \mhyph\ t' \tpcheck T}
     {
       \begin{array}{ccc}
-        \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \tpsynthleads \{t_1' \simeq t_2\}
-        & \textit{FV}(t_1) \subseteq \textit{dom}(\Gamma_1,X_1:\kappa_1,\Gamma_2)
-        & |t_1'| =_{\beta\eta} |t_1|
-        \\ \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash [t_1/x] T' \tpsynth \star
-        & \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \tpcheck [t_1/x]T'
-        & [t_2/x]T' \cong T_3
+        \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \tpsynthleads \{t_1 \simeq t_2'\}
+        & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma_1,X_1:\kappa_1,\Gamma_2)
+        & |t_2'| =_{\beta\eta} |t_2|
+        \\ \Gamma_1,X_1:\kappa_2,\Gamma_2 \vdash [t_2/x] T' \tpsynth \star
+        & \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t' \tpcheck [t_2/x]T'
+        & [t_1/x]T' \cong T
       \end{array}
     }
+    % \infer{\Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash \rho\ t\ @x\langle t_1 \rangle.T'\ \mhyph\ t' \tpcheck T_3}
+    % {
+    %   \begin{array}{ccc}
+    %     \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \tpsynthleads \{t_1' \simeq t_2\}
+    %     & \textit{FV}(t_1) \subseteq \textit{dom}(\Gamma_1,X_1:\kappa_1,\Gamma_2)
+    %     & |t_1'| =_{\beta\eta} |t_1|
+    %     \\ \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash [t_1/x] T' \tpsynth \star
+    %     & \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t \tpcheck [t_1/x]T'
+    %     & [t_2/x]T' \cong T_3
+    %   \end{array}
+    % }
   \]
@@
   \tpsynthleads \{t_1 \simeq t_2'\}\).
-  By the IH on the second premise of the second row,
-  \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash [t_1/x]\ T \tpsynth \star\)
+  By the IH on the first premise of the second row,
+  \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash [t_2/x]\ T \tpsynth \star\)
   (\(\star\) is only convertible with itself).
   By the IH the second premise of the second row,
-  \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t \tpcheck [t_1/x]T'\).
+  \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash t \tpcheck [t_2/x]T'\).
   By assumption, the third premise of the second row, and transitivity of
-  convertibility, we have \([t_2/x]T' \cong T_4\) (\(T_4\) is the type we must
+  convertibility, we have \([t_1/x]T' \cong T_4\) (\(T_4\) is the type we must
   check the entire expression against).
@@
   \[
-    \infer{\Gamma_1,x_1:T_1,\Gamma_2 \vdash \rho\ t\ @x\langle t_1 \rangle.T'\ \mhyph\ t' \tpcheck T_3}
+  \infer{\Gamma_1,x_1:T_1,\Gamma_2 \vdash \rho\ t\ @x\langle t_2 \rangle.T'\ \mhyph\ t' \tpcheck T}
   {
     \begin{array}{ccc}
-        \Gamma_1,x_1:T_1,\Gamma_2 \vdash t \tpsynthleads \{t_1' \simeq t_2\}
-        & \textit{FV}(t_1) \subseteq \textit{dom}(\Gamma_1,x_1:T_1,\Gamma_2)
-        & |t_1'| =_{\beta\eta} |t_1|
-        \\ \Gamma_1,x_1:T_1,\Gamma_2 \vdash [t_1/x] T' \tpsynth \star
-        & \Gamma_1,x_1:T_1,\Gamma_2 \vdash t \tpcheck [t_1/x]T'
-        & [t_2/x]T' \cong T_3
+      \Gamma_1,x_1:T_1,\Gamma_2 \vdash t \tpsynthleads \{t_1 \simeq t_2'\}
+      & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma_1,x_1:T_1,\Gamma_2)
+      & |t_2'| =_{\beta\eta} |t_2|
+      \\ \Gamma_1,x_1:T_1,\Gamma_2 \vdash [t_2/x] T' \tpsynth \star
+      & \Gamma_1,x_1:T_1,\Gamma_2 \vdash t' \tpcheck [t_2/x]T'
+      & [t_1/x]T' \cong T
     \end{array}
@@
   By the IH on the first premise, \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash t \tpsynth
-  T''\) (with no greater depth) for some \(T'' \cong \{t_1' \simeq t_2\}\).
+  T''\) (with no greater depth) for some \(T'' \cong \{t_1 \simeq t_2'\}\).
   That means \(T'' \leadsto_\beta^* \{t_3 \simeq t_4\}\) for some \(t_3,t_4\)
-  such that \(|t_3| =_{\beta\eta} |t_1'|\) and \(|t_4| =_{\beta\eta} |t_2|\).
-  So we further obtain that \(|t_3| =_{\beta\eta} |t_1|\).
+  such that \(|t_3| =_{\beta\eta} |t_1|\) and \(|t_4| =_{\beta\eta} |t_2'|\).
+  So we further obtain that \(|t_4| =_{\beta\eta} |t_2|\).
   By the IH on the first premise of the second row, \(\Gamma_1,x_1:T_2,\Gamma_2
-  \vdash [t_1/x]T \tpsynth \star\).
+  \vdash [t_2/x]T \tpsynth \star\).
   By the IH on the second premise of the second row, \(\Gamma,x_1:T_2,\Gamma_2
-  \vdash t \tpcheck [t_1/x]T'\).
+  \vdash t \tpcheck [t_2/x]T'\).
   By assumption, the third premise of the second row, and transitivity of
-  convertibility, we have \([t_2/x]T' \cong T_4\).
+  convertibility, we have \([t_1/x]T' \cong T_4\).
   We conclude with the rule, preserving the measure.
@@
   \[
-    \infer{\Gamma_1,x:T,\Gamma_2 \vdash \rho\ t''\ @x_1\langle t_1 \rangle.T_2\ \mhyph\ t' \tpcheck T_1}
+    \infer{\Gamma_1,x:T,\Gamma_2 \vdash \rho\ t''\ @x_1\langle t_2 \rangle.T_2\ \mhyph\ t' \tpcheck T_1}
     {
       \begin{array}{ccc}
-        \Gamma_1,x:T,\Gamma_2 \vdash t'' \tpsynthleads \{t_1' \simeq t_2\}
-        & \textit{FV}(t_1) \subseteq \textit{dom}(\Gamma_1,x:T,\Gamma_2)
-        & |t_1'| =_{\beta\eta} |t_1|
-        \\ \Gamma_1,x:T,\Gamma_2 \vdash [t_1/x_1] T_2 \tpsynth \star
-        & \Gamma_1,x:T,\Gamma_2 \vdash t' \tpcheck [t_1/x_1]T_2
-        & [t_2/x_1]T_2 \cong T_1
+        \Gamma_1,x:T,\Gamma_2 \vdash t'' \tpsynthleads \{t_1 \simeq t_2'\}
+        & \textit{FV}(t_2) \subseteq \textit{dom}(\Gamma_1,x:T,\Gamma_2)
+        & |t_2'| =_{\beta\eta} |t_2|
+        \\ \Gamma_1,x:T,\Gamma_2 \vdash [t_2/x_1] T_2 \tpsynth \star
+        & \Gamma_1,x:T,\Gamma_2 \vdash t' \tpcheck [t_2/x_1]T_2
+        & [t_1/x_1]T_2 \cong T_1
       \end{array}
     }
-    % \infer{\Gamma_1,x:T,\Gamma_2 \vdash \rho\ t''\ @x_1.T_1\ \mhyph\ t' \tpsynth [t_2/x_1] T_1}
-    % {
-    %   \begin{array}{cc}
-    %     \Gamma_1,x:T,\Gamma_2 \vdash t'' \tpsynth \{t_1 \simeq t_2\}
-    %     & \quad \Gamma_1,x:T,\Gamma_2 \vdash [t_2/x_1]\ T_1 \tpsynth \star
-    %     \\ \Gamma_1,x:T,\Gamma_2 \vdash t' \tpsynth T'
-    %     & T' \cong [t_1/x_1]T_1
-    %   \end{array}
-    % }
-  \]
-  From the IH, we have that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]t'' \tpsynth \{[t/x]t_1'
-  \simeq [t/x]t_2\}\), that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x][t_1/x_1]T_2
+  \]
+  From the IH, we have that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]t'' \tpsynth \{[t/x]t_1
+  \simeq [t/x]t_2'\}\), that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x][t_2/x_1]T_2
   \tpsynth \star\), and that \(\Gamma_1,[t/x]\Gamma_2 \vdash [t/x]t' \tpcheck
-  [t/x][t_1/x_1]T_2\).
+  [t/x][t_2/x_1]T_2\).
   From the last premise, we may conclude that \([t/x]T_1 \cong
-  [t/x][t_2/x_1]T_1\).
-  We also see from the premises that \(\textit{FV}([t/x]t_1) \subseteq
-  \textit{dom}(\Gamma_1,[t/x]\Gamma_2)\) and that \(|[t/x]t_1'| =_{\beta\eta}
-  |[t/x]t_1|\).
+  [t/x][t_1/x_1]T_1\).
+  We also see from the premises that \(\textit{FV}([t/x]t_2) \subseteq
+  \textit{dom}(\Gamma_1,[t/x]\Gamma_2)\) and that \(|[t/x]t_2'| =_{\beta\eta}
+  |[t/x]t_2|\).
   Applying the rule and permuting substitutions gives us the desired result
-  (note for example that \([t/x][t_1/x_1]T_1 = [[t/x]t_1/x_1][t/x]T_1\)).
+  (note for example that \([t/x][t_2/x_1]T_1 = [[t/x]t_2/x_1][t/x]T_1\)).
 \end{proof}

### 2021-04-13 docs/semantics/paper.tex
docs: semantics: typo
docs: semantics: make delta subterm synthesizing

@@
 \newcommand{\interp}[1]{\llbracket #1 \rrbracket} 
+\newcommand{\leadstoc}[0]{\ensuremath{\leadsto_{\mathbf{n}}}}
+\newcommand{\leadstocs}[0]{\ensuremath{\leadsto_{\mathbf{n}}^*}}
+\newcommand{\nleadstoc}[0]{\ensuremath{\nleadsto_{\mathbf{n}}}}
 \newcommand{\tpcheck}[0]{\Leftarrow}
 \newcommand{\tpsynth}[0]{\Rightarrow}
-\newcommand{\tpsynthleads}[0]{\ensuremath{\overset{\leadsto_\beta^*}{\Rightarrow}}}
+\newcommand{\tpsynthleads}[0]{\ensuremath{\overset{\leadstocs}{\Rightarrow}}}
 \newcommand{\cbe}[0]{c\beta\eta}
@@
 In terms, when the type of an introduction form is checked or the type of an
-elimination form is synthesized, we use call-by-name normalization for types
-\(\leadsto_\beta^*\) to put types in weak head normal form, revealing type
-constructors.
+elimination form is synthesized, we use call-by-name normalization, written
+\(\leadstoc\) and \(\leadstocs\) for the reflexive transitive closure, to put
+types in weak head normal form, revealing type constructors.
 We abbreviate the conjuction ``term \(t\) synthesizes some type, and that type
@@
       \Gamma \vdash t \tpsynthleads T\ =\ \exists T'.\ \Gamma
-      \vdash t \tpsynth T' \land T' \leadsto_\beta^* T
+      \vdash t \tpsynth T' \land T' \leadstocs T
       \\ \\ \\
@@
     \infer{\Gamma\vdash t\tpcheck T}{\Gamma\vdash t\tpsynth T' & T' \cong T} \\ \\
-    \infer{\Gamma\vdash \absu{\lambda}{x}{t} \tpcheck T}{T \leadsto_\beta^* \abs{\Pi}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t\tpcheck T_2} &
+    \infer{\Gamma\vdash \absu{\lambda}{x}{t} \tpcheck T}{T \leadstocs \abs{\Pi}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t\tpcheck T_2} &
     \infer{\Gamma\vdash t\ t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynthleads \abs{\Pi}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\
@@
     \infer{\Gamma\vdash \absu{\Lambda}{X}{t} \tpcheck T'}
-          {T' \leadsto_\beta^* \abs{\forall}{X}{\kappa}{T} & \Gamma,X:\kappa\vdash t \tpcheck T} &
+          {T' \leadstocs \abs{\forall}{X}{\kappa}{T} & \Gamma,X:\kappa\vdash t \tpcheck T} &
     \infer{\Gamma\vdash t \cdot T' \tpsynth [T'/X]T}
@@
     \infer{\Gamma\vdash \absu{\Lambda}{x}{t} \tpcheck T}
-          {T \leadsto_\beta^* \abs{\forall}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t \tpcheck T_2 & x\not\in\textit{FV}(|t|)} &
+          {T \leadstocs \abs{\forall}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t \tpcheck T_2 & x\not\in\textit{FV}(|t|)} &
     \infer{\Gamma\vdash t\ \mhyph t' \tpsynth [t'/x]^{T'}T}{\Gamma\vdash t \tpsynthleads \abs{\forall}{x}{T'}{T} & \Gamma\vdash t' \tpcheck T'} \\ \\
@@
            \begin{array}{cc}
-             T \leadsto_\beta^* \abs{\iota}{x}{T_1}{T_1}
+             T \leadstocs \abs{\iota}{x}{T_1}{T_1}
              & \Gamma \vdash t_1 \tpcheck T_1
@@
     \infer{\Gamma\vdash \beta\{t'\} \tpcheck T}
-          {T \leadsto_\beta^* \{t_1 \simeq t_2\} \quad \textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}  \\ \\    
+          {T \leadstocs \{t_1 \simeq t_2\} \quad \textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}  \\ \\    
     \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}
-          {\Gamma\vdash t\tpcheck \{\absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}} &
+          {
+            \Gamma \vdash t \tpsynth T'
+            \quad T' \cong \{\absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}
+          } &
     \infer{\Gamma \vdash \rho\ t\ @x\langle t_2 \rangle.T'\ \mhyph\ t' \tpcheck T}
@@
       }{
-       T_1 \leadsto^*_\beta T_1' \nleadsto_\beta
-       \quad T_2 \leadsto^*_\beta T_2' \nleadsto_\beta
+       T_1 \leadstocs T_1' \nleadstoc
+       \quad T_2 \leadstocs T_2' \nleadstoc
        \quad T_1' \cong^{\text{t}} T_2'
@@
 \item If $\Gamma\vdash t \tpcheck T$ and $\interp{T}_{\sigma,\rho}\in \mathcal{R}$, then
-    $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho}\in \mathcal{R}$.
+    $[\sigma |t|]_{\cbe}\in\interp{T}_{\sigma,\rho}$.
 \item If $T \cong T'$ or $T \cong^{\text{t}} T'$ and $\interp{T}_{\sigma,\rho}$ and $\interp{T'}_{\sigma,\rho}$ are both defined, then they are equal.
@@
   \label{thm:syntactic-kind-pres}
-  If \(\Gamma \vdash T \tpsynth \kappa\) and \(T \leadsto_{\beta} T'\) then
+  If \(\Gamma \vdash T \tpsynth \kappa\) and \(T \leadstoc T'\) then
   \(\Gamma \vdash T' \tpsynth \kappa'\) for some \(\kappa'\) such that \(\kappa
@@
   \label{lem:interppres}
-  If $T \leadsto^*_{\beta} T'$ and $\interp{T}_{\sigma,\rho}$ is defined, then $\interp{T'}_{\sigma,\rho}$ is also defined and equals $\interp{T}_{\sigma,\rho}$.
+  If $T \leadstocs T'$ and $\interp{T}_{\sigma,\rho}$ is defined, then $\interp{T'}_{\sigma,\rho}$ is also defined and equals $\interp{T}_{\sigma,\rho}$.
 \end{lemma}
@@
 \[
-  \infer{\Gamma\vdash \absu{\lambda}{x}{t} \tpcheck T}{T \leadsto_\beta^* \abs{\Pi}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t\tpcheck T_2}
+  \infer{\Gamma\vdash \absu{\lambda}{x}{t} \tpcheck T}{T \leadstocs \abs{\Pi}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t\tpcheck T_2}
 \]
@@
   \infer{\Gamma\vdash \absu{\Lambda}{X}{t} \tpcheck T'}
-  {T' \leadsto_\beta^* \abs{\forall}{X}{\kappa}{T} & \Gamma,X:\kappa\vdash t \tpcheck T}
+  {T' \leadstocs \abs{\forall}{X}{\kappa}{T} & \Gamma,X:\kappa\vdash t \tpcheck T}
 \]
@@
   \infer{\Gamma\vdash \absu{\Lambda}{x}{t} \tpcheck T}
-  {T \leadsto_\beta^* \abs{\forall}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t \tpcheck T_2 & x\not\in\textit{FV}(|t|)}
+  {T \leadstocs \abs{\forall}{x}{T_1}{T_2} & \Gamma,x:T_1\vdash t \tpcheck T_2 & x\not\in\textit{FV}(|t|)}
 \]
@@
     \begin{array}{cc}
-      T \leadsto_\beta^* \abs{\iota}{x}{T_1}{T_1}
+      T \leadstocs \abs{\iota}{x}{T_1}{T_1}
       & \Gamma \vdash t_1 \tpcheck T_1
@@
   \infer{\Gamma\vdash \beta\{t'\} \tpcheck T}
-  {T \leadsto_\beta^* \{t_1 \simeq t_2\} \quad \textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}
+  {T \leadstocs \{t_1 \simeq t_2\} \quad \textit{FV}(t')\subseteq \textit{dom}(\Gamma) \quad |t_1| =_{\beta\eta} |t_2|}
 \]
@@
 \[
-    \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpsynth \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  
-\]
-By the semantics of equality types, $[\sigma|t|]_{\cbe}$ cannot be in the interpretation of the equation in the premise,
-since the two terms in question are closed and not $\beta\eta$-equal.  By the IH applied to the first premise, however,
-$[\sigma|t|]_{\cbe}$ is in the interpretation of that equation.  This is a contradiction.
+  \infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}
+  {
+    \Gamma \vdash t \tpsynth T'
+    \quad T' \cong \{\absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}
+  }
+    %infer{\Gamma\vdash \delta\ \mhyph\ t \tpcheck T}{\Gamma\vdash t\tpsynth \{ \absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}}  
+\]
+By the inductive hypothesis, \([\sigma|t|]_{\cbe} \in \interp{T'} \in
+\mathcal{R}\).
+It is easy to see that the meaning of \(\{\absu{\lambda}{x}{\absu{\lambda}{y}{x}
+\simeq \absu{\lambda}{x}{\absu{\lambda}{y}}}\}\) exists and is in
+\(\mathcal{R}\), and is in fact the empty set.
+By mutual induction on the second premise we have \([\sigma|t|]_{\cbe} \in
+\interp{\{\absu{\lambda}{x}{\absu{\lambda}{y}{x} \simeq
+    \absu{\lambda}{x}{\absu{\lambda}{y}}}\}}_{\sigma,\rho}\), a contradiction.
 
@@
   }{
-    T_1 \leadsto^*_\beta T_1' \nleadsto_\beta
-    \quad T_2 \leadsto^*_\beta T_2' \nleadsto_\beta
+    T_1 \leadstocs T_1' \nleadstoc
+    \quad T_2 \leadstocs T_2' \nleadstoc
     \quad T_1' \cong^{\text{t}} T_2'
@@
 
-First a few lemmas (easy proofs omitted):
+First a few lemmas (easy cases omitted):
 
@@
   \end{itemize}
-  Furthermore, the resulting typing derivations derivations have depths no
+  Furthermore, the resulting typing derivations have depths no
   larger than the assumed ones.
@@
     }{
-      T_3 \leadsto_\beta^* \abs{\Pi}{x}{T_1}{T_2}
+      T_3 \leadstocs \abs{\Pi}{x}{T_1}{T_2}
       & \Gamma_1,X_1:\kappa_1,\Gamma_2,x:T_1\vdash t\tpcheck T_2
@@
   \abs{\Pi}{x}{T_1}{T_2}\).
-  This means \(T_4 \leadsto_\beta^* \abs{\Pi}{x}{T_1'}{T_2'}\) for some \(T_1'
+  This means \(T_4 \leadstocs \abs{\Pi}{x}{T_1'}{T_2'}\) for some \(T_1'
   \cong T_1\) and \(T_2' \cong T_2\).
@@
     }{
-      T_3 \leadsto_\beta^* \abs{\forall}{X}{\kappa}{T}
+      T_3 \leadstocs \abs{\forall}{X}{\kappa}{T}
       & \Gamma_1,X_1:\kappa_1,\Gamma_2,X:\kappa\vdash t \tpcheck T
@@
   \abs{\forall}{X}{\kappa}{T}\).
-  This means \(T_4 \leadsto_\beta^* \abs{\forall}{X}{\kappa'}{T'}\) for some
+  This means \(T_4 \leadstocs \abs{\forall}{X}{\kappa'}{T'}\) for some
   \(\kappa' \cong \kappa\) and \(T' \cong T\).
@@
     }{
-      T_3 \leadsto_\beta^* \{t_1 \simeq t_2\}
+      T_3 \leadstocs \{t_1 \simeq t_2\}
       & \textit{FV}(t')\subseteq \textit{dom}(\Gamma_1,X_1:\kappa_1,\Gamma_2)
@@
   \cong \{t_1 \simeq t_2\}\).
-  This means that \(T_4 \leadsto_\beta^* \{t_3 \simeq t_4\}\) for some
+  This means that \(T_4 \leadstocs \{t_3 \simeq t_4\}\) for some
   \(t_3,t_4\) such that \(|t_3| =_{\beta\eta} |t_1|\) and \(|t_4| =_{\beta\eta}
@@
   \(\Gamma_1,X_1:\kappa_1',\Gamma_2 \vdash [t_2/x]\ T \tpsynth \star\)
-  (\(\star\) is only convertible with itself).
+  (as \(\star\) is only convertible with itself).
   By the IH the second premise of the second row,
@@
       \begin{array}{cc}
-        T_3 \leadsto_\beta^* \abs{\iota}{x}{T_1}{T_1}
+        T_3 \leadstocs \abs{\iota}{x}{T_1}{T_1}
         & \Gamma_1,X_1:\kappa_1,\Gamma_2 \vdash t_1 \tpcheck T_1
@@
   relation, \(T_4 \cong \abs{\iota}{x}{T_1}{T_2}\).
-  This means that \(T_4 \leadsto_\beta^* \abs{\iota}{x}{T_1'}{T_2'}\) for some
+  This means that \(T_4 \leadstocs \abs{\iota}{x}{T_1'}{T_2'}\) for some
   \(T_1' \cong T_1\) and \(T_2' \cong T_2\).
@@
     }{
-      T \leadsto_\beta^* \abs{\Pi}{x}{T_5}{T_6}
+      T \leadstocs \abs{\Pi}{x}{T_5}{T_6}
       & \Gamma_1,x_1:T_1,\Gamma_2,x:T_5\vdash t\tpcheck T_6
@@
   \abs{\Pi}{x}{T_5}{T_6}\).
-  This means that \(T_4 \leadsto_\beta^* \abs{\Pi}{x}{T_5'}{T_6'}\) for some
+  This means that \(T_4 \leadstocs \abs{\Pi}{x}{T_5'}{T_6'}\) for some
   \(T_5' \cong T_5\) and \(T_6' \cong T_6\).
@@
     }{
-      T' \leadsto_\beta^* \abs{\forall}{X}{\kappa}{T}
+      T' \leadstocs \abs{\forall}{X}{\kappa}{T}
       & \Gamma_1,x_1:T_1,\Gamma_2,X:\kappa \vdash t \tpcheck T
@@
   \abs{\forall}{X}{\kappa}{T}\).
-  This means that \(T_4 \leadsto_\beta^* \abs{\forall}{X}{\kappa''}{T''}\) for
+  This means that \(T_4 \leadstocs \abs{\forall}{X}{\kappa''}{T''}\) for
   some \(\kappa'' \cong \kappa\) and \(T'' \cong T\).
@@
       \begin{array}{cc}
-        T_3 \leadsto_\beta^* \abs{\iota}{x}{T}{T'}
+        T_3 \leadstocs \abs{\iota}{x}{T}{T'}
         & \Gamma_1,x_1:T_1,\Gamma_2 \vdash t \tpcheck T
@@
   \abs{\iota}{x}{T}{T'}\).
-  This means \(T_4 \leadsto \abs{\iota}{x}{T''}{T'''}\) for some \(T'' \cong T\)
+  This means \(T_4 \leadstocs \abs{\iota}{x}{T''}{T'''}\) for some \(T'' \cong T\)
   and \(T''' \cong T'\).
@@
     }{
-      T_3 \leadsto_\beta^* \{t_1 \simeq t_2\}
+      T_3 \leadstocs \{t_1 \simeq t_2\}
       \quad \textit{FV}(t')\subseteq \textit{dom}(\Gamma_1,x_1:T_1,\Gamma_2)
@@
   \{t_1 \simeq t_2\}\).
-  This means \(T_4 \leadsto_\beta^* \{t_3 \simeq t_4\}\) for some \(t_3,t_4\)
+  This means \(T_4 \leadstocs \{t_3 \simeq t_4\}\) for some \(t_3,t_4\)
   such that \(|t_3| =_{\beta\eta} |t_1|\) and \(|t_4| =_{\beta\eta} |t_2|\).
@@
   T''\) (with no greater depth) for some \(T'' \cong \{t_1 \simeq t_2'\}\).
-  That means \(T'' \leadsto_\beta^* \{t_3 \simeq t_4\}\) for some \(t_3,t_4\)
+  That means \(T'' \leadstocs \{t_3 \simeq t_4\}\) for some \(t_3,t_4\)
   such that \(|t_3| =_{\beta\eta} |t_1|\) and \(|t_4| =_{\beta\eta} |t_2'|\).
@@
 
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma_1,x_1:T_1,\Gamma_2\vdash \delta\ \mhyph\ t \tpcheck T_3}
+    {
+      \Gamma_1,x_1 : T_1,\Gamma_2 \vdash t \tpsynth T'
+      \quad T' \cong \{\absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}
+    }
+  \]
+  By the IH on the first premise, \(\Gamma_1,x:T_2,\Gamma_2 \vdash t \tpsynth
+  T''\) for some \(T'' \cong T'\).
+  By transitivity of convertibility and the second premise, \(T'' \cong
+  \{\absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq
+  \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}\).
+  Using the \(\delta\) rule, we conclude that \(\Gamma_1,x_1:T_2,\Gamma_2 \vdash
+  \delta\ \mhyph\ t \tpcheck T_4\).
+
   % type synthesis
@@
     }{
-      (x : T)\in\Gamma_1,x_1:T_1,\Gamma_2
+      (x : T)\in\Gamma_1,x_1 : T_1,\Gamma_2
     }
@@
   derivation is no greater than that of the first premise.
-  From this we obtain that \(T_3 \leadsto^*_\beta \abs{\Pi}{x}{T_4'}{T_4}\) for
+  From this we obtain that \(T_3 \leadstocs \abs{\Pi}{x}{T_4'}{T_4}\) for
   some \(T_4' \cong T'\) and \(T_4 \cong T\).
@@
   depth.
-  This means \(T_3 \leadsto_\beta^* \abs{\forall}{X}{\kappa''}{T''}\) for some
+  This means \(T_3 \leadstocs \abs{\forall}{X}{\kappa''}{T''}\) for some
   \(\kappa'' \cong \kappa\) and \(T'' \cong T\).
@@
   (note for example that \([t/x][t_2/x_1]T_1 = [[t/x]t_2/x_1][t/x]T_1\)).
+
+  \startcase{.2cm}
+  \[
+    \infer{\Gamma_1,x:T,\Gamma_2 \vdash \delta\ \mhyph\ t \tpcheck T_2}
+    {
+      \Gamma_1,x:T,\Gamma_2 \vdash t \tpsynth T_2'
+      \quad T_2' \cong \{\absu{\lambda}{x}{\absu{\lambda}{y}{x}} \simeq \absu{\lambda}{x}{\absu{\lambda}{y}{y}}\}
+    }
+  \]
+
+  From the IH, we have that \(\Gamma_1[t'/x]\Gamma_2 \vdash [t'/x] t \tpsynth
+  [t'/x]T_2'\).
+  From the second premise, we have \([t'/x]T_2' \cong
+  \{\absu{\lambda}{x}{\absu{\lambda}{y}{x} \simeq
+    \absu{\lambda}{x}{\absu{\lambda}{y}{y}}}\}\).
+  From the \(\delta\) rule, we conclude that
+  \(\Gamma_1,[t'/x]\Gamma_2 \vdash \delta\ \mhyph\ [t'/x]t \tpcheck [t'/x]T_2\).
 \end{proof}
@@
   We may rule out the cases where \(T\) is a variable or formed by a type
-  constructor, as this would contradict the assumption that \(T \leadsto_{\beta}
-  T'\) for some \(T'\) (recall that \(\leadsto_\beta\) is call-by-name reduction
-  for types).
+  constructor, as this would contradict the assumption that \(T \leadstoc
+  T'\) for some \(T'\).
   We omit type annotations from substitutions when they are clear from the context.
@@
 
-  There are two subcases to consider for the derivation of \(T\ t \leadsto_\beta
-  T'\). In the first case, we have \(T \leadsto_\beta T''\) for some \(T''\) (so
+  There are two subcases to consider for the derivation of \(T\ t \leadstoc
+  T'\). In the first case, we have \(T \leadstoc T''\) for some \(T''\) (so
   \(T' = T''\ t\)). By the IH, we have \(\Gamma \vdash T'' \tpsynth
@@
   There are two subcases to consider for the derivation of \(T_1 \cdot T_2
-  \leadsto_\beta T'\). In the first subcase, we have \(T_1 \leadsto_\beta
+  \leadstoc T'\). In the first subcase, we have \(T_1 \leadstoc
   T_1''\) for some \(T_1''\) (so \(T' = T_1'' \cdot T_2\)). By the IH, we have
@@
   \label{cor:syntactic-kind-pres}
-  If \(\Gamma \vdash T \tpsynth \kappa\) and \(T \leadsto_\beta^* T'\) then
+  If \(\Gamma \vdash T \tpsynth \kappa\) and \(T \leadstocs T'\) then
   \(\Gamma \vdash T' \tpsynth \kappa'\) for some \(\kappa' \cong \kappa\).

### 2021-04-14 docs/semantics/paper.tex
docs: semantics: final revisions

@@
+% -*- eval: (flyspell-mode); -*-
 \documentclass{article}
@@
 Eliminations (CDLE).  This document presents the version of CDLE as of
-December 1, 2020.  We have made many changes from the first paper on CDLE~\cite{stump17},
+April 13, 2021.  We have made many changes from the first paper on CDLE~\cite{stump17},
 mostly in the form of dropping constructs we discovered (to our
@@
 constructs, which allow deriving induction principles within the
-theory for inductive datatypes.  The goal is to support usual
-idioms of dependently typed programming and proving as in Agda or
+theory for lambda encodings of inductive datatypes.  The goal is to support
+usual idioms of dependently typed programming and proving as in Agda or
 similar tools, but using pure lambda encodings for all data, and
-requiring a much smaller core theory.
-
-The current Cedille implementation of CDLE extends the system
-described below with a number of features intended to make programming
-in the system more convenient and with less redundancy.  These
-features all compile away to a slightly simplified version of the
+requiring a much smaller core theory suitable for formal verification.
+
+The current Cedille implementation of CDLE extends with a number of features
+intended to make programming in the system more convenient and with less
+redundancy.
+These features all compile away to a slightly simplified version of the
 theory presented in this document, called Cedille Core, described
 here: \url{https://github.com/astump/cedille-core-spec}.
+At the time of writing, the Cedille implementation lags behind the formulation
+of CDLE in this document on one point: the treatment of the rewriting construct
+\(\rho\).
+In the implementation, the fully-annotated form for this construct is \(\rho\ t\
+@x.T'\ \mhyph\ t'\), but in this document it is \(\rho\ t\ @x\langle
+t_2 \rangle.T'\ \mhyph\ t'\).
+The implementation infers the additional subterm \(t_2\).
 
@@
 The typing rules (Figure~\ref{fig:tp}) are
-bidirectional~\cite{pierce+00}, while the kinding and superkinding
-rules (Figure~\ref{fig:knd} and~\ref{fig:superknd}) are unidirectional
-(i.e., synthesizing only).
+bidirectional~\cite{pierce+00}, while the kinding rules (Figure~\ref{fig:knd})
+are unidirectional (synthesizing only) and the kind-formation rules
+(Figure~\ref{fig:superknd}) have no notion of directionality.
 We write $\Leftrightarrow$ to range over
@@
 
-In terms, when the type of an introduction form is checked or the type of an
+In typing rules, when the type of an introduction form is checked or the type of an
 elimination form is synthesized, we use call-by-name normalization, written
@@
 types in weak head normal form, revealing type constructors.
-We abbreviate the conjuction ``term \(t\) synthesizes some type, and that type
+We abbreviate the conjunction ``term \(t\) synthesizes some type, and that type
 call-by-name reduces to another type'' with the symbol \(\tpsynthleads\),
@@
 term, such as \((\abs{\lambda}{x}{T_1}{T_2})\ t_1\), the reduction uses
-substitution with annotations to \([t_1/x]^{T_1}\ T_2\).
+substitution with annotations: \([t_1/x]^{T_1}\ T_2\).
 
@@
 equivalence relation \(\cong\) for types and kinds, which is full
-\(\beta\)-equivalence (for types) and \(\beta\eta\)-equivalence (for terms)
-modulo erasure of annotations in terms.
+\(\beta\)-equivalence (for types) and \(\beta\eta\)-equivalence (for terms),
+both of which are modulo erasure of annotations in terms.
 The erasure operation is defined in Figure~\ref{fig:eraser}, and is essentially
@@
     \begin{array}{c}
-      \Gamma \vdash t \tpsynthleads T\ =\ \exists T'.\ \Gamma
-      \vdash t \tpsynth T' \land T' \leadstocs T
+      \Gamma \vdash t \tpsynthleads T\ =\ \exists T'.\ (\Gamma
+      \vdash t \tpsynth T') \land (T' \leadstocs T)
       \\ \\ \\
@@
            \begin{array}{cc}
-             T \leadstocs \abs{\iota}{x}{T_1}{T_1}
+             T \leadstocs \abs{\iota}{x}{T_1}{T_2}
              & \Gamma \vdash t_1 \tpcheck T_1
@@
     |\varphi\ t\ \mhyph\ t'\ \{t''\}| & = & |t''| &\ &
-    |\chi\ T\ \mhyph\ t'| & = & |t'|
+    |\chi\ T\ \mhyph\ t| & = & |t|
   \end{array}
@@
 CDLE has as a subsystem the extrinsic Calculus of
-Constructions (CC).  We have dependent types
+Constructions (CC).  We have dependent function types
 $\abs{\Pi}{x}{T}{T'}$ and kinds $\abs{\Pi}{x}{T}{\kappa}$, as well as
@@
 
-To Curry-style CC, CDLE adds: implicit products, introduced orginially
+To Curry-style CC, CDLE adds: implicit products, introduced originally
 by Miquel~\cite{miquel01}; a primitive equality type $\{ t \simeq
@@
 derive induction for Church-encoded natural numbers~\cite{stump18}.
-We will see below further uses of this construct.
 
 The typing rules include conversion checks in a few places, e.g., as is
-standard when switching from checking to synthesizing mode.  As mentioned
-earlier, two rules near the top of Figure~\ref{fig:tp} state that one may
-(nondeterministically) call-by-name reduce the type one is synthesizing or
-checking, before proceeding.
-We also include the construct $\chi\ T\ \mhyph\ t$ which allows a term whose
-type can be checked to be ascribed a user-provided type so that the whole
-expression may occur where a term whose type can be synthesized is expected.
+standard when switching from checking to synthesizing mode.
+For the introduction forms for types, the checked type first is call-by-name
+reduced to weak head normal form, which must be formed from the appropriate connective
+for the term construct (e.g., a \(\Pi\)-type for a lambda abstraction).
+Similarly, for the elimination forms the major premise has its type synthesized
+and then call-by-name normalized to reveal the correct connective.
+As is standard for a bidirectional type system, we also include the construct
+$\chi\ T\ \mhyph\ t$ for type ascription (allowing a term whose type can be
+checked to be given a user-provided type so that the whole 
+expression synthesizes its type) and a judgmental (or \emph{subsumption}) rule
+stating that terms whose types can be synthesized may be checked against a
+convertible type.
 
@@
   with the critical idea that $|\beta\{t'\}|$ erases to (the possibly unrelated) $|t'|$.
-  We call this the \textbf{Kleene trick}, because it goes back to Kleene's numeric
+  We call this the \textbf{Kleene trick} because it goes back to Kleene's numeric
   realizability~\cite{Kle65_Classical-Extensions-of-Intuitionistic-Mathematics},
@@
   of a true equation.  Here, we accept any term $t'$ as a realizer of $\{ t_1
-  \simeq t_2\}$, provided the free variables of \(t'\) are declared in the
-  context.
+  \simeq t_2\}$ when \(t_1\) and \(t_2\) are definitionally equal, provided the
+  free variables of \(t'\) are declared in the context.
 
@@
   the type of the subexpression \(t'\).
-  The version presented here requires a type annotations (``guide'') \(@x\langle
+  The version presented here requires a type annotation (``guide'') \(@x\langle
   t_2 \rangle.T\).
-  The first subexpression \(t\) must synthesize (possibly after some
-  normalization) an equation type \(\{t_1 \simeq t_2'\}\), and the user provided
-  term \(t_2\) must be \(\beta\eta\)-convertible (modulo erasure) to \(t_2'\).
+
+  In \(\rho\ t\ @x\langle t_2 \rangle.T\ \mhyph\ t'\), the first subexpression
+  \(t\) must synthesize (possibly after some normalization) an equation type of
+  the form \(\{t_1 \simeq t_2'\}\), and the user provided term \(t_2\) must be
+  \(\beta\eta\)-convertible (modulo erasure) to \(t_2'\).
   The user provided type \(T'\) is then checked to have kind \(\star\) after
@@
   checked against this type.
+  Since equality is untyped, it may be that \([t_2'/x]T\) is not a well-kinded
+  type, so in this explicit form we require a definitionally equal term \(t_2\)
+  from the user, which may involve more typing annotations.
+
   Finally, \([t_1/x]T'\) (which need not be well-kinded) must be convertible
@@
   this document, is used to produce a resulting type that is well-kinded.
+  The current implementation of Cedille (as of April 13, 2021), does not support
+  specifying \(t_2\) in the guide \(@x\langle t_2 \rangle.T\).
+  This additional specification of the term \(t_2\) was required to prove
+  Theorem~\ref{thm:syntactic-kind-pres}, which did not appear in earlier
+  versions of this document (and its absence in prior versions does not affect
+  the semantic proofs).
 
@@
   erases to $t''$.
-\item Where the previous version of CDLE uses $\beta$-equivalence for (erased) terms, we here adopt $\beta\eta$-equivalence.  This
-  allows us to observe in many cases that retyping functions are actually $\beta\eta$-equivalent to $\absu{\lambda}{x}{x}$.
+\item Where the previous version of CDLE uses $\beta$-equivalence for (erased)
+  terms, we here adopt $\beta\eta$-equivalence.
+  This allows us to observe in many cases that retyping functions are actually
+  $\beta\eta$-equivalent to $\absu{\lambda}{x}{x}$. 
   While $\beta\eta$-equivalence takes more work to incorporate into intrinsic
-  type theory~\cite{geuvers92},
-  it raises no difficulties for our extrinsic one.
-\item In this version, we add an explicit axiom $\delta$ saying that Church-encoded boolean \emph{true} is different from
-  \emph{false}.
+  type theory~\cite{geuvers92}, it raises no difficulties for our extrinsic one.
+\item We add an explicit axiom $\delta$ saying that
+  Church-encoded boolean \emph{true} is different from \emph{false}.
   In the implementation of Cedille, this is generalized to the rule where the
@@
 
-The equality type remains \textbf{intensional}: we equate terms iff they are $\beta\eta$-equal.
+The equality type remains \textbf{intensional}: we equate closed terms iff they
+are $\beta\eta$-equal.
 
@@
 \end{theorem}
+
+Type checking \((\Gamma \vdash t \tpcheck T)\) is not covered in
+Theorem~\ref{thm:judge-valid}, since the convention for a bidirectional system
+is that there \(T\) is already \emph{assumed} to have type \(\star\) under a
+typing context.
 
@@
 Note that Lemma~\ref{lem:termsubstinterp} also applies to typed substitution: if
-\(t' =_{\cbe} \sigma|t|\) then by erasure so does \(\chi\ T\ \mhyph\ t'\).
+\(t' =_{\cbe} \sigma|t|\) then by erasure it is equal to \(\sigma|\chi\ T\
+\mhyph\ t'|\).
 
@@
   \end{itemize}
-\end{lemma}
+x\end{lemma}
 
@@
   The following proof is adapted from~\cite{stump17}.  It proceeds by
-  mutual induction on the assumed typing, kinding, or superkinding
+  mutual induction on the assumed typing, kinding, or kind formation
   derivation, for each part of the lemma.  We prove the parts
@@
   congruent with the original type in the conclusion.
-  We see the the measure is preserved.
+  We see that the measure is preserved.
 
@@
   \end{itemize}
-  Furthermore, the resulting typing derivations derivations have depths no
-  larger than the assumed ones.
+  Furthermore, the resulting typing derivations have depths no larger than the assumed ones.
 \end{lemma}
@@
   \label{lem:subst-class}
-  Below, each statement seperately universally quantifies over meta-variables,
-  and it is assumed that typing contexts occuring in assumed derivations are
+  Below, each statement separately universally quantifies over meta-variables,
+  and it is assumed that typing contexts occurring in assumed derivations are
   well-formed.
@@
   \textit{FV}(\kappa_1)\) (which we obtain from the assumption that the typing
-  context is well formed), or else \((X_1 : \kappa_1) \in \Gamma_2\). 
+  context is well-formed), or else \((X_1 : \kappa_1) \in \Gamma_2\). 
   Either way, we have \(\Gamma_1,[T/X]\Gamma_2 \vdash X_1 \tpsynth
@@
   \(\abs{\Pi}{X}{\kappa_2}{\kappa_1}\)).
-  By transitivity we can conclude \(\kappa_2'' \cong \kappa_2'\).
+  By transitivity, we can conclude \(\kappa_2'' \cong \kappa_2'\).
   By the rule, we have \(\Gamma \vdash T_1'' \cdot T_2 \tpsynth
@@
   By the IH, we have \(\Gamma,x:T \vdash \kappa\).
-  Thus we obtain \(\Gamma \vdash \abs{\Pi}{x}{T}{\kappa}\).
+  Thus, we obtain \(\Gamma \vdash \abs{\Pi}{x}{T}{\kappa}\).
 
@@
   \mhyph\ t \tpsynth T\).
-  Finally by Lemma~\ref{lem:subst-class}, \(\Gamma \vdash [t/x]^{T} \kappa\).
+  Finally, by Lemma~\ref{lem:subst-class}, \(\Gamma \vdash [t/x]^{T} \kappa\).
 
@@
 % LocalWords:  nondeterministically Kleene's realizability realizer kinded CDLE
-% LocalWords:  Cedille subcases subcase kinding
+% LocalWords:  Cedille subcases subcase kinding datatypes encodings Agda redex
+% LocalWords:  directionality Kopylov Miquel definitionally untypable Kleene
+% LocalWords:  untyped Nuprl's boolean subexpression intensional decidability
+% LocalWords:  Reducibility definedness reducibility congruential subterm

### 2021-12-09 docs/semantics/paper.tex
docs: semantics: paper: update author emails
Signed-off-by: Chris Jenkins <cwjnkins@uiowa.edu>
docs: semantic: first sketch of refinement types proposal
Signed-off-by: Chris Jenkins <cwjnkins@uiowa.edu>

@@
   The University of Iowa \\
-\texttt{aaron-stump@uiowa.edu}}
+  \texttt{aaron-stump@uiowa.edu},
+  \texttt{christopher-jenkins@uiowa.edu}}
 
@@
         \vdash \Gamma,x:T
-        }{\Gamma \vdash T \tpsynth \star}
+        }{\vdash \Gamma
+          \quad \Gamma \vdash T \tpsynth \star}
       & \infer{
@@
         }{
-        \Gamma \vdash \kappa
+        \vdash \Gamma
+        \quad \Gamma \vdash \kappa
         }
@@
 
+\section{Refinement types proposal}
+
+The following is a proposal to replace dependent intersections with refinement
+types.
+
+\begin{figure}[h!]
+  \centering
+  \[
+    \begin{array}{c}
+      \begin{array}{cc}
+        \infer{
+        \Gamma \vdash \{ x : T_1\ |\ T_2\} \tpsynth \star
+        }{
+        \Gamma \vdash T_1 \tpsynth \star
+        \quad \Gamma,x : T_1 \vdash T_2 \tpsynth \star
+        }
+        &
+          \infer{
+           \{ x : T_1\ |\ T_2\} \cong^{\text{t}} \{ x : S_1\ |\ S_2\}
+          }{
+           T_1 \cong S_1 \quad T_2 \cong S_2
+          }
+      \\ \\
+        \infer{
+         \Gamma \vdash [ t_1\ |\ t_2] \tpcheck \{ x : T_1\ |\ T_2\}
+        }{
+         \Gamma \vdash t_1 \tpcheck T_1
+         \quad \Gamma \vdash t_2 \tpcheck [t_1/x]T_2
+        }
+        &
+          \infer{
+           \Gamma \vdash \uparrow t \tpsynth T_1
+          }{
+           \Gamma \vdash t \tpsynth \{ x : T_1\ |\ T_2\}
+          }
+      \end{array}
+      \\ \\
+      \infer{
+       \Gamma \vdash \epsilon\ t_1\ -\ y.t_2 \Leftrightarrow T
+      }{
+      \begin{array}{c}
+        \Gamma \vdash t_1 \tpsynth \{ x : T_1\ |\ T_2\}
+        \quad y \notin \mathit{FV}(|t_2|)
+        \\
+        y \notin \mathit{FV}(T)
+        \quad \Gamma, y : T_2d[\uparrow t_1/x] \vdash t_2 \Leftrightarrow T
+      \end{array}
+      }
+      \\ \\
+      \begin{array}{lllllll}
+        |[ t_1\ |\ t_2]| & = & |t_1| &\
+        & |\uparrow t| & = & |t|
+        \\ |\epsilon\ t_1\ -\ y.t_2| & = & |t_2|
+      \end{array}
+      \\ \\
+      \begin{array}{lll}
+        \interp{\{ x : T_1\ |\ T_2 \}}_{\sigma,\rho}
+        & =
+        & \{E_1 \in \interp{T_1}_{\sigma,\rho}\ |\ \exists E_2 \in
+          \interp{T_2}_{\sigma[x \mapsto \choice(E)],\rho}\}
+      \end{array}
+    \end{array}
+  \]
+  \caption{Refinement types}
+  \label{fig:refinement-types}
+\end{figure}
+
+\begin{proof}[Theorem~\ref{thm:snd}]
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma \vdash \{ x : T_1\ |\ T_2\} \tpsynth \star
+    }{
+      \Gamma \vdash T_1 \tpsynth \star
+      \quad \Gamma, x : T_1 \vdash T_2 \tpsynth \star
+    }
+  \]
+
+  The set $\interp{\{x : T_1\ |\ T_2\}}_{\sigma,\rho}$ is explicitly defined to
+be a subset of $\interp{T_1}_{\sigma,\rho}$, which is in $\mathcal{R}$,
+by the IH applied to the first premise.  Since for any
+$A\subseteq\mathcal{L}$, $[A]_{\cbe}$ is in $\mathcal{R}$, to show that
+$\interp{\{x : T_1\ |\ T_2\}}_{\sigma,\rho}$ is also in $\mathcal{R}$ it suffices
+to show definedness of $\interp{T_2}_{\sigma[x\mapsto \choice(E)],\rho}\}$
+(which is used in the predicate picking out the
+particular subset of $\interp{T_1}_{\sigma,\rho}$), for
+$E\in\interp{T_1}_{\sigma,\rho}$.  For such $E$, $\choice(E)$ is defined
+(since $\interp{T_1}_{\sigma,\rho}\in\mathcal{R}$ and hence $E\in\interp{T_1}_{\sigma,\rho}$ is nonempty)
+and in $E$,
+so $\sigma[x\mapsto\choice(E)]\in\interp{\Gamma,x:T_1}$.
+So by the IH for the second premise,
+$\interp{T_2}_{\sigma[x\mapsto\choice(E),\rho]}$ is defined.
+
+\startcase{.2cm}
+\[
+  \infer{\Gamma\vdash [ t_1\ |\ t_2 ] \tpcheck \{ x : T_1\ |\ T_2\}}
+  {
+    \Gamma \vdash t_1 \tpcheck T_1
+    \quad \Gamma \vdash t_2 \tpcheck [t/x]^{T_1}\ T_2
+  }
+\]
+By the IH, we have $[\sigma |t_1|]_{\cbe}\in\interp{T_1}_{\sigma,\rho}$ and
+$[\sigma |t_2|]_{\cbe}\in\interp{[t_1/x]T_2}_{\sigma,\rho}$.
+By Lemma~\ref{lem:termsubstinterp}, the latter is equivalent to
+$[\sigma |t_1|]_{\cbe}\in\interp{T_2}_{\sigma[x\mapsto\choice([\sigma
+  |t_1|]_{\cbe}),\rho}$.
+Thus, we exhibit \([\sigma |t|]_{\cbe}\) for the existentially quantified
+\(E_2\) in the definition of \(\interp{\{x : T_1\ |\ T_2\}}_{\sigma,\rho}\).
+So, we see that \([\sigma|t_1|] = [\sigma|[ t_1\ |\ t_2]|] \in \interp{\{x :
+  T_1\ |\ T_2\}}_{\sigma,\rho}\), as desired.
+
+\startcase{.2cm}
+\[
+  \infer{\Gamma\vdash \uparrow t \tpsynth T}{\Gamma\vdash t \tpsynth \{ x : T_1\
+    |\ T_2\}}
+\]
+The desired conclusion follows easily from the IH and the semantics of
+refinement types.
+
+\startcase{.2cm}
+\[
+  \infer{
+    \Gamma \vdash \epsilon\ t_1\ -\ y.t_2 \Leftrightarrow T
+  }{
+    \begin{array}{c}
+      \Gamma \vdash t_1 \tpsynth \{ x : T_1\ |\ T_2\}
+      \quad y \notin \mathit{FV}(|t_2|)
+      \\
+      y \notin \mathit{FV}(T)
+      \quad \Gamma, y : [\uparrow t_1/x]T_2 \vdash t_2 \Leftrightarrow T
+    \end{array}
+  }
+\]
+
+By the IH on the first premise, \([\sigma|t_1|] \in \interp{\{x : T_1\ |\
+  T_2\}}\).
+This means that there exists \(E_2 \in \interp{T_2}_{\sigma[x \mapsto
+  \choice([\sigma|t_1|])]}\).
+By Lemma~\ref{lem:termsubstinterp}, \(E_2 \in \interp{[\uparrow
+  t_1/x]T_2}_{\sigma,\rho}\).
+Thus, \((\sigma \uplus [y \mapsto \choice(E_2),\rho]) \in \interp{\Gamma,y :
+  [\uparrow t_1/x]T_2}\), and by the IH on the fourth premise \(\sigma[y \mapsto
+\choice(E_2)]|t_2| \in \interp{T}_{\sigma[y \mapsto \choice(E_2)],\rho}\).
+Since \(y \notin \mathit{FV}(|t_2|)\) and \(y \notin \mathit{FV}(T)\), we have
+\(\sigma|t_2| = \sigma|\epsilon\ t_1\ -\ y.t_2| \in \interp{T}_{\sigma,\rho}\)
+as desired.
+
+\end{proof}
+
+\subsection{Dependent intersections}
+
+\begin{figure}[h]
+  \centering
+  \[
+    \begin{array}{lll}
+      \abs{\iota}{x}{T_1}{T_2}
+      & =
+      & \{ x : T_1\ |\ \{y : T_2\ |\ \{y \simeq x \}\} \}
+      \\ {[ t_1 , t_2 ]}
+      & =
+      & {[ t_1\ |\ [t_2\ |\ \beta]]}
+      % & =
+      % & [ t_1\ |\ [ t_2\ |\ \beta ]]
+      \\ t.1
+      & =
+      & \uparrow t
+      \\ t.2
+      & =
+      & \epsilon\ t\ -\ y. \epsilon\ y\ - z. \varphi\ z\ -\ (\uparrow y) \{t\}
+    \end{array}
+  \]
+  \caption{Dependent intersections}
+  \label{fig:refinement-iota}
+\end{figure}
+
+Figure~\ref{fig:refinement-iota} shows an encoding of dependent intersections
+using refinement types.
+Abusing notation, let \(\mathcal{E}\) be the natural extension of these rules to
+types, kinds, contexts, and terms.
+
+\begin{theorem}
+  \label{thm:iota-as-refinement}
+  \ \\
+  \begin{itemize}
+
+  \item Let \(u\) be an untyped \(\lambda\)-calculus term.
+    If \(|t| = u\) then \(|\mathcal{E}(t)| = u\)
+
+  \item \(\mathcal{E}([t/x]T) = [\mathcal{E}(t)/x]\mathcal{E}(T)\)
+  \item 
+    \begin{enumerate}
+    \item If \(\vdash \Gamma\) then \(\vdash \mathcal{E}(\Gamma)\)
+    \item If \(\Gamma \vdash \kappa\) then \(\mathcal{E}(\Gamma) \vdash \mathcal{E}(\kappa)\)
+    \item If \(\Gamma \vdash T \tpsynth \kappa\) then \(\mathcal{E}(\Gamma) \vdash
+      \mathcal{E}(T) \tpsynth \mathcal{E}(\kappa)\)
+    \item If \(\Gamma \vdash t \Leftrightarrow T\) then \(\mathcal{E}(\Gamma) \vdash
+      \mathcal{E}(t) \Leftrightarrow \mathcal{E}(T)\)
+    \end{enumerate}
+  \end{itemize}
+\end{theorem}
+
+\begin{proof}
+  We show a few interesting cases.
+
+  \startcase{.2cm}
+  \[
+    \mathit{If}\ \Gamma \vdash [ t_1 , t_2 ] \tpcheck \abs{\iota}{x}{T_1}{T_2}\ 
+    \mathit{then}\ \mathcal{E}(\Gamma) \vdash \mathcal{E}([ t_1 , t_2]) \tpcheck
+    \mathcal{E}(\abs{\iota}{x}{T_1}{T_2})
+  \]
+
+  We have \(|t_1| =_{\beta\eta} |t_2|\) from the fourth premise of the assumed
+  derivation.
+  By the erasure lemma, this means \(|\mathcal{E}(t_1)| =_{\beta\eta}
+  |\mathcal{E}(t_2)|\).
+  Appeal to the IH on the two typing premises of the derivation to obtain
+  \(\mathcal{D}_1 :: \mathcal{E}(\Gamma) \vdash \mathcal{E}(t_1) \tpcheck
+  \mathcal{E}(T_1)\) and
+  \(\mathcal{D}_2 :: \mathcal{E}(\Gamma) \vdash \mathcal{E}(t_2) \tpcheck
+  \mathcal{E}([t_1/x]T_2)\).
+  By the substitution lemma, this means \(\mathcal{E}(t_2)\) checks against type
+  \([\mathcal{E}(t_1)/x]\mathcal{E}(T_2)\).
+  We build the following derivation.
+  \[
+    \infer{
+      \mathcal{E}(\Gamma) \vdash [ \mathcal{E}(t_1)\ |\ [\mathcal{E}(t_2)\ |\
+      \beta ]] \tpcheck \{ x : \mathcal{E}(T_1)\ |\ \{y : \mathcal{E}(T_2)\ |\
+      \{y \simeq x\}\}\}
+    }{
+      \mathcal{D}_1 :: \mathcal{E}(\Gamma) \vdash \mathcal{E}(t_1) \tpcheck
+      \mathcal{E}(T_1)
+      \quad
+      \infer{
+        \mathcal{E}(\Gamma) \vdash [ t_2\ |\ \beta ] \tpcheck
+        \{ y : T_2\ |\ \{y \simeq x\}\}
+      }{
+        \mathcal{D}_2 :: \mathcal{E}(\Gamma) \vdash \mathcal{E}(T_2) \tpcheck
+        [\mathcal{E}(t_1)/x]\mathcal{E}(T_2)
+        \quad
+        \infer{
+          \mathcal{E}(\Gamma) \vdash \beta \tpcheck \{\mathcal{E}(t_1) \simeq \mathcal{E}(t_2)\}
+        }{
+          |\mathcal{E}(t_1)| = |\mathcal{E}(t_2)|
+        }
+      }
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \mathit{If}\ \Gamma \vdash t.1 \tpsynth T\ \mathit{then}\ \mathcal{E}(\Gamma) \vdash
+    \mathcal{E}(t.1) \tpsynth \mathcal{E}(T)
+  \]
+
+  Appeal to the IH on the premise of
+  the assumed typing derivation to obtain \(\mathcal{D} :: \mathcal{E}(\Gamma) \vdash
+  \mathcal{E}(t) \tpsynth \mathcal{E}(\abs{\iota}{x}{T_1}{T_2})\).
+  By the definition of \(\mathcal{E}\) we therefore have that \(\mathcal{E}(t)\)
+  synthesizes type \(\{ x : \mathcal{E}(T_1)\ |\ \{ y : \mathcal{E}(T_2)\ |\ \{
+  y \simeq x\}\}\}\).
+  The goal is to derive \(\mathcal{E}(\Gamma) \vdash \mathcal{E}(t.1) \tpsynth
+  \mathcal{E}(T_1)\).
+  \[
+    \infer{
+      \mathcal{E}(\Gamma) \vdash\ \uparrow \mathcal{E}(t) \tpsynth \mathcal{E}(T_1)
+    }{
+      \mathcal{D} :: \mathcal{E}(\Gamma) \vdash t \tpsynth
+      \{ x : \mathcal{E}(T_1)\ |\ \{ y : \mathcal{E}(T_2)\ |\ \{ y \simeq x \}\}\}
+    }
+  \]
+\end{proof}
+
 \end{document}

### 2023-03-16 docs/semantics/paper.tex
docs: semantics: strong normalization for types (sans reduction of terms)
Signed-off-by: Chris Jenkins <cwjnkins@uiowa.edu>

--- 2021-04-14 docs/semantics/paper.tex
@@
 \newtheorem{observation}[theorem]{Observation}
+\newtheorem{proposition}[theorem]{Proposition}
 
@@
 
-\author{Aaron Stump, Chris Jenkins \\
+\author{Aaron Stump, Christa Jenkins \\
   Computer Science \\
   The University of Iowa \\
-\texttt{aaron-stump@uiowa.edu}}
+  \texttt{aaron-stump@uiowa.edu},
+  \texttt{christa-jenkins@uiowa.edu}}
 
@@
 can be simulated other ways within the system, though the limits of
-this are still under investigation.  We also include a construct
+this are still under investigation (see
+\cite{JMS21_Simulating-Large-Elims-Cedille}).  We also include a construct 
 $\delta$, for deriving a contradiction from a proof that
@@
 types in weak head normal form, revealing type constructors.
+Note that with \(\leadsto\) we do not reduce under binders
+(\(\lambda\),\(\Pi\),\(\forall\)) or reduce terms.
 We abbreviate the conjunction ``term \(t\) synthesizes some type, and that type
@@
 term, such as \((\abs{\lambda}{x}{T_1}{T_2})\ t_1\), the reduction uses
-substitution with annotations: \([t_1/x]^{T_1}\ T_2\).
+substitution with annotations: \(T_2[t_1/x]^{T_1}\).
 
@@
 both of which are modulo erasure of annotations in terms.
+
 The erasure operation is defined in Figure~\ref{fig:eraser}, and is essentially
@@
     \infer{\Gamma\vdash\abs{\lambda}{X}{\kappa}{T'} \tpsynth \abs{\Pi}{X}{\kappa}{\kappa'}}{\Gamma \vdash \kappa & \Gamma,X:\kappa\vdash T'\tpsynth\kappa'} \\ \\
-    \infer{\Gamma\vdash T\ t \tpsynth [t/x]^{T'}\kappa}{\Gamma\vdash T \tpsynth \abs{\Pi}{x}{T'}{\kappa} & \Gamma\vdash t \tpcheck T'} &
-    \infer{\Gamma\vdash T_1 \cdot T_2 \tpsynth [T_2/X]\kappa_1}{\Gamma\vdash T_1 \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1} & \Gamma\vdash T_2 \tpsynth \kappa_2' & \kappa_2 \cong \kappa_2'} \\ \\
+    \infer{\Gamma\vdash T\ t \tpsynth \kappa[t/x]^{T'}}{\Gamma\vdash T \tpsynth \abs{\Pi}{x}{T'}{\kappa} & \Gamma\vdash t \tpcheck T'} &
+    \infer{\Gamma\vdash T_1 \cdot T_2 \tpsynth \kappa_1[T_2/X]}{\Gamma\vdash T_1 \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1} & \Gamma\vdash T_2 \tpsynth \kappa_2' & \kappa_2 \cong \kappa_2'} \\ \\
     \infer{\Gamma\vdash\abs{\iota}{x}{T}{T'} \tpsynth \star}{\Gamma \vdash T \tpsynth \star & \Gamma,x:T\vdash T' \tpsynth \star} &
@@
   \label{fig:eraser}
+\end{figure}
+
+\begin{figure}
+  \[
+    \begin{array}{c}
+      \fbox{\(T \rightsquigarrow T'\)}
+      \quad \fbox{\(\kappa \rightsquigarrow \kappa'\)}
+      \\ \\
+      \begin{array}{cc}
+        \infer{
+         (\abs{\lambda}{x}{T_1}{T_2})\ t
+         \rightsquigarrow
+         T_2 [t/x]^{T_1}
+        }{}
+        &
+          \infer{
+           (\abs{\lambda}{X}{\kappa}{T})\ \cdot T'
+           \rightsquigarrow
+           T [T'/X]
+          }{}
+        \\ \\
+        \infer{
+         T_1\ t
+         \rightsquigarrow
+         T_1'\ t
+        }{
+         T_1 \rightsquigarrow T_1'
+        }
+        \\ \\
+          \infer{
+           T_1 \cdot T_2
+           \rightsquigarrow
+           T_1' \cdot T_2
+          }{
+           T_1 \rightsquigarrow T_1'
+          }
+        & 
+        \infer{
+         T_1 \cdot T_2
+         \rightsquigarrow T_1 \cdot T_2'
+        }{
+         T_2 \rightsquigarrow T_2'
+        }
+        \\ \\
+        \infer{
+         \Gamma \vdash \abs{\lambda}{x}{T_1}{T_2} \leadsto \abs{\lambda}{x}{T_1'}{T_2}
+        }{
+         T_1 \leadsto T_1'
+        }
+        &
+        \infer{
+         \abs{\lambda}{x}{T_1}{T_2}
+         \rightsquigarrow
+         \abs{\lambda}{x}{T_1}{T_2'}
+        }{
+         T_2 \rightsquigarrow T_2'
+        }
+        \\ \\
+        \infer{
+         \abs{\lambda}{X}{\kappa}{T}
+         \leadsto \abs{\lambda}{X}{\kappa'}{T}
+        }{
+         \kappa \leadsto \kappa'
+        }
+        &
+        \infer{
+         \abs{\lambda}{X}{\kappa}{T}
+         \rightsquigarrow
+         \abs{\lambda}{X}{\kappa}{T'}
+        }{
+         T \rightsquigarrow T'
+        }
+        \\ \\
+          \infer{
+           \abs{\mathcal{Q}}{x}{T_1}{T_2}
+           \rightsquigarrow
+           \abs{\mathcal{Q}}{x}{T_1'}{T_2}
+          }{
+           T_1 \rightsquigarrow T_1'
+           \quad \mathcal{Q} \in \{\Pi,\forall,\iota\}
+          }
+        &
+        \infer{
+         \abs{\mathcal{Q}}{x}{T_1}{T_2}
+         \rightsquigarrow
+         \abs{\mathcal{Q}}{x}{T_1}{T_2'}
+        }{
+         T_2 \rightsquigarrow T_2'
+         \quad \mathcal{Q} \in \{\Pi,\forall,\iota\}
+        }
+        \\ \\
+          \infer{
+           \abs{\forall}{X}{\kappa}{T}
+           \rightsquigarrow
+           \abs{\forall}{X}{\kappa'}{T}
+          }{
+           \kappa \rightsquigarrow \kappa'
+          }
+        & 
+        \infer{
+         \abs{\forall}{X}{\kappa}{T}
+         \rightsquigarrow
+         \abs{\forall}{X}{\kappa}{T'}
+        }{
+         T \rightsquigarrow T'
+        }
+        \\ \\ \\
+        \infer{
+         \abs{\Pi}{x}{T}{\kappa}
+         \rightsquigarrow
+         \abs{\Pi}{x}{T'}{\kappa}
+        }{
+         T \rightsquigarrow T'
+        }
+        &
+          \infer{
+           \abs{\Pi}{x}{T}{\kappa}
+           \rightsquigarrow
+           \abs{\Pi}{x}{T}{\kappa'}
+          }{
+           \kappa \rightsquigarrow \kappa'
+          }
+        \\
+        \infer{
+         \abs{\Pi}{X}{\kappa_1}{\kappa_2}
+         \rightsquigarrow
+         \abs{\Pi}{X}{\kappa_1'}{\kappa_2}
+        }{
+         \kappa_1 \rightsquigarrow \kappa_1'
+        }
+        &
+          \infer{
+           \abs{\Pi}{X}{\kappa_1}{\kappa_2}
+           \rightsquigarrow
+           \abs{\Pi}{X}{\kappa_1}{\kappa_2'}
+          }{
+           \kappa_2 \rightsquigarrow \kappa_2'
+          }
+      \end{array}
+    \end{array}
+  \]
+  \caption{Term-free reduction of types and kinds}
+  \label{fig:tp-reduce}
 \end{figure}
@@
 
+Lack of normalization for terms in general notwithstanding, it is still
+important to confirm that the process of reducing types to weak head normal form
+is itself terminating.
+\begin{theorem}[Normalization of types under weak head normal form reduction]
+  \label{thm:type-norm}
+  \
+  \begin{enumerate}
+  \item If \(\Gamma \vdash \kappa\) then there exists \(\kappa'\) such that
+    \(\kappa \leadsto^* \kappa' \nleadsto\)
+  \end{enumerate} If \(\Gamma \vdash T : \kappa\) then there exists \(T'\) such
+  that \(T \leadsto^* T' \nleadsto\).
+\end{theorem}
+
 \begin{figure}
@@
 
+\section{Proof of Theorem~\ref{thm:type-norm}}
+The proof proceeds by translation of the type language into a bespoke simply
+typed lambda calculus which preserves, almost on-the-nose, the computation steps of
+interest.
+The typing rules are given in Figure~\ref{fig:bespoke-stlc}, which are to be
+taken as implicitly defining the grammar, though note that term variables may be
+both upper and lowercase.
+The calculus has two base types: the unital \(\mathit{Top}\) with constructor
+\(\mathit{top}\), and \(\mathit{\star}\) with constructor \(\mathit{con}\).
+It also has two type constructors: the standard arrow \(\to\) for function types
+and Cartesian products with \(\times\).
+The reduction rules are standard, and so it is clear that this calculus enjoys
+strong normalization and subject reduction.
+
+\begin{figure}
+  \[
+    \begin{array}{ccc}
+      \infer{
+       \Gamma \vdash X : \tau
+      }{
+       (X\!:\!\tau) \in \Gamma
+      }
+      &
+        \infer{
+         \Gamma \vdash \mathit{top} : \mathit{Top}
+        }{}
+      &
+        \infer{
+         \Gamma \vdash \mathit{con}(x\!:\!\tau,t_1,t_2) : \star
+        }{
+         \Gamma \vdash t_1 : \tau_1
+         \quad \Gamma,x\!:\!\tau \vdash t_2 : \tau_2
+        }
+      \\ \\
+      \infer{
+       \Gamma \vdash \abs{\lambda}{x}{\tau_1}{t} : \tau_1 \to \tau_2
+      }{
+       \Gamma, x\!:\!\tau_1 \vdash t : \tau_2
+      }
+      & &
+        \infer{
+         \Gamma \vdash t_1\ t_2 : \tau_2
+        }{
+         \Gamma \vdash t_1 : \tau_1 \to \tau_2
+         \quad \Gamma \vdash t_2 : \tau_1
+        }
+      \\ \\
+      \infer{
+       \Gamma \vdash (t_1,t_2) : \tau_1 \times \tau_2
+      }{
+       \Gamma \vdash t_1 : \tau_1
+       \quad \Gamma \vdash t_2 : \tau_2
+      }
+      & &
+        \infer{
+         \Gamma \vdash \mathit{proj}_i\ t : \tau_i
+        }{
+         i \in \{1,2\}
+         \quad \Gamma \vdash t : \tau_1 \times \tau_2
+        }
+    \end{array}
+  \]
+  \caption{Bespoke STLC: typing}
+  \label{fig:bespoke-stlc}
+\end{figure}
+
+The translation of CDLE's kind and type language to this calculus is shown in
+Figure~\ref{fig:bespoke-trans}, and is inspired by the translation of CC to
+System F\(_\omega\) by Geuvers and
+Nederhof~\cite{GN91_Modular-Proof-of-Strong-Normalization-CC}.
+Since reduction can also occur in kinds, we need two translations: one
+into the a type of this calculus with \(\interp{-}_{\text{K}}\), and one into
+the language of terms with \(\interp{-}\).
+For the translation of CDLE kinds into types of the calculus, \(\Pi\)-kinds are
+mapped to arrows, with quantification over terms translating to a domain of
+\(\mathit{Top}\); for the translation of kinds into terms, we use the
+constructor \(\mathit{con}\), translating the classifier of the bound variable
+as both a type and a term of the calculus.
+
+In types, type variables are preserved, lambda abstractions are translated to
+lambda abstractions, where we introduce an additional redex to model reductions
+in annotations on the bound variable and where again term parameters of any type
+become parameters of type \(\mathit{Top}\), applications are translated to
+applications where term arguments are replaced by the constant \(\mathit{top}\).
+All other type constructs are translated to expressions involving
+\(\mathit{con}\), where again the classifier ascribed to the quantified term (if
+applicable) is translated as both a type and a term of the calculus.
+Contexts are translated in the obvious way.
+
+\begin{figure}
+  \[
+    \begin{array}{lcl}
+      \interp{\star}_{\text{K}}
+      & = 
+      & \star
+      \\ \interp{\abs{\Pi}{x}{T}{\kappa}}_{\text{K}}
+      & =
+      & \mathit{Top} \to \interp{\kappa}_{\text{K}}
+      \\ \interp{\abs{\Pi}{X}{\kappa_1}{\kappa_2}}_{\text{K}}
+      & =
+      & \interp{\kappa_1}_{\text{K}} \to \interp{\kappa_2}_{\text{K}}
+      \\ \\
+      \interp{\star}
+      & =
+      & \mathit{con}(x\!:\!\mathit{Top},\mathit{top},\mathit{top})
+      \\ \interp{\abs{\Pi}{x}{T}{\kappa}}
+      & =
+      & \mathit{con}(x\!:\!\mathit{Top},\interp{T},\interp{\kappa})
+      \\ \interp{\abs{\Pi}{X}{\kappa_1}{\kappa_2}}
+      & =
+      & \mathit{con}(X\!:\!\interp{\kappa_1}_{\text{K}}, \interp{\kappa_1},\interp{\kappa_2})
+      \\ \\
+      \interp{X}
+      & =
+      & X
+      \\ \interp{\abs{\forall}{X}{\kappa}{T}}
+      & =
+      & \mathit{con}(X\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa},\interp{T})
+      \\ \interp{\abs{\forall}{x}{T_1}{T_2}}
+      & =
+      & \mathit{con}(x\!:\!\mathit{Top},\interp{T_1},\interp{T_2})
+      \\ \interp{\abs{\Pi}{x}{T_1}{T_2}}
+      & =
+      & \mathit{con}(x\!:\!\mathit{Top},\interp{T_1},\interp{T_2})
+      \\ \interp{\abs{\iota}{x}{T_1}{T_2}}
+      & =
+      & \mathit{con}(x\!:\!\mathit{Top},\interp{T_1},\interp{T_2})
+      \\ \interp{\{ t_1 \simeq t_2 \}}
+      & =
+      & \mathit{con}(x\!:\!\mathit{Top},\mathit{top},\mathit{top})
+      \\ \interp{\abs{\lambda}{x}{T_1}{T_2}}
+      & =
+      & \abs{\lambda}{x}{\mathit{Top}}{\mathit{proj}_2\ (\interp{T_1},\interp{T_2})}
+      \\ \interp{\abs{\lambda}{X}{\kappa}{T}}
+      & =
+      & \abs{\lambda}{X}{\interp{\kappa}_{\text{K}}}{\mathit{proj_2}\ (\interp{\kappa},\interp{T})}
+      \\ \interp{T\ t}
+      & =
+      & \interp{T}\ \mathit{top}
+      \\ \interp{T_1 \cdot T_2}
+      & =
+      & \interp{T_1}\ \interp{T_2}
+      \\ \\
+      \interp{\cdot}
+      & =
+      & \cdot
+      \\ \interp{\Gamma,X\!:\!\kappa}
+      & =
+      & \interp{\Gamma},X\!:\!\interp{\kappa}_{\text{K}}
+      \\ \interp{\Gamma,x\!:\!T}
+      & =
+      & \interp{\Gamma},x\!:\!\mathit{Top}
+    \end{array}
+  \]
+  \caption{Translation of kinds, types, and contexts}
+  \label{fig:bespoke-trans}
+\end{figure}
+
+Straightaway, we see that, when translated as types, all CDLE kinds (well-formed
+or not) are translated to well-formed types of this calculus, and
+well-formedness of contexts is preserved by this translation.
+It is also not hard to see that convertible kinds are translated to equal types,
+and that substitution in kinds is a no-op.
+\begin{proposition}
+  \label{thm:stlc-kinds}
+  \ \\
+  \begin{enumerate}
+  \item \(\interp{\kappa}_{\text{K}}\ \mathit{wf}\) for all \(\kappa\)
+
+  \item If \(\vdash \Gamma\) then \(\mathit{DV}(\Gamma) =
+    \mathit{DV}(\interp{\Gamma})\) and for all \((X\!:\!\kappa) \in \Gamma\),
+    \((X\!:\!\interp{\kappa}) \in \interp{\Gamma}\)
+  \item If \(\kappa_1 \cong \kappa_2\) then \(\interp{\kappa_1}_{\text{K}} = \interp{\kappa_2}_{\text{K}}\)
+  \item \(\interp{\kappa[T/X]}_{\text{K}} = \interp{\kappa}_{\text{K}} = \interp{\kappa}_{\text{K}}
+    [\interp{T}/X]\)
+  \item \(\interp{\kappa[t/x]}_{\text{K}} = \interp{\kappa}_{\text{K}} = \interp{\kappa}_{\text{K}}[\mathit{top}/x]\)
+  \end{enumerate}
+\end{proposition}
+\begin{proof}
+  Each is provable by a simple inductive argument.
+\end{proof}
+
+We can also confirm that substitution is homomorphic for the translation.
+\begin{lemma}
+  \label{thm:stlc-subst-type}
+  For all \(T\),\(T'\),\(\kappa\), and \(X\):
+  \begin{enumerate}
+  \item  \(\interp{T[T'/X]} = \interp{T}[\interp{T'}/X]\)
+
+  \item \(\interp{\kappa[T'/X] = \interp{\kappa}[\interp{T'}/X]}\)
+  \end{enumerate}
+
+\end{lemma}
+\begin{proof}
+  By mutual induction. Proceed by cases of \(T\) and \(\kappa\).
+
+  \startcase{.2cm} \(\kappa = \star\). Immediate.
+
+  \startcase{.2cm} \(\kappa = \abs{\Pi}{x}{T}{\kappa'}\).
+  Calculate
+  \[
+    \begin{array}{lclr}
+      \interp{\kappa[T'/X]}
+      & =
+      & \interp{(\abs{\Pi}{x}{T}{\kappa'})[T'/X]}
+      &
+      \\
+      & =
+      & \interp{\abs{\Pi}{x}{T[T'/X]}{\kappa'[T'/X]}}
+      & \text{def.\ subst}
+      \\
+      & =
+      & \mathit{con}(x\!:\!\mathit{Top}, \interp{T[T'/X]},
+        \interp{\kappa'[T'/X]})
+      & \text{def.\ translation}
+      \\
+      & =
+      & \mathit{con}(x\!:\!\mathit{Top},\interp{T}[\interp{T'}/X],\interp{\kappa'}[\interp{T'}/X])
+      & \text{IH}
+      \\
+      & =
+      &
+        \mathit{con}(x\!:\!\mathit{Top},\interp{T},\interp{\kappa'})[\interp{T'}/X]
+      & \text{def.\ subst}
+      \\
+      & =
+      & \interp{\abs{\Pi}{x}{T}{\kappa'}}[\interp{T'}/X]
+      & \text{def.\ translation}
+      \\
+      &
+      &
+      & \qed
+    \end{array}
+  \]
+
+  \startcase{.2cm} \(\kappa = \abs{\Pi}{Y}{\kappa_1}{\kappa_2}\).
+  Calculate
+  \[
+    \begin{array}{lclr}
+      \interp{\kappa[T'/X]}
+      & =
+      & \interp{(\abs{\Pi}{Y}{\kappa_1}{\kappa_2})[T'/X]}
+      &
+      \\
+      & =
+      & \interp{\abs{\Pi}{Y}{\kappa_1[T'/X]}{\kappa_2[T'/X]}}
+      & \text{def.\ subst}
+      \\
+      & =
+      & \mathit{con}(Y\!:\!\interp{\kappa_1[T'/X]}_{\text{K}}, \interp{\kappa_1[T'/X]},
+        \interp{\kappa_2[T'/X]})
+      & \text{def.\ translation}
+      \\
+      & =
+      & \mathit{con}(Y\!:\!\interp{\kappa_1}_{\text{K}}, \interp{\kappa_1[T'/X]},
+        \interp{\kappa_2[T'/X]})
+      & \text{Proposition~\ref{thm:stlc-kinds}(4)}        
+      \\
+      & =
+      & \mathit{con}(Y\!:\!\interp{\kappa_1}_{\text{K}}, \interp{\kappa_1}[\interp{T'}/X],
+        \interp{\kappa_2}[\interp{T'}/X])
+      & \text{IH}
+      \\
+      & =
+      &
+        \mathit{con}(Y\!:\!\interp{\kappa_1}_{\text{K}},\interp{\kappa_1},\interp{\kappa_2})[\interp{T'}/X]
+      & \text{def.\ subst}
+      \\
+      & =
+      & \interp{\abs{\Pi}{Y}{\kappa_1}{\kappa_2}}[\interp{T'}/X]
+      & \text{def.\ translation}
+      \\
+      &
+      &
+      & \qed
+    \end{array}
+  \]
+
+  \startcase{.2cm} \(T = X\). Immediate.
+
+  \startcase{.2cm} \(T = Y\) with \(Y \neq X\). Immediate.
+
+  \startcase{.2cm} \(T = \abs{\forall}{Y}{\kappa}{T''}\).
+  Calculate
+  \[
+    \begin{array}{lclr}
+      \interp{T[T'/X]}
+      & =
+      & \interp{(\abs{\forall}{Y}{\kappa}{T''})[T'/X]}
+      &
+      \\
+      & =
+      & \interp{\abs{\forall}{Y}{\kappa[T'/X]}{T''[T'/X]}}
+      & \text{def.\ subst}
+      \\
+      & =
+      & \mathit{con}(Y\!:\!\interp{\kappa[T'/X]}_{\text{K}},\interp{\kappa[T'/X]},\interp{T''[T'/X]})
+      & \text{def.\ translation}
+      \\
+      & =
+      & \mathit{con}(Y\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa[T'/X]},\interp{T''[T'/X]})
+      & \text{Proposition~\ref{thm:stlc-kinds}(4)}
+      \\
+      & =
+      & \mathit{con}(Y\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa}[\interp{T'}/X],\interp{T''}[\interp{T'}/X])
+      & \text{IH}
+      \\
+      & =
+      &
+        \mathit{con}(Y\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa},\interp{T''})[\interp{T'}/X]
+      & \text{def.\ subst}
+      \\
+      & =
+      & \interp{\abs{\forall}{Y}{\kappa}{T''}}[\interp{T'}/X]
+      & \text{def.\ translation}
+      \\
+      &
+      &
+      & \qed
+    \end{array}
+  \]
+
+  \startcase{.2cm} \(T = \abs{\forall}{x}{T_1}{T_2}\).
+  Calculate
+  \[
+    \begin{array}{lclr}
+      \interp{T[T'/X]}
+      & =
+      & \interp{(\abs{\forall}{x}{T_1}{T_2})[T'/X]}
+      &
+      \\
+      & =
+      & \interp{\abs{\forall}{x}{T_1[T'/X]}{T_2[T'/X]}}
+      & \text{def.\ subst}
+      \\
+      & =
+      & \mathit{con}(x\!:\!\mathit{Top},\interp{T_1[T'/X]},\interp{T_2[T'/X]})
+      & \text{def.\ translation}
+      \\
+      & =
+      & \mathit{con}(x\!:\!\mathit{Top},\interp{T_1}[\interp{T'}/X],\interp{T_2}[\interp{T'}/X])
+      & \text{IH}
+      \\
+      & =
+      & \mathit{con}(x\!:\!\mathit{Top},\interp{T_1},\interp{T_2})[\interp{T'}/X]
+      & \text{def.\ subst}
+      \\
+      & =
+      & \interp{\abs{\forall}{x}{T_1}{T_2}}[\interp{T'}/X]
+      & \text{def.\ translation}
+      \\
+      &
+      &
+      & \qed
+    \end{array}
+  \]
+
+  \startcase{.2cm} \(T = \abs{\Pi}{x}{T_1}{T_2}\). Similar to the above case.
+
+  \startcase{.2cm} \(T = \abs{\iota}{x}{T_1}{T_2}\). Similar to the above case.
+
+  \startcase{.2cm} \(T = \{t_1 \simeq t_2\}\). Immediate.
+
+  \startcase{.2cm} \(T = \abs{\lambda}{x}{T_1}{T_2}\).
+  Calculate
+  \[
+    \begin{array}{lclr}
+      \interp{T[T'/X]}
+      & =
+      & \interp{(\abs{\lambda}{x}{T_1}{T_2})[T'/X]}
+      &
+      \\
+      & =
+      & \interp{\abs{\lambda}{x}{T_1[T'/X]}{T_2[T'/X]}}
+      & \text{def.\ subst}
+      \\
+      & =
+      & \abs{\lambda}{x}{\mathit{Top}}{\mathit{proj}_2\ (\interp{T_1[T'/X]},\interp{T_2[T'/X]})}
+      & \text{def.\ translation}
+      \\
+      & =
+      & \abs{\lambda}{x}{\mathit{Top}}{\mathit{proj}_2\ (\interp{T_1}[\interp{T'}/X],\interp{T_2}[\interp{T'}/X])}
+      & \text{IH}
+      \\
+      & =
+      & \interp{T}[\interp{T'}/X]
+      & \text{def.\ subst}
+    \end{array}
+  \]
+
+  \startcase{.2cm} \(T = \abs{\lambda}{X}{\kappa}{T_2}\).
+  Similar to the above case, appealing to Proposition~\ref{thm:stlc-kinds}(4)
+
+  \startcase{.2cm} \(T = T_1\ t\), calculate
+  \[
+    \begin{array}{lclr}
+      \interp{T[T'/X]}
+      & = 
+      & \interp{(T_1\ t)[T'/X]}
+      &
+      \\
+      & =
+      & \interp{T_1[T'/X]\ t[T'/X]}
+      & \text{def.\ subst}
+      \\
+      & =
+      & \interp{T_1[T'/X]}\ \mathit{top}
+      & \text{def.\ translation}
+      \\
+      & =
+      & \interp{T_1}[\interp{T'}/X]\ \mathit{top}
+      & \text{IH}
+      \\
+      & 
+      & \interp{T}[\interp{T'}/X]
+      & \text{def.\ subst}
+    \end{array}
+  \]
+
+  \startcase{.2cm} \(T = T_1 \cdot T_2\).
+  Similar to the above case.
+\end{proof}
+
+\begin{lemma}
+  \label{thm:stlc-subst-term}
+  \
+  \begin{enumerate}
+  \item \(\interp{\kappa[t/x]} = \interp{\kappa}[\mathit{top}/x]\)
+  \item \(\interp{T[t/x]} = \interp{T}[\mathit{top}/x]\) for all \(T\), \(t\), and \(x\).
+  \end{enumerate}
+
+\end{lemma}
+\begin{proof}
+  By inspection, the translation removes all occurrences of free term variables
+  from \(T\) and \(\kappa\).
+\end{proof}
+
+The main effort left is to show the translation preserves the static and
+term-skipping dynamic semantics of CDLE's type language.
+
+\begin{lemma}
+  \label{thm:stlc-static-sound}
+  \ 
+  \begin{enumerate}
+  \item If \(\Gamma \vdash \kappa\) then \(\interp{\Gamma} \vdash
+    \interp{\kappa} : \star\)
+
+  \item If \(\Gamma \vdash T \tpsynth \kappa\) then \(\interp{\Gamma} \vdash
+    \interp{T} : \interp{\kappa}_{\text{K}}\)
+  \end{enumerate}
+\end{lemma}
+\begin{proof}
+  By mutual induction on the structure of the typing derivations for \(\kappa\) and \(T\).
+
+  \startcase{.2cm} \(\kappa = \star\).
+  Deduce
+  \[
+    \infer{
+      \Gamma \vdash \mathit{con}(x\!:\!\mathit{Top},\mathit{top},\mathit{top}) : \star
+    }{
+      \infer{\Gamma \vdash \mathit{top} : \mathit{Top}}{}
+      \quad \infer{\Gamma,x\!:\!\mathit{Top} \vdash \mathit{top} : \mathit{Top}}{}
+    }
+  \]
+
+  \startcase{.2cm} \[
+    \infer{
+      \Gamma \vdash \abs{\Pi}{x}{T}{\kappa}
+    }{
+      \Gamma \vdash T \tpsynth \star
+      \quad \Gamma,x\!:\!T \vdash \kappa
+    }
+  \]
+
+  By the IH, \(\interp{\Gamma} \vdash \interp{T} : \star\) and
+  \(\interp{\Gamma},x\!:\!\mathit{Top} \vdash \interp{\kappa} :
+  \interp{\star}\).
+  Deduce
+
+  \[
+    \infer{
+      \interp{\Gamma} \vdash \mathit{con}(x\!:\!\mathit{Top},\interp{T},\interp{\kappa})
+      : \star
+    }{
+      \interp{\Gamma} \vdash \interp{T} : \star
+      \quad \interp{\Gamma},x\!:\!\mathit{Top} \vdash \interp{\kappa} : \star
+    }
+  \]
+
+  \startcase{.2cm} \[
+    \infer{
+      \Gamma \vdash \abs{\Pi}{X}{\kappa_1}{\kappa_2}
+    }{
+      \Gamma \vdash \kappa_1
+      \quad \Gamma,X\!:\!\kappa_1 \vdash \kappa_2
+    }
+  \]
+
+  By the IH, \(\interp{\Gamma} \vdash \interp{\kappa_1} : \star\) and
+  \(\interp{\Gamma},X\!:\!\interp{\kappa_1}_{\text{K}} \vdash \interp{\kappa_2} :
+  \interp{\star}\).
+  Deduce
+
+  \[
+    \infer{
+      \interp{\Gamma} \vdash \mathit{con}(X\!:\!\interp{\kappa_1}_{\text{K}},\interp{\kappa_1},\interp{\kappa_2})
+    }{
+      \interp{\Gamma} \vdash \interp{\kappa_1} : \star
+      \quad \interp{\Gamma},X\!:\!\interp{\kappa_1}_{\text{K}} \vdash
+      \interp{\kappa_2} : \star
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma \vdash X \tpsynth \kappa
+    }{
+      (X\!:\!\kappa) \in \Gamma
+    }
+  \]
+
+  By Proposition~\ref{thm:stlc-kinds}(2), \((X\!:\!\interp{\kappa}) \in
+  \interp{\Gamma}\).
+  Deduce
+  \[
+    \infer{
+      \interp{\Gamma} \vdash X : \interp{\kappa}
+    }{
+      (X\!:\!\interp{\kappa}) \in \interp{\Gamma}
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma \vdash \abs{\forall}{X}{\kappa}{T} \tpsynth \star
+    }{
+      \Gamma \vdash \kappa
+      \quad \Gamma,X\!:\!\kappa \vdash T \tpsynth \star
+    }
+  \]
+
+  By the IH, \(\interp{\Gamma} \vdash \interp{\kappa}: \star\) and
+  \(\interp{\Gamma},X\!:\!\interp{\kappa}_{\text{K}} \vdash \interp{T} : \star\).
+  Deduce
+  \[
+    \infer{
+      \interp{\Gamma} \vdash
+      \mathit{con}(X\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa},\interp{T})
+      : \star
+    }{
+      \interp{\Gamma} \vdash \interp{\kappa} : \star
+      \quad \interp{\Gamma},X\!:\!\interp{\kappa}_{\text{K}} \vdash \interp{T} : \star
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma \vdash \abs{\forall}{x}{T_1}{T_2} \tpsynth \star
+    }{
+      \Gamma \vdash T_1 \tpsynth \star
+      \quad \Gamma,x\!:\!T_1 \vdash T_2 \tpsynth \star
+    }
+  \]
+
+  By the IH, \(\interp{\Gamma} \vdash \interp{T_1}: \star\) and
+  \(\interp{\Gamma},X\!:\!\mathit{Top} \vdash \interp{T_2} : \star\).
+  Deduce
+  \[
+    \infer{
+      \interp{\Gamma} \vdash
+      \mathit{con}(x\!:\!\mathit{Top},\interp{T_1},\interp{T_2})
+      : \star
+    }{
+      \interp{\Gamma} \vdash \interp{T_1} : \star
+      \quad \interp{\Gamma},x\!:\!\mathit{Top} \vdash \interp{T_2} : \star
+    }
+  \]
+
+  \startcase{.2cm} \(T = \abs{\Pi}{x}{T_1}{T_2}\).
+  Similar to the above.
+
+  \startcase{.2cm} \(T = \abs{\iota}{x}{T_1}{T_2}\).
+  Similar to the above.
+
+  \startcase{.2cm} \(T = \{t_1 \simeq t_2\}\).
+  Similar to the case for \(\kappa = \star\).
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma \vdash \abs{\lambda}{x}{T_1}{T_2} \tpsynth \abs{\Pi}{x}{T_1}{\kappa}
+    }{
+      \Gamma \vdash T_1 \tpsynth \star
+      \quad \Gamma,x\!:\!T_1 \vdash T_2 \tpsynth \kappa
+    }
+  \]
+
+  By the IH,
+  \(\interp{\Gamma} \vdash \interp{T} : \star\)
+  and \(\interp{\Gamma},x\!:\!\mathit{Top} \vdash \interp{T_2} :
+  \interp{\kappa}_{\text{K}}\).
+  Deduce
+  \[
+    \infer{
+      \interp{\Gamma}
+      \vdash \abs{\lambda}{x}{\mathit{Top}}{\mathit{proj}_2\ (\interp{T_1},\interp{T_2})}
+      : \mathit{Top} \to \interp{\kappa}_{\text{K}}
+    }{
+      \infer{
+        \interp{\Gamma},x\!:\!\mathit{Top}
+        \vdash \mathit{proj}_2\ (\interp{T_1},\interp{T_2})
+        : \interp{\kappa}_{\text{K}}
+      }{
+        \infer{
+          \interp{\Gamma},x\!:\!\mathit{Top}
+          \vdash (\interp{T_1},\interp{T_2})
+          : \star \times \interp{\kappa}_{\text{K}}
+        }{
+          \infer[\text{Weaken}]{
+            \interp{\Gamma},x\!:\!\mathit{Top}
+            \vdash \interp{T_1}
+            : \star
+          }{
+            \interp{\Gamma}
+            \vdash \interp{T_1}
+            : \star
+          }
+            \quad
+            \interp{\Gamma},x\!:\!\mathit{Top}
+            \vdash \interp{T_2}
+            : \interp{\kappa}_{\text{K}}
+          }
+        }
+      }
+      %\Gamma,x\!:\!\mathit{Top} \vdash \interp{T_2} : \interp{\kappa}
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma \vdash \abs{\lambda}{X}{\kappa}{T_1} \tpsynth \abs{\Pi}{X}{\kappa}{\kappa'}
+    }{
+      \Gamma \vdash \kappa
+      \quad \Gamma,X\!:\!\kappa \vdash T_1 \tpsynth \kappa'
+    }
+  \]
+
+  Similar to the above case.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma \vdash T_1 \cdot T_2 \tpsynth \kappa_1[T_2/X]
+    }{
+      \Gamma \vdash T_1 \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1}
+      \quad \Gamma \vdash T_2 \tpsynth \kappa_2'
+      \quad \kappa_2 \cong \kappa_2'
+    }
+  \]
+
+  By the IH, \(\interp{\Gamma} \vdash \interp{T_1} : \interp{\kappa_2}_{\text{K}} \to
+  \interp{\kappa_1}_{\text{K}}\)
+  and \(\interp{\Gamma} \vdash \interp{T_2} : \interp{\kappa_2'}_{K}\).
+  By Proposition~\ref{thm:stlc-kinds}(3), \(\interp{\kappa_2}_{\text{K}} =
+  \interp{\kappa_2'}_{\text{K}}\).
+  By Proposition~\ref{thm:stlc-kinds}(4), \(\interp{\kappa_1[T_2/X]}_{\kappa} =
+  \interp{\kappa_1}_{\kappa}\).
+  Deduce
+  \[
+    \infer{
+      \interp{\Gamma} \vdash \interp{T_1}\ \interp{T_2} : \interp{\kappa_1}_{\text{K}}
+    }{
+      \interp{\Gamma} \vdash \interp{T_1} : \interp{\kappa_2}_{\text{K}} \to
+      \interp{\kappa_1}_{\text{K}}
+      \quad \interp{\Gamma} \vdash \interp{T_2} : \interp{\kappa_2}_{\text{K}}
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \Gamma \vdash T_1\ t \tpsynth \kappa [t/x]^{T'}
+    }{
+      \Gamma \vdash T \tpsynth \abs{\Pi}{x}{T'}{\kappa}
+      \quad \Gamma \vdash t \tpcheck T'
+    }
+  \]
+
+  Similar to the above case, noting that \(\interp{T\ t} = \interp{T}\ \mathit{top}\) and using
+  Lemma~\ref{thm:stlc-subst-term}.
+\end{proof}
+
+\begin{lemma}
+  \label{thm:stlc-dynamic-sound}
+  \
+  \begin{enumerate}
+  \item If \(\kappa \leadsto \kappa'\) then \(\interp{\kappa} \leadsto^+
+    \interp{\kappa'}\)
+
+  \item If \(T \leadsto T'\) then \(\interp{T} \leadsto^+ \interp{T'}\)
+  \end{enumerate}
+\end{lemma}
+\begin{proof}
+  By mutual induction on the derivations of \(\kappa \leadsto \kappa'\) and \(T
+  \leadsto T'\).
+  We implicitly make use of the property that if \(T \leadsto^+ T'\) (or
+  \(\kappa \leadsto^+ \kappa'\)), then this reduction can be lifted to any
+  contextual closure of the classifier.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\Pi}{x}{T}{\kappa} \leadsto \abs{\Pi}{x}{T'}{\kappa}
+    }{
+      T \leadsto T'
+    }
+  \]
+
+  By the IH, \(\interp{T} \leadsto^+ \interp{T'}\).
+  Deduce
+  \[
+    \infer{
+      \mathit{con}(x\!:\!\mathit{Top},\interp{T},\interp{\kappa})
+      \leadsto^+ \mathit{con}(x\!:\!\mathit{Top},\interp{T'},\interp{\kappa})
+    }{
+      \interp{T} \leadsto^+ \interp{T'}
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\Pi}{X}{\kappa_1}{\kappa_2} \leadsto \abs{\Pi}{X}{\kappa_1'}{\kappa_2}
+    }{
+      \kappa_1 \leadsto \kappa_1'
+    }
+  \]
+
+  Similar to the above case.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\Pi}{x}{T}{\kappa} \leadsto \abs{\Pi}{x}{T}{\kappa'}
+    }{
+      \kappa \leadsto \kappa'
+    }
+  \]
+
+  By the IH, \(\interp{\kappa} \leadsto^+ \interp{\kappa'}\).
+  Deduce
+  \[
+    \infer{
+      \mathit{con}(X\!:\!\interp{Top},\interp{T},\interp{\kappa})
+      \leadsto^+ \mathit{con}(x\!:\!\mathit{Top},\interp{T},\interp{\kappa'})
+    }{
+      \interp{\kappa} \leadsto^+ \interp{\kappa'}
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\Pi}{X}{\kappa_1}{\kappa_2} \leadsto \abs{\Pi}{X}{\kappa_1}{\kappa_2'}
+    }{
+      \kappa_2 \leadsto \kappa_2'
+    }
+  \]
+
+  Similar to the above case.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\forall}{X}{\kappa}{T} \leadsto \abs{\forall}{X}{\kappa'}{T}
+    }{
+      \kappa \leadsto \kappa'
+    }
+  \]
+
+  By the IH, \(\interp{\kappa} \leadsto \interp{\kappa'}\).
+  Deduce
+  \[
+    \infer{
+      \mathit{con}(X\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa},\interp{T})
+      \leadsto \mathit{con}(X\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa'},\interp{T})
+    }{
+      \interp{\kappa} \leadsto \interp{\kappa'}
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\mathcal{Q}}{x}{T_1}{T_2}
+      \leadsto
+      \abs{\mathcal{Q}}{x}{T_1'}{T_2}
+    }{
+      T_1 \leadsto T_1'
+      & \mathcal{Q} \in \{\Pi,\forall,\iota\}
+    }
+  \]
+
+  Similar to the above case.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\forall}{X}{\kappa}{T} \leadsto \abs{\forall}{X}{\kappa}{T'}
+    }{
+      T \leadsto T'
+    }
+  \]
+
+  By the IH, \(\interp{T} \leadsto^+ \interp{T'}\).
+  Deduce
+  \[
+    \infer{
+      \mathit{con}(X\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa},\interp{T})
+      \leadsto^+
+      \mathit{con}(X\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa},\interp{T'})
+    }{
+      \interp{T} \leadsto^+ \interp{T'}
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\mathcal{Q}}{x}{T_1}{T_2} \leadsto \abs{\mathcal{Q}}{x}{T_1}{T_2'}
+    }{
+      T_2 \leadsto T_2'
+      & \mathcal{Q} \in \{\Pi,\forall,\iota\}
+    }
+  \]
+
+  Similar to the above case.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      (\abs{\lambda}{x}{T_1}{T_2})\ t \leadsto T_2 [t/x]^{T_1}
+    }{}
+  \]
+
+  By definition of the translation, \(\interp{T} =
+  (\abs{\lambda}{x}{\mathit{Top}}{\mathit{proj}_2\
+    (\interp{T_1},\interp{T_2})})\ \mathit{top}\) and 
+  \(\interp{T'} = \interp{T_2[t/x]^{T_1}}\), which by
+  Lemma~\ref{thm:stlc-subst-term} is equal to \(\interp{T_2}[\mathit{top}/x]\).
+  Deduce
+  \[
+    \infer{
+      (\abs{\lambda}{x}{\mathit{Top}}{\mathit{proj}_2\ (\interp{T_1},\interp{T_2})})\ \mathit{top}
+      \leadsto^+ \interp{T_2}[\mathit{top}/x]
+    }{
+      (\abs{\lambda}{x}{\mathit{Top}}{\mathit{proj}_2\ (\interp{T_1},\interp{T_2})})\ \mathit{top}
+      \leadsto
+      \mathit{proj}_2\
+      (\interp{T_1}[\mathit{top}/x],\interp{T_2}[\mathit{top}/x])
+      &
+      \infer{
+        \ldots
+        \leadsto^+
+        \interp{T_2}[\mathit{top}/x]
+      }{
+        \infer{
+          \ldots \leadsto \interp{T_2}[\mathit{top}/x]
+        }{}
+      }
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      (\abs{\lambda}{X}{\kappa}{T_1}) \cdot T_2 \leadsto T_1[T_2/X]
+    }{}
+  \]
+
+  By the definition of the translation, \(\interp{T} =
+  \abs{\lambda}{X}{\interp{\kappa}_{\text{K}}}{\mathit{proj}_2\ (\interp{\kappa},\interp{T_1})}\) and \(\interp{T'} =
+  \interp{T_1[T_2/X]}\), which by Lemma~\ref{thm:stlc-subst-type} is equal to
+  \(\interp{T_1}[\interp{T_2}/X]\).
+  Deduce
+  \[
+    \infer{
+      (\abs{\lambda}{X}{\interp{\kappa}}{\mathit{proj}_2\ (\interp{\kappa},\interp{T_1})})\ \interp{T_2} \leadsto^+
+      \interp{T_1}[\interp{T_2}/X]
+    }{
+      (\abs{\lambda}{X}{\interp{\kappa}}{\mathit{proj}_2\
+        (\interp{\kappa},\interp{T_1})})\ \interp{T_2}
+      \leadsto
+      \mathit{proj}_2\
+      (\interp{\kappa}[\interp{T_2}/X],\interp{T_1}[\interp{T_2}/X])
+      &
+      \infer{
+        \ldots \leadsto^+
+        \interp{T_1}[\interp{T_2}/X]
+      }{
+        \ldots \leadsto \interp{T_1}[\interp{T_2}/X]
+      }
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      T_1 \cdot T_2 \leadsto T_1' \cdot T_2
+    }{
+      T_1 \leadsto T_1'
+    }
+  \]
+
+  By the IH, \(\interp{T_1} \leadsto^+ \interp{T_1'}\).
+  Deduce
+  \[
+    \infer{
+      \interp{T_1}\ \interp{T_2} \leadsto^+ \interp{T_1'}\ \interp{T_2}
+    }{
+      \interp{T_1} \leadsto^+ \interp{T_1'}
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      T_1\ t \leadsto T_1'\ t
+    }{
+      T_1 \leadsto T_1'
+    }
+  \]
+
+  Similar to the above case.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      T_1 \cdot T_2 \leadsto T_1 \cdot T_2'
+    }{
+      T_2 \leadsto T_2'
+    }
+  \]
+
+  By the IH, \(\interp{T_2} \leadsto^+ \interp{T_2'}\).
+  Deduce
+  \[
+    \infer{
+      \interp{T_1}\ \interp{T_2} \leadsto^+ \interp{T_1}\ \interp{T_2'}
+    }{
+      \interp{T_2} \leadsto^+ \interp{T_2'}
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\lambda}{x}{T_1}{T_2} \leadsto \abs{\lambda}{x}{T_1'}{T_2}
+    }{
+      T_1 \leadsto T_1'
+    }   
+  \]
+
+  By the IH, \(\interp{T_1} \leadsto^+ \interp{T_1'}\).
+  Deduce
+  \[
+    \infer{
+      \abs{\lambda}{x}{\mathit{Top}}{\mathit{proj}_2\
+        (\interp{T_1},\interp{T_2})}
+      \leadsto^+
+      \abs{\lambda}{x}{\mathit{Top}}{\mathit{proj}_2\ (\interp{T_1'}, \interp{T_2})}
+    }{
+      \infer{
+        \mathit{proj}_2\ (\interp{T_1},\interp{T_2})
+        \leadsto^+
+        \mathit{proj}_2\ (\interp{T_1'},\interp{T_2})
+      }{
+        \infer{
+          (\interp{T_1},\interp{T_2})
+          \leadsto^+
+          (\interp{T_1'},\interp{T_2})
+        }{
+          \interp{T_1} \leadsto^+ \interp{T_1'}
+        }
+      }
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\lambda}{X}{\kappa}{T}
+      \leadsto
+      \abs{\lambda}{X}{\kappa'}{T}
+    }{
+      \kappa \leadsto \kappa'
+    }
+  \]
+
+  Similar to the above case.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\lambda}{x}{T_1}{T_2} \leadsto \abs{\lambda}{x}{T_1}{T_2'}
+    }{
+      T_2 \leadsto T_2'
+    }
+  \]
+
+  By the IH, \(\interp{T_2} \leadsto^+ \interp{T_2'}\).
+  Deduce.
+  \[
+    \infer{
+      \abs{\lambda}{x}{\mathit{Top}}{\mathit{proj}_2\ (\interp{T_1},\interp{T_2})}
+      \leadsto^+
+      \abs{\lambda}{x}{\mathit{Top}}{\mathit{proj}_2\ (\interp{T_1},\interp{T_2'})}
+    }{
+      \infer{
+        \mathit{proj}_2\ (\interp{T_1},\interp{T_2})
+        \leadsto^+
+        \mathit{proj}_2\ (\interp{T_1},\interp{T_2'})
+      }{
+        \infer{
+          (\interp{T_1},\interp{T_2})
+          \leadsto^+
+          (\interp{T_1},\interp{T_2'})
+        }{
+          \interp{T_2} \leadsto^+ \interp{T_2'}
+        }
+      }
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\lambda}{X}{\kappa}{T} \leadsto \abs{\lambda}{X}{\kappa}{T'}
+    }{
+      T \leadsto T'
+    }
+  \]
+
+  Similar to the above case.
+\end{proof}
+
+We can now that reduction of CDLE types and kinds, minus reduction of terms, is
+strongly normalizing.
+\begin{proof}[Proof (of Theorem~\ref{thm:type-norm})]
+  By contradiction. Assume we have an infinite reduction sequence starting from
+  \(T\).
+  By Lemma~\ref{thm:stlc-static-sound}, \(\interp{\Gamma} \vdash \interp{T} :
+  \interp{\kappa}\), and by coiteratively applying
+  Lemma~\ref{thm:stlc-dynamic-sound} we have constructed an infinite reduction
+  sequence starting from the well-typed STLC term \(\interp{T}\), contradicting
+  strong normalization of the calculus.
+
+  The case for kinds \(\kappa\) is analogous.
+\end{proof}
+
 \end{document}
@@
 % LocalWords:  Reducibility definedness reducibility congruential subterm
+% LocalWords:  formedness

### 2023-10-23 docs/semantics/paper.tex
docs: semantics: some old unpushed changes
Signed-off-by: Christa Jenkins <cwjnkins@uiowa.edu>

@@
   \label{thm:syntactic-kind-pres}
-  If \(\Gamma \vdash T \tpsynth \kappa\) and \(T \leadstoc T'\) then
+  \
+  \begin{itemize}
+  \item If \(\Gamma \vdash \kappa\) and \(\kappa \rightsquigarrow \kappa'\) then
+    \(\Gamma \vdash \kappa'\)
+
+  \item If \(\Gamma \vdash T \tpsynth \kappa\) and \(T \leadstoc T'\) then
     \(\Gamma \vdash T' \tpsynth \kappa'\) for some \(\kappa'\) such that \(\kappa
     \cong \kappa'\).
+  \end{itemize}
+
 \end{theorem}
@@
 Theorem~\ref{thm:judge-valid}, since the convention for a bidirectional system
-is that there \(T\) is already \emph{assumed} to have type \(\star\) under a
-typing context.
+is that there \(T\) is already \emph{assumed} to have type \(\star\) under the
+given typing context.
 
@@
 is itself terminating.
-\begin{theorem}[Normalization of types under weak head normal form reduction]
+\begin{theorem}[Normalization of types and kinds under term-free reduction]
   \label{thm:type-norm}
@@
     \(\kappa \leadsto^* \kappa' \nleadsto\)
-  \end{enumerate} If \(\Gamma \vdash T : \kappa\) then there exists \(T'\) such
+   \item If \(\Gamma \vdash T : \kappa\) then there exists \(T'\) such
      that \(T \leadsto^* T' \nleadsto\).
+ \end{enumerate}
 \end{theorem}
@@
 \begin{corollary}
-  \ \\
   \begin{itemize}
@@
 \begin{proof}[Proof (of Theorem~\ref{thm:syntactic-kind-pres})]
-  We may rule out the cases where \(T\) is a variable or formed by a type
-  constructor, as this would contradict the assumption that \(T \leadstoc
-  T'\) for some \(T'\).
+  By mutual induction on the derivation of \(\kappa \rightsquigarrow \kappa'\)
+  or \(T \rightsquigarrow T'\), depending on the case.
   We omit type annotations from substitutions when they are clear from the context.
@@
   \[
-    \infer{\Gamma\vdash T\ t \tpsynth [t/x]^{T_1}\kappa_1}{\Gamma\vdash T \tpsynth
-      \abs{\Pi}{x}{T_1}{\kappa_1} & \Gamma\vdash t \tpcheck T_1}
-  \]
-
-  There are two subcases to consider for the derivation of \(T\ t \leadstoc
-  T'\). In the first case, we have \(T \leadstoc T''\) for some \(T''\) (so
-  \(T' = T''\ t\)). By the IH, we have \(\Gamma \vdash T'' \tpsynth
-  \abs{\Pi}{x}{T_1'}{\kappa_1'}\) for some \(T_1' \cong T_1\) and \(\kappa_1'
-  \cong \kappa_1\) (we have from that IH that kind of \(T''\) must be
-  convertible with \(\abs{\Pi}{x}{T_1}{\kappa_1}\)). By
-  Lemma~\ref{lem:ctxt-conv-class2}, \(\Gamma \vdash t \tpcheck T_1'\).
-  By the rule, \(\Gamma \vdash T''\ t \tpsynth [t/x]\kappa_1'\), and the
-  synthesized kind is clearly convertible with \([t/x]\kappa_1\).
-
-  In the other subcase, the subject of kinding is of the form
-  \((\abs{\lambda}{x}{T_1}{T''})\ t\) and our assumed reduction is to
-  \([t/x]T''\). By inversion of the kinding derivation, we have \(\Gamma,x:T_1
-  \vdash T'' \tpsynth \kappa_1\). By Lemma~\ref{lem:subst-class}, we have
-  \(\Gamma \vdash [t/x]T'' \tpsynth [t/x]\kappa_1\).
-
-  \startcase{.2cm}
-  \[
-    \infer{\Gamma\vdash T_1 \cdot T_2 \tpsynth [T_2/X]\kappa_1}{\Gamma\vdash T_1
-      \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1} & \Gamma\vdash T_2 \tpsynth
-      \kappa_2' & \kappa_2 \cong \kappa_2'}
-  \]
-
-  There are two subcases to consider for the derivation of \(T_1 \cdot T_2
-  \leadstoc T'\). In the first subcase, we have \(T_1 \leadstoc
-  T_1''\) for some \(T_1''\) (so \(T' = T_1'' \cdot T_2\)). By the IH, we have
-  \(\Gamma \vdash T_1'' \tpsynth \abs{\Pi}{X}{\kappa_2''}{\kappa_1''}\) for some
-  \(\kappa_2'' \cong \kappa_2\) and \(\kappa_1'' \cong \kappa_1\) (we have from
-  the IH that the kind of \(T_1''\) must be convertible with
-  \(\abs{\Pi}{X}{\kappa_2}{\kappa_1}\)).
-  By transitivity, we can conclude \(\kappa_2'' \cong \kappa_2'\).
-  By the rule, we have \(\Gamma \vdash T_1'' \cdot T_2 \tpsynth
-  [T_2/X]\kappa_1''\), and the synthesized kind is clearly convertible with
-  \([T_2/X]\kappa_1\).
-
-  In the other subcase, the subject of kinding is of the form
-  \((\abs{\lambda}{X}{\kappa_2}{T_1''}) \cdot T_2\) and our assumed reduction is
-  to \([T_2/X]T_1''\).
-  By inversion of the kinding derivation, we have \(\Gamma,X:\kappa_2 \vdash
-  T_1'' \tpsynth \kappa_1\).
-  By Lemma~\ref{lem:ctxt-conv-class}, we have \(\Gamma,X:\kappa_2' \vdash T_1''
-  \tpsynth \kappa_1'\) for some \(\kappa_1' \cong \kappa_1\), and from this and
-  Lemma~\ref{lem:subst-class} we have \(\Gamma \vdash [T_2/X]T_1'' \tpsynth
-  [T_2/X]\kappa_1'\).
-  This is clearly convertible with \([T_2/X]\kappa_1\), as desired.
+    \infer{
+      (\abs{\lambda}{x}{T_1}{T_2})\ t \rightsquigarrow T_2[t/x]
+    }{}
+  \]
+
+  By inversion of assumed the typing derivation,
+  \[
+    \infer{
+      \Gamma \vdash (\abs{\lambda}{x}{T_1}{T_2})\ t \tpsynth \kappa[t/x]
+    }{
+      \infer{
+        \Gamma \vdash \abs{\lambda}{x}{T_1}{T_2} \tpsynth \abs{\Pi}{x}{T_1}{\kappa}
+      }{
+        \Gamma,x\!:\!T_1 \vdash T_2 \tpsynth \kappa
+      }
+      &
+      \Gamma \vdash t \tpcheck T_1
+    }
+  \]
+
+  By Lemma~\ref{lem:subst-class}, \(\Gamma \vdash T_2[t/x] \tpsynth
+  \kappa[t/x]\), as desired.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      (\abs{\lambda}{X}{\kappa}T) \cdot T' \rightsquigarrow T[T'/X]
+    }{}
+  \]
+
+  Similar to the above case
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      T_1\ t_2 \rightsquigarrow T_1'\ t_2
+    }{
+      T_1 \rightsquigarrow T_1'
+    }
+  \]
+
+  By inversion of the assumed typing derivation,
+  \[
+    \infer{
+      \Gamma \vdash T_1\ t_2 \tpsynth \kappa[t/x]
+    }{
+      \Gamma \vdash T_1 \tpsynth \abs{\Pi}{x}{T_2}{\kappa}
+      & \Gamma \vdash t_2 \tpcheck T_2
+    }
+  \]
+
+  By the IH and inversion of kind conversion, \(\Gamma \vdash T_1' \tpsynth
+  \abs{\Pi}{x}{T_2'}{\kappa'}\) with \(T_2 \cong T_2'\) and \(\kappa \cong
+  \kappa'\).
+  As it is clear that \(\kappa \cong \kappa'\) implies \(\kappa[t_2/x]
+  \cong \kappa[t_2/x]\), and also by Lemma~\ref{lem:ctxt-conv-class2}   that
+  \(\Gamma \vdash t_2 \tpcheck T_2\) implies \(\Gamma \vdash t_2 \tpcheck
+  T_2'\), we deduce
+  % TODO
+  \[
+    \infer{
+      \Gamma \vdash T_1'\ t_2 \tpsynth \kappa'[t_2/x]
+    }{
+      \Gamma \vdash T_1 \tpsynth \abs{\Pi}{x}{T_2'}{\kappa'}
+      \quad \Gamma \vdash t_2 \tpcheck T_2
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      T_1 \cdot T_2 \rightsquigarrow T_1' \cdot T_2
+    }{
+      T_1 \rightsquigarrow T_1'
+    }
+  \]
+
+  Similar to the above case, using transitivity of \(\cong\) for kinds.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      T_1 \cdot T_2 \rightsquigarrow T_1 \cdot T_2'
+    }{
+      T_2 \rightsquigarrow T_2'
+    }
+  \]
+
+  By inversion of the typing derivation,
+  \[
+    \infer{
+      \Gamma \vdash T_1 \cdot T_2 \tpsynth \kappa_1[T_2/X]
+    }{
+      \Gamma \vdash T_1 \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1}
+      & \Gamma \vdash T_2 \tpsynth \kappa_2'
+      & \kappa_2 \cong \kappa_2'
+    }
+  \]
+
+  By the IH, \(\Gamma \vdash T_2 \tpsynth \kappa_2''\) for some \(\kappa_2''
+  \cong \kappa_2'\).
+  By transitivity of \(\cong\), \(\kappa_2 \cong \kappa_2''\).
+  Deduce
+  \[
+    \infer{
+      \Gamma \vdash T_1 \cdot T_2' \tpsynth \kappa_1[T_2'/X]
+    }{
+      \Gamma \vdash T_1 \tpsynth \abs{\Pi}{X}{\kappa_2}{\kappa_1}
+      & \Gamma \vdash T_2 \tpsynth \kappa_2''
+      & \kappa_2 \cong \kappa_2''
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\mathcal{Q}}{x}{T_1}{T_2} \rightsquigarrow \abs{\mathcal{Q}}{x}{T_1'}{T_2}
+    }{
+      T_1 \rightsquigarrow T_1'
+      & \mathcal{Q} \in \{\Pi,\forall,\iota\}
+    }
+  \]
+
+  By inversion of the kinding judgment,
+  \[
+    \infer{
+      \Gamma \vdash \abs{\mathcal{Q}}{x}{T_1}{T_2} \tpsynth \star
+    }{
+      \Gamma \vdash T_1 \tpsynth \star
+      & \Gamma,x\!:\!T_1 \vdash T_2 \tpsynth \star
+    }
+  \]
+
+  By the IH and inversion of kind conversion, \(\Gamma \vdash T_1' \tpsynth
+  \star\).
+  By Lemma~\ref{lem:ctxt-conv-class2}, \(\Gamma,x\!:\!T_1' \vdash T_2 \tpsynth
+  \star\) (\(T_1 \rightsquigarrow T_1'\) implies \(T_1 \cong T_1'\)).
+  Deduce
+  \[
+    \infer{
+      \Gamma \vdash \abs{\mathcal{Q}}{x}{T_1'}{T_2} \tpsynth \star
+    }{
+      \Gamma \vdash T_1' \tpsynth \star
+      & \Gamma, x\!:\!T_1' \vdash T_2 \tpsynth \star
+      & \mathcal{Q} \in \{\Pi,\forall,\iota\}
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\forall}{X}{\kappa}{T} \rightsquigarrow \abs{\forall}{X}{\kappa'}{T}
+    }{
+      \kappa \rightsquigarrow \kappa'
+    }
+  \]
+
+  By inversion of the kinding judgment,
+  \[
+    \infer{
+      \Gamma \vdash \abs{\forall}{X}{\kappa}{T} \tpsynth \star
+    }{
+      \Gamma \vdash \kappa
+      & \Gamma,X\!:\!\kappa \vdash T \tpsynth \star
+    }
+  \]
+
+  By the IH, \(\Gamma \vdash \kappa'\).
+  By Lemma~\ref{lem:ctxt-conv-class}, \(\Gamma,X\!:\kappa' \vdash T \tpsynth \star\).
+  Deduce
+  \[
+    \infer{
+      \Gamma \vdash \abs{\forall}{X}{\kappa'}{T} \tpsynth \star
+    }{
+      \Gamma \vdash \kappa'
+      & \Gamma,X\!:\!\kappa' \vdash T \tpsynth \star
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\mathcal{Q}}{x}{T_1}{T_2}
+      \rightsquigarrow
+      \abs{\mathcal{Q}}{x}{T_1}{T_2'}
+    }{
+      T_2 \rightsquigarrow T_2'
+    }
+  \]
+
+  By inversion of the kinding rule,
+  \[
+    \infer{
+      \Gamma \vdash \abs{\mathcal{Q}}{x}{T_1}{T_2} \tpsynth \star
+    }{
+      \Gamma \vdash T_1 \tpsynth \star
+      & \Gamma,x\!:\!T_1 \vdash T_2 \tpsynth \star
+    }
+  \]
+
+  By the IH and inversion of kind conversion, \(\Gamma,x\!:\!T_1 \vdash T_2'
+  \tpsynth \star\).
+  Deduce
+  \[
+    \infer{
+      \Gamma \vdash \abs{\mathcal{Q}}{x}{T_1}{T_2'} \tpsynth \star
+    }{
+      \Gamma \vdash T_1 \tpsynth \star
+      & \Gamma,x\!:\!T_1 \vdash T_2' \tpsynth \star
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\forall}{X}{\kappa}{T} \rightsquigarrow \abs{\forall}{X}{\kappa}{T'}
+    }{
+      T \rightsquigarrow T'
+    }
+  \]
+
+  Similar to the above case.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\Pi}{x}{T}{\kappa} \rightsquigarrow \abs{\Pi}{x}{T'}{\kappa'}
+    }{
+      T \rightsquigarrow T'
+    }
+  \]
+
+  By inversion of the kinding derivation,
+  \[
+    \infer{
+      \Gamma \vdash \abs{\Pi}{x}{T}{\kappa}
+    }{
+      \Gamma \vdash T \tpsynth \star
+      \quad \Gamma,x\!:\!T \vdash \kappa
+    }
+  \]
+
+  By the IH, \(\Gamma \vdash T' \tpsynth \star\).
+  By Lemma~\ref{lem:ctxt-conv-class2}, \(\Gamma,x\!:\!T' \vdash \kappa\).
+  Deduce
+  \[
+    \infer{
+      \Gamma \vdash \abs{\Pi}{x}{T'}{\kappa}
+    }{
+      \Gamma \vdash T' \tpsynth \star
+      \quad \Gamma,x\!:\!T' \vdash \kappa
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\Pi}{X}{\kappa_1}{\kappa_2} \rightsquigarrow \abs{\Pi}{X}{\kappa_1'}{\kappa_2}
+    }{
+      \kappa_1 \rightsquigarrow \kappa_1'
+    }
+  \]
+
+  Similar to the above case, appealing to Lemma~\ref{lem:ctxt-conv-class}.
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\Pi}{x}{T}{\kappa} \rightsquigarrow \abs{\Pi}{x}{T}{\kappa'}
+    }{
+      \kappa \rightsquigarrow \kappa'
+    }
+  \]
+
+  By inversion of the kinding derivation,
+  \[
+    \infer{
+      \Gamma \vdash \abs{\Pi}{x}{T}{\kappa}
+    }{
+      \Gamma \vdash T \tpsynth \star
+      \quad \Gamma,x\!:\!T \vdash \kappa
+    }
+  \]
+
+  By the IH, \(\Gamma,x\!:\!T \vdash \kappa'\).
+  Deduce
+  \[
+    \infer{
+      \Gamma \vdash \abs{\Pi}{x}{T}{\kappa'}
+    }{
+      \Gamma \vdash T \tpsynth \star
+      \quad \Gamma,x\!:\!T \vdash \kappa'
+    }
+  \]
+
+  \startcase{.2cm}
+  \[
+    \infer{
+      \abs{\Pi}{X}{\kappa_1}{\kappa_2} \rightsquigarrow \abs{\Pi}{X}{\kappa_1}{\kappa_2'}
+    }{
+      \kappa_2 \rightsquigarrow \kappa_2'
+    }
+  \]
+
+  Similar to the above case.
 \end{proof}
@@
   \label{thm:stlc-kinds}
-  \ \\
+  \ 
   \begin{enumerate}
@@
   \end{enumerate}
+
+  Where \(\rightsquigarrow^+ =_{\textbf{df}}\ (\rightsquigarrow)\ ;\ (\rightsquigarrow^*)\)
 \end{lemma}
@@
       \mathit{con}(X\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa},\interp{T})
-      \leadsto \mathit{con}(X\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa'},\interp{T})
-    }{
-      \interp{\kappa} \leadsto \interp{\kappa'}
+      \leadsto^+
+      \mathit{con}(X\!:\!\interp{\kappa}_{\text{K}},\interp{\kappa'},\interp{T})
+    }{
+      \interp{\kappa} \leadsto^+ \interp{\kappa'}
     }
@@
   By Lemma~\ref{thm:stlc-static-sound}, \(\interp{\Gamma} \vdash \interp{T} :
-  \interp{\kappa}\), and by coiteratively applying
+  \interp{\kappa}_{\text{K}}\), and by coiteratively applying
   Lemma~\ref{thm:stlc-dynamic-sound} we have constructed an infinite reduction
